### 2013-04-23 pnsac/apr13/article_ContactInfo.tex
Initial commit

% Template PNSAC newsletter - Contact information
% Language: Latex

\title{Board Members' Contact Information}

\maketitle

\section{PNSAC Executive}
%\label{sec:misc_pnsac_executive}

%\begin{minipage}{\columnwidth}

\address{Richard Lodge\\
Director, President; Fund Raising\\
613-612-4920\\
\email{rslodgeca@gmail.com}}

\address{Bill Tate\\
Director, Vice President; Special Events, Quarterly Meetings \\
613-523-8817\\
\email{billtate@bell.net}}

\address{Bruce Gemmill\\
Director, Project Manager; Membership\\
613-841-7248\\
\email{dbgemmill@rogers.com}}

\address{Garry Dupont\\
Director, Deputy Project Manager\\
\email{gkdupont@magma.ca}}

%\end{minipage}

%\columnbreak

%\begin{minipage}{\columnwidth}

\address{Bruce Grant\\
Director, NStar Chronicle Editor\\
\email{r\_b\_grant@yahoo.ca}}

\columnbreak

\address{Drew Hodge\\
Director, newsletter typesetting, website, social media\\
\email{ldhodge@gmail.com}}

\address{Phil Chrysler\\
Merchandise\\
\email{chrysler937@rogers.com}}

\address{Roger Button, BA. MA. LLB.\\
Corporate Secretary\\
\email{rgrbttn@gmail.com}}

\address{Paul Labranche\\
Treasurer\\
\email{plabranche@andrews.ca}}

%\end{minipage}

\subsection{Newsletter}

\address{Editor: Bruce Grant}

\address{Typesetter: Drew Hodge}

%\end{multicols}

%\vspace{5mm}

%\hspace{-10mm}
%\setlength\fboxrule{2pt}%
%\framebox{%
%\begin{minipage}{1\columnwidth}%
%\begin{center}
%If you would like to contribute an article to the NStar Chronicle,
%please contact Bruce Grant at 
%\par\end{center}

%\begin{center}
%\email{r_b_grant@yahoo.ca} 
%\par\end{center}

%\begin{center}
%Your comments on the contents of this issue are also appreciated.
%\par\end{center}%
%\end{minipage}}

%\begin{multicols}{2}

\address{PNSAC Newsletter\footnote{This newsletter is typeset using \LaTeX.  The style package used for the newsletter (PNSAC.sty) is a modification of GRASSnews.sty belonging to the Geographic Analysis Resources Support System (GRASS). The modification was made possible by kind permission of the Editor-in-Chief of GRASS-News.} email address:\\
  \email{info@projectnorthstar.ca} -- Attention: Editor}\\
\begin{em}Web site:\end{em} {\normalfont\color{blue}\texttt{\url{www.projectnorthstar.ca}}}\\
%% \begin{em}Facebook page:\end{em} {\normalfont\color{blue}\texttt{\url{www.facebook.com/PNSAC}}}\\
%% \begin{em}Twitter:\end{em} {\normalfont\texttt{@pnsaccomms}}

\begin{figure}
    \centering
    \mbox{\subfigure{ \href{http://www.facebook.com/PNSAC/}{\includegraphics[scale=0.3]{../images/facebook_128x128.eps}}}\quad
    \subfigure{ \href{http://twitter.com/PnsacComms/}{\includegraphics[scale=0.3]{../images/twitter_128x128.eps}}}}
\end{figure}

%\begin{footnotesize}
%This newsletter is typeset using \LaTeX
%  \end{footnotesize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-04-23 pnsac/apr13/article_Feature2.tex
Initial commit

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{RAPCAN visits CASM}
\subtitle{Retired Airline Pilots of Canada Visit the Canada Aviation and
  Space Museum}
\author{Bill Tate}

\maketitle

At a monthly luncheon meeting of Retired Airline Pilots of Canada,
Captain Jim Strang, retired A-340 Captain, advised me the annual
general meeting for RAPCAN would be held in Ottawa in September
2012. Jim asked me if Project North Star Association of Canada could
assist the Ottawa Chapter of RAPCAN in making sure all would have a
good time.

Our guests would want to visit the North Star, and they would also
like to see the DC-9 and the F-86 in the storage hangar. Many of them
started their early jet experience flying the Sabres and had their
command promotion on the DC-9. These visits were quickly approved,
along with the booking of the Bush Theatre with the assistance of
Stephen Quick, Director General of the CASM.

RAPCAN had no idea how many people would be coming to Ottawa. Based on
previous annual general meeting attendance, from a low of 30 to a high
of 300, a figure of 125 was suggested for planning purposes. Next
question, how many people would want to come to the CASM and see the
North Star versus the Vintage Wings of Canada Air Show?

As we got closer to the date, the North Star visit was chosen by 71
visitors and 75 attended the Air Show. Our own Mike Irvin made sure
that the DC-9 was shown in the best possible way, and PNSAC volunteers
assisted through the day. Among our guests there were several former
North Star pilots for Trans Canada Airlines who added interesting
insights into that aircraft.

Ron Lemieux, former Director of Protocol at Rideau Hall, was
instrumental in ensuring a private tour of the Governor General's
Residence . All members of that tour were impressed with the history
and beauty, and with the kindness of staff who assisted those with
mobility issues.

The day started with Richard Lodge welcoming the group in the Bush
Theatre and an additional welcome by Stephen Quick. Next, a Power
Point presentation highlighted the work to date, followed by the
drawing of a 50/50 draw.

Having some insight of the pilot psyche and knowing there would be a
two hour beer call at the RCAF Mess before dinner I made an
off-the-cuff suggestion. That suggestion was - rather than accept the
50/50 prize and "have fifty of your closest friends" hit you up for
free beer at the Mess, the winner might make a cash donation back to
PNSAC. It was most appreciated that the winner did so, which in turn
made Paul Labranche, our ``guardian of the vault'' very happy as well.

After the tours of the North Star, DC-9, engine shop and restoration
area were finished, eight former F-86 pilots were allowed an up-close
and personal view of their former aircraft. It was amazing to watch
fifty years evaporating on their faces as they went up the step ladder
to see their old airplane. Much to the relief of Jim Strang, the tours
were finished in time to allow for a visitation of the CASM exhibits
and still make beer call!

In conversation with various members of RAPCAN they all seemed
consistent that this was an incredibly detailed restoration that has
far exceeded anything else they have seen. A well deserved ``well done''
for those who work on the North Star as well as the ambassadors of
PNSAC who gave their time to show what we have done on the restoration
of this iconic aircraft.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-04-23 pnsac/apr13/article_ForthEvents.tex
Initial commit

% Template PNSAC newsletter - Miscellany
% Language: Latex

\title{Calendar of Events}

\maketitle

\end{multicols}

\noindent {\bf Note}: The date of the Annual General meeting has been
changed and will now he held on Saturday, June 8, 2013.

\begin{tabbing}
May 30, 2013     \hspace{20mm}\= Board of Directors' Meeting\\
 {\normalfont\color{red}June 8, 2013}     \>
 {\normalfont\color{red}Annual General Meeting}\\
June 8, 2013      \> Board of Directors' Meeting (post AGM)\\
June 27, 2013     \> Board of Directors' Meeting (first meeting of new Board)\\
September 12, 2013     \> Board of Directors' Meeting (to be confirmed)\\
September 21, 2013     \> Members' Quarterly Meeting (to be confirmed)\\
November 28, 2013      \> Board of Directors' Meeting (to be confirmed)\\
December 7, 2013       \> Members' Quarterly Meeting (to be confirmed)\\
March 30, 2014         \> Board of Directors' Meeting (to be confirmed)\\
March 29, 2014         \> Members' Quarterly Meeting (to be confirmed)

\end{tabbing}

\begin{multicols}{2}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-04-23 pnsac/apr13/article_ProgReport.tex
Initial commit

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Project Manager's Progress Report}
\subtitle{April 2013}
\author{Bruce Gemmill}

\maketitle

During the last three months we have continued working on our two top
priorities -- Engine \#3 and the interior of the North Star.  Some of
the external skin has also been polished.  Volunteers were also
involved in preparing the North Star and Argus for an early move
outside, to accommodate an exhibit in the Storage Wing.

\section{Engine and QEC}
\label{sec:engines}

The engine rebuild is nearing completion, with magnetos and a few
other accessories to be installed.  The completed engine will soon be
installed in the engine frame, then work will begin on the
supercharger and intercooler.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{engine3.eps}
   %caption of the figure 
   \caption*{\small \em North Star engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:engine3}
\end{figure}

The basic engine frame is back together.  The three radiator sections
were cleaned and painted, then attached to the front of the frame.
The rear cowl ring has been disassembled, and new mounting hardware is
being fitted before the parts are painted and reassembled.  The front
cowl assembly (radiator cover) has also been cleaned, repaired and
polished, ready to install.  Other cowl panels will each be cleaned,
repaired and polished, ready to install once the complete engine is in
place.  Work is also underway on a host of pipes, cables, accessories
and fittings needed to complete the assembly of the engine and QEC
prior to installation on the aircraft.

As mentioned, volunteers worked to prepare the North Star for an early
move outside.  Besides installing windows and closing other openings
for weather protection, propeller \#2 was installed on engine \#2,
rather than keeping it stored in the hangar.

\section{Fuselage and Crew Area}
\label{empennage}

The crew lounge and forward washroom are looking more finished,
although much work still remains.  New Plexiglas windows were
installed in the crew lounge, washroom and APU closet.  The water tank
was installed in the ceiling, and the water lines hooked to the newly
refurbished sink.

The air scoop that supplied cold air to the Janitrol heaters was badly
corroded and removed for repair.  This involved cutting away damaged
aluminum and riveting carefully formed patches in place.  The assembly
was primed, then riveted back in place over the crew lounge.  New
insulation was installed in the ceiling, including a layer of heat
resistant mica.  Some pipes and wires have been installed, and the
heating ducts were cleaned and recovered with fabric.  These will be
installed once the Janitrol heaters are in place.

The new galley is almost complete.  After the original doors are
repaired they will be attached to the new assembly, which can then be
installed in the crew lounge, just behind the equipment rack.  New
wiring legends were produced for the inside of the electrical
cabinets.  These will be laminated before installation, to prevent
deterioration from humidity.

The ceiling and wall panels in the forward and rear belly compartments
have been stencilled and installed.  The floor panels were badly
scratched and in some cases stretched, due to the heavy load of
baggage often carried in these compartments.  Each panel is being
assessed, and several have already been remade.  These will be painted
and stencilled, then installed later this spring.

One of the new de-icing boots has been successfully manufactured.
This will be an on-going project, as time permits.  Completed boots
will be stored until other repair work on wings and stabilizers has
been completed.

Work has begun manufacturing a full set of troop seats for the
aircraft.  A total of 33 seats will be sewn and fitted to new mounting
frames.  A design has been completed and fabric work has started.
Some of the fittings have also been made.  This work will continue for
some time and all completed items will be stored until the main cabin
has been repainted.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{sink_vanity.eps}
   %caption of the figure 
   \caption*{\small \em Forward vanity.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:vanity}
\end{figure}

The forward washroom has most equipment returned, including the toilet
and the window, minus the Plexiglas insert that is on order.
Likewise, the vanity mirror and various fixtures were repainted and
re-installed.  The door will not be installed until the new flooring
is glued down, one of the last tasks in the crew area.

The APU compartment, which had been converted to luggage and coat
storage, has been re-assembled.  The power inverter and cabinet were
re-installed, along with the luggage shelf and floor.

% The heater and ducting has become a major item of work.  All the ducts
% were removed and cleaned, but new woven fibreglass coverings will need
% to be attached.  Material is on order.  The main air scoop that feeds
% the heater was removed and found to be badly corroded.  This item is
% being rebuilt and must be installed before the aircraft goes outside
% this spring.

% Work also continues on the floor, ceiling and wall panels from the
% under floor belly compartment.  Some damaged panels may need to be
% replaced.  Progress is also being made on the de-icing boots for the
% wings

% We made good progress on our plan to build a set of troop seats for
% the aircraft.  Our appeal for donations was highly successful, and
% most of the material needed for the seats has now been received.
% Numerous fittings were manufactured and painted, and are stored until
% the seats are ready to install.

\section{Planned Restoration Work--2013}
\label{sec:plannedwork}

This winter we will replace all the items removed from the crew areas
and continue the rebuild of engine \#3.

Sewing of the seat fabric will begin soon and will continue until all
the fabric has been stitched together.  The frames will be
manufactured later this year

Next we expect to begin work to restore the main cabin.  This will
require removal of the wood cargo floor for a thorough cleaning and
inspection.  We expect to find some corrosion damage that will require
repair.

%% \begin{figure}[htbp]
%%    \vspace{2em}
%%    \centering
%%    %name of the graphic, without the path AND in EPS format:
%%    \includegraphics[scale=0.5]{MerlinCylinderHead.eps}
%%    %caption of the figure 
%%    \caption*{Richard Lodge and Garry Dupont, Merlin Cylinder Head}
%%    %label of the figure, which has to correspond to \ref{}:
%%    \label{fig:eng_one}
%% \end{figure}

\begin{footnotesize}
  \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-04-28 pnsac/apr13/article_Feature1.tex
Final revisions for Apr.2013 newsletter

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{My Days at the Home of the Merlin Engine}
\author{Richard Lodge}

\maketitle

I am a bean counter and I always wanted to be an engineer. For 5 plus
years in the late 1950s I had struggled to overcome the boredom of
training to be a Chartered Accountant in Leeds in Northern England. We
worked long hours for free because it was said that our bosses would
be giving us a valuable training.

\end{multicols}

\begin{figure*}[ht!]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=1.0]{richard_lodge.eps}
   %caption of the figure 
   \caption*{\small \em Richard Lodge in the 1970s.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:richard}
\end{figure*}

\begin{multicols}{2}

Most of my friends were at university and enjoyed long summer holidays
when they would spend time tinkering with unreliable prewar old cars
while I toiled away in an office. Weekends were supposed to be spent
studying for accountancy exams. I always did the minimum amount of
study and would move into my father's big old garage at the earliest
opportunity to tinker with my 1934 Daimler limousine. The most common
tools I had to use were my hands, Whitworth spanners, feeler gauges, a
good sized hammer and Swarfega for cleaning up before doing a little
lightweight courting in the evenings.

Eventually the prized certificate arrived in the mail saying that I
had qualified as a Chartered Accountant. I was now ready to leave the
accountants' office behind and get a job as an accountant in
industry. I could hardly believe my luck when I was offered a job as
an assistant accountant in the Aero Engine Division of Rolls-Royce in
Derby.

\end{multicols}

\begin{figure*}[ht!]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.75]{rolls_royce.eps}
   %caption of the figure 
   \caption*{\small \em Nightingale Road office before closure in 2008.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:rollsroyce}
\end{figure*}

\begin{multicols}{2}

As a young accountant I was very proud of myself to have got a job
with such a prestigious company and I duly reported to work on my
first Monday to be told that I would be given an induction period
under the guidance of an experienced employee. I found I was put with
two other green accountants to be guided by a man who was an ex WW2
regular army sergeant. He was about a foot shorter than me with a
pencil moustache but had a typical sergeant's voice and manner. Les
Hart knew very little about accounting and almost nothing about
engineering but he certainly knew the company and the world. One of
the best bits of advice he gave me was ``Do not let the buggers get you
down''. I needed this advice on many occasions while at Rolls-Royce.

Following a bewildering first week of being shown round some of the
R-R factories in Derby with my fellow new accountants, Les said we
were ready to start work. I was told to report to the Overheads
Department in the main accounting building near the Nightingale Road
head office. Les introduced me to the supervisor of the department and
then went on his way to deal with some other task. At that time R-R
employed about 30,000 people in several factories around the UK and I
felt very far removed from actual aero engines while sitting in this
vast sea of clerical people. The building was a converted warehouse
and my office was in the middle of the huge second floor housing about
200 people. There was no air conditioning and very little
ventilation. As the day wore on the atmosphere became putrid to say
the least of it. My pleasure at being offered a job at Rolls-Royce was
rapidly evaporating.

Derby in the 1960s was a two industry town, Rolls-Royce and the main
railway workshops of the Midland Region of British Railways. For many
people there were few choices other than to work at one of these two
employers. Many of the employees at R-R had been with the company
since leaving school at 16. They really did not like young
inexperienced accountants being foisted upon them but I was new, full
of enthusiasm and did not realize this.

The old sweats in the Overheads Department soon let me know that I
might be more technically qualified than them but that was only an
illusion. After a month or two, I made a bad mistake and my section
leader came to me and said ``You made a bad drop-off there''. I had no
idea what a drop-off was. He then called over the supervisor to
discuss the mistake and said ``Lodge here doesn't know what a drop-off
is and I guess he doesn't know whether he is punched, bored or
countersunk either.'' By that time I knew exactly what my mistake was
but what neither of these men explained was the meaning of their other
comments. It took me some time to find Les Hart who told me that a
drop-off referred to a careless person in the factory allowing an
expensive part to drop onto the floor and break and secondly that
punched, bored or countersunk described the method of making a certain
part of my anatomy. I was beginning to realize that to survive at R-R
I had a lot to learn.

Shortly afterwards I got my own back on these gentlemen when, to their
horror, I was appointed supervisor of the Overheads department.

In the next article I will describe how I came to enjoy the rest of my
time at R-R and learnt about life in the real world of traditional
British engineering.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-04-28 pnsac/apr13/article_PresReport.tex
Final revisions for Apr.2013 newsletter

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Notes from the President}
\author{Richard Lodge}

\maketitle

As I write these notes in the middle of April, I see a fresh fall of
snow on the ground.  For the first time in many years, the North Star,
already out in the open, will be covered in snow.  This unusual
situation has been caused by the Star Wars exhibition to be located in
the Storage Hangar this summer.  The larger Argus airplane has also
been moved outside, in a position which reduces visibility of the
North Star for the public.  This will present a challenge to us in
showing off our work to Museum visitors.

The Star Wars exhibition has also affected our plans for commemorating
the 10th anniversary of the founding of the Project North Star
Association.  We are now planning to celebrate the anniversary when we
open the North Star for the public on Canada Day.  That will be the
only time this summer when we can invite the general public to see the
North Star and offer our members conducted tours inside the plane.  We
will ensure that there are good directions to the North Star and
volunteers located in the main Museum will assist visitors and members
in finding it.

All this has had only a minor effect on the plane restoration.  We are
very actively working both inside the restoration shop and inside the
plane itself.  As usual our Project Manager has provided elsewhere in
the Chronicle a comprehensive report of our progress.

In the next Chronicle, we are planning to start a Letters to the
Editor section.  Many of our members are not located in Ottawa and may
well have interesting questions or comments on our work, or stories or
pictures about North Star in general.  We will welcome any letters or
stories, which should be sent to the Editor
(\email{editor@projectnorthstar.ca}) by email or by mail to the
address shown on the last page of this newsletter.  As with any
publication, our editor will decide whether an item should be
published.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-04-28 pnsac/apr13/article_RecentEvents.tex
Final revisions for Apr.2013 newsletter

% Template PNSAC newsletter
% Language: Latex

\title{Special Events}
%\subtitle{Visit to the Canada Aviation and Space Museum}
%\author{Bill Tate}

\maketitle

\section{Project North Star Visit to RCAF 426 Squadron ``Thunderbirds''
  -- Friday Jun 14, 2013'}
\label{sec:trenton}

Project North Star is pleased to announce a tour of 426 Squadron at
RCAF Trenton on Friday, June 14th, 2013.  This visit came up at very
short notice, through the help of Tim Timmins (retired Colonel CAF and
member of PNSAC) and Lieutenant Colonel Damon Perrault, Commanding
Officer of 426 Squadron.

Due to the fantastic demand that was generated for our trip to
Montreal, I would strongly recommend an early booking to avoid
disappointment as seats are limited for this fun day of activities.

Our visit to 426 Squadron, which does all the training for the new
C-130J, will involve a tour of their brand new state of the art
training facilities that includes the two airframes that are used for
technician training, another airframe for training loadmasters, a
procedure trainer, and full motion simulators with the usual caveat
``operational requirements.''

After our visit, we will attend the RCAF Memorial Museum which has
undergone a major upgrade to its display area along with their work
shops.

Afterwards we will have a debriefing at Rumour's Restaurant for
schnitzel before returning to Ottawa.

The cost per person is \$70:00. Please send two cheques including a
non-refundable deposit of \$30:00 and a post- dated cheque in the
amount of \$40:00 dated May 14, 2013.

Please note ``Trenton Trip'' on your cheques in addition to sending me
(Bill Tatae) an e-mail message (\email{pnsac.specialevents@gmail.com})
to confirm your participation.  Please include your full name and date
of birth in your e-mail message (this is a requirement of RCAF
Security) as well as your preferred contact information, email or
telephone.

Active volunteers can bring your cheques to the PNSAC office at the
CASM or alternatively mail to:

\address{PNSAC\\
P.O.Box 44005\\
514 Montreal Road\\
Ottawa, ON\\
K1K 4P8}

\section{Third Annual Golf Day at Loch March -- Wednesday August 14, 2013}
\label{sec:golf}

Our Second planned Special Event is for Wednesday August 14th which is
Third Annual Golf Day at Loch March in Carp Ontario.

The day will start off with breakfast at 09:00 with golf starting
shortly after 10:00.

The club professional, as last year, has graciously given PNSAC two
free passes for 18 holes at Loch March, which will be awarded for the
longest drive (staying in bounds of course) and closest to the
pin. Format will be best ball.

Afterwards on the deck, we will discuss the day's game and the
Vice-President will give the winner a ``free beer'' for their effort.

Please email \email{pnsac.specialevents@gmail.com} to confirm you are coming
and call the starter at Loch March 613-839-5885.

Please visit the web site
{\normalfont\color{blue}\texttt{\url{www.lochmarch.com}}} for more
information such as facilities and dress code.

\section{Montreal Area Control Centre and the Bombardier Aerospace  --
Friday Otober 4, 2013}
\label{sec:montreal}

Our Third planned Special Event, on Friday October 4th, is a trip to
the Montreal Area Control Centre and the Bombardier Aerospace factory
in Dorval.

The Montreal Area Control Centre is a secure facility responsible for
Radar Air Traffic Control in the Quebec region including northern
Labrador up to Baffin Island. This tour will have us divided into
smaller groups due to the requirement for quiet so as not to disturb
the Air Traffic Controllers. For operational considerations this tour
could be cancelled on short notice if there is a declared emergency or
an emergency in progress.

After a lunch break we will travel to the Bombardier factory in Dorval
where the CRJ series jet aircraft are built.

For dinner we will proceed to the Willow Inn in Hudson before
proceeding back to Ottawa.

As this tour is now sold out we can accept bookings on a cancellation
basis only.

To make a cancellation booking please email
\email{pnsac.specialevents@gmail.com} to confirm you are willing to standby
for this trip, and mail a non-refundable cheque payable to PNSAC for
\$60:00 payable to PNSAC and in the cheque memo please put in Montreal
trip. Please note this cheque will only be cashed if space becomes
available and if no space becomes available the cheque will be
destroyed.

Please mail the completed booking sheet and cheques to:

\address{PNSAC\\
P.O.Box 44005\\
514 Montreal Road\\
Ottawa, ON\\
K1K 4P8}

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-07-14 pnsac/jul13/article_ContactInfo.tex
Files for July 2013 newsletter.

--- 2013-04-23 pnsac/apr13/article_ContactInfo.tex
@@
 Director, NStar Chronicle Editor\\
-\email{r\_b\_grant@yahoo.ca}}
+\email{editor@projectnorthstar.ca}}
 
@@
 Treasurer\\
-\email{plabranche@andrews.ca}}
+\email{treasurer@projectnorthstar.ca}}
 
@@
 
-\subsection{Newsletter}
+\subsection{Newsletter\protect\footnote{This newsletter is typeset using
+    \LaTeX.  The style package used for the newsletter (PNSAC.sty) is
+    a modification of GRASSnews.sty belonging to the Geographic
+    Analysis Resources Support System (GRASS). The modification was
+    made possible by kind permission of the Editor-in-Chief of
+    GRASS-News.}}
 
-\address{Editor: Bruce Grant}
+\small\address{Editor: Bruce Grant\\
+\small\email{editor@projectnorthstar.ca}}
 
-\address{Typesetter: Drew Hodge}
+%\address{Typesetter: Drew Hodge}
 
@@
 
-\address{PNSAC Newsletter\footnote{This newsletter is typeset using \LaTeX.
-The style package used for the newsletter (PNSAC.sty) is a modification of GRASSnews.sty belonging to the Geographic Analysis Resources Support System (GRASS).
-The modification was made possible by kind permission of the Editor-in-Chief of GRASS-News.} email address:\\
-  \email{info@projectnorthstar.ca} -- Attention: Editor}\\
-\begin{em}Web site:\end{em} {\normalfont\color{blue}\texttt{\url{www.projectnorthstar.ca}}}\\
-%% \begin{em}Facebook page:\end{em} {\normalfont\color{blue}\texttt{\url{www.facebook.com/PNSAC}}}\\
-%% \begin{em}Twitter:\end{em} {\normalfont\texttt{@pnsaccomms}}
+%\address{PNSAC Newsletter postal and Web site addresses:}
+
+\scriptsize\address{PNSAC\\
+P.O.Box 44005\\
+514 Montreal Road\\
+Ottawa, ON\\
+K1K 4P8}\\
+
+\scriptsize\address{Web site: {\normalfont\color{blue}\texttt{\url{www.projectnorthstar.ca}}}}\\
 

### 2013-07-14 pnsac/jul13/article_ForthEvents.tex
Files for July 2013 newsletter.

% Template PNSAC newsletter - Miscellany
% Language: Latex

\title{Calendar of Events}

\maketitle

\end{multicols}

\noindent {\bf Note}: The date of the Annual General meeting has been
changed and will now he held on Saturday, June 8, 2013.

\begin{tabbing}
May 30, 2013     \hspace{20mm}\= Board of Directors' Meeting\\
 {\normalfont\color{red}June 8, 2013}     \>
 {\normalfont\color{red}Annual General Meeting}\\
June 8, 2013      \> Board of Directors' Meeting (post AGM)\\
June 27, 2013     \> Board of Directors' Meeting (first meeting of new Board)\\
September 12, 2013     \> Board of Directors' Meeting (to be confirmed)\\
September 21, 2013     \> Members' Quarterly Meeting (to be confirmed)\\
November 28, 2013      \> Board of Directors' Meeting (to be confirmed)\\
December 7, 2013       \> Members' Quarterly Meeting (to be confirmed)\\
March 20, 2014         \> Board of Directors' Meeting (to be confirmed)\\
March 29, 2014         \> Members' Quarterly Meeting (to be confirmed)

\end{tabbing}

\begin{multicols}{2}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-07-14 pnsac/jul13/article_Special.tex
Files for July 2013 newsletter.

--- 2013-04-28 pnsac/apr13/article_Feature1.tex
@@
 
-\title{My Days at the Home of the Merlin Engine}
-\author{Richard Lodge}
+\title{Special Presentation}
+%% \author{Drew Hodge}
 
@@
 
-I am a bean counter and I always wanted to be an engineer. For 5 plus
-years in the late 1950s I had struggled to overcome the boredom of
-training to be a Chartered Accountant in Leeds in Northern England. We
-worked long hours for free because it was said that our bosses would
-be giving us a valuable training.
-
 \end{multicols}
 
-\begin{figure*}[ht!]
+\begin{figure}[ht!]
    \vspace{2em}
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=1.0]{richard_lodge.eps}
+   \includegraphics[scale=1.0]{roger_button.eps}
    %caption of the figure 
-   \caption*{\small \em Richard Lodge in the 1970s.}
+   \caption*{\small \em "Do I really have to wear it?"}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:richard}
-\end{figure*}
+   \label{fig:rbutton}
+\end{figure}
 
-\begin{multicols}{2}
-
-Most of my friends were at university and enjoyed long summer holidays
-when they would spend time tinkering with unreliable prewar old cars
-while I toiled away in an office. Weekends were supposed to be spent
-studying for accountancy exams. I always did the minimum amount of
-study and would move into my father's big old garage at the earliest
-opportunity to tinker with my 1934 Daimler limousine. The most common
-tools I had to use were my hands, Whitworth spanners, feeler gauges, a
-good sized hammer and Swarfega for cleaning up before doing a little
-lightweight courting in the evenings.
-
-Eventually the prized certificate arrived in the mail saying that I
-had qualified as a Chartered Accountant. I was now ready to leave the
-accountants' office behind and get a job as an accountant in
-industry. I could hardly believe my luck when I was offered a job as
-an assistant accountant in the Aero Engine Division of Rolls-Royce in
-Derby.
-
-\end{multicols}
-
-\begin{figure*}[ht!]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.75]{rolls_royce.eps}
-   %caption of the figure 
-   \caption*{\small \em Nightingale Road office before closure in 2008.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:rollsroyce}
-\end{figure*}
-
-\begin{multicols}{2}
-
-As a young accountant I was very proud of myself to have got a job
-with such a prestigious company and I duly reported to work on my
-first Monday to be told that I would be given an induction period
-under the guidance of an experienced employee. I found I was put with
-two other green accountants to be guided by a man who was an ex WW2
-regular army sergeant. He was about a foot shorter than me with a
-pencil moustache but had a typical sergeant's voice and manner. Les
-Hart knew very little about accounting and almost nothing about
-engineering but he certainly knew the company and the world. One of
-the best bits of advice he gave me was ``Do not let the buggers get you
-down''. I needed this advice on many occasions while at Rolls-Royce.
-
-Following a bewildering first week of being shown round some of the
-R-R factories in Derby with my fellow new accountants, Les said we
-were ready to start work. I was told to report to the Overheads
-Department in the main accounting building near the Nightingale Road
-head office. Les introduced me to the supervisor of the department and
-then went on his way to deal with some other task. At that time R-R
-employed about 30,000 people in several factories around the UK and I
-felt very far removed from actual aero engines while sitting in this
-vast sea of clerical people. The building was a converted warehouse
-and my office was in the middle of the huge second floor housing about
-200 people. There was no air conditioning and very little
-ventilation. As the day wore on the atmosphere became putrid to say
-the least of it. My pleasure at being offered a job at Rolls-Royce was
-rapidly evaporating.
-
-Derby in the 1960s was a two industry town, Rolls-Royce and the main
-railway workshops of the Midland Region of British Railways. For many
-people there were few choices other than to work at one of these two
-employers. Many of the employees at R-R had been with the company
-since leaving school at 16. They really did not like young
-inexperienced accountants being foisted upon them but I was new, full
-of enthusiasm and did not realize this.
-
-The old sweats in the Overheads Department soon let me know that I
-might be more technically qualified than them but that was only an
-illusion. After a month or two, I made a bad mistake and my section
-leader came to me and said ``You made a bad drop-off there''. I had no
-idea what a drop-off was. He then called over the supervisor to
-discuss the mistake and said ``Lodge here doesn't know what a drop-off
-is and I guess he doesn't know whether he is punched, bored or
-countersunk either.'' By that time I knew exactly what my mistake was
-but what neither of these men explained was the meaning of their other
-comments. It took me some time to find Les Hart who told me that a
-drop-off referred to a careless person in the factory allowing an
-expensive part to drop onto the floor and break and secondly that
-punched, bored or countersunk described the method of making a certain
-part of my anatomy. I was beginning to realize that to survive at R-R
-I had a lot to learn.
-
-Shortly afterwards I got my own back on these gentlemen when, to their
-horror, I was appointed supervisor of the Overheads department.
-
-In the next article I will describe how I came to enjoy the rest of my
-time at R-R and learnt about life in the real world of traditional
-British engineering.
+At a Board meeting on March 20, the directors presented a PNSAC golf
+shirt to our hard-working Secretary, Roger Button. Our Vice President
+Bill Tate made the presentation; the joy and gratitude are evident on
+Roger's face.
 
@@
 \end{footnotesize}
+
+\begin{multicols}{2}
 

### 2013-07-15 pnsac/jul13/article_Feature1.tex
Changes and images for Jul 2013 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{My Days at the Home of the Merlin Engine}
\author{Richard Lodge}

\maketitle

As a new employee, part of my Rolls-Royce induction process was to get
to know some of the production and other R-R facilities around
Derby. After a few weeks of working in the Overheads Department, it
was suggested that it was time for me to go visiting. I was happy to
do this because any opportunity to leave the main accounting office
for a time was to be jumped at.

Rolls-Royce had an internal bus service to take employees between
sites. To catch the bus, an employee went down to the door of the
building and waited for a bus to turn up -- no booking or written
schedule.  I remember my first trip on a company bus. It took me about
half an hour of asking people around how to get to the site I
wanted. I was beginning to realize that much of the operations of R-R
were not written and were passed around by word of mouth and the long
time employees passing on information to the newcomers. Years later,
when working in Northern Canada in native communities I once again
came across this way of handing down unwritten information from the
Elders to the younger people. Before starting my job at R-R I had
believed that the company would be highly organized and disciplined.

I finally managed to find my way to the building I was due to visit.
It was the Foundry. I was introduced to the men who worked in this
evil smelling place. Each of the major engine parts was individually
cast -- there was no automation and a good foundry man could put
together the sand and other parts needed for the casting, mostly from
memory. During WW2 the foundry had concentrated entirely on the
production of military engines, mostly Merlins.  By the time I was
shown round the place there was a bewildering conglomeration of parts
being made for piston, turbo prop and jet engines.

As a young accountant I was expected to understand how the paperwork
and accounting for all this worked. It was a daunting thought and I
left the Foundry feeling overwhelmed.  Throughout my time at R-R I
never became comfortable dealing with the Foundry -- everything seemed
to be so imprecise and up to the skill on the foundry man. Each time I
went there I tried to find out how the men knew how to produce the
same casting every time and the answer was always the same: "Oh, we
just know what to do".  Many of these comments were probably made to
try to make a young guy in a suit from the office look foolish but
they were also largely true!

Sitting in the Overheads Department I saw many of the invoices for
purchases. I could hardly believe it when I saw invoices going through
for payment for dozens of chickens. The delivery was never to the
works canteen but always to the R-R site at Hucknall, a short distance
outside Derby. My supervisor told me that the chickens were fired at
engines on test. I did not believe him until I managed to find a
pretext to visit Hucknall. After another challenge finding the works
bus to get me there, I arrived at the site and was shown round and all
was revealed. Dead chickens ending their time on earth being used to
test the strength of compressor blades in an engine on full power
seemed to me to be a very imprecise way of testing a jet
engine. Another lesson learned -- not everybody was trying to make me
look foolish all the time.

The visits to the various R-R sites around Derby took several months
to complete. I had to have a good reason to leave my office and go
visiting, particularly as I was now working in a departmental
position. Eventually another bus trip to the large new complex at
Sinfin was required or I somehow managed to talk my way into finding
it necessary to go there. Sinfin, among many other things, housed the
main engine test beds and I was curious to find out what happened
there. Unlike the evil smelling Foundry, Sinfin was clean, new and had
plenty of windows.

The men in the test shop told me about the engine testing process. I
was still new to R-R and had a lot to learn. After a while I was asked
if I would like to see an engine on test. Of course I said yes; it
seemed interesting. With little experience I did not realize that an
aero engine on test is nothing much to see. One of the men said he
would take me to a Conway engine on test. He said he would open the
test bed door to let me see the engine and that I would find it
somewhat noisy. He stood behind the door, opened it and let me look
inside at the engine on full power. I now wear hearing aids. Health
and safety were two words not used in those days when many of the men
were used to serving in the war. This was 1960s Britain.

I have always enjoyed travelling and would jump at any opportunity to
travel to a new place. After a few months it was suggested to me that
I should visit the Scottish factory at Hillington near Glasgow. I
asked how I should get there. ``Oh. You might as well fly. There is a
flight every day that leaves in the morning and comes back in the
afternoon.'' I had never flown before.

The UK being a small country, everybody went around by train.  As
usual, I was given very little information and told to get myself to
East Midlands airport early on the appointed morning, where a ticket
would be waiting for me. On this occasion I had the good sense to take
my own car to the airport and did not get lost. I expected to find a
plane waiting, painted in some sort of Rolls-Royce colours. Instead,
after some questioning, I found out that I would be travelling on a
small airline, called British Midland, that ran a daily service
between the Derby and Glasgow factories. Far from being in some
interesting plane connected with R-R, I found that our plane was an
old DC3 which had seen better days.I duly got on the plane and was
somewhat nervous flying on my first flight.

An hour later we arrived in Scotland, also my first trip over the
Border, and I spent the day being shown around the factories and
offices.  I remember very little about the day but vividly remember
the flight back to Derby. It must have been in the winter because it
was getting dark as I boarded the plane around 5.00 pm. Seating was
first come first served and all the old sweats travelling knew what to
do. I ended up on the back row near the cargo area. The main thing I
remember was the crew having a fierce argument about the load on the
plane which was full. The discussion centred around the amount of
cargo which could be taken on the flight. I was quite relieved to get
back to Derby in one piece.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-07-15 pnsac/jul13/article_ForthEvents.tex
Changes and images for Jul 2013 newsletter.

@@
 \begin{tabbing}
-May 30, 2013     \hspace{20mm}\= Board of Directors' Meeting\\
- {\normalfont\color{red}June 8, 2013}     \>
- {\normalfont\color{red}Annual General Meeting}\\
-June 8, 2013      \> Board of Directors' Meeting (post AGM)\\
-June 27, 2013     \> Board of Directors' Meeting (first meeting of new Board)\\
-September 12, 2013     \> Board of Directors' Meeting (to be confirmed)\\
+%May 30, 2013     \hspace{20mm}\= Board of Directors' Meeting\\
+% {\normalfont\color{red}June 8, 2013}     \>
+% {\normalfont\color{red}Annual General Meeting}\\
+%June 8, 2013      \> Board of Directors' Meeting (post AGM)\\
+%June 27, 2013     \> Board of Directors' Meeting (first meeting of new Board)\\
+September 12, 2013     \hspace{20mm}\= Board of Directors' Meeting (to be confirmed)\\
 September 21, 2013     \> Members' Quarterly Meeting (to be confirmed)\\

### 2013-07-15 pnsac/jul13/article_PresReport.tex
Changes and images for Jul 2013 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Notes from the President}
\author{Richard Lodge}

\maketitle

Canada Day is now behind us.  We had an excellent day with a good
number of our volunteers showing off the North Star to the general
public. Judging by the activity around the plane all the day, we have
now developed a display format which is interesting to visitors.
Canada Day was the only day this year when the plane was on display to
the general public.

On June 27, in co-operation with the Museum, who hosted the event, we
showed off the plane to a group of donors attending the Donor
Appreciation Evening.  This was a successful evening which is likely
to be repeated.

In my last Notes from the President, I referred to plans for
celebrating the 10th anniversary of the founding of the Project North
Star Association.  This should have taken place around June 24.
Unfortunately, the event passed without us being able mark the
day. This was mainly due to situations beyond our control.

Our 2013 AGM was held on June 8, and all the existing directors were
re-elected.  This will enable the directors, who have now become
experienced, to concentrate on long-term matters.

The actual work on the restoration of the plane is going very well, as
reported elsewhere in this newsletter. The major issue facing the
Board in the next year will be to consider how to secure the future of
PNSAC, both in terms of volunteers and the funding of restoration
work.

We have now reached a point where it is necessary to update our
Memorandum of Understanding (MOU) with the Museum and our work plan
for completing the restoration of the plane.  Our volunteers have
already worked over 60,000 hours on the project and we expect to work
a similar number of hours before the restoration is completed.

There have been several informal meetings to discuss our future
operations and the opportunities available for new volunteers.  Within
the next few weeks we are expecting to meet with the Director General
of the Aviation Museum and members of his staff to work on the MOU
update.

Later this month we are meeting with senior management of the Museum
Foundation, the body charged with corporate fundraising for the Canada
Science and Technology Museums Corporation, to discuss how PNSAC can
work with the Foundation to raise funds for the North Star restoration
and the museums in general.

With the initiatives above, we are hoping in the next few months to
consolidate our position as an important organization working with the
Aviation Museum

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-07-15 pnsac/jul13/article_RecentEvents.tex
Changes and images for Jul 2013 newsletter.

% Template PNSAC newsletter
% Language: Latex

\title{Project North Star Trip to  426 Squadron}
\author{Bill Tate}

\maketitle

Friday June 14th dawned bright and clear for our visit to 426 Squadron
at 8 Wing Royal Canadian Air Force Station Trenton.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{trenton2013-herc.eps}
   %caption of the figure 
   \caption*{\small \em National Air Force Museum -- Hercules.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:herc}
\end{figure}

Once again scheduling allowed us to have Jean Pierre as our driver
which meant JP had as much fun as we did. We departed right on time at
07:00 with Richard Lodge giving everyone on board a welcome and a
brief explanation of the day's events.

With the kind assistance of Garry and Jim, all on board were offered
our normal complimentary coffee and donuts to help kick start the
early day. Following a brief stop in Mallorytown on the 401 we
proceeded for what we thought would be an on-time arrival at 426
Squadron but a freight train had other plans for us. After skillful
driving by JP we arrived only 20 minutes late at the Sedley
S. Blanchard Air Mobility Training Center which is the home of 426
Squadron.

Our group was warmly greeted by Lieutenant Colonel Perrault the
Commanding Officer of 426 Squadron who gave us a briefing on what
responsibilities 426 Squadron has in its training role .

After our welcome we were split into groups to start our planned tours
of this very modern high tech facility. One of the many highlights of
the tour was the Fuselage Trainer, a full size fuselage from an old
airframe which is used for training load masters. The Fuselage Trainer
can be integrated with the Tactical Flight Training Device, a
simulator that has full motion with a wrap around visual system. In a
training scenario the Fuselage Trainer can be integrated with the
simulator to offer real time integrated training for both crews. I was
impressed with the new capabilities of the wrap around visual system
which is superior to the older systems that I trained on. While we all
had a chance to hand fly the simulator I was impressed with how
advanced the simulator visual display is, in which you can pick out
the difference of a deciduous tree to a coniferous tree. Another
upgrade is the Heads Up Display or the HUD in which you look through a
screen in front of the cockpit window which has all the flight
instruments projected onto a small screen. The will allow a pilot the
enhanced situational awareness of looking outside the aircraft and
still have the flight instruments for guidance.

Another training aid was the Integrated Procedures Trainer which
allows maintenance and flight crews to work through issues with system
failures. The other part of this was another full scale fuselage with
partial wing where maintenance can be done on a full scale engine with
propeller as well as part of the wing and fuselage.

The other training module was the Night Vision Integrated Simulator
where observers train in tandem with pilots on detection of missiles ,
anti aircraft artillery aka "Triple A" using the latest in restricted
military use only night vision goggles which cost the price of a mid
sized car.

The tour was highly organized and we were all impressed with high
degree of enthusiastic professionalism displayed by all members of 426
Squadron.

After the visit to 426 Squadron we went to the Yukon Galley for a
light lunch before proceeding to the National Air Force Museum for a
tour of the restoration shops, their displays indoors and outdoors.

The day concluded with dinner at Tomasso's Italian Grill before we
drove back to Ottawa to end the day.

\end{multicols}

\begin{figure*}[ht!]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.75]{trenton2013_group.eps}
   %caption of the figure 
   \caption*{\small \em After Tomasso's -- ready for the drive home.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:rollsroyce}
\end{figure*}

\begin{multicols}{2}

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-07-15 pnsac/jul13/article_SpecialEvents.tex
Changes and images for Jul 2013 newsletter.

--- 2013-04-28 pnsac/apr13/article_RecentEvents.tex
@@
 
-\section{Project North Star Visit to RCAF 426 Squadron ``Thunderbirds''
-  -- Friday Jun 14, 2013'}
-\label{sec:trenton}
-
-Project North Star is pleased to announce a tour of 426 Squadron at
-RCAF Trenton on Friday, June 14th, 2013.  This visit came up at very
-short notice, through the help of Tim Timmins (retired Colonel CAF and
-member of PNSAC) and Lieutenant Colonel Damon Perrault, Commanding
-Officer of 426 Squadron.
-
-Due to the fantastic demand that was generated for our trip to
-Montreal, I would strongly recommend an early booking to avoid
-disappointment as seats are limited for this fun day of activities.
-
-Our visit to 426 Squadron, which does all the training for the new
-C-130J, will involve a tour of their brand new state of the art
-training facilities that includes the two airframes that are used for
-technician training, another airframe for training loadmasters, a
-procedure trainer, and full motion simulators with the usual caveat
-``operational requirements.''
-
-After our visit, we will attend the RCAF Memorial Museum which has
-undergone a major upgrade to its display area along with their work
-shops.
-
-Afterwards we will have a debriefing at Rumour's Restaurant for
-schnitzel before returning to Ottawa.
-
-The cost per person is \$70:00. Please send two cheques including a
-non-refundable deposit of \$30:00 and a post- dated cheque in the
-amount of \$40:00 dated May 14, 2013.
-
-Please note ``Trenton Trip'' on your cheques in addition to sending me
-(Bill Tatae) an e-mail message (\email{pnsac.specialevents@gmail.com})
-to confirm your participation.  Please include your full name and date
-of birth in your e-mail message (this is a requirement of RCAF
-Security) as well as your preferred contact information, email or
-telephone.
-
-Active volunteers can bring your cheques to the PNSAC office at the
-CASM or alternatively mail to:
-
-\address{PNSAC\\
-P.O.Box 44005\\
-514 Montreal Road\\
-Ottawa, ON\\
-K1K 4P8}
-
 \section{Third Annual Golf Day at Loch March -- Wednesday August 14, 2013}
@@
 
-Our Second planned Special Event is for Wednesday August 14th which is
-Third Annual Golf Day at Loch March in Carp Ontario.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=1.5]{loch_march.eps}
+   %caption of the figure 
+   %\caption*{\small \em North Star engine 3.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:logo}
+\end{figure}
 
-The day will start off with breakfast at 09:00 with golf starting
-shortly after 10:00.
+A two week break of the restoration of the aircraft will occur in
+August, as Mike Irvin takes a well-deserved holiday.
 
-The club professional, as last year, has graciously given PNSAC two
-free passes for 18 holes at Loch March, which will be awarded for the
-longest drive (staying in bounds of course) and closest to the
-pin. Format will be best ball.
+On Wednesday, August 14th, PNSAC will have its third annual golf day
+at the beautiful Loch March golf course starting at 10:00 A.M. with a
+nine minute separation between flights.  The format for the 18 holes
+will be a best ball game.
 
-Afterwards on the deck, we will discuss the day's game and the
-Vice-President will give the winner a ``free beer'' for their effort.
+The club house at Loch March has an excellent breakfast menu so to
+make your day more enjoyable it is recommended we meet for breakfast
+before the game at 9:00 A.M.
 
-Please email \email{pnsac.specialevents@gmail.com} to confirm you are coming
-and call the starter at Loch March 613-839-5885.
+Danielle Nadon, the Club Professional at Loch March has graciously
+given us two 18 hole passes for the winner of longest drive and
+closest to the pin.
 
-Please visit the web site
-{\normalfont\color{blue}\texttt{\url{www.lochmarch.com}}} for more
-information such as facilities and dress code.
+You are encouraged to wear your distinctive PNSAC golf shirt and ball
+cap.  As before, guests are welcome.
+
+Please email Danielle at \email{danielle@lochmarch.com} for booking, advising
+her that you are part of the Project North Star Day.  Please remember
+to send a copy of your e-mail to: \email{pnsac.specialevents@gmail.com}.
+
+Please review the web site\\
+{\color{blue}\texttt{\url{http://www.lochmarch.com/about}}} to
+familiarize yourself with directions and dress code restrictions to
+make your day a ``Great Day at Loch March''.
+
+\begin{itemize}
+  \item Seniors mid week rate is \$48:00 and under 60 is \$58:00 plus
+    H.S.T
+  \item Power cart rental is \$15:49 per person for 18 holes plus
+    H.S.T.
+  \item Pull cart rental \$5:31 plus H.S.T.
+\end{itemize}
 

### 2013-07-18 pnsac/jul13/article_Feature1.tex
Final content for July 2013 newsletter.

@@
 
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{bird_gun.eps}
+   %caption of the figure 
+   \caption*{\small \em A 1960s chicken gun as used at Hucknall.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:birdgun}
+\end{figure}
+
 The visits to the various R-R sites around Derby took several months
@@
 after some questioning, I found out that I would be travelling on a
-small airline, called British Midland, that ran a daily service
-between the Derby and Glasgow factories. Far from being in some
-interesting plane connected with R-R, I found that our plane was an
-old DC3 which had seen better days.I duly got on the plane and was
-somewhat nervous flying on my first flight.
+small airline, called British Midland, which ran a service, mainly for
+R-R personnel, between the Derby East Midlands Airport, near Derby and
+Glasgow factories Airport. Far from being in some interesting plane
+connected with R-R, I found that our plane was an old DC3 which had
+seen better days.I duly got on the plane and was somewhat nervous
+flying on my first flight.
 
@@
 
+In the next article I will describe how I worked hard to climb the
+hierarchy at R-R and learnt in an uncomfortable way some lessons on
+people management.
+
 \begin{footnotesize}

### 2013-07-18 pnsac/jul13/article_ProgReport.tex
Final content for July 2013 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Project Manager's Progress Report}
\subtitle{July 2013}
\author{Bruce Gemmill}

\maketitle

Looking back on the last ten years of restoration work, we have
achieved a great deal, but much work still needs to be done.  Our
volunteer workforce has slimmed down somewhat, but those who continue
to work on the North Star have much to be proud of.

\section{Nr 2 Engine}
\label{sec:engines_2}

We reported last year that Engine Nr 2 was finished and installed on
the aircraft in April.  The spinner was repainted after it was
discovered that the paint did not stand up to the harsh outdoor
environment.  The propeller and spinner were installed on the aircraft
before it was moved outside early this year.

\section{Nr 3 Engine}
\label{sec:engines_3}

The third engine was disassembled and most of the major assemblies
were cleaned, restored and re-installed over the past year.  This
included the engine block, pistons, cylinder heads, valve train and
crankcase.  The work has gone very quickly, thanks in part to a well
organized and experienced engine shop crew, led by Garry Dupont.  A
lot of detail work has gone into some small and intricate assemblies
that no one will see once the engine is completely assembled, but that
does not mean we don't take the time to get the job done right.  The
built up engine was recently installed on the engine frame.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{engine3_reduction_gear.eps}
   %caption of the figure 
   \caption*{\small \em Garry Dupont inspecting engine nr 3's
     reduction gearbox.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:engine3gearbox}
\end{figure}

Items still requiring restoration are the supercharger and
intercooler, and the auxiliary drive gearbox.  As always, there are
numerous pipes and hoses and fittings that also need attention.  The
engine should be completed by the end of this year.

\section{Engine Frame}
\label{engineframe}

The engine frame was completely disassembled, stripped of paint and
oil, and then repainted and the main components reassembled on the QEC
stand.  The three radiators (water, oil, and intercooler) were
thoroughly cleaned and then repainted and installed.  The cowl support
frame is currently being worked on, but some parts have been added to
the engine frame, along with the fire detectors and fire suppression
system pipes.  The large chin cowl and some of the lower cowls have
been restored.  The top and side cowls still require extensive
restoration.  This work is somewhat hampered by the small number of
volunteers available to work on these items.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{engine3_prop_cowl.eps}
   %caption of the figure 
   \caption*{\small \em John Thibert with the propeller cowl for
     engine nr 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:engine3propcowl}
\end{figure}

\section{Crew Lounge, Galley,  and Forward Washroom}
\label{crewlounge}

Last spring we removed the equipment from the crew lounge, galley and
forward washroom.  While the aircraft was outside these areas were
cleaned, stripped of paint, then repainted.  The flooring was removed
so the underfloor could be cleaned and primed.  There is some
corrosion damage that must be repaired before new flooring can be
installed.  Most items removed from these areas were restored over the
winter.  This included the washroom door, toilet and washroom
fittings, vanity sink and mirror, and the power inverter installed
opposite the forward washroom.  The old galley was removed and
inspected.  Due to heavy corrosion it was decided to build a
completely new galley, using the old one as a template.  Only the
original doors were retained.  The new galley is complete and will be
stored until it can be installed in the crew lounge.

The heaters had been removed from the ceiling in 2006 and fully
restored, but the air ducts, fuel lines and intake and exhaust scoops
required extensive restoration.  The air scoop was removed from the
aircraft, repaired and re-installed.  The ducts were cleaned, repaired
and new protective heat covers sewn over the old covering.  The water
tank and fittings were removed, cleaned and a new cover sewn over the
tank before installation.  New insulation was installed in the
ceiling, then most heating ducts, cables and pipes were installed.
Several fuel lines had been cut to allow removal of the Janitrol
heaters, and these need to be replaced before the heaters can be
installed.  The doors on the main electrical panel were restored and
installed.  New legends were produced to indicate the circuit breaker
and junction box connections.  These will be laminated and installed
shortly.

\section{Fuselage and Empennage}
\label{empenage}

The rear baggage compartment was cleaned and repainted.  Several floor
panels had stretched and cracked, so new panels were fabricated.  All
panels were stencilled with their respective part numbers before being
installed.  The belly compartment hatch was removed, repaired, painted
and installed.  Recently, the two battery elevators were removed, the
compartment painted, and the elevators will be repainted and installed
shortly.  Batteries were removed when the aircraft arrived at the
museum.  There has been some discussion about obtaining new
(non-functional) batteries to complete this item or work.

Most of the main fuselage was polished while inside the storage
hangar.  The underside and wings still require a lot of work.

Work is progressing on fabricating a complete set of troop seats to
fit up the interior of the main cabin, once this is restored.

\section{Planned Restoration Work--2013}
\label{sec:plannedwork}

Over the next year, we hope to have engine Nr 3 completed and engine
Nr 4 removed.  We will complete the reassembly of the crew lounge and
galley.  We then plan to work on the main cargo compartment, including
refurbishing the main heater duct and other fittings, and removing
floors to begin cleaning and repairs under the cargo floor.  We may
yet get to work on the four engine nacelles.

\section{Membership Report 2012--2013}
\label{sec:membership}

In 2012 we recorded 100 paid members and provided lifetime memberships
to two long serving volunteers and PNSAC Directors, Tim Timmins and
Jim Riddoch.

So far in 2013, we have a total of 66 new or renewed memberships.
This is slightly lower than the same time last year.  Normally, we
would expect to receive a large number of new memberships during
outside displays in the summer, but because of programming changes at
the Museum, our ability to attract new members will be greatly
reduced.

\begin{footnotesize}
  \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-09-19 pnsac/newsletter/article_PresReport.tex
Changed directory name. Primary newsletter files now in newsletter/

--- 2013-07-15 pnsac/jul13/article_PresReport.tex
@@
 the general public.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{canadaday_visitor.eps}
+   %caption of the figure 
+   \caption*{\small \em Showing off the North Star on Canada Day -- an
+   impressed visitor examines the cockpit.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:canadaday}
+\end{figure}
 
@@
 consolidate our position as an important organization working with the
-Aviation Museum
+Aviation Museum.
 

### 2013-10-16 pnsac/newsletter/article_Feature1.tex
Changed directory name. Primary newsletter files now in newsletter/
October 2013 newsletter (DRAFT)

--- 2013-07-15 pnsac/jul13/article_Feature1.tex
@@
 \title{My Days at the Home of the Merlin Engine}
+\subtitle{Part 3}
 \author{Richard Lodge}
@@
 
-As a new employee, part of my Rolls-Royce induction process was to get
-to know some of the production and other R-R facilities around
-Derby. After a few weeks of working in the Overheads Department, it
-was suggested that it was time for me to go visiting. I was happy to
-do this because any opportunity to leave the main accounting office
-for a time was to be jumped at.
+``Lodge, the expenses on the charity wagon go to Miss Prod''. What was a
+charity wagon and who was Miss Prod? I was working in the Overheads
+Department of the Rolls-Royce Aero Engine Division in Derby. As a
+young, ambitious accountant, I did not want to admit to the people in
+the department that I did not know what they were talking about. I was
+conscious that I had spent several weeks of initiation under the
+guidance of the old timer, Les Hart, and should now know something
+about the company. I was sitting in front of a huge leather bound
+ledger where accounting information was recorded. Eventually, by
+listening to other people talking around me, I established that a
+charity wagon was a much prized company car and that Miss Prod was not
+a person but referred to ``miscellaneous production'' work at R-R. Put
+in plain English, the sentence meant that the cost of repairing this
+individual's car was to be charged to a special account in the
+miscellaneous production section of the overheads. There was so much
+to learn!  I was completely bewildered. Each day was a challenge and
+the thought that aero engines were produced in the surrounding
+factories seemed very far away.
 
-Rolls-Royce had an internal bus service to take employees between
-sites. To catch the bus, an employee went down to the door of the
-building and waited for a bus to turn up -- no booking or written
-schedule.  I remember my first trip on a company bus. It took me about
-half an hour of asking people around how to get to the site I
-wanted. I was beginning to realize that much of the operations of R-R
-were not written and were passed around by word of mouth and the long
-time employees passing on information to the newcomers. Years later,
-when working in Northern Canada in native communities I once again
-came across this way of handing down unwritten information from the
-Elders to the younger people. Before starting my job at R-R I had
-believed that the company would be highly organized and disciplined.
+On being placed in the Overheads Department, I was given an internal
+mail address. The company was so large that a worldwide internal mail
+system had been established and I rapidly found out that I was at the
+bottom of the heap. My address was JD/JS/LOD which meant that JS was
+my immediate supervisor and JD was his boss. If you were important in
+the company, your mail would come to you with just your initials such
+as JD. Everybody with any ambition wanted to get a single initials
+mailing address. If you were really close to God, such as a director
+of the company, your mailing address would be Psn for the CEO, Sir
+Denning Pearson, or DPH for David Huddie, the Director of Engineering.
 
-I finally managed to find my way to the building I was due to visit.
-It was the Foundry. I was introduced to the men who worked in this
-evil smelling place. Each of the major engine parts was individually
-cast -- there was no automation and a good foundry man could put
-together the sand and other parts needed for the casting, mostly from
-memory. During WW2 the foundry had concentrated entirely on the
-production of military engines, mostly Merlins.  By the time I was
-shown round the place there was a bewildering conglomeration of parts
-being made for piston, turbo prop and jet engines.
+There were continuous battles between the engine development engineers
+and the accountants. When the engineers spent money, they tried
+wherever possible to avoid having it charged against their
+budget. Their favourite game was to charge the cost into the overheads
+where it would be lost in a mass of other expenditure such as emery
+paper and toilet rolls. Our task was to catch them at their game and
+reallocate the expense back to them. The development engineers were
+very skilled at this part of their work and became so good at it that
+they were able to hide major cost overruns, sometimes
+indefinitely. Little did I realize that I was helping to perpetuate
+the culture that was one of the causes of R-R going bust a few years
+later.
 
-As a young accountant I was expected to understand how the paperwork
-and accounting for all this worked. It was a daunting thought and I
-left the Foundry feeling overwhelmed.  Throughout my time at R-R I
-never became comfortable dealing with the Foundry -- everything seemed
-to be so imprecise and up to the skill on the foundry man. Each time I
-went there I tried to find out how the men knew how to produce the
-same casting every time and the answer was always the same: "Oh, we
-just know what to do".  Many of these comments were probably made to
-try to make a young guy in a suit from the office look foolish but
-they were also largely true!
+In the 1960s R-R was a paternalistic organization, steeped in
+tradition, still operating with many cost-plus government contracts,
+vast amounts of paper and virtually no computers. The Ministry of
+Aviation would regularly send auditors into the company to check how
+public money was being spent. These gentlemen had a permanent office
+allocated to them and would regularly attack the expenses recorded by
+the Overheads Department. When this happened, we and the development
+engineers became great buddies for a short time. Our unspoken job was
+to confuse the government auditors as much as possible so that they
+would not report back to their masters that things were not as they
+should be. We became quite good at our confusion duties. It was not a
+difficult job because we were frequently confused by all the paper and
+numbers as well.
 
-Sitting in the Overheads Department I saw many of the invoices for
-purchases. I could hardly believe it when I saw invoices going through
-for payment for dozens of chickens. The delivery was never to the
-works canteen but always to the R-R site at Hucknall, a short distance
-outside Derby. My supervisor told me that the chickens were fired at
-engines on test. I did not believe him until I managed to find a
-pretext to visit Hucknall. After another challenge finding the works
-bus to get me there, I arrived at the site and was shown round and all
-was revealed. Dead chickens ending their time on earth being used to
-test the strength of compressor blades in an engine on full power
-seemed to me to be a very imprecise way of testing a jet
-engine. Another lesson learned -- not everybody was trying to make me
-look foolish all the time.
+R-R continued to deliver new aero engines to customers and I gradually
+started to understand what I was doing. After a few months, my
+supervisor was promoted and I was appointed supervisor of the
+Overheads Department, much to the disgust of many of the people in the
+department. Although my pleasure was to be short-lived, I was
+delighted to get this promotion. By the time of my promotion, I had
+been around many of the facilities of R-R both in the Derby area and
+in Scotland. I felt that I knew my way around the company.
 
-The visits to the various R-R sites around Derby took several months
-to complete. I had to have a good reason to leave my office and go
-visiting, particularly as I was now working in a departmental
-position. Eventually another bus trip to the large new complex at
-Sinfin was required or I somehow managed to talk my way into finding
-it necessary to go there. Sinfin, among many other things, housed the
-main engine test beds and I was curious to find out what happened
-there. Unlike the evil smelling Foundry, Sinfin was clean, new and had
-plenty of windows.
+Almost everyone working in the Overheads Department, including the
+three section leaders under me, were older and more experienced than
+I. They did not like this young upstart accountant becoming their
+boss. Underground warfare started between us. Several of the men had
+fought through WW2 and were well practiced at the gentle art of ``dumb
+insolence''. The three section leaders came to me one day and told me
+they really disliked me being their boss. This was not good for my
+ego. I suggested to them that the four of us should go out after work
+and have a beer and that on this occasion I would just be one of them
+and not their boss. I bought the first round of beer and they really
+tore into me. After several more rounds of beer, we agreed to a truce
+and decided on a new strategy. I would stop trying to micromanage them
+and they would stop their dumb insolence. Of course, they never
+admitted that they were practicing dumb insolence. Life became much
+better after our beers. Miss Prods, charity wagons, drop-offs and all
+the other R-R jargon became normal for me with the active co-operation
+of my three section leaders.
 
-The men in the test shop told me about the engine testing process. I
-was still new to R-R and had a lot to learn. After a while I was asked
-if I would like to see an engine on test. Of course I said yes; it
-seemed interesting. With little experience I did not realize that an
-aero engine on test is nothing much to see. One of the men said he
-would take me to a Conway engine on test. He said he would open the
-test bed door to let me see the engine and that I would find it
-somewhat noisy. He stood behind the door, opened it and let me look
-inside at the engine on full power. I now wear hearing aids. Health
-and safety were two words not used in those days when many of the men
-were used to serving in the war. This was 1960s Britain.
+Apart from the pay rise upon my promotion I had now moved up in the
+company mailing system and my address had become JD/LOD. This probably
+helped my head to swell until the three section leaders brought me
+rapidly back to earth.
 
-I have always enjoyed travelling and would jump at any opportunity to
-travel to a new place. After a few months it was suggested to me that
-I should visit the Scottish factory at Hillington near Glasgow. I
-asked how I should get there. ``Oh. You might as well fly. There is a
-flight every day that leaves in the morning and comes back in the
-afternoon.'' I had never flown before.
-
-The UK being a small country, everybody went around by train.  As
-usual, I was given very little information and told to get myself to
-East Midlands airport early on the appointed morning, where a ticket
-would be waiting for me. On this occasion I had the good sense to take
-my own car to the airport and did not get lost. I expected to find a
-plane waiting, painted in some sort of Rolls-Royce colours. Instead,
-after some questioning, I found out that I would be travelling on a
-small airline, called British Midland, that ran a daily service
-between the Derby and Glasgow factories. Far from being in some
-interesting plane connected with R-R, I found that our plane was an
-old DC3 which had seen better days.I duly got on the plane and was
-somewhat nervous flying on my first flight.
-
-An hour later we arrived in Scotland, also my first trip over the
-Border, and I spent the day being shown around the factories and
-offices.  I remember very little about the day but vividly remember
-the flight back to Derby. It must have been in the winter because it
-was getting dark as I boarded the plane around 5.00 pm. Seating was
-first come first served and all the old sweats travelling knew what to
-do. I ended up on the back row near the cargo area. The main thing I
-remember was the crew having a fierce argument about the load on the
-plane which was full. The discussion centred around the amount of
-cargo which could be taken on the flight. I was quite relieved to get
-back to Derby in one piece.
+I must have learnt a few of life's lessons about people management
+because when I was again promoted, my three section leaders tried to
+encourage me to refuse the promotion. No chance. I was once again
+moving up the mailing system and would in future have my mail
+addressed to LOD, without a more senior person's initials appearing
+before mine. I was going in the right direction.
 

### 2013-10-16 pnsac/newsletter/article_PresReport.tex
October 2013 newsletter (DRAFT)

--- 2013-07-15 pnsac/jul13/article_PresReport.tex
@@
 
-Canada Day is now behind us.  We had an excellent day with a good
-number of our volunteers showing off the North Star to the general
-public. Judging by the activity around the plane all the day, we have
-now developed a display format which is interesting to visitors.
-Canada Day was the only day this year when the plane was on display to
-the general public.
+During the past three months we have moved steadily forward with the
+restoration of the North Star.  Our progress has been somewhat held up
+by lack of available volunteers.
 
-On June 27, in co-operation with the Museum, who hosted the event, we
-showed off the plane to a group of donors attending the Donor
-Appreciation Evening.  This was a successful evening which is likely
-to be repeated.
+Our Association has two distinct and very important sides.  First and
+foremost we are a group whose interest is in the North Star and the
+restoration and preservation of the plane at the Museum.  Secondly,
+over the years, we have developed a strong social side to the
+Association where volunteers and other members can join in interesting
+events centered around aviation. Elsewhere in this issue of the
+Chronicle you will find a description by Bill Tate our Vice President,
+of the very successful trip on October 4 to the Navcan Air Control
+Centre and to the Bombardier factory in Montr\'{e}al.
 
-In my last Notes from the President, I referred to plans for
-celebrating the 10th anniversary of the founding of the Project North
-Star Association.  This should have taken place around June 24.
-Unfortunately, the event passed without us being able mark the
-day. This was mainly due to situations beyond our control.
+After 10 years of restoration work on the North Star we recognize that
+we must review our progress and develop a plan for the rest of the
+work on the plane.  There will be three aspects to this review.  Our
+first task will be to identify the major sections of the project and
+develop ways of actually carrying out the restoration.  The
+undercarriage of the plane is a particularly complicated challenge.
+We then need to prepare a plan with projected timelines for completing
+each part of the work.  Our third task will be to estimate costs and
+to develop ideas for financing it.  During this process we must also
+ensure that we will have sufficient qualified volunteers to get it
+done.
 
-Our 2013 AGM was held on June 8, and all the existing directors were
-re-elected.  This will enable the directors, who have now become
-experienced, to concentrate on long-term matters.
+Before the end of November we are hoping to have meetings with the
+senior management of the Museum to update planning of all the tasks
+above.  The end result of this should be a documented project plan and
+a new Memorandum of Understanding between the Canada Aviation and
+Space Museum and the Project North Star Association of Canada.
 
-The actual work on the restoration of the plane is going very well, as
-reported elsewhere in this newsletter. The major issue facing the
-Board in the next year will be to consider how to secure the future of
-PNSAC, both in terms of volunteers and the funding of restoration
-work.
+At times our progress on the restoration of the North Star may appear
+to Association members to be somewhat slow.  Unlike other
+organizations in Canada such as Vintage Wings in Gatineau or the
+Canada Warplane Heritage Museum in Hamilton, PNSAC cannot make
+decisions about the future of the North Star on its own.  As an
+Association working with a Federal Government institution we are often
+affected by decisions not related to the work we do.  This can at
+times provide interesting challenges.
 
-We have now reached a point where it is necessary to update our
-Memorandum of Understanding (MOU) with the Museum and our work plan
-for completing the restoration of the plane.  Our volunteers have
-already worked over 60,000 hours on the project and we expect to work
-a similar number of hours before the restoration is completed.
-
-There have been several informal meetings to discuss our future
-operations and the opportunities available for new volunteers.  Within
-the next few weeks we are expecting to meet with the Director General
-of the Aviation Museum and members of his staff to work on the MOU
-update.
-
-Later this month we are meeting with senior management of the Museum
-Foundation, the body charged with corporate fundraising for the Canada
-Science and Technology Museums Corporation, to discuss how PNSAC can
-work with the Foundation to raise funds for the North Star restoration
-and the museums in general.
-
-With the initiatives above, we are hoping in the next few months to
-consolidate our position as an important organization working with the
-Aviation Museum
+In conclusion, I would say that the Directors are working hard with
+the Museum management to move our project forward even though at times
+it may appear to be very difficult.  I hope that I will be able to
+report significant progress in the next issue of the Chronicle.  In
+the meantime we have some interesting events in the planning stages on
+the social side of the Association.  Stay tuned.
 

### 2013-10-16 pnsac/newsletter/article_ProgReport.tex
October 2013 newsletter (DRAFT)

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Project Manager's Progress Report}
\subtitle{October 2013}
\author{Bruce Gemmill}

\maketitle

\section{Nr 3 Engine}
\label{sec:engines_3}

Engine 3 is almost completely reassembled, with only the supercharger
and intercooler preheat assemblies left to install on the engine,
which was moved to the QEC engine frame in early summer.  The work has
slowed somewhat, due to summer down time for staff and volunteers, but
we continue to make steady progress.

The auxiliary gearbox was recently removed from the engine firewall,
and is now being disassembled.  This is the last major assembly that
must be overhauled before the engine can be installed on the aircraft.

\section{Engine Frame}
\label{engineframe}

Now that the engine and frame are joined, the major work still
required is rebuilding the many cowl panels and the radiator flaps
that surround the engine.  There is a lot of corrosion on the pieces,
and each must be disassembled, cleaned, then repaired before painting
and reassembly.  This work must be done carefully to ensure the cowl
panels fit properly when installed on the QEC.

This work has gone slowly due to the small number of volunteers who
are able to do this type of repair.

\section{Crew Lounge, Galley,  and Forward Washroom}
\label{crewlounge}

The new galley has now been installed in the crew lounge, and work
overhead in the ceiling is also complete.  This involved installing
new insulation and the Janitrol heaters, along with refurbished
heating ducts, and some new fuel lines.  The ceiling panels were also
installed.

The doors on the main electrical panel were restored and installed.
New circuit legend cards were produced to indicate the circuit breaker
and junction box connections.  These were laminated and installed
inside the electrical panels.  Some additional investigation is needed
to complete the electrical circuit connection diagrams that were badly
faded.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{galley_interior.eps}
   %caption of the figure 
   \caption*{\small \em New galley -- interior view.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:galley_interior}
\end{figure}

One section of floor was patched after removing a corroded area.  One
additional repair needs to be done under the floor of the crew lounge.
Until this is complete, the new cushion flooring cannot be
installed. This has also delayed the installation of all of the crew
lounge accessories, such as the table, seats and secure storage bin,
all of which are complete but waiting in storage.

\section{Fuselage and Empennage}
\label{empenage}

The forward and rear belly compartments are now complete.  The two
battery elevators that were removed have been restored and installed
behind the nose wheel.  The battery compartment covers were repaired
but need painting and stencilling.

The firewall on nacelle 3 is being cleaned in preparation for the
installation of the engine sometime this winter.

Further progress has been made on the set of troop seats being
fabricated for the main cabin.

\section{Planned Restoration Work 2013/14}
\label{sec:plannedwork}

Over the next year, we hope to have engine Nr 3 completed and engine
Nr 4 removed.  We will complete the reassembly of the crew lounge and
galley.  We then plan to work on the main cargo compartment, including
refurbishing the main heater duct and other fittings, and removing
floors to begin cleaning and repairs under the cargo floor.  We may
yet get to work on the four engine nacelles.

\begin{footnotesize}
  \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-10-16 pnsac/newsletter/article_Special.tex
October 2013 newsletter (DRAFT)

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{PNSAC Merchandise}
%% \author{Drew Hodge}

\maketitle

%\end{multicols}

\begin{figure}[ht!]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=1.0]{fleece_YOW7578.eps}
   %caption of the figure 
   \caption*{\small \em PNSAC winter apparel}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:merchandise}
\end{figure}

The first frost warnings on the Weather Channel are reminders that
fall and winter are just around the corner. Keep warm on those frosty
fall days with a PNSAC toque and sweatshirt. We can also take
individual orders for the micro fleece vests shown in the photo on the
left, and if you're looking for early Christmas presents, how about a
coffee mug or a t-shirt?

Have a look at our Facebook page,
{\normalfont\color{blue}\texttt{\href{https://www.facebook.com/media/set/?set=a.610706705637878.1073741851.284829008225651&type=1}{PNSAC
      Merchandise}}}, to see what we're offering, and place your
orders by sending an email message to the following address
\email{pnsac.merchandise@gmail.com}.

For members in the Ottawa-Gatineau region, you can arrange to pick up
your merchandise at the Canada Aviation and Space museum any Monday or
Tuesday, between 10:00 AM and 3:00 PM. We can also ship your purchases
to you for an extra fee. Please specify museum pickup or shipping with
your order.

Merchandise is also available during our Quarterly Meetings
at CAVM.

Look forward to hearing from you.

``The Merchandise Committee''

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

%\begin{multicols}{2}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-10-16 pnsac/newsletter/article_SpecialEvents.tex
October 2013 newsletter (DRAFT)

% Template PNSAC newsletter
% Language: Latex

\title{Recent Special Events}
%\subtitle{Visit to Nav Canada Montreal A.C.C. and Bombardier}
\author{Bill Tate}

\maketitle

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=1.5]{loch_march.eps}
%    %caption of the figure 
%    %\caption*{\small \em North Star engine 3.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:logo}
% \end{figure}

\section{Loch March Golf Day}
\label{golf}

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=1.5]{loch_march.eps}
   %caption of the figure 
   %\caption*{\small \em North Star engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:logo}
\end{figure}

The PNSAC 2013 Annual Golf Day at Loch March started with breakfast in
the Club House, 14 members and friends in attendance.

Prior to breakfast, start times were arranged in such a way that there
was a mixing of participants, so that those who were not aware of what
PNSAC does had the opportunity to talk to members involved in the
project.

The first group started out in light rain showers and thirty minutes
later the skies cleared for a bright sunny day. With the extra
rainfall this year the course was found to be in a very lush
condition, and as usual for Loch March the greens were very fast.

Danielle Nadon the Club Professional graciously gave PNSAC two 18 hole
certificates as prizes, one for the closest to the pin and the other
for the longest drive, which were both won by Gary Hall. He also came
in with low gross for the day at 79. Our own Ron Lemieux was paired
with Gary and in Ron's own words ``I had nothing to do with the score;
I was riding on Gary's coat tails''.

Loch March is a very wonderful facility for golf and has an excellent
club house as well. The course can be challenging, but there is
something for everyone who plays golf. Even if you do not golf, it is
well worth the drive to enjoy the patio. For details please visit the
web site: {\normalfont\color{blue}\texttt{\url{http://www.lochmarch.com}}}

This year we had another prize which was given to us by New York New
York Hair Salon, - a gift certificate of \$100:00 for hair
services. So a new category of winner had to be chosen, which was for
the best dressed in our distinctive PNSAC clothing. To ensure fairness
in the award I had Alfredo Porco (who has been my barber for 30 years)
make the award. It went to our "special seamstress" Karen Lochhead.  I
should mention there are now three generations of Porcos at New York
New York, - Alfredo, his son Vince and Vince's daughter Marina.

For information on New York New York please visit their web site:
 {\normalfont\color{blue}\texttt{\url{http://www.nynyhairsalon.com}}}

Looking forward to our fourth golf day next year, we will work on more
prizes for our members and guests in 2014.

\section{Trip to the Montreal Area Control Centre and Bombardier Aerospace}
\label{yultrip}

On Friday October 4th, 33 members and guests departed the Canada
Aviation Space Museum (CASM) for another planned special event. First
off, special thanks to Bryon Mask (retired Air Canada Captain), one of
our members who assisted me with contact information at Nav Canada
which was required to have our visit approved. Bryon is a highly
respected individual in Flight Safety and Accident Investigation in
Canada.

We were met and welcomed by Richard Snider, Duty Manager at the
Montreal ACC. We visited the Training Centre where training is
provided to new Air Traffic Controllers and required recurrent
training is offered. We were also shown the tower simulator which has
a day/night 360 degree display where various airport scenarios can be
brought up to assist in training. My personal favourite was the
display showing a landing as seen from the cockpit, which even showed
it as a 360 degree display (first time I have experienced the rear
view mirror look for a landing).

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{yul_tca3.eps}
   %caption of the figure 
   \caption*{\small \em In the 360 degree tower simulator.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:tca3}
\end{figure}

Our next stop was the Technical Operations Centre where staff monitor
all radar and computers. ACC has dual systems with one system picking
up automatically in case the other system fails. Communications has a
dual back-up digital and analogue system.  Communications to aircraft
are by conventional VHF with some interesting back ups such as
Controller Pilot Data Link Communications (CPDLC). The easiest way of
explaining this is Text Messaging Sat Com in which either the aircraft
can call the phone number of the ACC or the ACC can call the dedicated
number for the aircraft (\$6:00 a minute cost) and as a back up by HF
radio through a Flight Service Station.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{yul_tca2.eps}
   %caption of the figure 
   \caption*{\small \em Air traffic controllers at work.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:tca2}
\end{figure}

In the Operations Room, we could hear some controllers talking very
quietly. Radar displays offer a great deal of information to the
Controller, which also helps keep talking to a minimum. A sense of
humour prevailed at the work station for Northern Quebec up to
Iqaluit, where a Motel 6 logo was mounted over the work station.

In total we were on site for close to 2 1/2 hours which I found to be
a fantastic experience, seeing the ATC environment from the other
side.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{yul_tca1.eps}
   %caption of the figure 
   \caption*{\small \em The Operations Room.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:tca1}
\end{figure}

With kind assistance in the form of a recommendation from Mae, my YUL
expert, we broke off for lunch at Le D\'{e}jeuner Cosmopolitain which
handled our group in a very friendly efficient manner. I should note,
this is the spot where Air Canada retirees meet for their monthly
luncheon dates.

After lunch we proceeded to the Bombardier Production plant where we
were met by three very knowledgeable tour guides. They were Alex
Ritchie, Fernand Richer, and Cristian Racine who together gave a very
informed hour and half tour of the production line. The tour was
divided into two areas, assembly and finishing. At the beginning of
the assembly we saw where the fuselage and wings are mated and
proceeded on to where systems are tested such as hydraulic systems,
then to the finishing area where interiors are installed. It was
something to see how clean and bright this plant is and how the many
sub assemblies come together. The tail assembly built in China comes
to the production line and meets with part of the fuselage built in
Mexico, just in time for the assembly.

We ended our visit in the delivery room where customers pick up their
aircraft all ready to go, all yours for approximately \$30 million
dollars.

We enjoyed an excellent dinner at the Auberge Willow Inn prior to our
return to the CASM at the end of a very enjoyable 14 1/2 hour day
together.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-10-17 pnsac/newsletter/article_Feature2.tex
Newsletter: final changes incorporated.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Building troop seats for North Star 17515}

\author{Rolf Geiger}

\maketitle

There are no original parts or drawings for the seats. Fortunately,
the tubing and brackets that connect the seats to the aircraft wall
were still in place in the aircraft. Also, two lower tubing support
brackets were found hiding behind the galley, allowing us to produce
exact copies of the original. Everything else concerning the seats is
missing and needs to  be replaced.

\end{multicols}

%\begin{figure*}[htbp]
\begin{figure*}[h]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.75]{ass.eps}
   %caption of the figure 
   \caption*{\small \em Troop seat construction detail.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:ass}
\end{figure*}

\begin{multicols}{2}

To find the size and the number of seats to be produced we measured
the space between the anchoring studs in the aircraft floor, which the
legs snap onto. This gave us the width of the seats. Thus, we will
produce (5) three-seaters plus (6) four-seaters, providing seating for
39 troops. No litters are planned.

Mike Irvin (our Museum coordinator) decided to use seats from a
helicopter in the Museum collection as an example for our seats. We
are more or less copying the parts of these seats, considering that we
have to adapt our methods to the equipment available to us.

To start the project, drawings of the parts were made. The list of
materials of these parts calls mainly for aluminum bar stock, aluminum
tubing, steel tubing, aluminum sheet and steel sheet.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{rolf.eps}
   %caption of the figure 
   \caption*{\small \em Rolf Geiger at work on the troop seats.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:rolf}
\end{figure}

Thus, we need to produce (15) different metal parts, including: lower
tubing plus support brackets and clamps, spreader bars with two shell
clamps each to clamp onto the lower support tubing plus the front
brackets holding the spreader bars to the frontal support tubes, leg
brackets connecting the legs to the frontal support tubes plus
secondary brackets (slipping over the leg brackets) holding a spring
loaded pin with which the seat fabric will be tautened, legs,
including the lower clamping mechanism fastening the legs to the
floor. Most parts need (22) of each, except the shell clamps for the
spreader bars (44), hooks (117), leg grippers (44).

The procedure in the production of these parts is as with any project:
Take the finished product and then work backwards to determine the
production stages.

In the case of a formed sheet metal part a drawing of a flattened
blank is produced by calculating bend radii and adding that to the
straight surfaces (in effect un-bending or stretching the finished
part).

Experimental operations may be needed to find the shape and dimensions
of the blanks.  Tooling may be needed to produce the formed sheet
metal parts.  The blanks are made by shearing and sawing, according
to the parts drawing.

In the case of a machined part, it is usually produced from bar stock:
Parts are cut to length and subsequently machined as per
drawing. Partial in-between drawings may be necessary to produce the
parts in progressive stages.

To avoid the higher costs associated with surface plating the metal
parts, (as per original parts) we chose painting them instead, which
will be acceptable given the limited use of the finished product.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-10-17 pnsac/newsletter/article_Special.tex
Newsletter: final changes incorporated.

@@
 
-Have a look at our Facebook page,
+See our Facebook page,
 {\normalfont\color{blue}\texttt{\href{https://www.facebook.com/media/set/?set=a.610706705637878.1073741851.284829008225651&type=1}{PNSAC
-      Merchandise}}}, to see what we're offering, and place your
+      Merchandise}}}, for details about what we're offering, and place your
 orders by sending an email message to the following address
@@
 For members in the Ottawa-Gatineau region, you can arrange to pick up
-your merchandise at the Canada Aviation and Space museum any Monday or
+your merchandise at the Canada Aviation and Space Museum any Monday or
 Tuesday, between 10:00 AM and 3:00 PM. We can also ship your purchases
@@
 Merchandise is also available during our Quarterly Meetings
-at CAVM.
+at CASM.
 

### 2013-12-14 pnsac/newsletter/article_EditorsNotes.tex
Preparation for December 2013 newsletter.

% Template PNSAC newsletter - Article % Language: Latex
%

% Head

\title{Editor's Notes} 
\author{Bruce Grant}

\maketitle

\end{multicols}

This issue of the Chronicle carries the last of four chapters in
Richard Lodge's story of his time at Rolls-Royce. Taken together, they
make an interesting ethnology of an extinct industrial culture. The
whole series is accessible through our website,
{\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}} (note that
the location of the neswletter archive will change when the new
website goes live) starting with the April 2013 issue of the
Chronicle.

We also have another hair-raising flying story from Tim Timmins. A
series of previous Timmins stories can be found through the same
website, starting with the December 2010 issue.

Motorheads among our readers should look into Ted Devey's extensive
writings on the Rolls-Royce Merlin, starting in the December 2006
issue.

All together, the archived issues are a valuable web resource for
readers with an interest in aviation history or the history of Project
North Star.

\begin{footnotesize} \raggedleft PNSAC\\
\end{footnotesize}

\begin{multicols}{2}

% End of text.

%%% Local Variables: %%% mode: latex %%% TeX-master: main_document.tex
%%% End:

### 2013-12-14 pnsac/newsletter/article_Feature1.tex
Preparation for December 2013 newsletter.

@@
 \title{My Days at the Home of the Merlin Engine}
-\subtitle{Part 3}
+\subtitle{Part 4}
 \author{Richard Lodge}
@@
 
-``Lodge, the expenses on the charity wagon go to Miss Prod''. What was a
-charity wagon and who was Miss Prod? I was working in the Overheads
-Department of the Rolls-Royce Aero Engine Division in Derby. As a
-young, ambitious accountant, I did not want to admit to the people in
-the department that I did not know what they were talking about. I was
-conscious that I had spent several weeks of initiation under the
-guidance of the old timer, Les Hart, and should now know something
-about the company. I was sitting in front of a huge leather bound
-ledger where accounting information was recorded. Eventually, by
-listening to other people talking around me, I established that a
-charity wagon was a much prized company car and that Miss Prod was not
-a person but referred to ``miscellaneous production'' work at R-R. Put
-in plain English, the sentence meant that the cost of repairing this
-individual's car was to be charged to a special account in the
-miscellaneous production section of the overheads. There was so much
-to learn!  I was completely bewildered. Each day was a challenge and
-the thought that aero engines were produced in the surrounding
-factories seemed very far away.
+One of the most mysterious departments at Rolls-Royce was the
+Technical Publications Department. Word processing was unheard of in
+the 1960s and every time an engine modification was made, it resulted
+in much work documenting even a small change and distributing the
+updated documentation to all engine users around the world. I never
+spent any time working with the people in this department even though
+there was a complete building devoted to the work. Compared with the
+resources allocated to engine development, the technical publication
+side of the business was very small and it always remained a mystery
+to me.
 
-On being placed in the Overheads Department, I was given an internal
-mail address. The company was so large that a worldwide internal mail
-system had been established and I rapidly found out that I was at the
-bottom of the heap. My address was JD/JS/LOD which meant that JS was
-my immediate supervisor and JD was his boss. If you were important in
-the company, your mail would come to you with just your initials such
-as JD. Everybody with any ambition wanted to get a single initials
-mailing address. If you were really close to God, such as a director
-of the company, your mailing address would be Psn for the CEO, Sir
-Denning Pearson, or DPH for David Huddie, the Director of Engineering.
+Since retiring from my career as an accountant in Canada, I have
+become active on the actual restoration work of the North
+Star. Understandably, I have a special interest in working on the
+Merlin engines. Nobody now working on the restoration has ever worked
+on a Merlin when it was in service. It is therefore necessary at all
+times to refer to the various publications and manuals we have managed
+to accumulate for the North Star.
 
-There were continuous battles between the engine development engineers
-and the accountants. When the engineers spent money, they tried
-wherever possible to avoid having it charged against their
-budget. Their favourite game was to charge the cost into the overheads
-where it would be lost in a mass of other expenditure such as emery
-paper and toilet rolls. Our task was to catch them at their game and
-reallocate the expense back to them. The development engineers were
-very skilled at this part of their work and became so good at it that
-they were able to hide major cost overruns, sometimes
-indefinitely. Little did I realize that I was helping to perpetuate
-the culture that was one of the causes of R-R going bust a few years
-later.
+It is only now, actually working on the engine and referring
+frequently to the exploded diagrams that I realize how complicated the
+work of the Technical Publications Department must have been. The men
+and women who worked there were largely unsung heroes in that it was
+very tedious and unglamorous work which was of great importance for
+the safe and trouble-free running of engines in service. I can
+remember that I would hear from time to time of the difficulties of
+ensuring that publication updates were actually delivered to many
+remote places in the world and the considerable logistical
+complications of ensuring that the packages arrived safely and
+furthermore that the updates were actually inserted in the relevant
+manuals, particularly as many of the engine users did not speak
+English. I cannot remember now if any of the documentation was
+translated into other languages although I have a recollection that we
+employed people who could translate into French and Spanish.
 
-In the 1960s R-R was a paternalistic organization, steeped in
-tradition, still operating with many cost-plus government contracts,
-vast amounts of paper and virtually no computers. The Ministry of
-Aviation would regularly send auditors into the company to check how
-public money was being spent. These gentlemen had a permanent office
-allocated to them and would regularly attack the expenses recorded by
-the Overheads Department. When this happened, we and the development
-engineers became great buddies for a short time. Our unspoken job was
-to confuse the government auditors as much as possible so that they
-would not report back to their masters that things were not as they
-should be. We became quite good at our confusion duties. It was not a
-difficult job because we were frequently confused by all the paper and
-numbers as well.
+After a couple of years I was given a new job in the Long Range
+Planning Department. I was one of three accountants working in this
+office under a very senior accountant. Here we looked at the costings
+of the development work, the projected sales  and production costs on
+new engine designs. This was the time when the Spey was going into
+service and the the company had signed a fixed price contract for the
+RB211 engines for the new Lockheed 1011. This was a major project,
+championed by David Huddie the Chief Engineer. Computer spreadsheets
+did not exist at that time - pencil and columnar paper were the order
+of the day, together with efficient erasers. Long range forecasts of
+25 years required many yearly columns and each time we produced
+figures that the engineers did not like, we were told to rework the
+numbers until we produced an acceptable answer. We were not limited in
+the number of erasers we could use. The usual way of cutting costs was
+to reduce the forecast development and testing time. Of course this
+did not work in real life.
 
-R-R continued to deliver new aero engines to customers and I gradually
-started to understand what I was doing. After a few months, my
-supervisor was promoted and I was appointed supervisor of the
-Overheads Department, much to the disgust of many of the people in the
-department. Although my pleasure was to be short-lived, I was
-delighted to get this promotion. By the time of my promotion, I had
-been around many of the facilities of R-R both in the Derby area and
-in Scotland. I felt that I knew my way around the company.
+We were often dealing with the directors of R-R who lived on the top
+floor of the Nightingale Road offices. Many of them were engineers and
+they had little time for young bean counters. It was still a time of
+optimism for the company. The huge contract for the Lockheed engines
+had been signed and the company was still enjoying its reputation as
+the company whose Merlin engines had saved Britain in the air
+war. This was in 1965/6 and it was several years before our
+projections of doom resulted in the bankruptcy of the
+company. Fortunately I left the company in 1966 and was not involved
+in what must have been a horrible process as a symbol of British
+engineering excellence slowly went to the wall.
 
-Almost everyone working in the Overheads Department, including the
-three section leaders under me, were older and more experienced than
-I. They did not like this young upstart accountant becoming their
-boss. Underground warfare started between us. Several of the men had
-fought through WW2 and were well practiced at the gentle art of ``dumb
-insolence''. The three section leaders came to me one day and told me
-they really disliked me being their boss. This was not good for my
-ego. I suggested to them that the four of us should go out after work
-and have a beer and that on this occasion I would just be one of them
-and not their boss. I bought the first round of beer and they really
-tore into me. After several more rounds of beer, we agreed to a truce
-and decided on a new strategy. I would stop trying to micromanage them
-and they would stop their dumb insolence. Of course, they never
-admitted that they were practicing dumb insolence. Life became much
-better after our beers. Miss Prods, charity wagons, drop-offs and all
-the other R-R jargon became normal for me with the active co-operation
-of my three section leaders.
-
-Apart from the pay rise upon my promotion I had now moved up in the
-company mailing system and my address had become JD/LOD. This probably
-helped my head to swell until the three section leaders brought me
-rapidly back to earth.
-
-I must have learnt a few of life's lessons about people management
-because when I was again promoted, my three section leaders tried to
-encourage me to refuse the promotion. No chance. I was once again
-moving up the mailing system and would in future have my mail
-addressed to LOD, without a more senior person's initials appearing
-before mine. I was going in the right direction.
+After the chaos caused by the demise of Rolls-Royce Limited, the
+British government formed a new nationalized company to take over the
+remains of the old one. This has been very much a success story and
+the newly privatized Rolls-Royce is now a highly profitable and
+innovative company. I look back on my days at R-R with nostalgia but I
+now realize that much of the time we were all living in a fool's
+paradise.
 

### 2013-12-14 pnsac/newsletter/article_Feature2.tex
Removed text from article_Feature2.tex to match wrtier's original content.
Preparation for December 2013 newsletter.

@@
 
-\title{Building troop seats for North Star 17515}
-
-\author{Rolf Geiger}
+\title{RCAF Overseas Ferry Operations}
+\author{AJS (Tim) Timmins, RAD/NAV/AI}
 
@@
 
-There are no original parts or drawings for the seats. Fortunately,
-the tubing and brackets that connect the seats to the aircraft wall
-were still in place in the aircraft. Also, two lower tubing support
-brackets were found hiding behind the galley, allowing us to produce
-exact copies of the original. Everything else concerning the seats is
-missing and needs to  be replaced.
-
-\end{multicols}
-
-%\begin{figure*}[htbp]
-\begin{figure*}[h]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.75]{ass.eps}
-   %caption of the figure 
-   \caption*{\small \em Troop seat construction detail.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:ass}
-\end{figure*}
-
-\begin{multicols}{2}
-
-To find the size and the number of seats to be produced we measured
-the space between the anchoring studs in the aircraft floor, which the
-legs snap onto. This gave us the width of the seats. Thus, we will
-produce (5) three-seaters plus (6) four-seaters, providing seating for
-39 troops. No litters are planned.
-
-Mike Irvin (our Museum coordinator) decided to use seats from a
-helicopter in the Museum collection as an example for our seats. We
-are more or less copying the parts of these seats, considering that we
-have to adapt our methods to the equipment available to us.
-
-To start the project, drawings of the parts were made. The list of
-materials of these parts calls mainly for aluminum bar stock, aluminum
-tubing, steel tubing, aluminum sheet and steel sheet.
+Over a three year period starting in 1952, with Operation Leap Frog,
+the deployment of 12 Sabre squadrons from Canada to Air Division in
+Europe, No 1 Overseas Ferry Unit (1OFU), based at St Hubert, Quebec,
+delivered over 800 aircraft. 426 Squadron North Stars flew support
+missions, carrying servicing personnel and providing enroute
+navigation assistance, airborne radio beacons known as duck
+butts. Besides the Sabre squadrons, four CF100 Squadrons were
+delivered to Air Division and 53 CF100 aircraft to the Belgium Air
+Force. Silver Stars and Expeditors were flown to Europe for delivery
+to NATO members under the Mutual Aid agreement. My personal experience
+with ferry operations was with Beechflight, the delivery of 25 C45s
+(Expeditors) to France and Portugal.
 
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{rolf.eps}
+   \includegraphics[scale=0.5]{rcaf_c45.eps}
    %caption of the figure 
-   \caption*{\small \em Rolf Geiger at work on the troop seats.}
+   \caption*{\small \em RCAF C45 Expeditor a military variant of the Beech 18.Purchased in large numbers during WW II and after for pilot training, navigator training, and as a utility transport.}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:rolf}
+   \label{fig:rcaf_c45}
 \end{figure}
 
-Thus, we need to produce (15) different metal parts, including: lower
-tubing plus support brackets and clamps, spreader bars with two shell
-clamps each to clamp onto the lower support tubing plus the front
-brackets holding the spreader bars to the frontal support tubes, leg
-brackets connecting the legs to the frontal support tubes plus
-secondary brackets (slipping over the leg brackets) holding a spring
-loaded pin with which the seat fabric will be tautened, legs,
-including the lower clamping mechanism fastening the legs to the
-floor. Most parts need (22) of each, except the shell clamps for the
-spreader bars (44), hooks (117), leg grippers (44).
+Beachflight was a Training Command operation, under the command of
+Wing Commander Harry Forbell. Twenty six expeditors were assembled at
+Trenton where they were fitted with a 100 gallon fuel tank in the main
+cabin, plus safety and navigation gear required for the planned
+route. A pilot and navigator were assigned to each aircraft. I flew
+with Squadron Leader Brian, Marfleet, the designated Deputy Operation
+Commander. The aircraft were to fly in formation, four sections of
+five and one of six, within block airspace.
 
-The procedure in the production of these parts is as with any project:
-Take the finished product and then work backwards to determine the
-production stages.
+On 20 May 1959 Operation Beachflight was launched. Sections departed
+Trenton for Goose Bay at designated intervals in order to maintain
+separation along the route. Big trouble, the weather within the
+assigned block airspace deteriorated to the point where aircraft could
+not maintain contact with each other and abandoned their
+formations. Now it was everybody for himself. The 26 Expeditors were
+up, down, right, left, within the block airspace and no doubt outside
+it. My section dispersed and we did not see any of them until we
+arrived at Goose Bay. Not a great start to the operation, but everyone
+arrived safely at Goose Bay, so a success in that sense. Our flight
+time was 6:45.
 
-In the case of a formed sheet metal part a drawing of a flattened
-blank is produced by calculating bend radii and adding that to the
-straight surfaces (in effect un-bending or stretching the finished
-part).
+The flight from Goose Bay to Frobisher (Iqualuit) the next day was
+uneventful. Our Flight time was 6:05.
 
-Experimental operations may be needed to find the shape and dimensions
-of the blanks.  Tooling may be needed to produce the formed sheet
-metal parts.  The blanks are made by shearing and sawing, according
-to the parts drawing.
+As I recall, we had a weather delay in Frobisher and did not depart
+until 24 May. The plan was for each section to proceed to Sondrestrom
+AFB (Kangerlussuag-big Fjord) located on the West coat of Greenland,
+refuel and carry on over the icecap to Keflavik. The Deputy Commander
+would lead the 5th and last section. Our aircraft was unserviceable so
+we had to take the spare, the 26th aircraft. The unserviceable
+aircraft was repaired and returned to Trenton. We were late into
+Sondrestrom, to find that section four was still on the ground because
+the three earlier sections had experienced heavy icing over the ice
+cap.  It was not known if the crews and aircraft were safe, a great
+concern as the Expeditor was a notoriously poor performer in icing
+conditions. We assembled in the bar to await word on the fate of our
+mates. All the aircraft made it over the ice cap and arrived safely in
+Keflavik, which was ample reason for us to celebrate another
+success. Our flight time to Sondrestrom was 3:50.
 
-In the case of a machined part, it is usually produced from bar stock:
-Parts are cut to length and subsequently machined as per
-drawing. Partial in-between drawings may be necessary to produce the
-parts in progressive stages.
+We departed Sondrestrom on 27 May after a two day delay waiting for
+improved weather over the ice cap. The flight to Keflavik was
+uneventful. Our flight time was 5:10.
 
-To avoid the higher costs associated with surface plating the metal
-parts, (as per original parts) we chose painting them instead, which
-will be acceptable given the limited use of the finished product.
+The first three sections waited for us in Keflavik. Every crew had a
+story to tell about their experience over the ice cap. Fifteen
+aircraft were loaded with ice and unable to maintain altitude. Their
+only salvation was to make it past the ice cap and out over the water
+where they could descend to warmer air and shed the ice. Thankfully,
+all of them did.
+
+The east coast of Greenland is very rugged with numerous high
+peaks. Several crews in their descent out of the icing conditions
+broke cloud between these peaks. These stories were oft repeated and
+all called for another round of drinks.
+
+The next day, 28 May, all the sections proceeded to Prestwick. The
+weather was perfect and I got a lot of stick time because my pilot was
+very tired after the debriefings in Keflavik. Our flight time was
+5:10.
+
+We remained overnight in Prestwick and flew to 1 Wing, Marville,
+arriving mid afternoon on 29 May.  The aircraft were prepared for
+delivery to the French and Portugese air forces. The cabin fuel tanks
+were removed as well as all the safety and navigation gear.
+
+On 3 June we delivered 19 Expeditors to the French Air Force at
+Chateaudun, located not far from Chartres. The flight crews were
+picked up by C47 Dakotas and returned to Marville. Nineteen crews
+boarded a 426 Squadron North Star for return to Trenton. Of course the
+Commander and his Deputy remained behind to deliver six expeditors to
+the Portugese Air Force in Lisbon.
+
+We departed Marville on 4 June for Bordeaux where we planned to refuel
+and then proceed along an air route over the Pyrenees Mountains to
+Lisbon. The weather along the planned route was very bad with
+intensive thunder storms over the mountains. Ever resourceful, our
+Commander decided to file a flight plan west over the Bay of Biscay,
+clear of Spain, then south to Lisbon. This would be a long overwater
+flight and we had no extra fuel, safety equipment or nav gear. We took
+off and were in the process of forming up in two sections of three,
+when several light military aircraft flew through our formation. No
+one saw them coming, it happened so fast there was no reaction. Not a
+very good start to the day.
+
+Our Commander led us west over the Bay of Biscay, maintaining VFR
+under a cloud layer. The cloud  base got lower and lower, we were down
+with the shipping and it was decided to climb above the cloud, VFR on
+top. This worked for a while but the cloud tops kept creeping up over
+10,000 feet and we had no oxygen. The Commander decided to descend and
+try for better conditions. The Deputy Commander, my pilot, decided to
+stay high; our six plane formation was split three high and three
+low. I was now doing the navigation for the high section. With nothing
+but a chart to work with, I managed to make use of the marine beacons
+along the coast to guide us west then south to Lisbon. We arrived at
+Lisbon before the low section and the Deputy Commander wisely elected
+to hold until the Commander arrived and landed. Our flight time was
+5:45.
+
+The next day we participated in a small handing over ceremony. Mission
+accomplished!
 

### 2013-12-14 pnsac/newsletter/article_Feature3.tex
Preparation for December 2013 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Troop Seats and Extras}
\author{Nelson Smith}

\maketitle

\section{Troop Seats}
\label{troopseats}

Seat back assemblies are completed for the troop seats. Thirty eight
units were produced. The assembling of the backs to the platforms to
produce triple and quad seats will be the next step. This will be done
over the winter.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{troopseats.eps}
   %caption of the figure 
   \caption*{\small \em New troop seats.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:troopseats}
\end{figure}

The prototype is progressing well. As previously stated, measurements
are a rarity with these projects and sometimes it takes longer to
figure out a plan than to do the job. A guesstimate for completion:
early December.

\section{Other Projects}
\label{otherporjects}

Proctective covers for the probes which protrude under the belly of
the North Star were made and installed. The foam pads previously
installed had done their time.

The First Officer's protective seat covers have been manufactured and
installed. This addition was required to shield the seat covers from
UV rays, dust and stains. Similar covers will be made for the
Captain's chair during the winter months.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2013-12-14 pnsac/newsletter/article_ProgReport.tex
Newsletter: final changes incorporated.
Preparation for December 2013 newsletter.

@@
 \title{Project Manager's Progress Report}
-\subtitle{October 2013}
+\subtitle{December 2013}
 \author{Bruce Gemmill}
@@
 
-Engine 3 is almost completely reassembled, with only the supercharger
-and intercooler preheat assemblies left to install on the engine,
-which was moved to the QEC engine frame in early summer.  The work has
-slowed somewhat, due to summer down time for staff and volunteers, but
-we continue to make steady progress.
-
-The auxiliary gearbox was recently removed from the engine firewall,
-and is now being disassembled.  This is the last major assembly that
-must be overhauled before the engine can be installed on the aircraft.
-
-\section{Engine Frame}
-\label{engineframe}
-
-Now that the engine and frame are joined, the major work still
-required is rebuilding the many cowl panels and the radiator flaps
-that surround the engine.  There is a lot of corrosion on the pieces,
-and each must be disassembled, cleaned, then repaired before painting
-and reassembly.  This work must be done carefully to ensure the cowl
-panels fit properly when installed on the QEC.
-
-This work has gone slowly due to the small number of volunteers who
-are able to do this type of repair.
-
-\section{Crew Lounge, Galley,  and Forward Washroom}
-\label{crewlounge}
-
-The new galley has now been installed in the crew lounge, and work
-overhead in the ceiling is also complete.  This involved installing
-new insulation and the Janitrol heaters, along with refurbished
-heating ducts, and some new fuel lines.  The ceiling panels were also
-installed.
-
-The doors on the main electrical panel were restored and installed.
-New circuit legend cards were produced to indicate the circuit breaker
-and junction box connections.  These were laminated and installed
-inside the electrical panels.  Some additional investigation is needed
-to complete the electrical circuit connection diagrams that were badly
-faded.
+Engine 3 is nearing completion.  The supercharger and intercooler
+preheat assemblies have been installed, along with the propeller
+reduction gearbox.  While this work was being carried out by the
+engine crew, the remainder of the volunteer work force busied
+themselves completing the numerous cowl panels, pipes and hoses and
+other pieces needed to complete the engine.  Two items that seemed to
+take forever were the large steel exhaust shrouds.  A lot of hammering
+was needed to pound out years of dents, then numerous rivets and all
+the Dzus fasteners and springs needed to be replaced.  The complex
+shape of the exhaust shroud made this challenging work.  Further
+delaying completion was the search for suitable clear coat to protect
+the steel from corrosion.  The old clearcoat we used successfully on
+engines 1 and 2 is no longer available.  Mike Irvin finally located a
+suitable replacement, and the shrouds were clear coated and serial
+numbers were stencilled on.
 
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{galley_interior.eps}
+   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
    %caption of the figure 
-   \caption*{\small \em New galley -- interior view.}
+   \caption*{\small \em Garry installing the intercooler on engine 3.}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:galley_interior}
+   \label{fig:Eng3Intercooler}
 \end{figure}
 
-One section of floor was patched after removing a corroded area.  One
-additional repair needs to be done under the floor of the crew lounge.
-Until this is complete, the new cushion flooring cannot be
-installed. This has also delayed the installation of all of the crew
-lounge accessories, such as the table, seats and secure storage bin,
-all of which are complete but waiting in storage.
+The auxiliary gearbox was disassembled, and significant corrosion was
+found on the bearings inside.  Bearings had to be ordered from
+England.  Meanwhile, the gearbox was cleaned and painted.  At the same
+time, the large DC generator was dismantled, repainted and
+reassembled, as well as the air pump and the tachometer generator and
+propeller synchronizer.  Once the new bearings were installed, the
+gearbox was completed and the generator and air pump were attached.
+Finally, the gearbox was installed on the firewall, which is now
+finished, awaiting the installation of engine \#3.
+
+\section{Crew Lounge, Galley,  and Forward Washroom}
+\label{crewlounge}
+
+Work has slowed on the forward section of the aircraft.  The repair
+needed under the floor of the crew lounge is on hold until a suitable
+piece of extruded aluminum can be found.  Until this is complete, the
+new cushion flooring cannot be installed. This has also delayed the
+installation of all of the crew lounge accessories, such as the table,
+seats and secure storage bin, all of which are complete but waiting in
+storage.
 
@@
 
-The forward and rear belly compartments are now complete.  The two
-battery elevators that were removed have been restored and installed
-behind the nose wheel.  The battery compartment covers were repaired
-but need painting and stencilling.
-
-The firewall on nacelle 3 is being cleaned in preparation for the
-installation of the engine sometime this winter.
-
-Further progress has been made on the set of troop seats being
-fabricated for the main cabin.
+The forward and rear belly compartments are now complete.  All the
+frame pieces and hardware needed for the troop seats have been made,
+and work is progressing on the seats and backs.  One section has
+undergone a trial fit, and all pieces seem to fit perfectly.
 
@@
 
-Over the next year, we hope to have engine Nr 3 completed and engine
-Nr 4 removed.  We will complete the reassembly of the crew lounge and
-galley.  We then plan to work on the main cargo compartment, including
-refurbishing the main heater duct and other fittings, and removing
-floors to begin cleaning and repairs under the cargo floor.  We may
-yet get to work on the four engine nacelles.
+The museum has advised all North Star volunteers that our work must be
+put on hold for several months while a major exhibit for the 100th
+anniversary of the beginning of the First World War is prepared.  Our
+work space in the restoration hangar is needed to assemble and prepare
+several aircraft that will be added to the display space in March.
+Until then our work on the aircraft is suspended, but we don't expect
+this delay to have a serious impact on the overall project.  We look
+forward to resuming work in the Spring.
 

### 2013-12-14 pnsac/newsletter/article_SpecialEvents.tex
Preparation for December 2013 newsletter.

@@
 
-\title{Recent Special Events}
-%\subtitle{Visit to Nav Canada Montreal A.C.C. and Bombardier}
-\author{Bill Tate}
+\title{Quarterly Meeting Report}
+\author{Drew Hodge}
 
@@
 
-% \begin{figure}[htbp]
-%    \vspace{2em}
-%    \centering
-%    %name of the graphic, without the path AND in EPS format:
-%    \includegraphics[scale=1.5]{loch_march.eps}
-%    %caption of the figure 
-%    %\caption*{\small \em North Star engine 3.}
-%    %label of the figure, which has to correspond to \ref{}:
-%    \label{fig:logo}
-% \end{figure}
+On Saturday 7th December 2013, Project North Star Association of
+Canada (PNSAC) held a quarterly meeting in the auditorium of the
+Canadian Aviation and Space Museum (CASM) in Ottawa. It was a
+memorable meeting with over 200 people in attendance. Those in the
+audience that day were able to hear the story of the Gimli Glider told
+in person by Captain (retd.) Robert (Bob) Pearson, the pilot of Air
+Canada Boeing 767-200, fin number 604, when it made an unscheduled
+stop at Gimli Manitoba, thirty years ago.
 
-\section{Loch March Golf Day}
-\label{golf}
+The meeting started with an introduction by Bill Tate, Project North
+Star's Vice President.  Bill began by welcoming the audience, which,
+as well as PNSAC volunteers, included members of Vintage Wings of
+Canada, Ottawa Airport Watch, the Canadian Aviation Historical
+Society, the College of Professional Pilots of Canada, CASM volunteers
+and staff, Air Canada and WestJet pilots, and visitors from the
+general public, one of whom had travelled to Ottawa from Vancouver for
+the meeting. After describing the goals of Project North Star and
+asking PNSAC volunteers to stand and be recognized, Bill invited the
+audience to visit the world class exhibits in the Museum after the
+meeting.
+
+Bill's acquaintance with Bob Pearson goes back thirty-five years, when
+Bill was Captain Pearson's Second Officer on Boeing 727s based in
+Montreal. His duties in the cockpit back then gave Bill ample
+opportunities to watch Captain Pearson at work. Bill described Captain
+Pearson as a "highly professional pilot" who, "wearing his command
+lightly", demonstrated "highly polished interpersonal skills and a
+delightful sense of humour".  He then invited Captain Pearson to tell
+the story of flying unscheduled into Gimli.
+
+Captain Pearson spoke for a little over an hour with no notes and with
+reference to a few appropriate slides.  He described eloquently how,
+because of errors in the conversion of metric measurements, his Boeing
+767-200 ran out of fuel and lost both engines over Manitoba on the way
+from Ottawa to Edmonton. The full story is well documented in the book
+"Freefall: From 41,000 feet to zero - a true story, William and
+Marilyn Hoffer, Simon \& Schuster, 1989", but to hear it told in
+person by the pilot in command that Saturday, 23rd July, 1983, was a
+rare privilege.
 
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=1.5]{loch_march.eps}
+   \includegraphics[scale=0.5]{Pearson004.eps}
    %caption of the figure 
-   %\caption*{\small \em North Star engine 3.}
+   \caption*{\small \em Captain (retd.) Robert Pearson.}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:logo}
+   \label{fig:pearson004}
 \end{figure}
 
-The PNSAC 2013 Annual Golf Day at Loch March started with breakfast in
-the Club House, 14 members and friends in attendance.
+The port engine stopped first at 41,000 feet prompting the decision to
+divert to Winnipeg.  At 28,000 feet the starboard engine ran down too,
+and Maurice Quintal, Captain Pearson's First Officer, made
+calculations that showed they couldn't reach Winnipeg.  Quintal
+advised Captain Pearson to try for the nearest runway on the former
+RCAF aerodrome at Gimili, twelve miles away.  Some eight or ten
+minutes after losing the first engine, Captain Pearson side-slipped
+the big Boeing down to an old runway that was being used as a racing
+track by members of the Winnipeg Sports Car Club. The nose wheel
+collapsed on landing, but the aircraft came to a stop and no one was
+seriously injured. The racers helped to put out a minor fire in the
+nose.
 
-Prior to breakfast, start times were arranged in such a way that there
-was a mixing of participants, so that those who were not aware of what
-PNSAC does had the opportunity to talk to members involved in the
-project.
+Captain Pearson talked about the inquiry that followed the incident
+and its eventual exoneration of both pilots.  The aircraft was flown
+to Winnipeg just two days after the Gimli landing and saw service with
+Air Canada for another twenty-five years.  When the Boeing was retired
+to the Mohave desert in 2008, Captain Pearson was at the controls to
+make that last landing in Arizona.
 
-The first group started out in light rain showers and thirty minutes
-later the skies cleared for a bright sunny day. With the extra
-rainfall this year the course was found to be in a very lush
-condition, and as usual for Loch March the greens were very fast.
+Following his talk, Captain Pearson answered questions from the
+audience.  Richard Lodge, PNSAC President, thanked Captain Pearson and
+presented him with a PNSAC golf shirt, a baseball cap, a coffee mug,
+and finally a certificate conferring upon him an Honorary Life
+Membership of the Project North Star Association.  Captain Pearson
+then graciously agreed to make presentations of volunteer-hours
+certificates and 10th Anniversary photos to PNSAC volunteers.  The
+following volunteers received certificates for the hours shown:
 
-Danielle Nadon the Club Professional graciously gave PNSAC two 18 hole
-certificates as prizes, one for the closest to the pin and the other
-for the longest drive, which were both won by Gary Hall. He also came
-in with low gross for the day at 79. Our own Ron Lemieux was paired
-with Gary and in Ron's own words ``I had nothing to do with the score;
-I was riding on Gary's coat tails''.
-
-Loch March is a very wonderful facility for golf and has an excellent
-club house as well. The course can be challenging, but there is
-something for everyone who plays golf. Even if you do not golf, it is
-well worth the drive to enjoy the patio. For details please visit the
-web site: {\normalfont\color{blue}\texttt{\url{http://www.lochmarch.com}}}
-
-This year we had another prize which was given to us by New York New
-York Hair Salon, - a gift certificate of \$100:00 for hair
-services. So a new category of winner had to be chosen, which was for
-the best dressed in our distinctive PNSAC clothing. To ensure fairness
-in the award I had Alfredo Porco (who has been my barber for 30 years)
-make the award. It went to our "special seamstress" Karen Lochhead.  I
-should mention there are now three generations of Porcos at New York
-New York, - Alfredo, his son Vince and Vince's daughter Marina.
-
-For information on New York New York please visit their web site:
- {\normalfont\color{blue}\texttt{\url{http://www.nynyhairsalon.com}}}
-
-Looking forward to our fourth golf day next year, we will work on more
-prizes for our members and guests in 2014.
-
-\section{Trip to the Montreal Area Control Centre and Bombardier Aerospace}
-\label{yultrip}
-
-On Friday October 4th, 33 members and guests departed the Canada
-Aviation Space Museum (CASM) for another planned special event. First
-off, special thanks to Bryon Mask (retired Air Canada Captain), one of
-our members who assisted me with contact information at Nav Canada
-which was required to have our visit approved. Bryon is a highly
-respected individual in Flight Safety and Accident Investigation in
-Canada.
-
-We were met and welcomed by Richard Snider, Duty Manager at the
-Montreal ACC. We visited the Training Centre where training is
-provided to new Air Traffic Controllers and required recurrent
-training is offered. We were also shown the tower simulator which has
-a day/night 360 degree display where various airport scenarios can be
-brought up to assist in training. My personal favourite was the
-display showing a landing as seen from the cockpit, which even showed
-it as a 360 degree display (first time I have experienced the rear
-view mirror look for a landing).
+\begin{itemize}
+  \item Bruce Gemmill (7000 hrs)
+  \item Ted Devey (4000 hrs)
+  \item Bill Tate (3000 hrs)
+  \item Richard Lodge (1000 hrs)
+  \item Bruce Grant (1000 hrs)
+  \item Robert D\'{e}sjardins (1000 hrs)
+  \item Peter Trowbridge (1000 hrs)
+  \item Charles Baril (1000 hrs)
+\end{itemize}
 
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{yul_tca3.eps}
+   \includegraphics[scale=0.5]{Pearson007.eps}
    %caption of the figure 
-   \caption*{\small \em In the 360 degree tower simulator.}
+   \caption*{\small \em Richard Lodge presenting Captain Pearon with a
+     certificate for Honorary Life Membership in PNSAC.}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:tca3}
+   \label{fig:engine3propcowl}
 \end{figure}
 
-Our next stop was the Technical Operations Centre where staff monitor
-all radar and computers. ACC has dual systems with one system picking
-up automatically in case the other system fails. Communications has a
-dual back-up digital and analogue system.  Communications to aircraft
-are by conventional VHF with some interesting back ups such as
-Controller Pilot Data Link Communications (CPDLC). The easiest way of
-explaining this is Text Messaging Sat Com in which either the aircraft
-can call the phone number of the ACC or the ACC can call the dedicated
-number for the aircraft (\$6:00 a minute cost) and as a back up by HF
-radio through a Flight Service Station.
+To conclude the meeting, Captain Pearson presented signed copies of
+the book "Freefall: From 41,000 feet to zero - a true story" to three
+audience members whose ticket numbers he had drawn at random.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{yul_tca2.eps}
-   %caption of the figure 
-   \caption*{\small \em Air traffic controllers at work.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:tca2}
-\end{figure}
-
-In the Operations Room, we could hear some controllers talking very
-quietly. Radar displays offer a great deal of information to the
-Controller, which also helps keep talking to a minimum. A sense of
-humour prevailed at the work station for Northern Quebec up to
-Iqaluit, where a Motel 6 logo was mounted over the work station.
-
-In total we were on site for close to 2 1/2 hours which I found to be
-a fantastic experience, seeing the ATC environment from the other
-side.
-
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{yul_tca1.eps}
-   %caption of the figure 
-   \caption*{\small \em The Operations Room.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:tca1}
-\end{figure}
-
-With kind assistance in the form of a recommendation from Mae, my YUL
-expert, we broke off for lunch at Le D\'{e}jeuner Cosmopolitain which
-handled our group in a very friendly efficient manner. I should note,
-this is the spot where Air Canada retirees meet for their monthly
-luncheon dates.
-
-After lunch we proceeded to the Bombardier Production plant where we
-were met by three very knowledgeable tour guides. They were Alex
-Ritchie, Fernand Richer, and Cristian Racine who together gave a very
-informed hour and half tour of the production line. The tour was
-divided into two areas, assembly and finishing. At the beginning of
-the assembly we saw where the fuselage and wings are mated and
-proceeded on to where systems are tested such as hydraulic systems,
-then to the finishing area where interiors are installed. It was
-something to see how clean and bright this plant is and how the many
-sub assemblies come together. The tail assembly built in China comes
-to the production line and meets with part of the fuselage built in
-Mexico, just in time for the assembly.
-
-We ended our visit in the delivery room where customers pick up their
-aircraft all ready to go, all yours for approximately \$30 million
-dollars.
-
-We enjoyed an excellent dinner at the Auberge Willow Inn prior to our
-return to the CASM at the end of a very enjoyable 14 1/2 hour day
-together.
+The Association's sincere thanks go to Bill Tate for organizing this
+quarterly meeting and the talk by Captain Pearson -- it was largely
+Bill's efforts that made the meeting the great success that it was.
 

### 2013-12-28 pnsac/newsletter/article_EditorsNotes.tex
Source files for Dec13 newsletter.

@@
 {\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}} (note that
-the location of the neswletter archive will change when the new
+the location of the newsletter archive will change when the new
 website goes live) starting with the April 2013 issue of the
 Chronicle.
+
+\vspace{12pt}
 
@@
 
+\vspace{12pt}
+
 Motorheads among our readers should look into Ted Devey's extensive
@@
 issue.
+
+\vspace{12pt}
 

### 2013-12-28 pnsac/newsletter/article_Feature1.tex
Source files for Dec13 newsletter.

@@
 very tedious and unglamorous work which was of great importance for
-the safe and trouble-free running of engines in service. I can
+the safe and trouble free running of engines in service. I can
 remember that I would hear from time to time of the difficulties of
@@
 had been signed and the company was still enjoying its reputation as
-the company whose Merlin engines had saved Britain in the air
-war. This was in 1965/6 and it was several years before our
-projections of doom resulted in the bankruptcy of the
+the company whose Merlin engines had helped to save Britain in the
+Second World War. This was in 1965/6 and it was several years before
+our projections of doom resulted in the bankruptcy of the
 company. Fortunately I left the company in 1966 and was not involved
@@
 remains of the old one. This has been very much a success story and
-the newly privatized Rolls-Royce is now a highly profitable and
-innovative company. I look back on my days at R-R with nostalgia but I
-now realize that much of the time we were all living in a fool's
-paradise.
+the privatized Rolls-Royce is now a highly profitable and innovative
+company. I look back on my days at R-R with nostalgia but I now
+realize that much of the time we were all living in a fool's paradise.
 

### 2013-12-28 pnsac/newsletter/article_Feature2.tex
Source files for Dec13 newsletter.

@@
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{rcaf_c45.eps}
-   %caption of the figure 
-   \caption*{\small \em RCAF C45 Expeditor a military variant of the Beech 18.Purchased in large numbers during WW II and after for pilot training, navigator training, and as a utility transport.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:rcaf_c45}
-\end{figure}
-
 Beachflight was a Training Command operation, under the command of
@@
 five and one of six, within block airspace.
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{rcaf_c45.eps}
+%    %caption of the figure 
+%    \caption*{\small \em RCAF C45 Expeditor a military variant of the Beech 18.Purchased in large numbers during WW II and after for pilot training, navigator training, and as a utility transport.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:rcaf_c45}
+% \end{figure}
 
@@
 all called for another round of drinks.
+
+\end{multicols}
+
+\begin{figure*}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=1.0]{BEECHFLIGHT0001.eps}
+   %caption of the figure 
+   \caption*{\small \em Photo, taken by Tim, of two RCAF C45
+     Expeditors (military variant of the Beech 18) during the ferry
+     flight he describes in this article. These aircraft were
+     purchased in large numbers during and after WW II for pilot
+     training, navigator training, and as utility transports.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:BEECHFLIGHT0001}
+\end{figure*}
+
+\begin{multicols}{2}
 

### 2013-12-28 pnsac/newsletter/article_PresReport.tex
Source files for Dec13 newsletter.

--- 2013-07-15 pnsac/jul13/article_PresReport.tex
@@
 
-Canada Day is now behind us.  We had an excellent day with a good
-number of our volunteers showing off the North Star to the general
-public. Judging by the activity around the plane all the day, we have
-now developed a display format which is interesting to visitors.
-Canada Day was the only day this year when the plane was on display to
-the general public.
+As we come to the end of 2013, I think of it as an interesting year
+with some disappointments and important successes. I believe we end
+the year in a very positive way and I look forward to 2014 with
+enthusiasm and renewed optimism.
 
-On June 27, in co-operation with the Museum, who hosted the event, we
-showed off the plane to a group of donors attending the Donor
-Appreciation Evening.  This was a successful evening which is likely
-to be repeated.
+First let me get the disappointments out of the way. 2013 was our 10th
+anniversary year. We had hoped to make the event an opportunity to
+show off our achievements during the first 10 years and create
+additional interest around the plane. Unfortunately, due to events
+beyond our control, it was not possible to stage a significant event
+marking the anniversary. We therefore did our usual opening of the
+plane to the public on Canada Day and marked the anniversary with our
+PNSAC photographer taking photographs of members as they visited the
+restored cockpit.
 
-In my last Notes from the President, I referred to plans for
-celebrating the 10th anniversary of the founding of the Project North
-Star Association.  This should have taken place around June 24.
-Unfortunately, the event passed without us being able mark the
-day. This was mainly due to situations beyond our control.
+The second disappointment is more serious. During 2013 we did not take
+on any new volunteers to work on the plane. We had several good
+applicants but were unable to fit any of them into the working
+schedule of volunteers at the Canada Aviation \& Space Museum
+(CASM). This has two downsides; the first being that we are now
+becoming short of active volunteers and the second being that we are
+not training any new and sometimes younger volunteers to work on
+pre-jet age aircraft. It is important to pass on the skills of the
+older volunteers to younger members of our Association.
 
-Our 2013 AGM was held on June 8, and all the existing directors were
-re-elected.  This will enable the directors, who have now become
-experienced, to concentrate on long-term matters.
+Funding the restoration of the aircraft is always a challenge and is
+becoming more so as the Federal government cuts back on the funds it
+appropriates to the CASM and other similar institutions..
 
-The actual work on the restoration of the plane is going very well, as
-reported elsewhere in this newsletter. The major issue facing the
-Board in the next year will be to consider how to secure the future of
-PNSAC, both in terms of volunteers and the funding of restoration
-work.
+You may well ask why I feel optimistic about 2014 after referring to
+disappointments and problems. In the last two months the Directors of
+PNSAC met with the Director General of CASM, Stephen Quick, and
+subsequently a small group of directors met again with him together
+with CASM conservation staff. The outcome of these meetings has been
+an excellent exchange of views, a very firm commitment by the CASM to
+continue to actively support the North Star restoration and a better
+understanding by both sides of the challenges we face and possible
+ways forward.
 
-We have now reached a point where it is necessary to update our
-Memorandum of Understanding (MOU) with the Museum and our work plan
-for completing the restoration of the plane.  Our volunteers have
-already worked over 60,000 hours on the project and we expect to work
-a similar number of hours before the restoration is completed.
+As many of you already know, volunteer work on the North Star will be
+suspended for approximately the first three months of 2014. This is to
+enable CASM to undertake a major reorganization of the Museum exhibits
+in preparation for the centenary of the start of the First World
+War. During this time the CASM professional staff will be occupied
+with the reorganization and will not be able to provide the necessary
+supervision of restoration work. During this time, though, we plan to
+update the work plan and overall budget for completing the North Star
+restoration, work on fundraising initiatives and finally update our
+Memorandum of Understanding with CASM.
 
-There have been several informal meetings to discuss our future
-operations and the opportunities available for new volunteers.  Within
-the next few weeks we are expecting to meet with the Director General
-of the Aviation Museum and members of his staff to work on the MOU
-update.
-
-Later this month we are meeting with senior management of the Museum
-Foundation, the body charged with corporate fundraising for the Canada
-Science and Technology Museums Corporation, to discuss how PNSAC can
-work with the Foundation to raise funds for the North Star restoration
-and the museums in general.
-
-With the initiatives above, we are hoping in the next few months to
-consolidate our position as an important organization working with the
-Aviation Museum
+In addition to our hands-on restoration work, the Association has a
+strong social side. Bill Tate, our Vice President, and his Special
+Events team have organized some memorable bus trips, the last of which
+was to Montreal in October. Bill has now turned his considerable
+organizational skills to further developing our quarterly Members'
+Meetings. In early December we had a great morning at CASM when
+Capt. Bob Pearson of the Gimli Glider fame made a presentation to 150
+of our members and members of other aviation organizations we invited
+to attend. Although our Vice President has retired as an airline
+captain, he certainly has not retired from anything else. Watch your
+emails, Facebook and Twitter for news of events Bill and his team are
+planning for 2014.
 

### 2014-05-17 pnsac/newsletter/article_EditorsNotes.tex
Updated files for Apr/May 2014 newsletter.

@@
 
-This issue of the Chronicle carries the last of four chapters in
-Richard Lodge's story of his time at Rolls-Royce. Taken together, they
-make an interesting ethnology of an extinct industrial culture. The
-whole series is accessible through our website,
-{\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}} (note that
-the location of the newsletter archive will change when the new
-website goes live) starting with the April 2013 issue of the
-Chronicle.
+While our project has been on hold, your editor still had to find some
+content for the Chronicle.
+
+%This issue of the Chronicle carries the last of four chapters in
+%Richard Lodge's story of his time at Rolls-Royce. Taken together, they
+%make an interesting ethnology of an extinct industrial culture. The
+%whole series is accessible through our website,
+%{\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}} (note that
+%the location of the newsletter archive will change when the new
+%website goes live) starting with the April 2013 issue of the
+%Chronicle.
 
@@
 
-We also have another hair-raising flying story from Tim Timmins. A
-series of previous Timmins stories can be found through the same
-website, starting with the December 2010 issue.
+Jim Riddoch came to the rescue with a story about a
+successful\/disastrous flight test.  We like these kinds of
+stories--to small and personal to turn up in the history books, but
+they do pull you into real human experiences. There will be more of these.
+
+%We also have another hair-raising flying story from Tim Timmins. A
+%series of previous Timmins stories can be found through the same
+%website, starting with the December 2010 issue.
 
@@
 
-Motorheads among our readers should look into Ted Devey's extensive
-writings on the Rolls-Royce Merlin, starting in the December 2006
-issue.
+Your editor became a writer too, and the writer submitted a tragic
+story about a French aviatice. The story was accepted by the editor in
+a blatant conflict of interest.
+
+%Motorheads among our readers should look into Ted Devey's extensive
+%writings on the Rolls-Royce Merlin, starting in the December 2006
+%issue.
 
@@
 
-All together, the archived issues are a valuable web resource for
-readers with an interest in aviation history or the history of Project
-North Star.
+And we bring you a poem by David Lambeth, an ode to the North Star
+that was first published in the Chronicle of November 2005. Volunteer
+Tim Timmins had found it in a hidey-hole in the aeroplane in the early
+days of our restoration project.  Enjoy!
+
+%All together, the archived issues are a valuable web resource for
+%readers with an interest in aviation history or the history of Project
+%North Star.
 

### 2014-05-17 pnsac/newsletter/article_Feature2.tex
Updated files for Apr/May 2014 newsletter.

--- 2013-10-17 pnsac/newsletter/article_Feature2.tex
@@
 
-\title{Building troop seats for North Star 17515}
-
-\author{Rolf Geiger}
+\title{The Whispering Giant}
+\author{Dave Lambeth}
 
@@
 
-There are no original parts or drawings for the seats. Fortunately,
-the tubing and brackets that connect the seats to the aircraft wall
-were still in place in the aircraft. Also, two lower tubing support
-brackets were found hiding behind the galley, allowing us to produce
-exact copies of the original. Everything else concerning the seats is
-missing and needs to  be replaced.
-
-\end{multicols}
-
-%\begin{figure*}[htbp]
-\begin{figure*}[h]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.75]{ass.eps}
-   %caption of the figure 
-   \caption*{\small \em Troop seat construction detail.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:ass}
-\end{figure*}
-
-\begin{multicols}{2}
-
-To find the size and the number of seats to be produced we measured
-the space between the anchoring studs in the aircraft floor, which the
-legs snap onto. This gave us the width of the seats. Thus, we will
-produce (5) three-seaters plus (6) four-seaters, providing seating for
-39 troops. No litters are planned.
-
-Mike Irvin (our Museum coordinator) decided to use seats from a
-helicopter in the Museum collection as an example for our seats. We
-are more or less copying the parts of these seats, considering that we
-have to adapt our methods to the equipment available to us.
-
-To start the project, drawings of the parts were made. The list of
-materials of these parts calls mainly for aluminum bar stock, aluminum
-tubing, steel tubing, aluminum sheet and steel sheet.
-
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{rolf.eps}
-   %caption of the figure 
-   \caption*{\small \em Rolf Geiger at work on the troop seats.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:rolf}
-\end{figure}
-
-Thus, we need to produce (15) different metal parts, including: lower
-tubing plus support brackets and clamps, spreader bars with two shell
-clamps each to clamp onto the lower support tubing plus the front
-brackets holding the spreader bars to the frontal support tubes, leg
-brackets connecting the legs to the frontal support tubes plus
-secondary brackets (slipping over the leg brackets) holding a spring
-loaded pin with which the seat fabric will be tautened, legs,
-including the lower clamping mechanism fastening the legs to the
-floor. Most parts need (22) of each, except the shell clamps for the
-spreader bars (44), hooks (117), leg grippers (44).
-
-The procedure in the production of these parts is as with any project:
-Take the finished product and then work backwards to determine the
-production stages.
-
-In the case of a formed sheet metal part a drawing of a flattened
-blank is produced by calculating bend radii and adding that to the
-straight surfaces (in effect un-bending or stretching the finished
-part).
-
-Experimental operations may be needed to find the shape and dimensions
-of the blanks.  Tooling may be needed to produce the formed sheet
-metal parts.  The blanks are made by shearing and sawing, according
-to the parts drawing.
-
-In the case of a machined part, it is usually produced from bar stock:
-Parts are cut to length and subsequently machined as per
-drawing. Partial in-between drawings may be necessary to produce the
-parts in progressive stages.
-
-To avoid the higher costs associated with surface plating the metal
-parts, (as per original parts) we chose painting them instead, which
-will be acceptable given the limited use of the finished product.
+Twinkle, twinkle, old North Star,\\
+With wrinkled skin and creaky
+spar,\\
+For years you’ve always been on
+sked,\\
+As many of our force you’ve sped\\
+to distant lands beyond the skies,\\
+you’ve never failed to see the rise\\
+of distant suns or moonrise glare.\\
+You’ve carried us with tender care\\
+-back you came to take us home,\\
+from old Japan or France or Rome\\
+from England, Goose, and Kef,\\
+We cannot hear you.\\
+(‘cause we’re deaf)\\
 

### 2014-05-17 pnsac/newsletter/article_ProgReport.tex
Updated files for Apr/May 2014 newsletter.

@@
 \title{Project Manager's Progress Report}
-\subtitle{December 2013}
+\subtitle{April 2014}
 \author{Bruce Gemmill}
@@
 
+Due to museum preparations for a display to mark the beginning of
+the First World War and other events in 2014, the restoration
+space used by Project North Star was taken over by other museum
+projects, and most work on the North Star was suspended.
+However, some limited work did continue.
+
 \section{Nr 3 Engine}
@@
 
-Engine 3 is nearing completion.  The supercharger and intercooler
-preheat assemblies have been installed, along with the propeller
-reduction gearbox.  While this work was being carried out by the
-engine crew, the remainder of the volunteer work force busied
-themselves completing the numerous cowl panels, pipes and hoses and
-other pieces needed to complete the engine.  Two items that seemed to
-take forever were the large steel exhaust shrouds.  A lot of hammering
-was needed to pound out years of dents, then numerous rivets and all
-the Dzus fasteners and springs needed to be replaced.  The complex
-shape of the exhaust shroud made this challenging work.  Further
-delaying completion was the search for suitable clear coat to protect
-the steel from corrosion.  The old clearcoat we used successfully on
-engines 1 and 2 is no longer available.  Mike Irvin finally located a
-suitable replacement, and the shrouds were clear coated and serial
-numbers were stencilled on.
+Work on Engine 3 was halted for several weeks in early 2014, but
+has now resumed on a reduced work schedule.  Most accessories are
+now installed, and soon the cowl panels will be fitted.  It is
+expected that Engine 3 will be installed late this spring or
+early summer, so that work can begin on removing and restoring
+Engine 4.
+
+This downtime also offered the opportunity for the engine crew to
+clean and re-organize the engine shop.
 
@@
 
-The auxiliary gearbox was disassembled, and significant corrosion was
-found on the bearings inside.  Bearings had to be ordered from
-England.  Meanwhile, the gearbox was cleaned and painted.  At the same
-time, the large DC generator was dismantled, repainted and
-reassembled, as well as the air pump and the tachometer generator and
-propeller synchronizer.  Once the new bearings were installed, the
-gearbox was completed and the generator and air pump were attached.
-Finally, the gearbox was installed on the firewall, which is now
-finished, awaiting the installation of engine \#3.
+\section{Cockpit}
+\label{cockpit}
+
+While restoration work in the shop could not be continued, we
+used the opportunity to install the crew seats and other restored
+items to the cockpit.  The pilot seat was installed (first
+officer seat was installed last year) then the railings and
+safety board were installed behind the seats.  Next we installed
+the radio officer and navigator tables, and the seats for these
+positions.  The radio officer seat proved to be a particular
+challenge, as the floor in this area is no longer flat, and some
+modifications to the seat rails were necessary to allow the seat
+to slide.  Finally the flight engineer’s jump seat was attached
+to the railing.  With five crew seats installed, the cockpit is
+very cramped.
+
+The cockpit installations were finished off by installing the
+curtains, crew oxygen hoses, and a Morse code key at the radio
+position.
 
@@
 
-Work has slowed on the forward section of the aircraft.  The repair
-needed under the floor of the crew lounge is on hold until a suitable
-piece of extruded aluminum can be found.  Until this is complete, the
-new cushion flooring cannot be installed. This has also delayed the
-installation of all of the crew lounge accessories, such as the table,
-seats and secure storage bin, all of which are complete but waiting in
-storage.
+The floor under this section was repaired by removing a section
+of badly corroded “I” beam and fabricating a new one from several
+pieces of aluminum which were riveted in place to support the
+floor.  Once this repair was completed, new cushion flooring was
+cut and glued to the crew lounge and the washroom floor, then
+various pieces of floor trim were installed.  To install the crew
+seat, it was necessary to remove the lower support section, since
+the fully assembled seat could not easily fit through the narrow
+doorway.  Once the seat was installed, we added the crew table
+and bunk bed, along with the secure storage box.  We also
+installed the curtain rod and new curtains to close off the bunk
+bed. The installation was finished off by re-installing the door
+between the lounge and the main cabin, so that this fully
+restored area can be protected from dust and debris while the
+main cabin undergoes restoration over the next two years.
 
-\section{Fuselage and Empennage}
-\label{empenage}
+The completion of the forward area of the aircraft marks a
+significant milestone in the restoration of the North Star.
 
-The forward and rear belly compartments are now complete.  All the
-frame pieces and hardware needed for the troop seats have been made,
-and work is progressing on the seats and backs.  One section has
-undergone a trial fit, and all pieces seem to fit perfectly.
-
-\section{Planned Restoration Work 2013/14}
+\section{Planned Restoration Work 2014}
 \label{sec:plannedwork}
 
-The museum has advised all North Star volunteers that our work must be
-put on hold for several months while a major exhibit for the 100th
-anniversary of the beginning of the First World War is prepared.  Our
-work space in the restoration hangar is needed to assemble and prepare
-several aircraft that will be added to the display space in March.
-Until then our work on the aircraft is suspended, but we don't expect
-this delay to have a serious impact on the overall project.  We look
-forward to resuming work in the Spring.
+Some limited work is being done to remove fittings and wall
+panels from the main cabin, in preparation for some major
+restoration work this summer.  Fabric work on the troop seats is
+nearing completion, although the seats will not be installed
+until all work on the main cabin is complete.  We hope to resume
+a full work schedule by June, with the aircraft outside once
+more.
 

### 2014-05-24 pnsac/newsletter/article_EditorsNotes.tex
Updated files for May 2014 newsletter.

@@
 Jim Riddoch came to the rescue with a story about a
-successful\/disastrous flight test.  We like these kinds of
-stories--to small and personal to turn up in the history books, but
+successful/disastrous flight test.  We like these kinds of
+stories\textemdash too small and personal to turn up in the history books, but
 they do pull you into real human experiences. There will be more of these.
@@
 %series of previous Timmins stories can be found through the same
-%website, starting with the December 2010 issue.
-
+%website, starting with the December 2010 issue
 \vspace{12pt}

### 2014-05-24 pnsac/newsletter/article_Feature1.tex
Updated files for May 2014 newsletter.

--- 2013-12-14 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{My Days at the Home of the Merlin Engine}
-\subtitle{Part 4}
-\author{Richard Lodge}
+\title{RAF Lightning with Over-Wing Tanks}
+%\subtitle{Part 4}
+\author{Jim Riddoch}
 
@@
 
-One of the most mysterious departments at Rolls-Royce was the
-Technical Publications Department. Word processing was unheard of in
-the 1960s and every time an engine modification was made, it resulted
-in much work documenting even a small change and distributing the
-updated documentation to all engine users around the world. I never
-spent any time working with the people in this department even though
-there was a complete building devoted to the work. Compared with the
-resources allocated to engine development, the technical publication
-side of the business was very small and it always remained a mystery
-to me.
+During the early 60's Britain's Defence Department decided that the
+new fleet of Lightning aircraft should have the capability to extend
+its current sortie of 22 minutes to permit ferry to the Middle
+East. In-flight refuelling had not worked successfully due to the
+offset probe positioned under one wing. Accordingly BAC (British
+Aircraft Corporation,Warton Division) was awarded a contract to
+develop extended wing tanks that could be jettisoned if necessary.
 
-Since retiring from my career as an accountant in Canada, I have
-become active on the actual restoration work of the North
-Star. Understandably, I have a special interest in working on the
-Merlin engines. Nobody now working on the restoration has ever worked
-on a Merlin when it was in service. It is therefore necessary at all
-times to refer to the various publications and manuals we have managed
-to accumulate for the North Star.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{jim_riddoch.eps}
+   %caption of the figure 
+   \caption*{\small \em Jim Riddoch.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:jim_riddoch.eps}
+\end{figure}
 
-It is only now, actually working on the engine and referring
-frequently to the exploded diagrams that I realize how complicated the
-work of the Technical Publications Department must have been. The men
-and women who worked there were largely unsung heroes in that it was
-very tedious and unglamorous work which was of great importance for
-the safe and trouble-free running of engines in service. I can
-remember that I would hear from time to time of the difficulties of
-ensuring that publication updates were actually delivered to many
-remote places in the world and the considerable logistical
-complications of ensuring that the packages arrived safely and
-furthermore that the updates were actually inserted in the relevant
-manuals, particularly as many of the engine users did not speak
-English. I cannot remember now if any of the documentation was
-translated into other languages although I have a recollection that we
-employed people who could translate into French and Spanish.
+%% \begin{footnotesize}
+%% [Image: {\normalfont\color{blue}\href{http://www.airteamimages.com/english-electric-lightning_XP693_united-kingdom---royal-air-force_43865.html}{English
+%%     Electric Lightning with over-wing tanks}}]
+%% \end{footnotesize}
 
-After a couple of years I was given a new job in the Long Range
-Planning Department. I was one of three accountants working in this
-office under a very senior accountant. Here we looked at the costings
-of the development work, the projected sales  and production costs on
-new engine designs. This was the time when the Spey was going into
-service and the the company had signed a fixed price contract for the
-RB211 engines for the new Lockheed 1011. This was a major project,
-championed by David Huddie the Chief Engineer. Computer spreadsheets
-did not exist at that time - pencil and columnar paper were the order
-of the day, together with efficient erasers. Long range forecasts of
-25 years required many yearly columns and each time we produced
-figures that the engineers did not like, we were told to rework the
-numbers until we produced an acceptable answer. We were not limited in
-the number of erasers we could use. The usual way of cutting costs was
-to reduce the forecast development and testing time. Of course this
-did not work in real life.
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Garry installing the intercooler on engine 3.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:Eng3Intercooler}
+%\end{figure}
 
-We were often dealing with the directors of R-R who lived on the top
-floor of the Nightingale Road offices. Many of them were engineers and
-they had little time for young bean counters. It was still a time of
-optimism for the company. The huge contract for the Lockheed engines
-had been signed and the company was still enjoying its reputation as
-the company whose Merlin engines had saved Britain in the air
-war. This was in 1965/6 and it was several years before our
-projections of doom resulted in the bankruptcy of the
-company. Fortunately I left the company in 1966 and was not involved
-in what must have been a horrible process as a symbol of British
-engineering excellence slowly went to the wall.
+Unfortunately the Lightning with its high speed wings had limited
+capacity for wing tip tanks and already with armament pylons under
+wing could not accommodate under-wing tanks, so the decision was made
+to develop over-wing tanks secured with pylons through the wing
+structure. These tanks would each carry 250 imperial gallons to
+provide initial fuel transfer after take off and be jettisoned once
+empty.
 
-After the chaos caused by the demise of Rolls-Royce Limited, the
-British government formed a new nationalized company to take over the
-remains of the old one. This has been very much a success story and
-the newly privatized Rolls-Royce is now a highly profitable and
-innovative company. I look back on my days at R-R with nostalgia but I
-now realize that much of the time we were all living in a fool's
-paradise.
+I was designated to be the Development Engineer to carry out the
+development and testing of this design about 1963/64. The design
+concept of these tanks had to include a balanced transfer system to
+maintain aircraft C of G during take off and climb out, this meant 5
+small interior sealed compartments pressurized with air to permit
+sequential transfer. Also an emergency fuel jettison system prior to
+tank wing ejection. As you can readily appreciate this was a very
+challenging design and even more demanding for testing and
+certification.
+
+Once the initial test tanks were built they were dispatched to
+Mechanical Test Department at Warton Lancashire, where I worked.
+Needless to say the RAF were keen to witness the development and
+testing of these over-wing tanks. A Wing Commander headed a team of
+three Senior Techs to keep track of our progress. Fortunately the
+Techs were very helpful and hands-on in working with us.  You can
+imagine the challenges we faced checking the transfer sequences at
+varying attitudes. Many a Saturday morning was spent testing and
+verifying these change in attitudes and minimizing fuel transfer loss,
+the end result being approximately 225 gallons of fuel transferred
+from each wing tank depending on aircraft attitude and roll.
+
+The next major design requirement was the emergency fuel jettison;
+this was to be accomplished by air pressure from centre compartment to
+rear of the tank where a jettisonable fairing was fastened by
+explosive bolts and attached by a cord to a bung on the end of the
+fuel ejection pipe. Talk about Heath Robinson device, this takes the
+biscuit! Suffice to report that on the ground test of this
+ill-conceived system with Chief Designer and my Chief Engineer and
+many observers with cameras, the rear fairing departed into oblivion
+and the cord broke, due to a miserable sub standard cord substituted
+instead of a parachute cord as specified. The bung remained firmly
+attached to the jettison pipe. It was about this time I first thought
+of emigrating to Canada, in fact anywhere outside Britain!
+
+Needless to say the cord was replaced, the ejection system proved
+satisfactory and the tanks were certified for flight test. This was to
+be done with one tank attached to the top of one wing and painted with
+black and white checkered squares for easy recognition when ejected. I
+always remember the Test Pilot, Peter Knight inquiring about the
+likely occurrence of the tank separating from the wing and hitting the
+tail of the aircraft. I assured him that ground testing of a model in
+the wind tunnel had worked, in fact the small tank had disappeared
+through the end of the tunnel and took some time to recover in the
+neighbouring field.
+
+To replicate the fuel system jettison some bright spark proposed
+filling the tank with coloured water for better visibility. The
+aircraft took off, circled Warton Aerodrome and made a low pass to
+jettison the coloured water over the attentive crowd including chiefs
+and dignitaries. Fortunately, the jettison of the tail fairing and
+bung release worked perfectly and the water emerged in a colourful
+shower.
+
+The weather gods not being with us, the wind blew all the water over
+the spectators covering their expensive clothing with a colourful dye!
+Finally, the tank was ejected safely over Morecombe Bay and eventually
+recovered and returned to BAC Warton amidst embarrassed and colourful
+faces. As one might expect, this completely successful test left the
+Wing Commander disgusted with the entire experiment. He cancelled the
+project.
+
+Any thoughts I had of joining the RAF completely disintegrated. Once
+again, immigration to Canada looked much more inviting.
 

### 2014-05-24 pnsac/newsletter/article_Feature2.tex
Updated files for May 2014 newsletter.

@@
 
-\title{The Whispering Giant}
-\author{Dave Lambeth}
+\title{Elise D\'{e}roche, Aviatrice}
+\author{Bruce Grant, Fl\^{a}neur}
 
@@
 
-Twinkle, twinkle, old North Star,\\
-With wrinkled skin and creaky
-spar,\\
-For years you’ve always been on
-sked,\\
-As many of our force you’ve sped\\
-to distant lands beyond the skies,\\
-you’ve never failed to see the rise\\
-of distant suns or moonrise glare.\\
-You’ve carried us with tender care\\
--back you came to take us home,\\
-from old Japan or France or Rome\\
-from England, Goose, and Kef,\\
-We cannot hear you.\\
-(‘cause we’re deaf)\\
+Wandering about in Paris recently I came upon a plaque, bolted to a
+wall at 61 Rue de la Verrerie in the 4th. Wandering about is my
+favourite activity in Paris. Let others wait in the long lines at the
+tourist sites (I've done all that) or plunder the shops in Faubourg
+Saint Honor\'{e} (I'll never do that); I wander about.
+
+So, who was this Elise D\'{e}roche a.k.a. Baronne Elisa Raymonde de
+Laroche, Aviatrice?  I took the picture and resolved to look her up at
+home.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{aviatrice.eps}
+   %caption of the figure 
+   %\caption*{\small \em Garry installing the intercooler on engine 3.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:aviatrice}
+\end{figure}
+
+Here's a brief outline of the story I encourage readers to look
+for her online.  Many sites relate her story though with some
+contradictions among the various accounts.
+
+Born into the working class family of a plumber, Elise D\'{e}roche
+became an artist and actress, took the stage name of Baronne Elisa
+Raymonde de Laroche. She also became the "amie intime" of the artist
+and aviator L\'{e}on Delagrange who introduced her to his friend Charles
+Voisin, a builder of airplanes.
+
+There is an interesting chronology that indicates a life full of
+tragedy, a life that she lived with intensity and audacity: L\'{e}on
+Delagrange died, January 1910 when the wing came off his Bl\'{e}riot
+airplane. Elise recovered quickly from her grief and appealed to her
+new ami intime Charles Voisin to teach her to fly. She won her pilot's
+licence in March 1910. A crash in July 1910 put her in hospital with
+18 fractures. She recovered and returned to flying.
+
+In a car crash in September 1912, Charles Voisin was killed but
+Elise survived. She maried the pilot Jacques Vial in February of
+1915, her first marriage. During the war, she was not allowed to
+fly but served in the military as a driver. In 1918 her only
+child Andr\'{e} (with L\'{e}on Delagrange) died aged 15 of the Spanish
+Flu. July 1919, she was killed in a plane crash while training to
+become a test pilot. Audace? Mais oui!
+
+La Baronne Elisa Raymonde de Laroche is buried in Cimeti\`{e}re
+P\`{e}re Lachaise in Paris.
+
+\vspace{10mm}
+
+\begin{centering}
+My candle burns at both ends;\\
+It will not last the night;\\
+But ah, my foes, and oh, my friends --\\
+It gives a lovely light!\\
+\vspace{2mm}
+\begin{footnotesize}
+Edna St. Vincent Millay
+\end{footnotesize}
+
+\end{centering}
 

### 2014-05-24 pnsac/newsletter/article_Feature3.tex
Updated files for May 2014 newsletter.

@@
 
-\title{Troop Seats and Extras}
-\author{Nelson Smith}
+\title{The Whispering Giant}
+\author{Dave Lambeth}
 
@@
 
-\section{Troop Seats}
-\label{troopseats}
-
-Seat back assemblies are completed for the troop seats. Thirty eight
-units were produced. The assembling of the backs to the platforms to
-produce triple and quad seats will be the next step. This will be done
-over the winter.
+\begin{centering}
+Twinkle, twinkle, old North Star,\\
+With wrinkled skin and creaky
+spar,\\
+For years you've always been on
+sked,\\
+As many of our force you've sped\\
+to distant lands beyond the skies,\\
+you've never failed to see the rise\\
+of distant suns or moonrise glare.\\
+You've carried us with tender care\\
+-- back you came to take us home,\\
+from old Japan or France or Rome\\
+from England, Goose, and Kef.\\
+We cannot hear you.\\
+('cause we're deaf)\\
+\end{centering}
 
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{troopseats.eps}
+   \includegraphics[scale=0.5]{jim_jung_2.eps}
    %caption of the figure 
-   \caption*{\small \em New troop seats.}
+   \caption*{\small \em A Whispering Giant (North Star 17515) at
+     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:troopseats}
+   \label{fig:jim_jung}
 \end{figure}
-
-The prototype is progressing well. As previously stated, measurements
-are a rarity with these projects and sometimes it takes longer to
-figure out a plan than to do the job. A guesstimate for completion:
-early December.
-
-\section{Other Projects}
-\label{otherporjects}
-
-Proctective covers for the probes which protrude under the belly of
-the North Star were made and installed. The foam pads previously
-installed had done their time.
-
-The First Officer's protective seat covers have been manufactured and
-installed. This addition was required to shield the seat covers from
-UV rays, dust and stains. Similar covers will be made for the
-Captain's chair during the winter months.
 

### 2014-05-24 pnsac/newsletter/article_PresReport.tex
Updated files for May 2014 newsletter.

--- 2013-09-19 pnsac/newsletter/article_PresReport.tex
@@
 
-Canada Day is now behind us.  We had an excellent day with a good
-number of our volunteers showing off the North Star to the general
-public. Judging by the activity around the plane all the day, we have
-now developed a display format which is interesting to visitors.
-Canada Day was the only day this year when the plane was on display to
-the general public.
+Since my last Notes from the President in December 2013, restoration
+work on the North Star has mostly been suspended. The Museum has
+needed extra working space so that exhibits could be made ready for
+the commemoration this year of the centenary of the start of the First
+World War and the 70th anniversary on June 6 of D-Day in the Second
+World War. These preparations have resulted in our normal working
+areas being unavailable for North Star restoration work.
 
@@
   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{canadaday_visitor.eps}
+  \includegraphics[scale=0.5]{richard_mtg.eps}
   %caption of the figure 
-   \caption*{\small \em Showing off the North Star on Canada Day -- an
-   impressed visitor examines the cockpit.}
+  \caption*{\small \em Richard at the Quarterly Meeting in April.}
   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:canadaday}
+  \label{fig:richard_mtg}
 \end{figure}
 
-On June 27, in co-operation with the Museum, who hosted the event, we
-showed off the plane to a group of donors attending the Donor
-Appreciation Evening.  This was a successful evening which is likely
-to be repeated.
+During the shut down period some limited work has been continuing. Two
+volunteers have been working in the cockpit of the plane and two other
+volunteers have been doing some work on \#3 engine reassembly.
 
-In my last Notes from the President, I referred to plans for
-celebrating the 10th anniversary of the founding of the Project North
-Star Association.  This should have taken place around June 24.
-Unfortunately, the event passed without us being able mark the
-day. This was mainly due to situations beyond our control.
+While the restoration work has been on hold, we have arranged three
+monthly Volunteer Lunches at the Museum to enable the people who
+normally work on the plane to meet and socialize. Although we would
+all like to be working on the plane, we understand why the Museum has
+had to ask us to temporarily suspend our activities. The Volunteer
+Lunches are a way of keeping the team together and up-to-date on what
+is happening.
 
-Our 2013 AGM was held on June 8, and all the existing directors were
-re-elected.  This will enable the directors, who have now become
-experienced, to concentrate on long-term matters.
+At the last meeting of the PNSAC Board of Directors we agreed to
+reschedule the Annual General Meeting from June each year to
+September. It has always been difficult to find a good weekend in June
+to hold the AGM and it is felt that a middle of September date would
+be much better. Our corporate secretary, Roger Button, is aware of
+this proposed change and has confirmed that the existing Board of
+Directors can extend its term until the new AGM date.
 
-The actual work on the restoration of the plane is going very well, as
-reported elsewhere in this newsletter. The major issue facing the
-Board in the next year will be to consider how to secure the future of
-PNSAC, both in terms of volunteers and the funding of restoration
-work.
+Although there has been little restoration work done since December,
+we are working with the Museum to get the project back on track again
+in early June. We are also working to enable us to recruit new
+volunteers to replace the several people who for various reasons have
+had to give up active work.
 
-We have now reached a point where it is necessary to update our
-Memorandum of Understanding (MOU) with the Museum and our work plan
-for completing the restoration of the plane.  Our volunteers have
-already worked over 60,000 hours on the project and we expect to work
-a similar number of hours before the restoration is completed.
+One of the recent discussion areas has been in connection with
+long-term funding of the restoration project. Everybody is aware that
+funding from the Federal Government has been substantially reduced. We
+have accepted that major  expenditures will have to be met from funds
+raised by donations from friends of the Museum and members of our
+Association. This is a significant challenge since everybody receives
+many appeals throughout the year for donations from many different
+deserving organizations.
 
-There have been several informal meetings to discuss our future
-operations and the opportunities available for new volunteers.  Within
-the next few weeks we are expecting to meet with the Director General
-of the Aviation Museum and members of his staff to work on the MOU
-update.
-
-Later this month we are meeting with senior management of the Museum
-Foundation, the body charged with corporate fundraising for the Canada
-Science and Technology Museums Corporation, to discuss how PNSAC can
-work with the Foundation to raise funds for the North Star restoration
-and the museums in general.
-
-With the initiatives above, we are hoping in the next few months to
-consolidate our position as an important organization working with the
-Aviation Museum.
+During the next few months we hope to develop a revised restoration
+work plan which will enable us to estimate the funds needed to
+complete the project. We will then be able to target appeals for funds
+to carry out specific parts of the restoration in the same way as we
+did for our successful appeal for funds to make troop seats for the
+interior of the plane.
 

### 2014-05-24 pnsac/newsletter/article_ProgReport.tex
Updated files for May 2014 newsletter.

@@
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
-   %caption of the figure 
-   \caption*{\small \em Garry installing the intercooler on engine 3.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:Eng3Intercooler}
-\end{figure}
-
 \section{Cockpit}
@@
 officer seat was installed last year) then the railings and
-safety board were installed behind the seats.  Next we installed
+safety board were installed behind the seats.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
+   %caption of the figure 
+   \caption*{\small \em Forward cockpit -- complete.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:fwd_cockpit_complete}
+\end{figure}
+
+Next we installed
 the radio officer and navigator tables, and the seats for these
@@
 modifications to the seat rails were necessary to allow the seat
-to slide.  Finally the flight engineer’s jump seat was attached
+to slide.  Finally the flight engineer's jump seat was attached
 to the railing.  With five crew seats installed, the cockpit is
 very cramped.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{radio_position.eps}
+   %caption of the figure 
+   \caption*{\small \em Radio Officer's position showing the morse
+     code key.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:radio_position}
+\end{figure}
 
@@
 
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{cockpit_curtain.eps}
+   %caption of the figure 
+   \caption*{\small \em Finishing touches -- the cockpit curtain.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:cockpit_curtain}
+\end{figure}
+
 \section{Crew Lounge, Galley,  and Forward Washroom}
@@
 The floor under this section was repaired by removing a section
-of badly corroded “I” beam and fabricating a new one from several
+of badly corroded I-beam and fabricating a new one from several
 pieces of aluminum which were riveted in place to support the
@@
 main cabin undergoes restoration over the next two years.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{new_galley_floor.eps}
+   %caption of the figure 
+   \caption*{\small \em New galley and floor.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:new_galley_floor}
+\end{figure}
 

### 2015-04-11 pnsac/newsletter/article_Feature1.tex
Prep for April 2015 issue.

--- 2013-12-14 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{My Days at the Home of the Merlin Engine}
-\subtitle{Part 4}
-\author{Richard Lodge}
+\title{Operation Hawk}
+%\subtitle{Part 4}
+\author{Hugh A. Halladay, Legion Magazine}
+
+\textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
+  Sqn, RCAF.}
 
@@
 
-One of the most mysterious departments at Rolls-Royce was the
-Technical Publications Department. Word processing was unheard of in
-the 1960s and every time an engine modification was made, it resulted
-in much work documenting even a small change and distributing the
-updated documentation to all engine users around the world. I never
-spent any time working with the people in this department even though
-there was a complete building devoted to the work. Compared with the
-resources allocated to engine development, the technical publication
-side of the business was very small and it always remained a mystery
-to me.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{jim_riddoch.eps}
+   %caption of the figure 
+   \caption*{\small \em Jim Riddoch.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:jim_riddoch.eps}
+\end{figure}
 
-Since retiring from my career as an accountant in Canada, I have
-become active on the actual restoration work of the North
-Star. Understandably, I have a special interest in working on the
-Merlin engines. Nobody now working on the restoration has ever worked
-on a Merlin when it was in service. It is therefore necessary at all
-times to refer to the various publications and manuals we have managed
-to accumulate for the North Star.
+In 1947, while Prime Minister Mackenzie King was in London, his
+Minister of Foreign Affairs (and chosen suc-cessor), Louis St-Laurent,
+consented to Canadian participation on the United Nations Temporary
+Commission of Korea (UNTOK). On his return to Canada, King furiously
+confronted St-Laurent. He wanted no part---even at the fringes—of
+Korean affairs, which he considered dangerous.
 
-It is only now, actually working on the engine and referring
-frequently to the exploded diagrams that I realize how complicated the
-work of the Technical Publications Department must have been. The men
-and women who worked there were largely unsung heroes in that it was
-very tedious and unglamorous work which was of great importance for
-the safe and trouble-free running of engines in service. I can
-remember that I would hear from time to time of the difficulties of
-ensuring that publication updates were actually delivered to many
-remote places in the world and the considerable logistical
-complications of ensuring that the packages arrived safely and
-furthermore that the updates were actually inserted in the relevant
-manuals, particularly as many of the engine users did not speak
-English. I cannot remember now if any of the documentation was
-translated into other languages although I have a recollection that we
-employed people who could translate into French and Spanish.
+St-Laurent, in turn, threatened to resign. Ultimately, the old prime
+minister backed down. The incident marked a phase in the changing of
+the guard in Ottawa, and a movement towards increasing Canadian
+involvement in international affairs.
 
-After a couple of years I was given a new job in the Long Range
-Planning Department. I was one of three accountants working in this
-office under a very senior accountant. Here we looked at the costings
-of the development work, the projected sales  and production costs on
-new engine designs. This was the time when the Spey was going into
-service and the the company had signed a fixed price contract for the
-RB211 engines for the new Lockheed 1011. This was a major project,
-championed by David Huddie the Chief Engineer. Computer spreadsheets
-did not exist at that time - pencil and columnar paper were the order
-of the day, together with efficient erasers. Long range forecasts of
-25 years required many yearly columns and each time we produced
-figures that the engineers did not like, we were told to rework the
-numbers until we produced an acceptable answer. We were not limited in
-the number of erasers we could use. The usual way of cutting costs was
-to reduce the forecast development and testing time. Of course this
-did not work in real life.
+Mackenzie King died on July 22, 1950. As he lay in state in the
+Parliament Buildings, six North Star aircraft of No. 426 Squadron flew
+over, saluting his passing as they were en route to Tacoma,
+Washington, to commence a Royal Canadian Air Force transoceanic
+airlift in sup-port of United Nations forces fighting in Korea.
 
-We were often dealing with the directors of R-R who lived on the top
-floor of the Nightingale Road offices. Many of them were engineers and
-they had little time for young bean counters. It was still a time of
-optimism for the company. The huge contract for the Lockheed engines
-had been signed and the company was still enjoying its reputation as
-the company whose Merlin engines had saved Britain in the air
-war. This was in 1965/6 and it was several years before our
-projections of doom resulted in the bankruptcy of the
-company. Fortunately I left the company in 1966 and was not involved
-in what must have been a horrible process as a symbol of British
-engineering excellence slowly went to the wall.
+It is not known if the former prime minister was spin-ning in his
+coffin.
 
-After the chaos caused by the demise of Rolls-Royce Limited, the
-British government formed a new nationalized company to take over the
-remains of the old one. This has been very much a success story and
-the newly privatized Rolls-Royce is now a highly profitable and
-innovative company. I look back on my days at R-R with nostalgia but I
-now realize that much of the time we were all living in a fool's
-paradise.
+Although the RCAF had considerable air transport experience during
+the Second World War, only the work of No. 168 (Heavy Transport)
+Sqdn., delivering service mail to Europe, constituted a sustained
+transatlantic operation. With the retirement of its Fortress and
+Liberator aircraft, which were converted from bombers to freighters,
+the RCAF briefly lost its long-range transport capability. This gap
+was soon filled by the Canadair North Star, built in Montreal for the
+RCAF, TCA (Trans-Canada Airlines and later Air Canada) and other civil
+carriers.
+
+The North Star was basically a Douglas DC-4, but with Rolls-Royce
+Merlin in-line engines instead of Pratt-andWhitney radials. It was a
+dependable aircraft, but generations of passengers and crew would
+remember the tremendous noise and vibration in flight. The full story
+of this is related by Larry Milberry in his 1982 book The Canadair
+North Star.
+
+Deliveries of the new aircraft to the RCAF were delayed when
+priority was given to TCA to help it establish a Canadian presence in
+transatlantic commercial flying. Crews of No. 426 (Thunderbird)
+Sqdn. finally began North Star training early in 1948, and actual
+operations on the type commenced that summer.
+
+While the RCAF did not participate in the July 1948 to May 1949 Berlin
+Airlift, a few individual aircrew did while on exchange duties with
+the Royal Air Force and United States Air Force. As it happened, the
+airlift seriously degraded the USAF Military Air Transport Service
+(MATS) after flour and coal dust penetrated the seams of every
+participating aircraft. When it was next marshalled to meet the Korean
+crisis, MATS needed all the help it could get—and that was forthcoming
+when the Canadian government committed No. 426 Sqdn.—its only
+long-range transport unit—to the Korean campaign.
+
+In recounting the RCAF’s transpacific operational story, one must pay
+tribute to Laurence Motiuk, author of two monumental volumes, namely
+Thunderbirds at War: Diary of a Bomber Squadron, and Thunderbirds for
+Peace: Diary of a Transport Squadron, which set out the history of
+No. 426 Sqdn. The latter work was particularly difficult to write,
+given that in the 1950s, RCAF units often submitted incomplete
+historical narratives and in some instances neglected to prepare any
+at all.
+
+Motiuk’s account of the postwar squadron was drawn from numerous
+unofficial records, including log books as well as the sketchy unit
+reports.
+
+Almost from the moment that North Korea launched its invasion of South
+Korea on June 25, 1950, No. 426 Sqdn. personnel were preparing for
+overseas operations. If Canada was going to be involved, that unit was
+certain to be mobilized. Cabinet approval came through on July 19,
+1950. The squadron cancelled all further transport work with the
+exception of vital northern resupply.
+
+Deployment to McChord Air Force Base, Tacoma, Washington, began on the
+25th. Six North Stars, 12 crews and 185 ground personnel were
+dispatched. Twothirds of the aircrew were veterans of the Second
+World War; the remainder were men who had joined the RCAF after
+1947. Three of the aircraft left Tacoma on the 27th, arriving at
+Haneda in Tokyo two days later. Operation Hawk was under way.
+
+It might be noted that No. 426 Sqdn. was not alone in supporting
+operations in Korea. Historian Carl Mills reminds us that Canadian
+Pacific Airlines provided charter services to MATS from August 1950 to
+March 1955, conducting 703 transpacific flights. Unlike the RCAF
+operation, these involved passengers only. They lost one aircraft
+(CF-CPC), a DC-4 with seven crew members and 31 passengers on July 2,
+1951. The aircraft, outbound from Vancouver, disappeared near
+Yakutat, Alaska; no trace was ever found.
+
+The North Stars followed a challenging route. From McChord to
+Elmendorf Air Force Base in Alaska was 1,490 miles. This was followed
+by a 1,537-mile run to Shemya Air Force Base in the Aleutian
+Islands. Next came the 2,104mile flight to Haneda in Tokyo, with the
+aircraft making landfall at Matsushima (Honshu, Japan).
+
+The Shemya-Matsushima leg was particularly challenging; care had to
+be taken not to drift into Russian airspace, a task complicated by
+Russian efforts to jam radio and navigational aids. Depending on the
+season, the aircraft might return by the same route, or take a more
+southerly central Pacific track from Haneda to Wake Island to Hickam
+Air Force Base in Honolulu. From there the aircraft would journey to
+Travis Air Force Base in San Francisco, and then on to McChord.
+
+Initially it was hoped the unit could dispatch one sortie per day, but
+this proved beyond the resources of a six-plane outfit. Eventually,
+No. 426 found it could send off five aircraft per week. Not
+surprisingly, its peak activity coincided with the buildup of United
+Nations forces in Korea and their maintenance during the period of
+mobile warfare.
+
+Between July 28 and Dec. 31, 1950, the squadron dispatched 123
+aircraft across the Pacific. In the whole of 1951 it was 193 missions,
+but in 1952 it dropped to 133. Not only were the Thunderbirds
+returning to more traditional duties in Canada, they were also
+assuming the job of providing transport, including personnel
+rotation, to Canada’s growing land and air presence in Europe. In
+1953—the last full year of Hawk—it was 85 missions.
+
+The final Hawk mission departed McChord on May 31, 1954, and ended at
+the squadron’s home base at Dorval on June 9, 1954. As it wound up, an
+RCAF dispatch reported that No. 426 Sqdn. had flown 599 round
+trips—four of them directly to Korea itself—logged 34,000 flying
+hours, carried 13,000 personnel, and airlifted 3,500 tons of
+freight. The report erred on two points. Six flights (not four) had
+gone directly to Korea. The other discrepancy was that there were
+only 584 missions; for some unknown reason the numbered listing of
+flights had jumped from “439” on Nov. 26, 1952, to “455” on Nov. 28,
+1952; the inaccurate 1954 press release continues to be quoted, ad
+nauseam, to this day.
+
+Operation Hawk was accomplished with no fatalities but some close
+calls and more than a few incidents. On Sept. 15, 1950, Wing Commander
+C.H. Mussells, Commanding Officer of No. 426 Sqdn., departed McChord
+for Elmendorf with American soldiers and anti-tank ammunition. Both
+were urgently needed in Korea, so the aircraft had been permitted an
+overload clearance.
+
+Three hours after takeoff, one of the Merlin engines began overheating
+through a coolant leak. Mussells shut down the engine, feathered the
+propeller, and prepared to return to McChord. Then a second engine on
+the same side overheated. It was also shut down. There was now no
+question about getting to McChord; Mussells prepared to make an
+emergency landing at Sandspit (Moresby Island).
+
+To lighten the aircraft, Mussells dumped excess fuel. Some passengers
+mistook the misty gasoline trail for smoke, panicked and crowded to
+the back of the fuselage, as if that would have offered safety. This
+complicated the process of trimming the aircraft for landing. A
+crewman ordered everyone back to their seats, but the soldiers could
+not have been reassured as the North Star flew over a crashed USAF
+aircraft, its tail still sticking out of the water, during the final
+approach into Sandspit. Happily, Mussells landed safely.
+
+On April 19, 1951, a North Star piloted by Flight Lieutenant J.A. Watt
+was on a route-training flight to an unfamiliar field—Ashiya—located
+between Osaka and Kobe, Japan. He was cleared by Ashiya tower to
+descend from 4,000 to 3,000 feet. At 3,400 feet the aircraft hit trees
+atop a hill that was not even indicated on the airfield map. No one
+was injured but the aircraft sustained considerable damage to its
+nose, oil cooler and exterior radio aerials. The pitot head was also
+wiped out, and with it the airspeed indicator. The aircraft also lost
+the use of an engine.
+
+Fortunately, the passengers included three of the most experienced
+North Star captains in the RCAF: Wing Cmdr. Mussells, Wing
+Cmdr. J.K. MacDonald, and Flying Officer Robert Edwards. After
+assessing the damage, Mussells ordered MacDonald and Edwards to com-
+plete the landing, which they did.
+
+On the night of Dec. 27, 1953, Squadron Leader E.L. Hare was homeward
+bound from Japan in North Star 17516, bound for Shemya. The weather
+was bad and the ceiling minimal. Visibility was a half mile with a
+savage crosswind blowing at a 90-degree angle. The runway surface was
+covered with wet snow; braking conditions were poor. On the third
+landing attempt, everything came together in the worst possible
+way. The North Star was blown off the runway into a gully. Everyone
+survived, but 17516 was a total “write-off”–the only RCAF aircraft
+lost in the course of Operation Hawk.
+
+Canadian military policy with respect to honours and awards was very
+restrictive from 1946 onwards, but rules were relaxed during the
+Korean War. Personnel involved in Hawk were awarded one Officer of the
+Order of the British Empire (OBE), one Member of the Order of the
+British Empire (MBE), four Air Force Crosses, two Air Force Medals and
+13 Queen’s Commendations. Most of these were to aircrew personnel, but
+Sqdn. Ldr. W.H. Lord was granted an MBE for his work in establishing
+and supervising the maintenance detachments that serviced the North
+Stars wherever they alighted along the 11,000 mile route.
 

### 2015-04-11 pnsac/newsletter/article_ProgReport.tex
Prep for April 2015 issue.

--- 2013-12-14 pnsac/newsletter/article_ProgReport.tex
@@
 \title{Project Manager's Progress Report}
-\subtitle{December 2013}
+\subtitle{April 2015}
 \author{Bruce Gemmill}
@@
 
-\section{Nr 3 Engine}
-\label{sec:engines_3}
+Since our last Project Report, our project has gone through much
+upheaval, and work has been delayed for a number of reasons.  However,
+I can report now that we are back on track and making steady progress
+on several parts of the North Star. This report tracks progress since
+April 2014.
 
-Engine 3 is nearing completion.  The supercharger and intercooler
-preheat assemblies have been installed, along with the propeller
-reduction gearbox.  While this work was being carried out by the
-engine crew, the remainder of the volunteer work force busied
-themselves completing the numerous cowl panels, pipes and hoses and
-other pieces needed to complete the engine.  Two items that seemed to
-take forever were the large steel exhaust shrouds.  A lot of hammering
-was needed to pound out years of dents, then numerous rivets and all
-the Dzus fasteners and springs needed to be replaced.  The complex
-shape of the exhaust shroud made this challenging work.  Further
-delaying completion was the search for suitable clear coat to protect
-the steel from corrosion.  The old clearcoat we used successfully on
-engines 1 and 2 is no longer available.  Mike Irvin finally located a
-suitable replacement, and the shrouds were clear coated and serial
-numbers were stencilled on.
+\section{Restoration Shop Shutdown}
+\label{sec:restoration_shop}
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
-   %caption of the figure 
-   \caption*{\small \em Garry installing the intercooler on engine 3.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:Eng3Intercooler}
-\end{figure}
+Engine \#3 was in the final stages of assembly when we had to shut down
+our project at the end of December 2013, to make way for other museum
+work, in preparation for events to mark the 100th anniversary of the
+beginning of the First World War in 2014.  In particular, new exhibits
+in the museum required all of the available floor space in the shop,
+as well as full time work by all museum conservation staff, meaning
+the volunteer crew could not be supervised as usual.
 
-The auxiliary gearbox was disassembled, and significant corrosion was
-found on the bearings inside.  Bearings had to be ordered from
-England.  Meanwhile, the gearbox was cleaned and painted.  At the same
-time, the large DC generator was dismantled, repainted and
-reassembled, as well as the air pump and the tachometer generator and
-propeller synchronizer.  Once the new bearings were installed, the
-gearbox was completed and the generator and air pump were attached.
-Finally, the gearbox was installed on the firewall, which is now
-finished, awaiting the installation of engine \#3.
+However, two volunteers, myself and Peter Trobridge, were allowed to
+work on the aircraft during the shutdown between January and July
+2014.  The focus was on completing the assembly of the cockpit and
+crew lounge.  More on this later.
 
-\section{Crew Lounge, Galley,  and Forward Washroom}
+\section{Engine Frame}
+\label{sec:engine_frame}
+
+Nr 3 engine frame was completed after the installation of the engine
+and supercharger.  Fitting the many cowl panels proved difficult, due
+to the extensive rework that many of these items went through, causing
+some alignment difficulties when fitting these panels onto the engine
+frame.  This work was done after work resumed in July 2014, and the
+completed engine was successfully installed on the aircraft last
+summer.
+
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Forward cockpit -- complete.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:fwd_cockpit_complete}
+%\end{figure}
+
+\section{Nr 4 Engine}
+\label{sec:engine_4}
+
+The last engine to undergo restoration was removed in July and work
+immediately began in the engine shop to disassemble the engine frame
+and move the engine to the engine stand for restoration.  All cowl
+panels, pipes, hoses and accessories have been removed and stored
+until space is available to restore these items.  The priority has
+been to disassemble the engine, which was completed in October, and
+restoration and reassembly of engine 4 is now well underway.  The
+refurbished crankshaft has been installed in the engine block, and the
+connecting rods and pistons cleaned and polished.  Work has now begun
+on the more complex cylinder heads and valve train assemblies.
+
+Normally, work would also be underway to restore the engine frame, but
+due to a lack of space, this will be delayed until the engine is back
+together and can be moved out of the engine shop.  Some cleaning and
+paint stripping has been done, but further work on the frame will
+likely have to wait until more space is available for full restoration
+of this large item.  This work will likely delay the completion of Nr
+4 engine until 2016.
+
+\section{Cocpit, Crew Lounge, Galley,  and Forward Washroom}
 \label{crewlounge}
 
-Work has slowed on the forward section of the aircraft.  The repair
-needed under the floor of the crew lounge is on hold until a suitable
-piece of extruded aluminum can be found.  Until this is complete, the
-new cushion flooring cannot be installed. This has also delayed the
-installation of all of the crew lounge accessories, such as the table,
-seats and secure storage bin, all of which are complete but waiting in
-storage.
+Since our last report, the crew lounge and galley equipment was fully
+restored.  A new wall was installed in the galley, as the old wall
+contained asbestos and needed to be removed for safety reasons.  The
+galley cupboards were badly corroded, so a new set was built, with
+only the original doors remaining.  This was installed in the galley,
+along with the auxiliary hydraulic reservoir, after two sections of
+the floor were patched to repair corrosion damage.  New flooring was
+cut and glued in the crew lounge and washroom, then the washroom
+equipment was installed, including the toilet and washroom door, and
+the vanity sink and mirror.
 
-\section{Fuselage and Empennage}
-\label{empenage}
+The five crew seats that had been removed early in the project were
+finally installed in the cockpit, after installing the radio operator
+and navigator tables, and the appropriate seat rails.  We also
+installed new cockpit curtains, and completed these crew positions by
+installing a donated headset and several microphones, along with a
+Morse code key at the radio position.
 
-The forward and rear belly compartments are now complete.  All the
-frame pieces and hardware needed for the troop seats have been made,
-and work is progressing on the seats and backs.  One section has
-undergone a trial fit, and all pieces seem to fit perfectly.
+The seat rails and seats for the crew lounge were installed, along
+with the crew table and secure stowage case.  The crew bed was
+installed above the crew seats, and then the curtain rod with a new
+set of curtains was installed to complete the crew lounge.  The access
+door to the crew area has not been fully restored, but it was
+re-installed with a new lock to secure the forward portion of the
+aircraft.  The cockpit, crew lounge and forward washroom are now
+complete.  This is a significant milestone for the project, and has
+been very popular with select museum visitors who have been allowed
+access.
 
-\section{Planned Restoration Work 2013/14}
+\section{Fuselage and Main Cabin}
+\label{sec:main_cabin}
+
+The baggage compartments are now fully restored, and the refurbished
+battery elevators were installed just aft of the nose wheel.  Work has
+now begun on the main cabin.  This started with removal of various
+fittings along the starboard side of the cabin, then removing the wood
+paneling below the windows, and then removing the wood cargo floor
+panels to expose the metal sub floor.  Most of the side panels and
+some floor panels are badly damaged and will need to be replaced.  The
+fittings for patient litters and troop seats were also removed.
+Finally, the starboard windows were removed.  Once all the fittings
+were removed, work began on repairing corrosion damage along the
+exposed ribs and sub floor.
+
+The rear washroom was also dismantled, after removing the toilet and
+fittings.  The floor under the toilet was badly corroded, and the old
+floor panel was removed.  This required drilling out approximately
+1000 rivets!  A new panel has been made and will be installed shortly.
+Portions of the corroded rear bulkhead were also removed, so that new
+panels can be installed.  Some items were also removed from the tail
+section, to allow for thorough cleaning and spot priming bare metal
+where necessary.  During this work damage was discovered on the torque
+tube that controls the movement of the rudder.  This was unexpected,
+and will require removal of the tail cone to allow for a new torque
+tube to be fitted.
+
+Several badly corroded sections of the starboard side of the cabin
+have been repaired, and work continues with corrosion repairs and
+other cleanup while the aircraft is inside.  Part of the forward cargo
+door has been removed for restoration.  The paratroop door insert is
+complete, and the remaining portion of the forward cargo door is being
+stripped and repaired prior to painting.  It is likely this will have
+to be re-installed on the aircraft before all painting and restoration
+is complete, so as not to delay it’s move outside.
+
+A major effort was launched to restore the large heater ducts removed
+from the main cabin.  These were covered with formed mica insulation
+sections, which were removed earlier.  Many of these sections fell
+apart and could not be reused.  After many repairs to the metal
+ductwork, new sections of mica were formed in a jig and glued to the
+ducts.  These duct sections are in storage waiting for completion of
+the main cabin so they can be installed on the ceiling.
+
+Work has now begun on the port windows and wall sections.  We hope to
+complete the windows soon, as these must be installed before the
+aircraft can be moved outside.
+
+\section{Cabin Liners and Troop Seats}
+\label{sec:liners_troop_seats}
+
+The original cabin liners were removed from the aircraft earlier in
+the project.  These had been washed and stored until repair work could
+be done.  All liners were unrolled, photographed and assessed for
+damage.  Some liners will be repaired, while others, particularly the
+cargo door liners, need to be replaced because the fabric is too badly
+deteriorated.  A sewing station has been set up in the storage hangar,
+and several of the new liners have been completed.  These will need to
+have new fasteners installed, then painted to match the original
+liners.
+
+The North Star was originally fitted with troop seats down both sides
+of the main cabin.  These could be stowed against the wall to make
+room for cargo, or unrolled and set up to carry military personnel.
+The original seats were removed by the RCAF, so a decision was made to
+make a new set, using funds donated by Project North Star members.
+Seat fabric, steel and aluminum tubing was ordered, and the troop seat
+sections were fabricated, based on available drawings.  All seat
+sections have been completed, and have been stored until the main
+cabin is completed.
+
+\section{Planned Restoration Work---2015}
 \label{sec:plannedwork}
 
-The museum has advised all North Star volunteers that our work must be
-put on hold for several months while a major exhibit for the 100th
-anniversary of the beginning of the First World War is prepared.  Our
-work space in the restoration hangar is needed to assemble and prepare
-several aircraft that will be added to the display space in March.
-Until then our work on the aircraft is suspended, but we don't expect
-this delay to have a serious impact on the overall project.  We look
-forward to resuming work in the Spring.
+Over the next year, we will continue to work on Engine 4 and the main
+cabin and tail section.  The aircraft will be moved outside this
+summer for some interior painting.  This means the windows and doors
+will need to be installed for weatherproofing and security.  During
+this time we also hope to begin fabricating new wall and floor
+sections.  One other work item that may be started is to clean and
+restore the four engine nacelles.  Pipes and fittings need to be
+removed, followed by a thorough cleaning and a fresh coat of paint.
+This work may be delayed until qualified volunteers become available.
 

### 2016-01-16 pnsac/newsletter/article_Feature1.tex
Adding files for April 2015 newsletter.

@@
 
-\textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
-  Sqn, RCAF.}
-
 \maketitle
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{jim_riddoch.eps}
-   %caption of the figure 
-   \caption*{\small \em Jim Riddoch.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:jim_riddoch.eps}
-\end{figure}
+\textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
+  Sqn, RCAF.}\\
 
 In 1947, while Prime Minister Mackenzie King was in London, his
-Minister of Foreign Affairs (and chosen suc-cessor), Louis St-Laurent,
+Minister of Foreign Affairs (and chosen successor), Louis St-Laurent,
 consented to Canadian participation on the United Nations Temporary
 Commission of Korea (UNTOK). On his return to Canada, King furiously
-confronted St-Laurent. He wanted no part---even at the fringes—of
+confronted St-Laurent. He wanted no part---even at the fringes of
 Korean affairs, which he considered dangerous.
@@
 Washington, to commence a Royal Canadian Air Force transoceanic
-airlift in sup-port of United Nations forces fighting in Korea.
+airlift in support of United Nations forces fighting in Korea.
 
-It is not known if the former prime minister was spin-ning in his
+It is not known if the former prime minister was spinning in his
 coffin.
+
+\begin{figure*}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{op-hawk-one.jpg}
+   %caption of the figure 
+   %\caption*{\small \em Jim Riddoch.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:op-hawk-one.eps}
+\end{figure*}
 
@@
 The North Star was basically a Douglas DC-4, but with Rolls-Royce
-Merlin in-line engines instead of Pratt-andWhitney radials. It was a
+Merlin in-line engines instead of Pratt-and-Whitney radials. It was a
 dependable aircraft, but generations of passengers and crew would
@@
 participating aircraft. When it was next marshalled to meet the Korean
-crisis, MATS needed all the help it could get—and that was forthcoming
-when the Canadian government committed No. 426 Sqdn.—its only
-long-range transport unit—to the Korean campaign.
+crisis, MATS needed all the help it could get---and that was forthcoming
+when the Canadian government committed No. 426 Sqdn.---its only
+long-range transport unit---to the Korean campaign.
 
-In recounting the RCAF’s transpacific operational story, one must pay
+In recounting the RCAF's transpacific operational story, one must pay
 tribute to Laurence Motiuk, author of two monumental volumes, namely
@@
 
-Motiuk’s account of the postwar squadron was drawn from numerous
+Motiuk's account of the postwar squadron was drawn from numerous
 unofficial records, including log books as well as the sketchy unit
@@
 25th. Six North Stars, 12 crews and 185 ground personnel were
-dispatched. Twothirds of the aircrew were veterans of the Second
+dispatched. Two-thirds of the aircrew were veterans of the Second
 World War; the remainder were men who had joined the RCAF after
@@
 Haneda in Tokyo two days later. Operation Hawk was under way.
+
+\begin{figure*}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{op-hawk-two.jpg}
+   %caption of the figure 
+   %\caption*{\small \em Jim Riddoch.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:op-hawk-two.eps}
+\end{figure*}
 
@@
 by a 1,537-mile run to Shemya Air Force Base in the Aleutian
-Islands. Next came the 2,104mile flight to Haneda in Tokyo, with the
+Islands. Next came the 2,104 mile flight to Haneda in Tokyo, with the
 aircraft making landfall at Matsushima (Honshu, Japan).
@@
 assuming the job of providing transport, including personnel
-rotation, to Canada’s growing land and air presence in Europe. In
-1953—the last full year of Hawk—it was 85 missions.
+rotation, to Canada's growing land and air presence in Europe. In
+1953---the last full year of Hawk---it was 85 missions.
 
 The final Hawk mission departed McChord on May 31, 1954, and ended at
-the squadron’s home base at Dorval on June 9, 1954. As it wound up, an
+the squadron's home base at Dorval on June 9, 1954. As it wound up, an
 RCAF dispatch reported that No. 426 Sqdn. had flown 599 round
-trips—four of them directly to Korea itself—logged 34,000 flying
+trips---four of them directly to Korea itself---logged 34,000 flying
 hours, carried 13,000 personnel, and airlifted 3,500 tons of
@@
 only 584 missions; for some unknown reason the numbered listing of
-flights had jumped from “439” on Nov. 26, 1952, to “455” on Nov. 28,
+flights had jumped from "439" on Nov. 26, 1952, to "455" on Nov. 28,
 1952; the inaccurate 1954 press release continues to be quoted, ad
@@
 On April 19, 1951, a North Star piloted by Flight Lieutenant J.A. Watt
-was on a route-training flight to an unfamiliar field—Ashiya—located
+was on a route-training flight to an unfamiliar field---Ashiya---located
 between Osaka and Kobe, Japan. He was cleared by Ashiya tower to
@@
 way. The North Star was blown off the runway into a gully. Everyone
-survived, but 17516 was a total “write-off”–the only RCAF aircraft
+survived, but 17516 was a total "write-off"---the only RCAF aircraft
 lost in the course of Operation Hawk.
@@
 British Empire (MBE), four Air Force Crosses, two Air Force Medals and
-13 Queen’s Commendations. Most of these were to aircrew personnel, but
+13 Queen's Commendations. Most of these were to aircrew personnel, but
 Sqdn. Ldr. W.H. Lord was granted an MBE for his work in establishing

### 2016-01-16 pnsac/newsletter/article_Feature2.tex
Adding files for April 2015 newsletter.

--- 2014-05-17 pnsac/newsletter/article_Feature2.tex
@@
-% Template PNSAC newsletter - Article
-% Language: Latex
+% Template PNSAC newsletter - Article % Language: Latex
 %
@@
 
-\title{The Whispering Giant}
-\author{Dave Lambeth}
+\title{Volunteers Needed} \author{Richard Lodge}
 
@@
 
-Twinkle, twinkle, old North Star,\\
-With wrinkled skin and creaky
-spar,\\
-For years you’ve always been on
-sked,\\
-As many of our force you’ve sped\\
-to distant lands beyond the skies,\\
-you’ve never failed to see the rise\\
-of distant suns or moonrise glare.\\
-You’ve carried us with tender care\\
--back you came to take us home,\\
-from old Japan or France or Rome\\
-from England, Goose, and Kef,\\
-We cannot hear you.\\
-(‘cause we’re deaf)\\
+The Project North Star Association has two types of volunteers. The
+first are known as Restoration volunteers and they are the group who
+get their hands dirty and physically work on the restoration of the
+aeroplane. The other group are known as Association volunteers and
+they work in other very important areas of our Association.
 
-\begin{footnotesize}
-    \raggedleft PNSAC\\
+At present we are not looking for any more Restoration volunteers but
+we do have a need for some Association volunteers. In particular we
+are looking for a Social Events Coordinator and a Merchandise Manager
+and volunteers who would work with those officers.  These volunteer
+posts have become open because the existing volunteers have held the
+positions for some years and in each case have decided that it is time
+to pursue other interests within the Association.
+
+Below are brief details of the requirements for each volunteer
+position. By volunteering for one of these functions, a volunteer will
+become more involved with Association activities and will enjoy the
+comradeship and rewards of working with a very interesting group of
+people who have many diverse life experiences.
+
+The Social Events Coordinator is the person responsible within the
+Association for overseeing and organizing events for
+members. Typically, he or she would, with the assistance of others,
+organize the Members Meetings which are mostly held every three months
+and such other events as he or she might agree to run as approved by
+the Board of Directors.
+
+The Merchandise Manager is responsible for managing the Association's
+stock of merchandise for sale, setting the ordering levels as agreed
+with the Board of Directors, placing orders for new stock and ensuring
+that sales and inventory levels are properly reported to the
+Treasurer. Other volunteers assist the Merchandise Manager in the
+work.
+
+If any of our members are interested in taking over one of these
+positions or working with these officers or know of a non-member who
+would be interested, please would he or she contact me by email at
+\email{info@projectnorthstar.ca} or by phone on (613) 612-4920.
+
+\begin{footnotesize} \raggedleft PNSAC\\
 \end{footnotesize}
@@
 
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: main_document.tex
+%%% Local Variables: %%% mode: latex %%% TeX-master: main_document.tex
 %%% End:

### 2016-01-16 pnsac/newsletter/article_PresReport.tex
Adding files for April 2015 newsletter.

--- 2013-10-16 pnsac/newsletter/article_PresReport.tex
@@
 
-During the past three months we have moved steadily forward with the
-restoration of the North Star.  Our progress has been somewhat held up
-by lack of available volunteers.
+It is a year since I last wrote Notes from the President for the NStar Chronicle. During this time there have been many changes within the Project North Star Association.
 
-Our Association has two distinct and very important sides.  First and
-foremost we are a group whose interest is in the North Star and the
-restoration and preservation of the plane at the Museum.  Secondly,
-over the years, we have developed a strong social side to the
-Association where volunteers and other members can join in interesting
-events centered around aviation. Elsewhere in this issue of the
-Chronicle you will find a description by Bill Tate our Vice President,
-of the very successful trip on October 4 to the Navcan Air Control
-Centre and to the Bombardier factory in Montr\'{e}al.
+In my last Notes, I referred to the temporary closure of restoration work by PNSAC volunteers on 
+the North Star. We effectively did not restart operations until the middle of last year which 
+resulted in a considerable loss of momentum for the Association as a whole together with the 
+resignation of some of our directors and officers.
 
-After 10 years of restoration work on the North Star we recognize that
-we must review our progress and develop a plan for the rest of the
-work on the plane.  There will be three aspects to this review.  Our
-first task will be to identify the major sections of the project and
-develop ways of actually carrying out the restoration.  The
-undercarriage of the plane is a particularly complicated challenge.
-We then need to prepare a plan with projected timelines for completing
-each part of the work.  Our third task will be to estimate costs and
-to develop ideas for financing it.  During this process we must also
-ensure that we will have sufficient qualified volunteers to get it
-done.
+Last November we held our Annual General Meeting, which will in future be held in the Fall rather 
+than in the summer when many of our members are on vacation. At this AGM two new directors were 
+elected to replace two out of the three directors who had resigned.
 
-Before the end of November we are hoping to have meetings with the
-senior management of the Museum to update planning of all the tasks
-above.  The end result of this should be a documented project plan and
-a new Memorandum of Understanding between the Canada Aviation and
-Space Museum and the Project North Star Association of Canada.
+We were very happy to welcome Roger Button and Phil Chrysler as new directors.
+Roger has been our corporate secretary.
+Phil has been our merchandise manager for several years and has also acted as assistant treasurer to the Association.
 
-At times our progress on the restoration of the North Star may appear
-to Association members to be somewhat slow.  Unlike other
-organizations in Canada such as Vintage Wings in Gatineau or the
-Canada Warplane Heritage Museum in Hamilton, PNSAC cannot make
-decisions about the future of the North Star on its own.  As an
-Association working with a Federal Government institution we are often
-affected by decisions not related to the work we do.  This can at
-times provide interesting challenges.
+Drew Hodge, our long time NStar Chronicle publisher and webmaster, has after ten years, decided 
+that he would like to spend more time on his other interests and did not stand for re-election as a director. Drew will continue to publish the NStar Chronicle and we are very pleased to 
+welcome Garth Pyper as our new webmaster.
 
-In conclusion, I would say that the Directors are working hard with
-the Museum management to move our project forward even though at times
-it may appear to be very difficult.  I hope that I will be able to
-report significant progress in the next issue of the Chronicle.  In
-the meantime we have some interesting events in the planning stages on
-the social side of the Association.  Stay tuned.
+Bill Tate has been our vice president and  highly regarded special events coordinator for several years but decided that it was time to hand over to someone else.
+
+Bruce Grant, who has ably editor the NS Chronicle for several years, also decided not to continue as a director and officer.
+Roger has agreed to take over the editor's  job where I am expecting that his word smithing abilities as a lawyer will come in handy.
+
+The table officers of the Association  are now:  myself as president; Bruce Gemmill ,our senior project manager, as vice president; Roger  as secretary; and Paul Labranche as treasurer.
+
+While the changes have been taking place with the executive of the Association, work has continued with our main reason for being; the restoration of the North Star.
+Elsewhere in this issue you will find a progress report on the work on the aircraft which as always is being done to a very high standard.
+
+The North Star restoration project, now in its twelve year, will shortly be entering a new phase. 
+Work on the fourth engine will be complete in a few months and we are now reworking our work plan 
+for the restoration as a whole, in conjunction with the conservation professionals at Museum. In 
+a future issue of the NStar Chronicle we hope to provide members with the broad outline of the 
+updated work plan, together with anticipated time scales and funding requirements.
+Since the Museum does not receive government funds for the North Star restoration, future major work on the aircraft will have to be financed by donations from members and other people interested in preserving a plane which contributed so much to Canada?s military transportation needs in the late 1940s through to the early 1960s.
+
+During the summer we hope to have two major displays of the North Star for Association members and the general public.
+The first will be during the weekend of June 6/7 when we will be taking part in Doors Open Ottawa and the second will be our usual display on Canada Day.
+As always, Association members will be particularly welcome at these displays and will be able to see the results of the excellent work being done by the volunteers on the plane.
 

### 2016-01-16 pnsac/newsletter/article_ProgReport.tex
Adding files for April 2015 newsletter.

@@
 been to disassemble the engine, which was completed in October, and
-restoration and reassembly of engine 4 is now well underway.  The
+restoration and reassembly of engine 4 is now well under way.  The
 refurbished crankshaft has been installed in the engine block, and the
@@
 
-Normally, work would also be underway to restore the engine frame, but
+Normally, work would also be under way to restore the engine frame, but
 due to a lack of space, this will be delayed until the engine is back
@@
 
-\section{Cocpit, Crew Lounge, Galley,  and Forward Washroom}
+\section{Cockpit, Crew Lounge, Galley,  and Forward Washroom}
 \label{crewlounge}
@@
 fittings along the starboard side of the cabin, then removing the wood
-paneling below the windows, and then removing the wood cargo floor
+panelling below the windows, and then removing the wood cargo floor
 panels to expose the metal sub floor.  Most of the side panels and
@@
 to be re-installed on the aircraft before all painting and restoration
-is complete, so as not to delay it’s move outside.
+is complete, so as not to delay its move outside.
 
@@
 make a new set, using funds donated by Project North Star members.
-Seat fabric, steel and aluminum tubing was ordered, and the troop seat
+Seat fabric, steel and aluminium tubing was ordered, and the troop seat
 sections were fabricated, based on available drawings.  All seat

### 2016-01-16 pnsac/newsletter/article_ProgReport_Sup.tex
Adding files for April 2015 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{North Star Merchandise}
\author{Phil Chrysler}

\maketitle

\end{multicols}

The idea for North Star merchandise started with a need to identify those volunteers that were assisting when the aircraft was put on display to the general public.  That requirement evolved into the blue member's only crested golf shirts (men and ladies).  Over time we have added baseball caps, toques sweatshirts, T-shirts and fleece vests (by special order) all with the North Star crest.  Also a 1:200 scale model of the aircraft was produced in limited numbers of which we still have a few.  With the exception of the members only golf shirt all the other items are available at the Aviation and Space Museum store.  The items are available to members of the North Star directly at a discount.

We also have a CD of the North Star operating manual and a North Star pin that are not available in the Museum store.

From the beginning our purpose was to bring more awareness to the general public of our restoration activities rather generate money on the sale of the items.  So we have made an effort to keep the quality high and profits modest.  All money earned is used to advance the restoration effort.  The North Star web site has prices and pictures of the available items.

Merchandise page: {\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca/menu-merchanidise}}}

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

\begin{multicols}{2}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2016-01-16 pnsac/newsletter/article_Volunteer.tex
Adding files for April 2015 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{My First Flight}
\author{John Corby, with Bruce Grant}

\maketitle

At the end of the war, the Australians were very anxious to get
British people to immigrate there, and so they had an offering at
\pounds10, almost a free passage. Of course, at the end of the war
there was a distinct lack of shipping, so much had been sunk, and some
of the ships that were left were not really equipped for passenger
carrying anyway.

I put my name down with Shaw Savill Line, and periodically I would go
up to London to Leadenhall Street to see how my name was progressing
down the list. The guy would say "Name? ah yes, Corby", thumb through
the pages and he'd say "About eighteen months---we'll give you a
call". Well, being young and impatient, I didn't like this very much.

The Americans during the war had a vast establishment in Piccadilly
called Rainbow Corner, probably the biggest PX in the world. End of
the war they all went home and the Ontario Government took it over as
an immigration centre.

There was a sign up: \textbf{Emigrate by Air to Ontario Now!
  \pounds70}.  That was a lot more than Australia's \pounds10 of
course, but then it was a lot closer. I thought if I don't like it, I
can come back.  I went in; that would probably be in August of 1947. I
went back for a medical, and on October 13th 1947 I landed in Toronto.

We flew from Heathrow. The aircraft was a North Star, but not in TCA
livery; it was---something else. I think the idea was---they were
trying to prove that they could maintain this schedule across the
Atlantic. They didn't want their name on the planes until they had
proven it. Just in case \ldots

There was at the time a tremendous controversy over these Rolls-Royce
engines. The opposition was all over the government. These engines
were designed for wartime service. In the war, we'd be losing ten to
fifteen planes every day, ninety-five in the Nuremberg raid; that was
the worst day. The engines were designed for maximum performance, not
for reliability and longevity.

When the airlines contracted for these engines, Rolls-Royce guaranteed
that they would do the job. Well, at first they didn't do the job. But
when I flew on them, they obviously did the job!

That was the first time I had ever been in a plane. It took 13 1/2
hours flying time from Heathrow to Toronto, plus a stop in
Gander. They refuelled there and they also fed us a big breakfast---of
pancakes and maple syrup, another first and my introduction to
Canada. When I got to Toronto, my ears rang for two days afterwards
until I got my hearing back.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{jim_jung_2.eps}
%   %caption of the figure 
%   \caption*{\small \em A Whispering Giant (North Star 17515) at
%     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:jim_jung}
%\end{figure}

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2016-01-17 pnsac/newsletter/article_Feature1.tex
Adding and updating content for Jan/Feb 2016 newsletter.

--- 2014-05-24 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{RAF Lightning with Over-Wing Tanks}
+\title{Volunteers Working on North Star Restoration}
 %\subtitle{Part 4}
-\author{Jim Riddoch}
+\author{Tom MacGregor, Legion Magazine}
 
@@
 
-During the early 60's Britain's Defence Department decided that the
-new fleet of Lightning aircraft should have the capability to extend
-its current sortie of 22 minutes to permit ferry to the Middle
-East. In-flight refuelling had not worked successfully due to the
-offset probe positioned under one wing. Accordingly BAC (British
-Aircraft Corporation,Warton Division) was awarded a contract to
-develop extended wing tanks that could be jettisoned if necessary.
+% \textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
+%  Sqn, RCAF.}\\
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{jim_riddoch.eps}
-   %caption of the figure 
-   \caption*{\small \em Jim Riddoch.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:jim_riddoch.eps}
-\end{figure}
+\begin{quotation}
+	You have to be careful opening up the panels covering the 
+	engines, you never know if a bird or some animal might have made
+	its home there.
+\end{quotation}
 
-%% \begin{footnotesize}
-%% [Image: {\normalfont\color{blue}\href{http://www.airteamimages.com/english-electric-lightning_XP693_united-kingdom---royal-air-force_43865.html}{English
-%%     Electric Lightning with over-wing tanks}}]
-%% \end{footnotesize}
+That’s one lesson a group of dedicated volunteers has learned in its
+10 years of painstaking work restoring a Royal Canadian Air Force
+North Star aircraft at the Canada Aviation and Space Museum in Ottawa.
 
-%\begin{figure}[htbp]
+%\begin{figure*}[htbp]
 %	\vspace{2em}
@@
 %	%name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
+%	\includegraphics[scale=0.5]{op-hawk-one.jpg}
 %	%caption of the figure 
-%   \caption*{\small \em Garry installing the intercooler on engine 3.}
+%	%\caption*{\small \em Volunteers put in countless hours restoring a
+%	North Star aircraft.}
 %	%label of the figure, which has to correspond to \ref{}:
-%   \label{fig:Eng3Intercooler}
-%\end{figure}
+%	\label{fig:op-hawk-one.eps}
+%\end{figure*}
 
-Unfortunately the Lightning with its high speed wings had limited
-capacity for wing tip tanks and already with armament pylons under
-wing could not accommodate under-wing tanks, so the decision was made
-to develop over-wing tanks secured with pylons through the wing
-structure. These tanks would each carry 250 imperial gallons to
-provide initial fuel transfer after take off and be jettisoned once
-empty.
+The aircraft had sat neglected outside the museum for
+years. “Essentially, the RCAF flew this one into the Rockcliffe Airport
+when it was decommissioned. It sat in a old until Robert Holmgren said
+someone should do something about it,” said project manager Bruce
+Gemmill.
 
-I was designated to be the Development Engineer to carry out the
-development and testing of this design about 1963/64. The design
-concept of these tanks had to include a balanced transfer system to
-maintain aircraft C of G during take off and climb out, this meant 5
-small interior sealed compartments pressurized with air to permit
-sequential transfer. Also an emergency fuel jettison system prior to
-tank wing ejection. As you can readily appreciate this was a very
-challenging design and even more demanding for testing and
-certification.
+The airplane had been stripped of its non-essential gear and turned
+over to what was then the National Aeronautical Collection, housed in
+hangars built during the Second World War at the former RCAF station
+at Rockcliffe in Ottawa. Time and weather took its toll. Birds and
+animals made nests where they could get inside the panelling.
 
-Once the initial test tanks were built they were dispatched to
-Mechanical Test Department at Warton Lancashire, where I worked.
-Needless to say the RAF were keen to witness the development and
-testing of these over-wing tanks. A Wing Commander headed a team of
-three Senior Techs to keep track of our progress. Fortunately the
-Techs were very helpful and hands-on in working with us.  You can
-imagine the challenges we faced checking the transfer sequences at
-varying attitudes. Many a Saturday morning was spent testing and
-verifying these change in attitudes and minimizing fuel transfer loss,
-the end result being approximately 225 gallons of fuel transferred
-from each wing tank depending on aircraft attitude and roll.
+“We approached the museum about restoring the aircraft. At first, they
+didn’t know how to take us,” said Gemmill. “The museum had its own
+professional conservation staff but it had never worked with a team of
+volunteers.”
 
-The next major design requirement was the emergency fuel jettison;
-this was to be accomplished by air pressure from centre compartment to
-rear of the tank where a jettisonable fairing was fastened by
-explosive bolts and attached by a cord to a bung on the end of the
-fuel ejection pipe. Talk about Heath Robinson device, this takes the
-biscuit! Suffice to report that on the ground test of this
-ill-conceived system with Chief Designer and my Chief Engineer and
-many observers with cameras, the rear fairing departed into oblivion
-and the cord broke, due to a miserable sub standard cord substituted
-instead of a parachute cord as specified. The bung remained firmly
-attached to the jettison pipe. It was about this time I first thought
-of emigrating to Canada, in fact anywhere outside Britain!
+Eventually the museum agreed to work with the volunteers who formed
+the Project North Star Association of Canada with the late Robert
+Holmgren as president. “We became the model for other volunteer groups
+now working with the museum,” said the association’s current
+president, Richard Lodge.
 
-Needless to say the cord was replaced, the ejection system proved
-satisfactory and the tanks were certified for flight test. This was to
-be done with one tank attached to the top of one wing and painted with
-black and white checkered squares for easy recognition when ejected. I
-always remember the Test Pilot, Peter Knight inquiring about the
-likely occurrence of the tank separating from the wing and hitting the
-tail of the aircraft. I assured him that ground testing of a model in
-the wind tunnel had worked, in fact the small tank had disappeared
-through the end of the tunnel and took some time to recover in the
-neighbouring field.
+Now the aircraft gets towed into a hangar where it is available to the
+volunteers between September and May before it is towed outside for
+the summer.
 
-To replicate the fuel system jettison some bright spark proposed
-filling the tank with coloured water for better visibility. The
-aircraft took off, circled Warton Aerodrome and made a low pass to
-jettison the coloured water over the attentive crowd including chiefs
-and dignitaries. Fortunately, the jettison of the tail fairing and
-bung release worked perfectly and the water emerged in a colourful
-shower.
+The North Star was built by Canadair in the late 1940s and 1950s. It
+was the RCAF version of the DC-4 civilian aircraft except that it used
+four Rolls-Royce Merlin engines. Canadair built 71 of the aircraft.
 
-The weather gods not being with us, the wind blew all the water over
-the spectators covering their expensive clothing with a colourful dye!
-Finally, the tank was ejected safely over Morecombe Bay and eventually
-recovered and returned to BAC Warton amidst embarrassed and colourful
-faces. As one might expect, this completely successful test left the
-Wing Commander disgusted with the entire experiment. He cancelled the
-project.
+The RCAF assigned the rst North Stars to 412 Squadron where they
+transported VIPs and were used in various transfigurations for
+reliable, long-range transport services.
 
-Any thoughts I had of joining the RAF completely disintegrated. Once
-again, immigration to Canada looked much more inviting.
+During the Korean War the North Star was used by No. 426 (Thunderbird)
+Sqdn. to ferry sup- plies across the Pacific Ocean to Japan (Operation
+Hawk: The Korean Airlift, July/August). They would y 599 round trips
+over the Pacific and deliver seven million pounds of cargo and 13,000
+personnel on return trips. All this was achieved without a fatal
+crash.
+
+Most of the air force’s North Stars were declared surplus in the
+1970s.
+
+In the 10 years that Project North Star has been in operation, the
+group has restored two of the four Merlin engines with a third one
+nearly completed.
+
+Each engine had to be taken o the aircraft, brought into the museum’s
+shops where they were taken apart. Each item was tagged, catalogued 
+and usually photographed. “We take a lot of photographs. That’s
+how we know how they go back together,” said Lodge.
+
+Engine project leader Garry Dupont said, “The first engine took us about
+four and a half years to restore. We’re almost finished the third engine
+which as taken us about two years. So we are getting faster.” The
+propellers were restored separately by Hope Aero Propeller and
+Components Inc. in Mississauga, Ont. All restoration work has to be up
+to the professional standards set by the museum.
+
+The group plans to restore the aircraft to its original look, not to
+have it y again.
+
+Lodge said that they have some RCAF manuals but a lot of time they do
+not help when it comes to putting 50-year-old parts together today.
+
+The aircraft itself, 17515, has seen countless hours of work put into
+re- storing the cockpit with all its switches and navigation
+equipment and a small galley for the crew of seven.
+
+Still to be done is work in the cargo section and painting the
+exterior of the aircraft.
+
+“We know this aircraft ew during the Korean War. In fact, we have 
+pictures showing it configured with litters to bring the wounded back to
+Canada,” said Gimmell. “We don’t want to put in the litters but we
+have people working on creating the rope seats that were used.”
+
+The plan is to paint the aircraft to its post-1965 markings, since it
+was painted with the Maple Leaf ag on it.
+
+Still the search goes on to nd new members for the association. “The
+important thing is to nd new people to whom we can pass on this 
+knowledge and skills we are developing,” said Lodge.
 

### 2016-01-17 pnsac/newsletter/article_MemberInterview.tex
Adding and updating content for Jan/Feb 2016 newsletter.

--- 2016-01-16 pnsac/newsletter/article_ProgReport_Sup.tex
@@
 
-\title{North Star Merchandise}
-\author{Phil Chrysler}
+\title{Our Members}
+\author{Interview with Bruce Gemmill}
 
@@
 
-The idea for North Star merchandise started with a need to identify those volunteers that were assisting when the aircraft was put on display to the general public.
-That requirement evolved into the blue member's only crested golf shirts (men and ladies).
-Over time we have added baseball caps, toques sweatshirts, T-shirts and fleece vests (by special order) all with the North Star crest.
-Also a 1:200 scale model of the aircraft was produced in limited numbers of which we still have a few.
-With the exception of the members only golf shirt all the other items are available at the Aviation and Space Museum store.
-The items are available to members of the North Star directly at a discount.
-
-We also have a CD of the North Star operating manual and a North Star pin that are not available in the Museum store.
-
-From the beginning our purpose was to bring more awareness to the general public of our restoration activities rather generate money on the sale of the items.
-So we have made an effort to keep the quality high and profits modest.
-All money earned is used to advance the restoration effort.
-The North Star web site has prices and pictures of the available items.
-
-Merchandise page: {\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca/menu-merchanidise}}}
+\textit{Bruce Gemmill has been involved with Project North Star since 2005 and
+	has exceeded 9,000 volunteer hours. Bruce is presently the Project
+	Manager, as well as a member of the board of directors.}
 

### 2016-01-17 pnsac/newsletter/article_MessageDGCASM.tex
Adding and updating content for Jan/Feb 2016 newsletter.

% Template PNSAC newsletter - Article
% Language: Latex
%

% Head

\title{Project Manager's Progress Report}
\subtitle{November 2015}
\author{Bruce Gemmill}

\maketitle

The aircraft spent considerable time outside this year, but since we
are short of volunteers, not as much work was done as hoped.  The
aircraft is still outside because we are waiting for space to be freed
up in the storage hangar.  We also expect the North Star to be moved
outside again in early spring, much sooner than normal, to accommodate
other uses of the storage facility.  This extended time outside has
delayed some work, and may result in additional weather damage.

\section{Nr 4 Engine}
\label{sec:engine_4}

The last engine to undergo restoration was disassembled last fall.
Since the engine shop crew have cleaned, painted and re-assembled much
of the engine on the engine assembly stand. The refurbished crankshaft
has been installed in the engine block, and the connecting rods and
pistons cleaned and polished.  Also completed are the more complex
cylinder heads and valve train assemblies, and work has begun to
restore many ancillary pieces, such as the magnetos and other
electrical components, and accessories.

The auxiliary gearbox was removed from Nacelle 4 and has been
completely rebuilt, along with the generator.  These won’t be
installed on the aircraft until shortly before engine 4 is ready to be
installed.

Work has also begun on the engine frame which holds the 12 cylinder
engine and supports the surrounding cowl panels.  Once the engine
frame is ready, the engine will be transferred to it from the engine
stand.  Because of a lack of space and shortage of qualified
volunteers, we don’t expect to have engine 4 completed before 2017.  A
great deal of work is also needed on the cowl panels.  We hope to
start work on these in the New Year.

\section{Fuselage and Main Cabin}
\label{sec:main_cabin}

Work has continued in the main cabin.  After removing all floor and
wall panels, as well as numerous fittings, the floor was cleaned and
all cargo tiedowns thoroughly treated to remove corrosion.  Several
required removal and repair. The starboard windows were repainted and
new Plexiglas installed, then the windows were re-installed on the
aircraft.  The port windows were then removed and repaired, but had to
be installed prior to completion so the aircraft could stay outside
this summer.  These will need to be removed again this winter and the
new glass installed before being returned to the aircraft.  Some
corrosion damage along the exposed ribs and sub floor has been
completed.  The rear bulkhead required considerable work.  This
involved removing a large corroded floor panel and three rear bulkhead
panels.  New panels were made and are being installed.

The forward bulkhead and forward half of the floor have been painted,
but further painting was stopped by the arrival of cold weather in
October.  The painting will have to be completed next summer.

The forward cargo door was cleaned, painted and new seals installed.
The door was put back on the aircraft to secure it while outside this
summer.  This winter we plan on removing the rear cargo door for
similar work.

The tail section and rear belly compartment underwent some cleaning
and repair.  This work will also continue this winter in the storage
hangar.

\section{Cabin Liners and Troop Seats}
\label{sec:liners_troop_seats}

Most original cabin liners have now been repaired or replaced with new
liners.  The liners still need to be painted and many stencils need to
be applied, as well as many new snap fasteners.

The troop seats were built last year and placed in storage.  Some work
has been done to restore the many fitting needed to install the troop
seats, along with litter straps and tiedowns for litters and cargo.

\section{Planned Restoration Work---2016}
\label{sec:plannedwork}

Over the next year, we will continue to work on Engine 4 and the main
cabin and doors.  The aircraft will be moved outside in the late
winter or early spring to allow setup for a planned Star Trek exhibit.
Work will not begin inside the aircraft until the weather warms up, so
further delays are inevitable.

We will be able to make new wall panels and either replace or restore
the wood floor panels.

We hope to have all main cabin painting done this summer, so that the
wall and floor panels can be installed.

One other work item that may be started is to clean and restore the
four engine nacelles.  Pipes and fittings need to be removed, followed
by a thorough cleaning and a fresh coat of paint.  This work may be
delayed until qualified volunteers become available.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
%   %caption of the figure 
%   \caption*{\small \em Forward cockpit -- complete.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:fwd_cockpit_complete}
%\end{figure}

\begin{footnotesize}
  \raggedleft PNSAC\\
\end{footnotesize}

% End of text.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2017-03-11 pnsac/newsletter/article_EditorsNotes.tex
Adding and updating content for Jan/Feb 2016 newsletter.
Updating all tex files, images.

--- 2013-12-28 pnsac/newsletter/article_EditorsNotes.tex
@@
 \title{Editor's Notes} 
-\author{Bruce Grant}
+\author{Roger Button}
 
@@
 
-This issue of the Chronicle carries the last of four chapters in
-Richard Lodge's story of his time at Rolls-Royce. Taken together, they
-make an interesting ethnology of an extinct industrial culture. The
-whole series is accessible through our website,
-{\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}} (note that
-the location of the newsletter archive will change when the new
-website goes live) starting with the April 2013 issue of the
-Chronicle.
-
 \vspace{12pt}
 
-We also have another hair-raising flying story from Tim Timmins. A
-series of previous Timmins stories can be found through the same
-website, starting with the December 2010 issue.
+This past year started off as a repeat of the difficult year the
+Association experienced in 2014. Consequently there was little news
+and less enthusiasm for publishing the Chronicle on any regular
+basis. However, as a result of changes to the management of the museum
+organization there is a renewed sense of hope that we will be able to
+move forward in 2016, as expressed in the Notes from the
+President. One of those changes is the appointment of Chris Kitzan as
+the new director general of CASM. You will find a message from Chris
+in this edition. Chris and his boss Alex Benay, the CEO of the museum
+corporation, are enthusiastic supporters of our organization. We are also
+commencing a new feature highlighting the members of our Association
+and starting off with Bruce Gemmill, our Project Manager. In addition,
+we are adding a "Miscellany" feature which lists items of interest to
+our members. If any of our readers wish to contribute an article, have
+suggestions for future articles, have suggestions for the "Miscellany"
+feature, or have any comments about the Chronicle please let me know.
 
-\vspace{12pt}
-
-Motorheads among our readers should look into Ted Devey's extensive
-writings on the Rolls-Royce Merlin, starting in the December 2006
-issue.
-
-\vspace{12pt}
-
-All together, the archived issues are a valuable web resource for
-readers with an interest in aviation history or the history of Project
-North Star.
+%\vspace{12pt}
 

### 2017-03-11 pnsac/newsletter/article_Feature1.tex
Updating all tex files, images.

@@
 \begin{quotation}
-	You have to be careful opening up the panels covering the 
+	\textit{You have to be careful opening up the panels covering the 
 	engines, you never know if a bird or some animal might have made
-	its home there.
+	its home there.}
 \end{quotation}
 
-That’s one lesson a group of dedicated volunteers has learned in its
+That's one lesson a group of dedicated volunteers has learned in its
 10 years of painstaking work restoring a Royal Canadian Air Force
@@
 The aircraft had sat neglected outside the museum for
-years. “Essentially, the RCAF flew this one into the Rockcliffe Airport
-when it was decommissioned. It sat in a old until Robert Holmgren said
-someone should do something about it,” said project manager Bruce
+years. "Essentially, the RCAF flew this one into the Rockcliffe Airport
+when it was decommissioned. It sat in a field until Robert Holmgren said
+someone should do something about it," said project manager Bruce
 Gemmill.
@@
 
-“We approached the museum about restoring the aircraft. At first, they
-didn’t know how to take us,” said Gemmill. “The museum had its own
+"We approached the museum about restoring the aircraft. At first, they
+didn't know how to take us," said Gemmill. "The museum had its own
 professional conservation staff but it had never worked with a team of
-volunteers.”
+volunteers."
 
@@
 the Project North Star Association of Canada with the late Robert
-Holmgren as president. “We became the model for other volunteer groups
-now working with the museum,” said the association’s current
+Holmgren as president. "We became the model for other volunteer groups
+now working with the museum," said the association's current
 president, Richard Lodge.
@@
 
-The RCAF assigned the rst North Stars to 412 Squadron where they
+The RCAF assigned the first North Stars to 412 Squadron where they
 transported VIPs and were used in various transfigurations for
@@
 During the Korean War the North Star was used by No. 426 (Thunderbird)
-Sqdn. to ferry sup- plies across the Pacific Ocean to Japan (Operation
-Hawk: The Korean Airlift, July/August). They would y 599 round trips
+Sqdn. to ferry supplies across the Pacific Ocean to Japan (Operation
+Hawk: The Korean Airlift, July/August). They would fly 599 round trips
 over the Pacific and deliver seven million pounds of cargo and 13,000
@@
 
-Most of the air force’s North Stars were declared surplus in the
+Most of the air force's North Stars were declared surplus in the
 1970s.
@@
 
-Each engine had to be taken o the aircraft, brought into the museum’s
+Each engine had to be taken o the aircraft, brought into the museum's
 shops where they were taken apart. Each item was tagged, catalogued 
-and usually photographed. “We take a lot of photographs. That’s
-how we know how they go back together,” said Lodge.
+and usually photographed. "We take a lot of photographs. That's
+how we know how they go back together," said Lodge.
 
-Engine project leader Garry Dupont said, “The first engine took us about
-four and a half years to restore. We’re almost finished the third engine
-which as taken us about two years. So we are getting faster.” The
+Engine project leader Garry Dupont said, "The first engine took us about
+four and a half years to restore. We're almost finished the third engine
+which as taken us about two years. So we are getting faster." The
 propellers were restored separately by Hope Aero Propeller and
@@
 The group plans to restore the aircraft to its original look, not to
-have it y again.
+have it fly again.
 
@@
 The aircraft itself, 17515, has seen countless hours of work put into
-re- storing the cockpit with all its switches and navigation
+restoring the cockpit with all its switches and navigation
 equipment and a small galley for the crew of seven.
@@
 
-“We know this aircraft ew during the Korean War. In fact, we have 
+"We know this aircraft flew during the Korean War. In fact, we have 
 pictures showing it configured with litters to bring the wounded back to
-Canada,” said Gimmell. “We don’t want to put in the litters but we
-have people working on creating the rope seats that were used.”
+Canada," said Gemmill. "We don't want to put in the litters but we
+have people working on creating the rope seats that were used."
 
 The plan is to paint the aircraft to its post-1965 markings, since it
-was painted with the Maple Leaf ag on it.
+was painted with the Maple Leaf flag on it.
 
-Still the search goes on to nd new members for the association. “The
-important thing is to nd new people to whom we can pass on this 
-knowledge and skills we are developing,” said Lodge.
+Still the search goes on to find new members for the association. "The
+important thing is to find new people to whom we can pass on this 
+knowledge and skills we are developing," said Lodge.
+
+\textit{Reprinted from the March/April 2014 edition of "Legion Magazine," with the kind permission of Jennifer Morse, General Manager of Canvet Publications, and Tom MacGregor.}
 

### 2017-03-11 pnsac/newsletter/article_MemberInterview.tex
Updating all tex files, images.

@@
 
-\textit{Bruce Gemmill has been involved with Project North Star since 2005 and
-	has exceeded 9,000 volunteer hours. Bruce is presently the Project
-	Manager, as well as a member of the board of directors.}
+\begin{quotation}
+	\textit{Bruce Gemmill has been involved with Project North Star since 2005 and	has exceeded 9,000 volunteer hours.
+Bruce is presently the Project Manager, as well as a member of the board of directors.}
+\end{quotation}
+
+\begin{multicols}{2}
+
+1.\textit{What is your background in aviation?}
+
+Before joining Project North Star, I served in the Canadian Forces for
+38 years as a communications and electronics officer.  Although I
+never flew as air crew, I worked on air force communications and radar
+systems, air traffic control, and NORAD air defence.  I served on
+several air force bases, including North Bay and Trenton, and my last
+air force position was senior communications adviser to the Commander
+of Canadian NORAD Region in Winnipeg.
+
+2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}
+
+After I retired from the Canadian Forces in 2005, I wanted to do
+something that was technically challenging and interesting.  I'd done
+some woodworking and motorcycle restoration, and one day I was riding
+past the Aviation Museum and decided on a whim I would ask if there
+was an opportunity to volunteer on an aircraft restoration project.
+Six months later I started on Project North Star.
+
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Forward cockpit -- complete.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:fwd_cockpit_complete}
+%\end{figure}
+
+3. \textit{What has been the history of your involvement to date?}
+
+Since I had no specific aircraft maintenance training or experience, I
+began working on electronic assemblies that had been removed from the
+aircraft cockpit during the early days of the project.  With no test
+equipment, manuals or spare parts, the work was mainly cleaning and
+repainting the cases and front panels.  As I got more familiar with
+the equipment, I learned how to spray paint, and did minor repairs to
+wiring.  I also worked on electrical components from the Merlin
+engine, such as the magnetos and starter motor.
+
+\begin{figure}[htbp]
+	\vspace{2em}
+	\centering
+	%name of the graphic, without the path AND in EPS format:
+	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
+	%caption of the figure 
+	\caption*{\small \em Propeller assembly.}
+	%label of the figure, which has to correspond to \ref{}:
+	\label{fig:propeller}
+\end{figure}
+
+The most demanding job was removing several of the propellers so they could be shipped for restoration.
+Years of corrosion had sealed the spinners onto the propeller hubs.
+Removal required the use of dental picks and a large sledge hammer!
+
+After helping with restoration work on Engine One, I began working on
+the cockpit as a major project.  I spent approximately five years
+working with other volunteers, beginning with removal of as much
+equipment as possible from the cockpit and crew lounge, including the
+control column, rudder pedals and center control console, seats and
+headliners.  A lot of time was spent cleaning and repairing corrosion
+damage before painting the cockpit and installing new insulation.  I
+found it very rewarding when we began installing previously restored
+equipment to the "new" cockpit, but also very challenging.
+
+\begin{figure}[htbp]
+	\vspace{2em}
+	\centering
+	%name of the graphic, without the path AND in EPS format:
+	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
+	%caption of the figure 
+	\caption*{\small \em Forward cockpit -- complete.}
+	%label of the figure, which has to correspond to \ref{}:
+	\label{fig:fwd_cockpit_complete}
+\end{figure}
+
+ A lot of equipment and fittings had been removed prior to me joining the project, and not all removal work had been documented.
+Detective work was required to search out photographs that showed where some equipment was installed.
+
+4. \textit{What has been the highlight of your involvement?}
+
+Completion of the cockpit and crew lounge has certainly been the most
+significant achievement for me personally. The project team has
+received high praise from numerous visitors who have commented on the
+high quality of the restoration work.  Although we consider the work
+done, I do still want to complete the restoration by finding a few
+missing radios and instruments that were removed by the RCAF.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
+   %caption of the figure 
+   \caption*{\small \em Navigator rack.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:nav-radio-rack}
+\end{figure}
+
+5. \textit{What has been the most challenging part of your involvement?}
+
+Finding spare and missing parts for the aircraft has always been a
+challenge.  Our ability to search out possible sources is limited, as
+are funds.  We have been lucky to get most parts needed for the Rolls
+Royce engines, but many items for the airframe are not easy to come
+by.  These include everything from very large and expensive items,
+such as the de-icing boots for the wings and tail, to door and window
+seals.  Rubber and fabric items are particularly hard to find, since
+these items deteriorate with age.  Finding the correct paint has
+recently become a major challenge because of stricter environmental
+laws.  Our goal of achieving an authentic restoration of the North
+Star will only become more difficult with the passage of time.
 
@@
 \end{footnotesize}
-
-\begin{multicols}{2}
 

### 2017-03-11 pnsac/newsletter/article_MessageDGCASM.tex
Updating all tex files, images.

@@
 
-\title{Project Manager's Progress Report}
-\subtitle{November 2015}
-\author{Bruce Gemmill}
+\title{Message from the CASM Director General}
+%\subtitle{November 2015}
+\author{Chris Kitzan}
 
@@
 
-The aircraft spent considerable time outside this year, but since we
-are short of volunteers, not as much work was done as hoped.  The
-aircraft is still outside because we are waiting for space to be freed
-up in the storage hangar.  We also expect the North Star to be moved
-outside again in early spring, much sooner than normal, to accommodate
-other uses of the storage facility.  This extended time outside has
-delayed some work, and may result in additional weather damage.
+\begin{quotation}
+	\textit{Chris Kitzan was appointed the new director general of CASM last
+	November. Previously he was the chief of staff at Library and Archives
+	Canada. Before that he worked as a curator and exhibit manager at what
+	was then known as the Canadian Museum of Civilization.}
+\end{quotation}
 
-\section{Nr 4 Engine}
-\label{sec:engine_4}
+My tenure at the Museum is just beginning, but already I have observed
+first-hand the contribution volunteers are making, from my meetings
+with members, to my perspective as a father watching you
+enthusiastically cater to my eight year old son's interests, to the
+communications I have received from you with ideas and offers of
+support, and finally from guests who have written thanking the Museum
+for experiences they have had interacting with you on the floor.  I
+have never worked in an organization that had so many individuals
+willing to donate anywhere near the amount of time and energy you do.
+Given that, is it fair to want more?  I hope so.  With your support, I
+would like to tap a little deeper into your energy and discuss ways we
+can further work together to innovate and inspire.
 
-The last engine to undergo restoration was disassembled last fall.
-Since the engine shop crew have cleaned, painted and re-assembled much
-of the engine on the engine assembly stand. The refurbished crankshaft
-has been installed in the engine block, and the connecting rods and
-pistons cleaned and polished.  Also completed are the more complex
-cylinder heads and valve train assemblies, and work has begun to
-restore many ancillary pieces, such as the magnetos and other
-electrical components, and accessories.
+I recognize that increased collaboration between passionate parties
+will sometimes create moments of turbulence.  I feel it is a small
+price to pay if we are all committed to the ultimate goal --
+maximizing the potential of this organization.  That is why, on the
+North Star front, Richard and CASM staff are currently trying to find
+solutions that allow us to meet our collective objectives.  And why I
+hope that these initial conversations lead to increased discussions,
+and volunteer/staff collaboration on a wider range of fronts.
 
-The auxiliary gearbox was removed from Nacelle 4 and has been
-completely rebuilt, along with the generator.  These won’t be
-installed on the aircraft until shortly before engine 4 is ready to be
-installed.
-
-Work has also begun on the engine frame which holds the 12 cylinder
-engine and supports the surrounding cowl panels.  Once the engine
-frame is ready, the engine will be transferred to it from the engine
-stand.  Because of a lack of space and shortage of qualified
-volunteers, we don’t expect to have engine 4 completed before 2017.  A
-great deal of work is also needed on the cowl panels.  We hope to
-start work on these in the New Year.
-
-\section{Fuselage and Main Cabin}
-\label{sec:main_cabin}
-
-Work has continued in the main cabin.  After removing all floor and
-wall panels, as well as numerous fittings, the floor was cleaned and
-all cargo tiedowns thoroughly treated to remove corrosion.  Several
-required removal and repair. The starboard windows were repainted and
-new Plexiglas installed, then the windows were re-installed on the
-aircraft.  The port windows were then removed and repaired, but had to
-be installed prior to completion so the aircraft could stay outside
-this summer.  These will need to be removed again this winter and the
-new glass installed before being returned to the aircraft.  Some
-corrosion damage along the exposed ribs and sub floor has been
-completed.  The rear bulkhead required considerable work.  This
-involved removing a large corroded floor panel and three rear bulkhead
-panels.  New panels were made and are being installed.
-
-The forward bulkhead and forward half of the floor have been painted,
-but further painting was stopped by the arrival of cold weather in
-October.  The painting will have to be completed next summer.
-
-The forward cargo door was cleaned, painted and new seals installed.
-The door was put back on the aircraft to secure it while outside this
-summer.  This winter we plan on removing the rear cargo door for
-similar work.
-
-The tail section and rear belly compartment underwent some cleaning
-and repair.  This work will also continue this winter in the storage
-hangar.
-
-\section{Cabin Liners and Troop Seats}
-\label{sec:liners_troop_seats}
-
-Most original cabin liners have now been repaired or replaced with new
-liners.  The liners still need to be painted and many stencils need to
-be applied, as well as many new snap fasteners.
-
-The troop seats were built last year and placed in storage.  Some work
-has been done to restore the many fitting needed to install the troop
-seats, along with litter straps and tiedowns for litters and cargo.
-
-\section{Planned Restoration Work---2016}
-\label{sec:plannedwork}
-
-Over the next year, we will continue to work on Engine 4 and the main
-cabin and doors.  The aircraft will be moved outside in the late
-winter or early spring to allow setup for a planned Star Trek exhibit.
-Work will not begin inside the aircraft until the weather warms up, so
-further delays are inevitable.
-
-We will be able to make new wall panels and either replace or restore
-the wood floor panels.
-
-We hope to have all main cabin painting done this summer, so that the
-wall and floor panels can be installed.
-
-One other work item that may be started is to clean and restore the
-four engine nacelles.  Pipes and fittings need to be removed, followed
-by a thorough cleaning and a fresh coat of paint.  This work may be
-delayed until qualified volunteers become available.
-
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   %name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
-%   %caption of the figure 
-%   \caption*{\small \em Forward cockpit -- complete.}
-%   %label of the figure, which has to correspond to \ref{}:
-%   \label{fig:fwd_cockpit_complete}
-%\end{figure}
+On the one hand, the Canadair North Star was a reliable, versatile
+Canadian aircraft that provided our country with excellent service on
+both the military and civilian fronts for decades.  On the other, it
+made such a bloody racket that wherever it went it could not be
+ignored.  To my mind it's a worthy symbol of the kind of Museum I
+would like to see us continue to develop.  Based on its collection,
+its staff and its volunteers CASM has already established a reputation
+as a world-class Museum.    That standard cannot be sacrificed.  At
+the same time, there is room for us to make more noise; to try new
+ways to reach Canadians, catch their attention, and further share our
+rich collection and stories with them.
 

### 2017-03-11 pnsac/newsletter/article_PresReport.tex
Updating all tex files, images.

--- 2014-05-24 pnsac/newsletter/article_PresReport.tex
@@
 
-Since my last Notes from the President in December 2013, restoration
-work on the North Star has mostly been suspended. The Museum has
-needed extra working space so that exhibits could be made ready for
-the commemoration this year of the centenary of the start of the First
-World War and the 70th anniversary on June 6 of D-Day in the Second
-World War. These preparations have resulted in our normal working
-areas being unavailable for North Star restoration work.
+As I write these words I am watching the falling snow and preparing for a nasty weather mix this evening and at the same time thinking of the spring and summer and our outdoor activities with the North Star aircraft.
+2016 has every chance of being an interesting year for the North Star volunteers.
+Although full details are not available yet, we are expecting to be able to show off the plane to the general public and to other groups several times this year.
 
-\begin{figure}[htbp]
-  \vspace{2em}
-  \centering
-  %name of the graphic, without the path AND in EPS format:
-  \includegraphics[scale=0.5]{richard_mtg.eps}
-  %caption of the figure 
-  \caption*{\small \em Richard at the Quarterly Meeting in April.}
-  %label of the figure, which has to correspond to \ref{}:
-  \label{fig:richard_mtg}
-\end{figure}
+We are continuing to work closely with the Museum conservation staff to develop plans for extended hours working which basically means running a restoration shift on one or more Saturdays each month.
+This should enable some members to become Restoration Volunteers even though they are not available for work Mondays to Fridays.
+We had hoped to start the shift before Easter but it looks now as if the start date will be in the fall.
+There are still some details of supervision, security and safety to be worked out.
+Running a Saturday shift is a new endeavour for the Museum and it is important for everybody that we take the necessary time to get the operation properly setup so that we have a successful start and a good model for any further out of hours working arrangements for PNSAC or any other volunteer organization working on a museum corporation artifact.
 
-During the shut down period some limited work has been continuing. Two
-volunteers have been working in the cockpit of the plane and two other
-volunteers have been doing some work on \#3 engine reassembly.
+We were very pleased to welcome Chris Kitzan, the new Director General of the Canada Aviation and Space Museum to our first Members' Meeting of 2016 which was held in the Bush Theatre on February 27th.
+Chris acknowledged the importance of the partnership between the museum and Project North Star.
+Following the meeting, I saw Chris having a lively and upbeat conversation with our members, over coffee and snacks.
 
-While the restoration work has been on hold, we have arranged three
-monthly Volunteer Lunches at the Museum to enable the people who
-normally work on the plane to meet and socialize. Although we would
-all like to be working on the plane, we understand why the Museum has
-had to ask us to temporarily suspend our activities. The Volunteer
-Lunches are a way of keeping the team together and up-to-date on what
-is happening.
+Prior to the Members' Meeting, several of us ran into Mike Charters, who is a volunteer maintenance mechanic at the Warplane Heritage Museum in Hamilton.
+Mike was one of the team who flew to England in the summer of 2014 for the tour of the UK by the Canadian Lancaster together with the British Lancaster.
+Mike generously agreed to speak to the members for a few minutes and had some very interesting stories about the trip.
 
-At the last meeting of the PNSAC Board of Directors we agreed to
-reschedule the Annual General Meeting from June each year to
-September. It has always been difficult to find a good weekend in June
-to hold the AGM and it is felt that a middle of September date would
-be much better. Our corporate secretary, Roger Button, is aware of
-this proposed change and has confirmed that the existing Board of
-Directors can extend its term until the new AGM date.
-
-Although there has been little restoration work done since December,
-we are working with the Museum to get the project back on track again
-in early June. We are also working to enable us to recruit new
-volunteers to replace the several people who for various reasons have
-had to give up active work.
-
-One of the recent discussion areas has been in connection with
-long-term funding of the restoration project. Everybody is aware that
-funding from the Federal Government has been substantially reduced. We
-have accepted that major  expenditures will have to be met from funds
-raised by donations from friends of the Museum and members of our
-Association. This is a significant challenge since everybody receives
-many appeals throughout the year for donations from many different
-deserving organizations.
-
-During the next few months we hope to develop a revised restoration
-work plan which will enable us to estimate the funds needed to
-complete the project. We will then be able to target appeals for funds
-to carry out specific parts of the restoration in the same way as we
-did for our successful appeal for funds to make troop seats for the
-interior of the plane.
+One of the best activities for our members is to become an active volunteer.
+Our volunteers fall into two categories, Restoration Volunteers and Association Volunteers.
+The former are those who work on the North Star restoration and have to be limited in number because it is not possible to have too many people working at any one time; this is mainly for supervision and safety reasons.
+Association Volunteers are those who volunteer for other activities in our Association.
+There are many different things which need to be done in a volunteer organization and we encourage our members to contact us to volunteer for such things as meeting the public on aircraft open days, helping to organize events and assisting with sales of our merchandise.
+Volunteering has many rewards, from comradeship with other volunteers, enjoying involvement with an interesting aviation artifact and personal satisfaction derived from doing something worthwhile.
+If you are interested in exploring volunteering opportunities, do contact any one of our directors whose contact information is shown at the end of this newsletter.
 

### 2017-03-11 pnsac/newsletter/article_ProgReport.tex
Updating all tex files, images.

--- 2014-05-24 pnsac/newsletter/article_ProgReport.tex
@@
 \title{Project Manager's Progress Report}
-\subtitle{April 2014}
+\subtitle{November 2015}
 \author{Bruce Gemmill}
@@
 
-Due to museum preparations for a display to mark the beginning of
-the First World War and other events in 2014, the restoration
-space used by Project North Star was taken over by other museum
-projects, and most work on the North Star was suspended.
-However, some limited work did continue.
+The aircraft spent considerable time outside this year, but since we
+are short of volunteers, not as much work was done as hoped.  The
+aircraft is still outside because we are waiting for space to be freed
+up in the storage hangar.  We also expect the North Star to be moved
+outside again in early spring, much sooner than normal, to accommodate
+other uses of the storage facility.  This extended time outside has
+delayed some work, and may result in additional weather damage.
 
-\section{Nr 3 Engine}
-\label{sec:engines_3}
+\section{Nr 4 Engine}
+\label{sec:engine_4}
 
-Work on Engine 3 was halted for several weeks in early 2014, but
-has now resumed on a reduced work schedule.  Most accessories are
-now installed, and soon the cowl panels will be fitted.  It is
-expected that Engine 3 will be installed late this spring or
-early summer, so that work can begin on removing and restoring
-Engine 4.
+The last engine to undergo restoration was disassembled last fall.
+Since the engine shop crew have cleaned, painted and re-assembled much
+of the engine on the engine assembly stand. The refurbished crankshaft
+has been installed in the engine block, and the connecting rods and
+pistons cleaned and polished.  Also completed are the more complex
+cylinder heads and valve train assemblies, and work has begun to
+restore many ancillary pieces, such as the magnetos and other
+electrical components, and accessories.
 
-This downtime also offered the opportunity for the engine crew to
-clean and re-organize the engine shop.
+The auxiliary gearbox was removed from Nacelle 4 and has been
+completely rebuilt, along with the generator.  These won't be
+installed on the aircraft until shortly before engine 4 is ready to be
+installed.
 
-\section{Cockpit}
-\label{cockpit}
+Work has also begun on the engine frame which holds the 12 cylinder
+engine and supports the surrounding cowl panels.  Once the engine
+frame is ready, the engine will be transferred to it from the engine
+stand.  Because of a lack of space and shortage of qualified
+volunteers, we don't expect to have engine 4 completed before 2017.  A
+great deal of work is also needed on the cowl panels.  We hope to
+start work on these in the New Year.
 
-While restoration work in the shop could not be continued, we
-used the opportunity to install the crew seats and other restored
-items to the cockpit.  The pilot seat was installed (first
-officer seat was installed last year) then the railings and
-safety board were installed behind the seats.
+\section{Fuselage and Main Cabin}
+\label{sec:main_cabin}
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
-   %caption of the figure 
-   \caption*{\small \em Forward cockpit -- complete.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:fwd_cockpit_complete}
-\end{figure}
+Work has continued in the main cabin.  After removing all floor and
+wall panels, as well as numerous fittings, the floor was cleaned and
+all cargo tiedowns thoroughly treated to remove corrosion.  Several
+required removal and repair. The starboard windows were repainted and
+new Plexiglas installed, then the windows were re-installed on the
+aircraft.  The port windows were then removed and repaired, but had to
+be installed prior to completion so the aircraft could stay outside
+this summer.  These will need to be removed again this winter and the
+new glass installed before being returned to the aircraft.  Some
+corrosion damage along the exposed ribs and sub floor has been
+completed.  The rear bulkhead required considerable work.  This
+involved removing a large corroded floor panel and three rear bulkhead
+panels.  New panels were made and are being installed.
 
-Next we installed
-the radio officer and navigator tables, and the seats for these
-positions.  The radio officer seat proved to be a particular
-challenge, as the floor in this area is no longer flat, and some
-modifications to the seat rails were necessary to allow the seat
-to slide.  Finally the flight engineer's jump seat was attached
-to the railing.  With five crew seats installed, the cockpit is
-very cramped.
+The forward bulkhead and forward half of the floor have been painted,
+but further painting was stopped by the arrival of cold weather in
+October.  The painting will have to be completed next summer.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{radio_position.eps}
-   %caption of the figure 
-   \caption*{\small \em Radio Officer's position showing the morse
-     code key.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:radio_position}
-\end{figure}
+The forward cargo door was cleaned, painted and new seals installed.
+The door was put back on the aircraft to secure it while outside this
+summer.  This winter we plan on removing the rear cargo door for
+similar work.
 
-The cockpit installations were finished off by installing the
-curtains, crew oxygen hoses, and a Morse code key at the radio
-position.
+The tail section and rear belly compartment underwent some cleaning
+and repair.  This work will also continue this winter in the storage
+hangar.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{cockpit_curtain.eps}
-   %caption of the figure 
-   \caption*{\small \em Finishing touches -- the cockpit curtain.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:cockpit_curtain}
-\end{figure}
+\section{Cabin Liners and Troop Seats}
+\label{sec:liners_troop_seats}
 
-\section{Crew Lounge, Galley,  and Forward Washroom}
-\label{crewlounge}
+Most original cabin liners have now been repaired or replaced with new
+liners.  The liners still need to be painted and many stencils need to
+be applied, as well as many new snap fasteners.
 
-The floor under this section was repaired by removing a section
-of badly corroded I-beam and fabricating a new one from several
-pieces of aluminum which were riveted in place to support the
-floor.  Once this repair was completed, new cushion flooring was
-cut and glued to the crew lounge and the washroom floor, then
-various pieces of floor trim were installed.  To install the crew
-seat, it was necessary to remove the lower support section, since
-the fully assembled seat could not easily fit through the narrow
-doorway.  Once the seat was installed, we added the crew table
-and bunk bed, along with the secure storage box.  We also
-installed the curtain rod and new curtains to close off the bunk
-bed. The installation was finished off by re-installing the door
-between the lounge and the main cabin, so that this fully
-restored area can be protected from dust and debris while the
-main cabin undergoes restoration over the next two years.
+The troop seats were built last year and placed in storage.  Some work
+has been done to restore the many fitting needed to install the troop
+seats, along with litter straps and tiedowns for litters and cargo.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{new_galley_floor.eps}
-   %caption of the figure 
-   \caption*{\small \em New galley and floor.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:new_galley_floor}
-\end{figure}
-
-The completion of the forward area of the aircraft marks a
-significant milestone in the restoration of the North Star.
-
-\section{Planned Restoration Work 2014}
+\section{Planned Restoration Work---2016}
 \label{sec:plannedwork}
 
-Some limited work is being done to remove fittings and wall
-panels from the main cabin, in preparation for some major
-restoration work this summer.  Fabric work on the troop seats is
-nearing completion, although the seats will not be installed
-until all work on the main cabin is complete.  We hope to resume
-a full work schedule by June, with the aircraft outside once
-more.
+Over the next year, we will continue to work on Engine 4 and the main
+cabin and doors.  The aircraft will be moved outside in the late
+winter or early spring to allow setup for a planned Star Trek exhibit.
+Work will not begin inside the aircraft until the weather warms up, so
+further delays are inevitable.
+
+We will be able to make new wall panels and either replace or restore
+the wood floor panels.
+
+We hope to have all main cabin painting done this summer, so that the
+wall and floor panels can be installed.
+
+One other work item that may be started is to clean and restore the
+four engine nacelles.  Pipes and fittings need to be removed, followed
+by a thorough cleaning and a fresh coat of paint.  This work may be
+delayed until qualified volunteers become available.
+
+\section{March 2016 Update}
+\label{sec:update}
+
+Since our report at the November AGM, the aircraft was moved inside,
+so that it could properly sealed against the weather, as it will spend
+much of 2016 outside. The aircraft was moved out again in March.
+During this time, very little work was done on the aircraft, except
+for replacement of the Plexiglas in the port windows, and the
+installation of the ARC-552 UHF radio, kindly transferred to the
+museum by the Department of National Defence.
+
+\subsection{Nr 4 Engine}
+\label{sec:update-engine}
+
+The engine frame has been refurbished, and the engine was transferred
+to the frame, so that ancillary equipment can be installed.  Work has
+also begun on the supercharger, as well as such items as the starter
+motor and header tank.  Some work is also being done on the cowl panels.
+
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Forward cockpit -- complete.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:fwd_cockpit_complete}
+%\end{figure}
 

### 2017-03-12 pnsac/newsletter/article_PresReport.tex
Draft version sent for review.

--- 2014-05-24 pnsac/newsletter/article_PresReport.tex
@@
 
-Since my last Notes from the President in December 2013, restoration
-work on the North Star has mostly been suspended. The Museum has
-needed extra working space so that exhibits could be made ready for
-the commemoration this year of the centenary of the start of the First
-World War and the 70th anniversary on June 6 of D-Day in the Second
-World War. These preparations have resulted in our normal working
-areas being unavailable for North Star restoration work.
+The last North Star Chronicle was published in March 2016 and since that time we
+have successfully continued our work on the aircraft while the Museum was making
+preparations for the Star Trek event in the Reserve Hangar. We were not asked to
+suspend our work, unlike in 2014 when preparations for the Star Wars event were
+being made. This has helped us to retain our Restoration volunteers and recruit
+two new ones. However, we are still short of Restoration volunteers and continue
+to actively work with the Museum to address this problem.
 
-\begin{figure}[htbp]
-  \vspace{2em}
-  \centering
-  %name of the graphic, without the path AND in EPS format:
-  \includegraphics[scale=0.5]{richard_mtg.eps}
-  %caption of the figure 
-  \caption*{\small \em Richard at the Quarterly Meeting in April.}
-  %label of the figure, which has to correspond to \ref{}:
-  \label{fig:richard_mtg}
-\end{figure}
+When Project North Star was originally formed in 2004, the Association was
+treated with considerable skepticism by the Museum. It was felt that restoring
+an aircraft of the size and complexity of a North Star, to the exacting
+standards of a national museum, was likely to be beyond the capabilities of a
+group of volunteers and that the project would probably fail.
 
-During the shut down period some limited work has been continuing. Two
-volunteers have been working in the cockpit of the plane and two other
-volunteers have been doing some work on \#3 engine reassembly.
+We will soon have completed 70,000 hours of volunteer work over a period of 14
+years, restoring everything we do, to what is acknowledged to be a very high
+standard. Skepticism is a thing of the past and we are now treated as a
+respected and ongoing part of the Museum operations.
 
-While the restoration work has been on hold, we have arranged three
-monthly Volunteer Lunches at the Museum to enable the people who
-normally work on the plane to meet and socialize. Although we would
-all like to be working on the plane, we understand why the Museum has
-had to ask us to temporarily suspend our activities. The Volunteer
-Lunches are a way of keeping the team together and up-to-date on what
-is happening.
+Respect has to be earned and comes through hard work and in the case of aircraft
+restoration work, attention to detail. Our team of volunteers needs guidance
+both from the Museum staff and from our own senior volunteers. One volunteer has
+been particularly important in helping us the get to the position we are now in.
+He has recently reached a milestone only reached on two other occasions by a
+volunteer and in each case the volunteer worked in the the Science and
+Technology Museum. Our volunteer has reached 10,000 volunteer hours of working
+on the North Star and throughout this time he also acted as our Project Manager
+for everything except the engines. Bruce Gemmill probably was not aware that he
+had reached 10,000 hours when he unfortunately had to announce that he would
+have to suspend his activities as a volunteer and director of PNSAC due to
+family circumstances. We very much hope that he will soon be able to rejoin us
+and continue to help us complete this large project. In the meantime Bruce will
+continue to be our Membership Secretary. During June, Bruce's remarkable
+contribution was recognized by the Museum and in a private event he was
+presented with a certificate and a gift by Cedric St Amour, the Museum
+Corporation Volunteer Coordinator. One of our volunteers, Robert Desjardins,
+also joined Cedric on the occasion to congratulate Bruce on behalf of PNSAC.
+Well done Bruce and we are all very proud of your achievement.
 
-At the last meeting of the PNSAC Board of Directors we agreed to
-reschedule the Annual General Meeting from June each year to
-September. It has always been difficult to find a good weekend in June
-to hold the AGM and it is felt that a middle of September date would
-be much better. Our corporate secretary, Roger Button, is aware of
-this proposed change and has confirmed that the existing Board of
-Directors can extend its term until the new AGM date.
+I have always seen our Association as having multiple roles. First and foremost
+we are restoring a major artifact with our volunteers; secondly we show off our
+restoration work to the public which in turn brings visitors and revenue into
+the Museum and thirdly through our experienced volunteers we help to transfer
+skills to younger and less experienced volunteers.
 
-Although there has been little restoration work done since December,
-we are working with the Museum to get the project back on track again
-in early June. We are also working to enable us to recruit new
-volunteers to replace the several people who for various reasons have
-had to give up active work.
+Planning our work in conjunction with the Museum management and the conservation
+staff is something we must always try to jointly do better. It is often a fine
+balancing act to decide when to do work on the aircraft in relation to when the
+Museum is having a major event such as Canada Day. I am hoping that in future we
+can regularly update our work schedule in conjunction with our project plan so
+that we can more readily have the aircraft available for public viewing at the
+times when the Museum would like us to do so, without having to delay our
+restoration work unnecessarily.
 
-One of the recent discussion areas has been in connection with
-long-term funding of the restoration project. Everybody is aware that
-funding from the Federal Government has been substantially reduced. We
-have accepted that major  expenditures will have to be met from funds
-raised by donations from friends of the Museum and members of our
-Association. This is a significant challenge since everybody receives
-many appeals throughout the year for donations from many different
-deserving organizations.
-
-During the next few months we hope to develop a revised restoration
-work plan which will enable us to estimate the funds needed to
-complete the project. We will then be able to target appeals for funds
-to carry out specific parts of the restoration in the same way as we
-did for our successful appeal for funds to make troop seats for the
-interior of the plane.
+I will conclude with an invitation to any of our members not volunteering at
+present to come and join us. Although we have a definite limit on the number of
+volunteers who can get their hands dirty working on restoration work, we always
+welcome others who would like to participate in the running of the organization.
+By becoming involved either as a restoration volunteer or an association
+volunteer you will meet other like minded people and have an interesting time
+talking to members of the public. You do not need to be an expert on aviation or
+on the North Star to make a contribution. I can be reached by email or phone if
+you would like to find out more about volunteering.
 

### 2017-03-12 pnsac/newsletter/article_ProgReport.tex
Draft version sent for review.

@@
 \title{Project Manager's Progress Report}
-\subtitle{November 2015}
+\subtitle{PNS 2016 Status Update}
 \author{Bruce Gemmill}
@@
 
-The aircraft spent considerable time outside this year, but since we
-are short of volunteers, not as much work was done as hoped.  The
-aircraft is still outside because we are waiting for space to be freed
-up in the storage hangar.  We also expect the North Star to be moved
-outside again in early spring, much sooner than normal, to accommodate
-other uses of the storage facility.  This extended time outside has
-delayed some work, and may result in additional weather damage.
-
 \section{Nr 4 Engine}
@@
 
-The last engine to undergo restoration was disassembled last fall.
-Since the engine shop crew have cleaned, painted and re-assembled much
-of the engine on the engine assembly stand. The refurbished crankshaft
-has been installed in the engine block, and the connecting rods and
-pistons cleaned and polished.  Also completed are the more complex
-cylinder heads and valve train assemblies, and work has begun to
-restore many ancillary pieces, such as the magnetos and other
-electrical components, and accessories.
+Work has continued on much of the ancillary equipment for the engine. The
+supercharger and intercooler have been completed and installed, as well as the
+reduction gearbox. The radiators and header tank have been restored and painted, and
+two of the three radiator sections have been installed. Restoration work has
+begun on the air intake system, and many of the fuel and oil lines.
 
-The auxiliary gearbox was removed from Nacelle 4 and has been
-completely rebuilt, along with the generator.  These won't be
-installed on the aircraft until shortly before engine 4 is ready to be
-installed.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{"PNS 007-im2".png}
+   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
+   \label{fig:engine_no_4_2}
+\end{figure}
 
-Work has also begun on the engine frame which holds the 12 cylinder
-engine and supports the surrounding cowl panels.  Once the engine
-frame is ready, the engine will be transferred to it from the engine
-stand.  Because of a lack of space and shortage of qualified
-volunteers, we don't expect to have engine 4 completed before 2017.  A
-great deal of work is also needed on the cowl panels.  We hope to
-start work on these in the New Year.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{"PNS 003-im3".png}
+   \caption*{\small \em Volunteers Robert Desjardins, Bruce Gemmill and Michel Cote hold the refinished coolant tank from engine \#4.
+Hours of polishing were required to get the tank to look like new.}
+   \label{fig:coolant_tank}
+\end{figure}
+
+The starter motor was found to be very corroded, and disassembly and repair not viable. A
+new unit has been ordered and will be installed once delivered. The fire
+suppression lines are complete and will be installed along with the cowls when
+ready. Some of the cowl panels have been restored, although it has been found
+that this engine has more corrosion than the other three engines, possibly it
+had been in service longer? This has meant that more work is required to
+refurbish each of the many panels. We still hope to have this engine ready for
+installation by next spring.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{"PNS 008-im1".png}
+   \caption*{\small \em Volunteer Garry Dupont working on the re-assembly of engine \#4.}
+   \label{fig:engine_no_4}
+\end{figure}
 
@@
 
-Work has continued in the main cabin.  After removing all floor and
-wall panels, as well as numerous fittings, the floor was cleaned and
-all cargo tiedowns thoroughly treated to remove corrosion.  Several
-required removal and repair. The starboard windows were repainted and
-new Plexiglas installed, then the windows were re-installed on the
-aircraft.  The port windows were then removed and repaired, but had to
-be installed prior to completion so the aircraft could stay outside
-this summer.  These will need to be removed again this winter and the
-new glass installed before being returned to the aircraft.  Some
-corrosion damage along the exposed ribs and sub floor has been
-completed.  The rear bulkhead required considerable work.  This
-involved removing a large corroded floor panel and three rear bulkhead
-panels.  New panels were made and are being installed.
+With the aircraft moved outside in March, little work was done in the aircraft
+over the winter. Once the temperatures warmed up, we continued work in the main
+cabin to prepare for completion of the painting started last year. Repairs to
+the rear bulkhead are now complete with the installation of several new pieces
+of aluminum to replace corroded ones. A new section of flooring was installed
+where the rear washroom was situated. This required replacing about 1000 rivets!
 
-The forward bulkhead and forward half of the floor have been painted,
-but further painting was stopped by the arrival of cold weather in
-October.  The painting will have to be completed next summer.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{"pns 043-im4".png}
+   \caption*{\small \em A single new floor panel at the rear of the main cabin required over 1000 rivets.  
+   Cleco clips are used to accurately align the panel during riveting.}
+   \label{fig:panel_revets}
+\end{figure}
 
-The forward cargo door was cleaned, painted and new seals installed.
-The door was put back on the aircraft to secure it while outside this
-summer.  This winter we plan on removing the rear cargo door for
-similar work.
+Cleaning and corrosion removal continued throughout the cabin. It seems whenever
+we feel we are near completion, we find more corrosion. We had also planned to
+finish priming the floor and painting the rear bulkhead before the aircraft
+returned to the hanger later in the year, but this was delayed by the shutdown
+of the project while a new Project Manager was hired by the museum. The wood
+floors and walls still need to be repaired, or new panels made, and all will
+require painting.
 
-The tail section and rear belly compartment underwent some cleaning
-and repair.  This work will also continue this winter in the storage
-hangar.
+\section{Cabin Liners}
+\label{sec:liners}
 
-\section{Cabin Liners and Troop Seats}
-\label{sec:liners_troop_seats}
+All the large pieces of liner have either been repaired or replaced, but with
+limitations on space, most remaining work on the cabin liners has been
+postponed. New window surrounds have also been made, but then all liners will
+require painting and stenciling, and this must wait until there is more room in
+the shop and the paint booth. Liners were also made for the top of the heater
+ducts.
 
-Most original cabin liners have now been repaired or replaced with new
-liners.  The liners still need to be painted and many stencils need to
-be applied, as well as many new snap fasteners.
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{P1020369-im5.png}
+   \caption*{\small \em One of the refurbished heater duct sections with a new fabric headliner secured to the top.
+The sides are covered with mica insulation that was removed, cleaned and reattached after repairs to the metal duct.
+These will be installed in the main cabin.}
+   \label{fig:heaters_liners}
+\end{figure}
 
-The troop seats were built last year and placed in storage.  Some work
-has been done to restore the many fitting needed to install the troop
-seats, along with litter straps and tiedowns for litters and cargo.
+\section{PNS 2017 Update}
+\label{sec:pnsupdate}
 
-\section{Planned Restoration Work---2016}
-\label{sec:plannedwork}
+We were able to restart work on the aircraft in late January, following the
+arrival of our new Project manager, Rej Demers. So far, we have completed the
+top cowl panels and one of the radiator flaps, which required a repair patch to
+the outer skin, and the electrical harnesses, which are now waiting for
+painting. Work continues on the air box and the remaining cowl panels.
 
-Over the next year, we will continue to work on Engine 4 and the main
-cabin and doors.  The aircraft will be moved outside in the late
-winter or early spring to allow setup for a planned Star Trek exhibit.
-Work will not begin inside the aircraft until the weather warms up, so
-further delays are inevitable.
-
-We will be able to make new wall panels and either replace or restore
-the wood floor panels.
-
-We hope to have all main cabin painting done this summer, so that the
-wall and floor panels can be installed.
-
-One other work item that may be started is to clean and restore the
-four engine nacelles.  Pipes and fittings need to be removed, followed
-by a thorough cleaning and a fresh coat of paint.  This work may be
-delayed until qualified volunteers become available.
-
-\section{March 2016 Update}
-\label{sec:update}
-
-Since our report at the November AGM, the aircraft was moved inside,
-so that it could properly sealed against the weather, as it will spend
-much of 2016 outside. The aircraft was moved out again in March.
-During this time, very little work was done on the aircraft, except
-for replacement of the Plexiglas in the port windows, and the
-installation of the ARC-552 UHF radio, kindly transferred to the
-museum by the Department of National Defence.
-
-\subsection{Nr 4 Engine}
-\label{sec:update-engine}
-
-The engine frame has been refurbished, and the engine was transferred
-to the frame, so that ancillary equipment can be installed.  Work has
-also begun on the supercharger, as well as such items as the starter
-motor and header tank.  Some work is also being done on the cowl panels.
-
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   %name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
-%   %caption of the figure 
-%   \caption*{\small \em Forward cockpit -- complete.}
-%   %label of the figure, which has to correspond to \ref{}:
-%   \label{fig:fwd_cockpit_complete}
-%\end{figure}
+We have decided not to move the aircraft outside this year. With our efforts
+concentrated on completing engine \#4, it is unlikely we would have the resources
+to also complete the repair work in the cabin needed to allow for priming and
+painting, which is the main reason for moving the aircraft outside in the summer.
 

### 2017-03-13 pnsac/newsletter/article_EditorsNotes.tex
Incorporated changes from first draft review.

@@
 
-This past year started off as a repeat of the difficult year the
-Association experienced in 2014. Consequently there was little news
-and less enthusiasm for publishing the Chronicle on any regular
-basis. However, as a result of changes to the management of the museum
-organization there is a renewed sense of hope that we will be able to
-move forward in 2016, as expressed in the Notes from the
-President. One of those changes is the appointment of Chris Kitzan as
-the new director general of CASM. You will find a message from Chris
-in this edition. Chris and his boss Alex Benay, the CEO of the museum
-corporation, are enthusiastic supporters of our organization. We are also
-commencing a new feature highlighting the members of our Association
-and starting off with Bruce Gemmill, our Project Manager. In addition,
-we are adding a "Miscellany" feature which lists items of interest to
-our members. If any of our readers wish to contribute an article, have
-suggestions for future articles, have suggestions for the "Miscellany"
-feature, or have any comments about the Chronicle please let me know.
+As the President has noted the Association has, through its volunteers,
+expended thousands of hours on the North Star's restoration since 2004. In the
+last couple of years, however, various developments at CASM have resulted in a
+diminution in the amount of restoration work we have been able to do. The result
+has been a decline in the number of volunteers working on the aircraft, and a
+corresponding decline in our membership. This in turn has led to an overall
+decline in the Association's activities, including the publication of this
+newsletter. CASM has made some major changes to its personnel, including those
+directly responsible for the restoration of the North Star. The Association is
+already experiencing a positive change in its relationship with the Museum which
+can only be to the good. We look forward to a renewed partnership with CASM to
+our mutual benefit.
 

### 2017-03-13 pnsac/newsletter/article_Feature1.tex
Incorporated changes from first draft review.

--- 2016-01-17 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{Volunteers Working on North Star Restoration}
+\title{Reunion of Giants}
 %\subtitle{Part 4}
-\author{Tom MacGregor, Legion Magazine}
+\author{Bruce Grant}
 
@@
 
-\begin{quotation}
-	You have to be careful opening up the panels covering the 
-	engines, you never know if a bird or some animal might have made
-	its home there.
-\end{quotation}
+The "Reunion of Giants" video came to the Canadian Aviation and Space Museum's
+theatre on May 14th, courtesy of the Canadian Warplane Heritage Museum in
+Hamilton. CWHM sent along the pilot Captain Leon Evans (retired Air Canada), and
+Dave Rohrer, CEO, as well as Morgan Elliott the videographer. Also present were
+two World War Two decorated Lancaster pilots Don McKechnie and Bill Button. With
+almost a full house of 200 it was a very successful Project North Star public
+event.
 
-That’s one lesson a group of dedicated volunteers has learned in its
-10 years of painstaking work restoring a Royal Canadian Air Force
-North Star aircraft at the Canada Aviation and Space Museum in Ottawa.
+The outstanding feature of the story is the audacity of CWHM to fly Canada's
+only airworthy Lancaster to England for a reunion with the only other flying
+Lancaster in the world. The Lanc is the flagship of the CWHM, its biggest
+attraction and an irreplaceable historic artifact. In a typical year it flies
+only about fifty hours, attending aviation events in Canada and every move is
+calculated some years in advance.
+
+In England, the arrival over the North Atlantic route of our Lanc was embraced
+as a tribute to the quality of the machine and the courage of its crew. It
+reinforced the link with Canada which provided so much support to Great Britain
+during the war.
+
+While on a final approach in England, the supercharger on engine \#3 went to
+pieces and the plane landed with a trail of smoke. A replacement engine would
+have to be found. The replacement that turned up was a Rolls Royce Merlin that
+would have to be substituted for the Packard Merlin in our Canadian-built
+Lancaster. This presented a challenge in sorting out the elements of QEC with
+incompatible Whitworth and SAE fasteners. The new engine would be a "bitsa"
+(bitsa this and bitsa that). They did it in about three days and got the plane
+back in the air.
+
+On the return flight, stronger than anticipated headwinds raised the risk of
+running out of gas over the Davis Strait. They diverted to Narsarsuaq in
+Greenland.
+
+Following the showing of the film there was a spirited question and answer
+session with CWHM's representatives moderately by Mike Pearson (retired Air
+Canada pilot of the Gimli glider fame). There is no doubt that all those in
+attendance enjoyed themselves thoroughly and gained some understanding of what
+it meant for those involved in this project.
+
+At Project North Star, we have some appreciation idea of the scale of CWHM's
+accomplishment. As our members know we are in the twelfth year of our project to
+restore the last surviving North Star. But, we are restoring it only as a museum
+artifact, not an airworthy machine. The video is available for sale at
+www.warplane.com and the trailer can be viewed ath www.suddenlyseemore.com .
+
+%\begin{quotation}
+%	\textit{You have to be careful opening up the panels covering the 
+%	engines, you never know if a bird or some animal might have made
+%	its home there.}
+%\end{quotation}
 
@@
 
-The aircraft had sat neglected outside the museum for
-years. “Essentially, the RCAF flew this one into the Rockcliffe Airport
-when it was decommissioned. It sat in a old until Robert Holmgren said
-someone should do something about it,” said project manager Bruce
-Gemmill.
-
-The airplane had been stripped of its non-essential gear and turned
-over to what was then the National Aeronautical Collection, housed in
-hangars built during the Second World War at the former RCAF station
-at Rockcliffe in Ottawa. Time and weather took its toll. Birds and
-animals made nests where they could get inside the panelling.
-
-“We approached the museum about restoring the aircraft. At first, they
-didn’t know how to take us,” said Gemmill. “The museum had its own
-professional conservation staff but it had never worked with a team of
-volunteers.”
-
-Eventually the museum agreed to work with the volunteers who formed
-the Project North Star Association of Canada with the late Robert
-Holmgren as president. “We became the model for other volunteer groups
-now working with the museum,” said the association’s current
-president, Richard Lodge.
-
-Now the aircraft gets towed into a hangar where it is available to the
-volunteers between September and May before it is towed outside for
-the summer.
-
-The North Star was built by Canadair in the late 1940s and 1950s. It
-was the RCAF version of the DC-4 civilian aircraft except that it used
-four Rolls-Royce Merlin engines. Canadair built 71 of the aircraft.
-
-The RCAF assigned the rst North Stars to 412 Squadron where they
-transported VIPs and were used in various transfigurations for
-reliable, long-range transport services.
-
-During the Korean War the North Star was used by No. 426 (Thunderbird)
-Sqdn. to ferry sup- plies across the Pacific Ocean to Japan (Operation
-Hawk: The Korean Airlift, July/August). They would y 599 round trips
-over the Pacific and deliver seven million pounds of cargo and 13,000
-personnel on return trips. All this was achieved without a fatal
-crash.
-
-Most of the air force’s North Stars were declared surplus in the
-1970s.
-
-In the 10 years that Project North Star has been in operation, the
-group has restored two of the four Merlin engines with a third one
-nearly completed.
-
-Each engine had to be taken o the aircraft, brought into the museum’s
-shops where they were taken apart. Each item was tagged, catalogued 
-and usually photographed. “We take a lot of photographs. That’s
-how we know how they go back together,” said Lodge.
-
-Engine project leader Garry Dupont said, “The first engine took us about
-four and a half years to restore. We’re almost finished the third engine
-which as taken us about two years. So we are getting faster.” The
-propellers were restored separately by Hope Aero Propeller and
-Components Inc. in Mississauga, Ont. All restoration work has to be up
-to the professional standards set by the museum.
-
-The group plans to restore the aircraft to its original look, not to
-have it y again.
-
-Lodge said that they have some RCAF manuals but a lot of time they do
-not help when it comes to putting 50-year-old parts together today.
-
-The aircraft itself, 17515, has seen countless hours of work put into
-re- storing the cockpit with all its switches and navigation
-equipment and a small galley for the crew of seven.
-
-Still to be done is work in the cargo section and painting the
-exterior of the aircraft.
-
-“We know this aircraft ew during the Korean War. In fact, we have 
-pictures showing it configured with litters to bring the wounded back to
-Canada,” said Gimmell. “We don’t want to put in the litters but we
-have people working on creating the rope seats that were used.”
-
-The plan is to paint the aircraft to its post-1965 markings, since it
-was painted with the Maple Leaf ag on it.
-
-Still the search goes on to nd new members for the association. “The
-important thing is to nd new people to whom we can pass on this 
-knowledge and skills we are developing,” said Lodge.
+%\textit{Reprinted from the March/April 2014 edition of "Legion Magazine," with the kind permission of Jennifer Morse, General Manager of Canvet Publications, and Tom MacGregor.}
 

### 2017-03-13 pnsac/newsletter/article_MemberInterview.tex
Incorporated changes from first draft review.

--- 2016-01-17 pnsac/newsletter/article_MemberInterview.tex
@@
 \title{Our Members}
-\author{Interview with Bruce Gemmill}
+\author{Interview with Jm Riddoch}
 
@@
 
-\textit{Bruce Gemmill has been involved with Project North Star since 2005 and
-	has exceeded 9,000 volunteer hours. Bruce is presently the Project
-	Manager, as well as a member of the board of directors.}
+\begin{quotation}
+	\textit{Jim Riddoch is one of the earliest and longest serving members of the
+Association. He has made a significant contribution to the operations of the
+Association including working on the aircraft as well as being an officer and
+member of the board of directors. Although officially retired from formal
+positions in the organization, he can still be counted on to help out when
+needed.}
+\end{quotation}
+
+\begin{multicols}{2}
+
+1.\textit{What is your background in aviation?}
+
+I started with English Electric Aviation Company in 1956 as an
+apprentice technician eventually being assigned to the Mechanical
+Engineering Department as a fuel systems technician. I remained there
+for 5 years until I was 28 years old and married with a daughter. After
+immigrating to Canada in March 1966 I joined DeHavilland Spar Division
+in Malton and then moved to Montreal with Jarry Hydraulics as a
+development engineer. I remained there for 2 years working chiefly on
+development proposals for the wing sweep actuators for F111 aircraft
+and Boeing's proposed supersonic Concorde.
+
+In 1968 I joined Air Canada as a junior engineer in the Mechanical
+Systems Department and assigned as a landing gear technologist until I
+achieved Professional Engineer status in 1974. I then moved around
+several other departments in engineering and maintenance until I was
+promoted to Superintendent of DC9 maintenance. In that capacity I was
+involved in the accident investigation into the loss of one aircraft at
+Cincinnati.
+
+I eventually returned to engineering as Director of Interior Systems and
+Equipment to supervise extensive fleet modifications and new aircraft
+acquisitions. I remained in that position until I retired in 1990.
+Following a short stint with First Air I was asked to join a newly formed
+Canadian Aviation Maintenance Council to standardize aircraft
+maintenance trades basic training. I was hired as the Accreditation
+Manager to approve various training establishments including schools,
+companies and military programs, I eventually also took on the job of
+Registration Manager to accept qualified trade technicians.
+
+2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}
+
+Following my retirement from the Canadian Aviation Maintenance
+Council in 2003 I was approached by Robert Holmgren to participate in
+a voluntary program to help restore an the Canadair North Star aircraft
+parked outside at the Canada Aviation Museum. Along with Tim
+Timmins and a few others we formed a steering committee to approach
+the Museum about forming a voluntary work force to assist the museum
+staff to restore the aircraft. At first we met with opposition from some
+staff members who felt their jobs were being compromised and possible
+staff reductions. Slowly we were approved by the Museum staff with
+limited supervision and controlled access and work direction. I was
+assigned by the Project North Star Association as Chief Engineer of the
+project although all work was controlled and supervised by the
+Museum's staff manager.
+
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Forward cockpit -- complete.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:fwd_cockpit_complete}
+%\end{figure}
+
+3. \textit{What has been the history of your involvement to date?}
+
+I worked on the Project for 10 years starting in 2003 both as a “hands
+on” volunteer as well as a member of the board of directors and officer
+of the Association. Generally I put in at least two days per week and
+sometimes more.
+
+%\begin{figure}[htbp]
+%	\vspace{2em}
+%	\centering
+%	%name of the graphic, without the path AND in EPS format:
+%	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
+%	%caption of the figure 
+%	\caption*{\small \em Propeller assembly.}
+%	%label of the figure, which has to correspond to \ref{}:
+%	\label{fig:propeller}
+%\end{figure}
+
+%\begin{figure}[htbp]
+%	\vspace{2em}
+%	\centering
+%	%name of the graphic, without the path AND in EPS format:
+%	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
+%	%caption of the figure 
+%	\caption*{\small \em Forward cockpit -- complete.}
+%	%label of the figure, which has to correspond to \ref{}:
+%	\label{fig:fwd_cockpit_complete}
+%\end{figure}
+
+4. \textit{What has been the highlight of your involvement?}
+
+The chief highlights have been the high level of workmanship of the
+volunteers and displaying the aircraft to the public and interested
+persons.
+
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
+%   %caption of the figure 
+%   \caption*{\small \em Navigator rack.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:nav-radio-rack}
+%\end{figure}
+
+5. \textit{What has been the most challenging part of your involvement?}
+
+Keeping the Museum staff and management on side with our goals
+and objectives and recruiting enough new volunteers to keep the
+project going.
 

### 2017-03-13 pnsac/newsletter/article_RecentEvents.tex
Incorporated changes from first draft review.

% Template PNSAC newsletter
% Language: Latex

\title{Canada Day 2016}
\author{Bill Tate}

\maketitle

Canada Day at the Museum always provides us with an opportunity to show off what
PNSAC is and does; and July 1, 2016 was no exception. Fortunately the weather
was ideal for most of the day. Our volunteers partially set up our display on
June 30th and the balance was done early in the morning of July1st.

Our display, unlike the picture shown below from 2011, was limited by the nose
section being shrink wrapped in plastic to prevent water damage. Consequently,
we unable to show off the cockpit because of lighting issues and ventilation.
This could have been rectified with additional fans and lights but with
restoration work being done on the main cabin floor, visits were limited to
viewing from just inside the aircraft rear cargo door.

Although not widely known we are also limited by the number of people. we can
have inside the aircraft at any one time. There is a reason for this as an empty
aircraft "generally" has the centre of gravity in the aft range limit and with
the engine \#4 off the aircraft this becomes even more pronounced. For this
reason sand bags have been placed in the forward cargo hold to help keep the
centre of gravity as far forward as possible. This limits us to having a maximum
of four adults or a family with children in at one time and yes we have to
include our volunteers so that we do not exceed the equivalent of six adults.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{trenton2013-herc.eps}
%   %caption of the figure 
%   \caption*{\small \em National Air Force Museum -- Hercules.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:herc}
%\end{figure}

\end{multicols}

\begin{figure*}[ht!]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.75]{canada-day-2016}
   %caption of the figure 
   \caption*{\small \em Canada Day 2016.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:canadaday}
\end{figure*}

\begin{multicols}{2}

The weather forecast that day was for thunder storm activity and as the day
progressed the skies darkened. For the safety of our volunteers and public we
shut down our display earlier than normal because as the wind increased we had
some of our story boards blown over by gusts of wind. As an aside, airports with
"Thor Guard" cease ramp operations when a thunderstorm is within 6 nautical
miles.

Our display attracted many visitors and there was a continuous line of people
waiting patiently for their turn to climb the steps and view the aircraft
interior. As in previous years we also took the covers off engine \#2 so we could
show an engine to our visitors after it had been fully restored.

I would like to thank all our volunteers who made Canada Day such a success and
to mention the very demanding work performed by Bruce Grant and Jim Riddoch who
handled crowd control, and John Thibert and Charles Baril who spent several
hours in the hot interior of the North Star talking to our visitors about the
work that went on inside the airplane.

Any suggestions to improve our display, who we are and what we do would be most
appreciated. Please send to Bill Tate.

\begin{footnotesize}
    \raggedleft PNSAC\\
\end{footnotesize}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: main_document.tex
%%% End:

### 2017-03-15 pnsac/newsletter/article_MemberInterview.tex
Draft 3--sent to Roger for review.

@@
 several other departments in engineering and maintenance until I was
-promoted to Superintendent of DC9 maintenance. In that capacity I was
+promoted to Superintendent of DC 9 maintenance. In that capacity I was
 involved in the accident investigation into the loss of one aircraft at
@@
 
-I worked on the Project for 10 years starting in 2003 both as a “hands
-on” volunteer as well as a member of the board of directors and officer
+I worked on the Project for 10 years starting in 2003 both as a "hands
+on" volunteer as well as a member of the board of directors and officer
 of the Association. Generally I put in at least two days per week and

### 2017-03-15 pnsac/newsletter/article_PresReport.tex
Draft 3--sent to Roger for review.

@@
 
-The last North Star Chronicle was published in March 2016 and since that time we
-have successfully continued our work on the aircraft while the Museum was making
-preparations for the Star Trek event in the Reserve Hangar. We were not asked to
-suspend our work, unlike in 2014 when preparations for the Star Wars event were
-being made. This has helped us to retain our Restoration volunteers and recruit
-two new ones. However, we are still short of Restoration volunteers and continue
-to actively work with the Museum to address this problem.
+The last NStar Chronicle was published in June 2016 and since that time we have
+continued our work on the aircraft, but at a greatly reduced pace. Last summer
+the Museum staged a Star Trek event in the Reserve Hanger which affected our
+ability to work on the North Star. We were not asked to suspend our work until
+the end of September, unlike in 2014 when preparations for the Star Wars event
+were being made. This helped us retain our Restoration volunteers and recruit
+two new ones.
 
-When Project North Star was originally formed in 2004, the Association was
-treated with considerable skepticism by the Museum. It was felt that restoring
+At the end of September, Mike Irvin, the Museum's Conservation Special Projects
+Manager, retired and we were asked to suspend work on the aircraft until a new
+manager was appointed. In January, R\'{e}jean (Rej) Demers was appointed as the new
+Special Projects Manager. Our work on the aircraft resumed on January 30th.
+
+When Project North Star was originally formed in 2003, the Association was
+treated with considerable scepticism by the Museum. It was felt that restoring
 an aircraft of the size and complexity of a North Star, to the exacting
@@
 
-We will soon have completed 70,000 hours of volunteer work over a period of 14
-years, restoring everything we do, to what is acknowledged to be a very high
-standard. Skepticism is a thing of the past and we are now treated as a
+We have now completed approximately 70,000 hours of volunteer work over a period
+of 14 years, restoring everything we do to what is acknowledged to be a very
+high standard. Scepticism is a thing of the past and we are now treated as a
 respected and ongoing part of the Museum operations.
 
-Respect has to be earned and comes through hard work and in the case of aircraft
+Respect must be earned and comes through hard work and in the case of aircraft
 restoration work, attention to detail. Our team of volunteers needs guidance
@@
 been particularly important in helping us the get to the position we are now in.
-He has recently reached a milestone only reached on two other occasions by a
-volunteer and in each case the volunteer worked in the the Science and
-Technology Museum. Our volunteer has reached 10,000 volunteer hours of working
-on the North Star and throughout this time he also acted as our Project Manager
-for everything except the engines. Bruce Gemmill probably was not aware that he
-had reached 10,000 hours when he unfortunately had to announce that he would
-have to suspend his activities as a volunteer and director of PNSAC due to
-family circumstances. We very much hope that he will soon be able to rejoin us
-and continue to help us complete this large project. In the meantime Bruce will
-continue to be our Membership Secretary. During June, Bruce's remarkable
-contribution was recognized by the Museum and in a private event he was
-presented with a certificate and a gift by Cedric St Amour, the Museum
-Corporation Volunteer Coordinator. One of our volunteers, Robert Desjardins,
-also joined Cedric on the occasion to congratulate Bruce on behalf of PNSAC.
-Well done Bruce and we are all very proud of your achievement.
+Last year he reached a milestone only achieved on two other occasions by a
+volunteer and in each case the volunteer worked at the Science and Technology
+Museum. Our volunteer has worked 10,000 volunteer hours on the North Star and
+for most of this time he also acted as our Project Manager for everything except
+the engines. Bruce Gemmill probably was not aware that he had reached 10,000
+hours when during June, his remarkable contribution was recognized by the Museum
+and in a private event he was presented with a certificate and a gift by Cedric
+St-Amour, the Museum Corporation Volunteer Co-ordinator.
 
-I have always seen our Association as having multiple roles. First and foremost
-we are restoring a major artifact with our volunteers; secondly we show off our
+I have always seen our Association as having multiple roles. First and foremost,
+we are restoring a major artifact with our volunteers. Secondly, we show off our
 restoration work to the public which in turn brings visitors and revenue into
-the Museum and thirdly through our experienced volunteers we help to transfer
+the Museum and thirdly through our experienced volunteers we help transfer
 skills to younger and less experienced volunteers.
 
-Planning our work in conjunction with the Museum management and the conservation
-staff is something we must always try to jointly do better. It is often a fine
-balancing act to decide when to do work on the aircraft in relation to when the
-Museum is having a major event such as Canada Day. I am hoping that in future we
-can regularly update our work schedule in conjunction with our project plan so
-that we can more readily have the aircraft available for public viewing at the
-times when the Museum would like us to do so, without having to delay our
-restoration work unnecessarily.
+Planning our work in conjunction with the Museum management and the Conservation
+staff is something we must continuously work at. It is often a fine balancing
+act to decide when to do work on the aircraft in relation to when the Museum is
+having a major event such as Open Doors Ottawa and Canada Day. I am hoping that
+in future we can regularly update our work schedule in conjunction with our
+project plan so that we can more readily have the aircraft available for public
+viewing at the times when the Museum would like us to do so, without having to
+delay our restoration work unnecessarily.
+
+During the last year, we were able to hold two of our successful Lecture Series
+events and are planning to hold another one in May of this year. More details of
+the next event will be available soon.
 
 I will conclude with an invitation to any of our members not volunteering at
-present to come and join us. Although we have a definite limit on the number of
-volunteers who can get their hands dirty working on restoration work, we always
-welcome others who would like to participate in the running of the organization.
-By becoming involved either as a restoration volunteer or an association
-volunteer you will meet other like minded people and have an interesting time
-talking to members of the public. You do not need to be an expert on aviation or
-on the North Star to make a contribution. I can be reached by email or phone if
-you would like to find out more about volunteering.
+present to come and join us. Although we have a limit on the number of
+volunteers who can get their hands dirty working on restoration, we always
+welcome others who would like to join us at our events. You will meet other
+likeminded people and have an interesting time talking to members of the public.
+You do not need to be an expert on aviation or on the North Star to contribute.
+You can contact me by email at the address at the end of this newsletter if you
+would like to find out more about volunteering.
 

### 2017-03-15 pnsac/newsletter/article_ProgReport.tex
Draft 3--sent to Roger for review.

@@
 
-\title{Project Manager's Progress Report}
+\title{Project Progress Report}
 \subtitle{PNS 2016 Status Update}
-\author{Bruce Gemmill}
+\author{Bruce Gemmill, Peter Trowbridge, Garry Dupont}
 
@@
    \includegraphics[scale=0.5]{"PNS 007-im2".png}
-   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
+   \caption*{\small \em Volunteer Peter Trowbridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
    \label{fig:engine_no_4_2}

### 2017-03-19 pnsac/newsletter/article_PresReport.tex
Incorporating latest changes after review of DRAFT #3.

@@
 
-The last NStar Chronicle was published in June 2016 and since that time we have
+The last NStar Chronicle was published in March 2016 and since that time we have
 continued our work on the aircraft, but at a greatly reduced pace. Last summer
@@
 the end of September, unlike in 2014 when preparations for the Star Wars event
-were being made. This helped us retain our Restoration volunteers and recruit
-two new ones.
+were being made.
 
-At the end of September, Mike Irvin, the Museum's Conservation Special Projects
+At the end of September 2016, Mike Irvin, the Museum's Conservation Special Projects
 Manager, retired and we were asked to suspend work on the aircraft until a new
@@
 
+The Annual General Meeting of the Association was held on March 11, 2017 at the 
+Museum. It was gratifying to see a good turnout of members, given the difficult
+times our organization has gone through in the past few years. Both the Museum's 
+Volunteer Coordinator, Cedric St. Amour, and the new Special Projects Manager 
+R\'{e}jean Demers gave up their Saturday mornings to come and address our members. 
+Their presentations were very well received and are further evidence of a new 
+era in our relationship with the Museum.
+
 I will conclude with an invitation to any of our members not volunteering at
 present to come and join us. Although we have a limit on the number of
-volunteers who can get their hands dirty working on restoration, we always
-welcome others who would like to join us at our events. You will meet other
-likeminded people and have an interesting time talking to members of the public.
-You do not need to be an expert on aviation or on the North Star to contribute.
-You can contact me by email at the address at the end of this newsletter if you
-would like to find out more about volunteering.
+volunteers who can get their hands dirty working on restoration, our organization 
+needs  volunteers to assist with other Association activities including merchandise
+sales, special events, fund raising, communications and administrative work. Whether
+you are working on the plane or doing other Association work you will meet likeminded
+people and have the opportunity to speak to the general public. You do not need to 
+be an expert on aviation or on the North Star to contribute. You can contact me by 
+email at the address at the end of this newsletter if you would like to find out 
+more about volunteering.
 

### 2018-01-14 pnsac/newsletter/article_EditorsNotes.tex
Draft 3--sent to Roger for review.
New content for Jan 2018 Vol 14, issue 1.

@@
 
-As the President has noted the Association has, through its volunteers,
-expended thousands of hours on the North Star's restoration since 2004. In the
-last couple of years, however, various developments at CASM have resulted in a
-diminution in the amount of restoration work we have been able to do. The result
-has been a decline in the number of volunteers working on the aircraft, and a
-corresponding decline in our membership. This in turn has led to an overall
-decline in the Association's activities, including the publication of this
-newsletter. CASM has made some major changes to its personnel, including those
-directly responsible for the restoration of the North Star. The Association is
-already experiencing a positive change in its relationship with the Museum which
-can only be to the good. We look forward to a renewed partnership with CASM to
-our mutual benefit.
+Our Association was incorporated in 2003. We are fortunate that a
+number of the original members of the organization, such as Tim
+Timmins, are still involved as they can provide an historical
+perspective. As Tim says, in the article "Our Members," there have been
+many bumps in the road, and the fact that the aircraft is still a long
+way from completion after 16 years speaks for itself. In January 2017,
+R\'{e}jean (Rej) Demers took up his position as CASM's Conservation Project
+Manager. Rej's first contribution to the Chronicle can be found in the
+"Conservator's Corner." The restoration of the North Star is a priority
+for Rej. He has spent the first year of his mandate gathering
+information and developing systems and procedures which will, in the
+long run, move the restoration of the aircraft forward in a coherent
+manner. I have previously stated the obvious and that is the health and
+vigour of the Association is dependent on the restoration status of the
+aircraft. In the last few years there have been as many bumps in the
+road as there are potholes in Ottawa in the spring. The President is
+very optimistic we have reached a real turning point (see the
+"President's Notes") and wouldn't that be great.
 
-%\vspace{12pt}
+\vspace{12pt}
 

### 2018-01-14 pnsac/newsletter/article_MemberInterview.tex
New content for Jan 2018 Vol 14, issue 1.

--- 2017-03-13 pnsac/newsletter/article_MemberInterview.tex
@@
 \title{Our Members}
-\author{Interview with Jm Riddoch}
+\author{Interview with Austin J Timmins}
 
@@
 
-\begin{quotation}
-	\textit{Jim Riddoch is one of the earliest and longest serving members of the
-Association. He has made a significant contribution to the operations of the
-Association including working on the aircraft as well as being an officer and
-member of the board of directors. Although officially retired from formal
-positions in the organization, he can still be counted on to help out when
-needed.}
-\end{quotation}
+%\begin{quotation}
+%	\textit{Jim Riddoch is one of the earliest and longest serving members of the
+%Association. He has made a significant contribution to the operations of the
+%Association including working on the aircraft as well as being an officer and
+%member of the board of directors. Although officially retired from formal
+%positions in the organization, he can still be counted on to help out when
+%needed.}
+%\end{quotation}
 
@@
 
-I started with English Electric Aviation Company in 1956 as an
-apprentice technician eventually being assigned to the Mechanical
-Engineering Department as a fuel systems technician. I remained there
-for 5 years until I was 28 years old and married with a daughter. After
-immigrating to Canada in March 1966 I joined DeHavilland Spar Division
-in Malton and then moved to Montreal with Jarry Hydraulics as a
-development engineer. I remained there for 2 years working chiefly on
-development proposals for the wing sweep actuators for F111 aircraft
-and Boeing's proposed supersonic Concorde.
+My aviation experience includes 36 years service in the RCAF and nine
+years with the Canadian Aeronautics and Space Institute. In the RCAF I
+completed radio navigator training, which included communications,
+navigation and air gunner. I later qualified as an airborne
+interception navigator . My operational experience includes air
+transport, search and rescue and training. My assignments included
+Navigation Leader, Aircrew Leader, Squadron Commander, Air Command
+Navigator and Deputy Commander (COO) of the Air Transport Group. I
+logged 9500 hours , mostly on the following aircraft: Beechcraft,
+Dakota, North Star , Mitchell, Yukon, Hercules, and Boeing 707. My
+staff assignments in Europe included movement control (traffic
+management) for 1 Air Division and a tour at The NATO Defense College
+in Rome.
 
-In 1968 I joined Air Canada as a junior engineer in the Mechanical
-Systems Department and assigned as a landing gear technologist until I
-achieved Professional Engineer status in 1974. I then moved around
-several other departments in engineering and maintenance until I was
-promoted to Superintendent of DC9 maintenance. In that capacity I was
-involved in the accident investigation into the loss of one aircraft at
-Cincinnati.
-
-I eventually returned to engineering as Director of Interior Systems and
-Equipment to supervise extensive fleet modifications and new aircraft
-acquisitions. I remained in that position until I retired in 1990.
-Following a short stint with First Air I was asked to join a newly formed
-Canadian Aviation Maintenance Council to standardize aircraft
-maintenance trades basic training. I was hired as the Accreditation
-Manager to approve various training establishments including schools,
-companies and military programs, I eventually also took on the job of
-Registration Manager to accept qualified trade technicians.
+On retirement from the RCAF I participated in a Transport
+Canada A Base review of several airports. I then secured full
+employment with Canadian Aeronautics and Space Institute as Executive
+Director, retiring in 1995.
 
@@
 
-Following my retirement from the Canadian Aviation Maintenance
-Council in 2003 I was approached by Robert Holmgren to participate in
-a voluntary program to help restore an the Canadair North Star aircraft
-parked outside at the Canada Aviation Museum. Along with Tim
-Timmins and a few others we formed a steering committee to approach
-the Museum about forming a voluntary work force to assist the museum
-staff to restore the aircraft. At first we met with opposition from some
-staff members who felt their jobs were being compromised and possible
-staff reductions. Slowly we were approved by the Museum staff with
-limited supervision and controlled access and work direction. I was
-assigned by the Project North Star Association as Chief Engineer of the
-project although all work was controlled and supervised by the
-Museum's staff manager.
-
+My involvement with efforts to conserve North Star 17515 predates
+Project North Star. Ottawa area members of the 426 Thunderbird Squadron
+Association, mostly retired North Star aircrew, met with the National
+Aviation Museum staff to express their concern about the state of the
+North Star and proposed volunteer involvement in a preservation effort.
+These expressions of concern and interest led to a meeting between the
+Association representatives, where General Adamson presented a proposal
+for the restoration of the aircraft. It was rejected but assurance was
+given that the North Star would be first priority when the new hangar
+was completed Also a North Star Trust Fund would be established. Robert
+Holmgren, an observer at the meeting, then proceeded to develop a
+proposal for volunteers to restore the North Star.
 %\begin{figure}[htbp]
@@
 
-I worked on the Project for 10 years starting in 2003 both as a “hands
-on” volunteer as well as a member of the board of directors and officer
-of the Association. Generally I put in at least two days per week and
-sometimes more.
+I have been involved from the beginning of efforts to establish PNS,
+first as a member of a Steering Committee, led by Robert Holmgren,
+later, Vice President and then President of PNSAC. My contacts in
+aviation provided much needed advice and support, including access to
+initial funding for PNSAC.
 
@@
 
-The chief highlights have been the high level of workmanship of the
-volunteers and displaying the aircraft to the public and interested
-persons.
+The highlight of my involvement has to be the establishment of PNS and
+PNSAC.
 
@@
 
-Keeping the Museum staff and management on side with our goals
-and objectives and recruiting enough new volunteers to keep the
-project going.
+PNS is now in its 16th year. It has been a work in progress with many
+bumps along the way. I spent a lot of time and effort, along with other
+members of the Executive, evening out these bumps. Establishing and
+maintaining effective channels of communication with CASM and CSTM
+required constant attention. Much effort was directed to fund raising,
+publicity, and, finding volunteers. In the beginning, every date
+brought a new challenge.
 

### 2018-01-14 pnsac/newsletter/article_PresReport.tex
New content for Jan 2018 Vol 14, issue 1.

--- 2017-03-15 pnsac/newsletter/article_PresReport.tex
@@
 
-The last NStar Chronicle was published in June 2016 and since that time we have
-continued our work on the aircraft, but at a greatly reduced pace. Last summer
-the Museum staged a Star Trek event in the Reserve Hanger which affected our
-ability to work on the North Star. We were not asked to suspend our work until
-the end of September, unlike in 2014 when preparations for the Star Wars event
-were being made. This helped us retain our Restoration volunteers and recruit
-two new ones.
+Since the last issue of the NStar Chronicle in March of this year we
+have continued to reorganize our volunteer operations under the
+leadership of the Museum's Special Projects Manager, R\'{e}jean (Rej)
+Demers. We have started a recruitment process for new restoration
+volunteers and to date two new people have started work. During the
+summer two more permanent Museum staff joined Rej to replace people who
+have retired. It is obviously taking time for the new Museum staff to
+get to know the complex operations of the Aviation Museum, to get to
+know the Project North Star volunteers and assess their various talents
+and areas of expertise. Consequently, our summer has been quiet but
+steadily productive.
 
-At the end of September, Mike Irvin, the Museum's Conservation Special Projects
-Manager, retired and we were asked to suspend work on the aircraft until a new
-manager was appointed. In January, R\'{e}jean (Rej) Demers was appointed as the new
-Special Projects Manager. Our work on the aircraft resumed on January 30th.
+For several reasons it became necessary to reorganize the location of
+several major Museum artifacts. This affected the North Star aircraft
+which spent the summer outside the Reserve Hanger resulting in some
+problems with water ingress and further bird activity. I am pleased to
+say that the aircraft is now back inside the hangar.
 
-When Project North Star was originally formed in 2003, the Association was
-treated with considerable scepticism by the Museum. It was felt that restoring
-an aircraft of the size and complexity of a North Star, to the exacting
-standards of a national museum, was likely to be beyond the capabilities of a
-group of volunteers and that the project would probably fail.
+The Museum is actively supporting Project North Star and trying to
+facilitate our restoration work wherever possible and is promoting our
+work and future activities. Our Treasurer, Paul Labranche, and I
+recently met with the Museum Foundation. The Foundation is the umbrella
+organization with responsibility for fundraising and business
+development at the three museums, comprising the Museum Corporation,
+(Aviation, Science \& Tech and Agriculture). We had a very productive
+meeting, which among other things has resulted in the Foundation
+deciding to highlight Project North Star in its Fall fundraising mail
+out. This type of collaboration is very good for our Association and we
+hope as good for the Foundation.
 
-We have now completed approximately 70,000 hours of volunteer work over a period
-of 14 years, restoring everything we do to what is acknowledged to be a very
-high standard. Scepticism is a thing of the past and we are now treated as a
-respected and ongoing part of the Museum operations.
+Our biggest challenge at present is to get our website properly
+operational again. A couple of our volunteers with website knowledge
+have indicated a willingness to do the necessary work and we are hoping
+to have the site working and up to date by the time this issue of the
+Nstar Chronicle is published.
 
-Respect must be earned and comes through hard work and in the case of aircraft
-restoration work, attention to detail. Our team of volunteers needs guidance
-both from the Museum staff and from our own senior volunteers. One volunteer has
-been particularly important in helping us the get to the position we are now in.
-Last year he reached a milestone only achieved on two other occasions by a
-volunteer and in each case the volunteer worked at the Science and Technology
-Museum. Our volunteer has worked 10,000 volunteer hours on the North Star and
-for most of this time he also acted as our Project Manager for everything except
-the engines. Bruce Gemmill probably was not aware that he had reached 10,000
-hours when during June, his remarkable contribution was recognized by the Museum
-and in a private event he was presented with a certificate and a gift by Cedric
-St-Amour, the Museum Corporation Volunteer Co-ordinator.
+On December 2nd we held a Members' meeting at the Museum when we were
+fortunate to have Jim Mason join us to talk about the restoration of an
+early Lockheed aircraft flown by TCA. The meeting was well attended in
+a new venue of the Avro Arrow room instead of the Bush Theatre.
+Following Jim's interesting presentation, the group moved to the
+Museum's Restoration Hangar where Rej talked about the progress of the
+North Star restoration and the plans for work in 2018.
 
-I have always seen our Association as having multiple roles. First and foremost,
-we are restoring a major artifact with our volunteers. Secondly, we show off our
-restoration work to the public which in turn brings visitors and revenue into
-the Museum and thirdly through our experienced volunteers we help transfer
-skills to younger and less experienced volunteers.
+One of the initiatives currently underway is to create a visual
+timeline of the work already completed on the North Star restoration
+and the planned future work to bring the project to completion. It is
+anticipated that this visual timeline will be mounted on a display
+board for our members and the public to view.
 
-Planning our work in conjunction with the Museum management and the Conservation
-staff is something we must continuously work at. It is often a fine balancing
-act to decide when to do work on the aircraft in relation to when the Museum is
-having a major event such as Open Doors Ottawa and Canada Day. I am hoping that
-in future we can regularly update our work schedule in conjunction with our
-project plan so that we can more readily have the aircraft available for public
-viewing at the times when the Museum would like us to do so, without having to
-delay our restoration work unnecessarily.
-
-During the last year, we were able to hold two of our successful Lecture Series
-events and are planning to hold another one in May of this year. More details of
-the next event will be available soon.
-
-I will conclude with an invitation to any of our members not volunteering at
-present to come and join us. Although we have a limit on the number of
-volunteers who can get their hands dirty working on restoration, we always
-welcome others who would like to join us at our events. You will meet other
-likeminded people and have an interesting time talking to members of the public.
-You do not need to be an expert on aviation or on the North Star to contribute.
-You can contact me by email at the address at the end of this newsletter if you
-would like to find out more about volunteering.
+In closing, I am optimistic that 2018 will see increased activity both
+in the aircraft restoration work and also within our Association.
 

### 2018-01-14 pnsac/newsletter/article_ProgReport.tex
Incorporating latest changes after review of DRAFT #3.
New content for Jan 2018 Vol 14, issue 1.

--- 2017-03-12 pnsac/newsletter/article_ProgReport.tex
@@
 
-\title{Project Manager's Progress Report}
-\subtitle{PNS 2016 Status Update}
-\author{Bruce Gemmill}
+\title{Conservator's Corner}
+%\subtitle{PNS 2016 Status Update}
+\author{R\'{e}jean Demers}
 
@@
 
-\section{Nr 4 Engine}
-\label{sec:engine_4}
+\textit{R\'{e}jean Demers (Rej, to those who work with him) is a Transport Canada
+licensed Aircraft Maintenance Engineer (AME). R\'{e}jean was a small
+business owner in Gatineau QC for over 10 years, working for a wide
+variety of private and commercial aircraft owners and operators.
+Following a stint in Northern Ontario working on turpoprop freighters
+and bulk fuel haulers, he took up his present position as a
+Conservator, Special Projects Manager with Ingenium, Canada's Museums
+of Science and Innovation in January 2017. His main project is the
+restoration of the Canadair North Star DC-4GM at the Canadian Aviation
+and Space Museum. He recounts below his introduction to his new job.}
 
-Work has continued on much of the ancillary equipment for the engine. The
-supercharger and intercooler have been completed and installed, as well as the
-reduction gearbox. The radiators and header tank have been restored and painted, and
-two of the three radiator sections have been installed. Restoration work has
-begun on the air intake system, and many of the fuel and oil lines.
+Leaving Pickle Lake was like getting shot out of a cannon in a snow
+storm. At that time, Christmas of 2016, I was the last person on my
+crew to leave the YPL base, having seen my colleagues off for the end
+of our tour and the holidays. After a full day of flying, the fleet had
+to be tucked away in the barn ready for service. On Boxing Day, a fresh
+crew would fly in and start another two week tour of line maintenance
+with the company we worked for. However, on that Christmas, I was
+quitting my job.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   \includegraphics[scale=0.5]{"PNS 007-im2".png}
-   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
-   \label{fig:engine_no_4_2}
-\end{figure}
+Nobody knew I was leaving until I got back to the bunkhouse and shared
+Christmas dinner with the remaining pilots and ops personnel. I refer
+to them in these terms now, but then and there, I could say they were
+my friends. In the weeks leading to my departure, no one knew I was
+leaving. No one speaks of breaking out of prison when they are behind
+bars. The difference being the only bars I saw were the Pickle Lake
+Hotel and a watering hole in Thunder Bay called the Apollo.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   \includegraphics[scale=0.5]{"PNS 003-im3".png}
-   \caption*{\small \em Volunteers Robert Desjardins, Bruce Gemmill and Michel Cote hold the refinished coolant tank from engine \#4.
+Driving down the 599 in that snow storm with a truck I had bought for
+\$100 was a white knuckle ride. Knowing this would probably be the last
+time I drove the 2000km distance between Northern Ontario and home in
+Gatineau. The stakes were high: A flight to catch in Ottawa (T-minus
+48hrs), New Year's eve in New York city with my family waiting there
+for me and a new job starting at CASM on the day of my wedding
+anniversary. Could 2017 start with anything less? Arriving to work that
+January 3rd morning was a good feeling, until I got startled by "Manny"
+the mannequin sitting in my predecessor's office chair. If a sense of
+humour was a job requirement, I hoped to get the joke. I would later
+find Manny in other compromising positions, some of which I refuse to
+admit my participation in writing.
+
+The momentum I have gained in the twelve months since my escape from
+Pickle Lake is only part of what keeps me moving forward. I also count
+on the support of others, namely co-workers and volunteers. My position
+as Project Manager of the North Star restoration relies on the good
+work and cooperation of all hands involved. As the work itself proves
+to be technical and highly detailed, with wide ranges of abilities
+involved, people are always the driving force.
+
+It was suggested my article for the North Star Chronicle be focused on
+the restoration of a certain four engined artifact \#1967.0645. For
+details on the work that has been carried out by our dedicated group of
+volunteers, I recommend you read my conservation report on the subject.
+It is available online: \textcolor{blue}{\texttt{\url{http://www.projectnorthstar.ca}}}
+or by email: \textcolor{blue}{\email{rdemers@techno-science.ca}}
+
+%\section{Nr 4 Engine}
+%\label{sec:engine_4}
+%
+%Work has continued on much of the ancillary equipment for the engine. The
+%supercharger and intercooler have been completed and installed, as well as the
+%reduction gearbox. The radiators and header tank have been restored and painted, and
+%two of the three radiator sections have been installed. Restoration work has
+%begun on the air intake system, and many of the fuel and oil lines. 
+%
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics[scale=0.5]{"PNS 007-im2".png}
+%   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
+%   \label{fig:engine_no_4_2}
+%\end{figure}
+%
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics[scale=0.5]{"PNS 003-im3".png}
+%   \caption*{\small \em Volunteers Robert Desjardins, Bruce Gemmill and Michel Cote hold the refinished coolant tank from engine \#4.
 Hours of polishing were required to get the tank to look like new.}
-   \label{fig:coolant_tank}
-\end{figure}
-
-The starter motor was found to be very corroded, and disassembly and repair not viable. A
-new unit has been ordered and will be installed once delivered. The fire
-suppression lines are complete and will be installed along with the cowls when
-ready. Some of the cowl panels have been restored, although it has been found
-that this engine has more corrosion than the other three engines, possibly it
-had been in service longer? This has meant that more work is required to
-refurbish each of the many panels. We still hope to have this engine ready for
-installation by next spring.
-
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   \includegraphics[scale=0.5]{"PNS 008-im1".png}
-   \caption*{\small \em Volunteer Garry Dupont working on the re-assembly of engine \#4.}
-   \label{fig:engine_no_4}
-\end{figure}
-
-\section{Fuselage and Main Cabin}
-\label{sec:main_cabin}
-
-With the aircraft moved outside in March, little work was done in the aircraft
-over the winter. Once the temperatures warmed up, we continued work in the main
-cabin to prepare for completion of the painting started last year. Repairs to
-the rear bulkhead are now complete with the installation of several new pieces
-of aluminum to replace corroded ones. A new section of flooring was installed
-where the rear washroom was situated. This required replacing about 1000 rivets!
-
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   \includegraphics[scale=0.5]{"pns 043-im4".png}
-   \caption*{\small \em A single new floor panel at the rear of the main cabin required over 1000 rivets.  
-   Cleco clips are used to accurately align the panel during riveting.}
-   \label{fig:panel_revets}
-\end{figure}
-
-Cleaning and corrosion removal continued throughout the cabin. It seems whenever
-we feel we are near completion, we find more corrosion. We had also planned to
-finish priming the floor and painting the rear bulkhead before the aircraft
-returned to the hanger later in the year, but this was delayed by the shutdown
-of the project while a new Project Manager was hired by the museum. The wood
-floors and walls still need to be repaired, or new panels made, and all will
-require painting.
-
-\section{Cabin Liners}
-\label{sec:liners}
-
-All the large pieces of liner have either been repaired or replaced, but with
-limitations on space, most remaining work on the cabin liners has been
-postponed. New window surrounds have also been made, but then all liners will
-require painting and stenciling, and this must wait until there is more room in
-the shop and the paint booth. Liners were also made for the top of the heater
-ducts.
-
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   \includegraphics[scale=0.5]{P1020369-im5.png}
-   \caption*{\small \em One of the refurbished heater duct sections with a new fabric headliner secured to the top.
+%   \label{fig:coolant_tank}
+%\end{figure}
+%
+%The starter motor was found to be very corroded, and disassembly and repair not viable. A
+%new unit has been ordered and will be installed once delivered. The fire
+%suppression lines are complete and will be installed along with the cowls when
+%ready. Some of the cowl panels have been restored, although it has been found
+%that this engine has more corrosion than the other three engines, possibly it
+%had been in service longer? This has meant that more work is required to
+%refurbish each of the many panels. We still hope to have this engine ready for
+%installation by next spring. 
+%
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics[scale=0.5]{"PNS 008-im1".png}
+%   \caption*{\small \em Volunteer Garry Dupont working on the re-assembly of engine \#4.}
+%   \label{fig:engine_no_4}
+%\end{figure}
+%
+%\section{Fuselage and Main Cabin}
+%\label{sec:main_cabin}
+%
+%With the aircraft moved outside in March, little work was done in the aircraft
+%over the winter. Once the temperatures warmed up, we continued work in the main
+%cabin to prepare for completion of the painting started last year. Repairs to
+%the rear bulkhead are now complete with the installation of several new pieces
+%of aluminum to replace corroded ones. A new section of flooring was installed
+%where the rear washroom was situated. This required replacing about 1000 rivets!
+%
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics[scale=0.5]{"pns 043-im4".png}
+%   \caption*{\small \em A single new floor panel at the rear of the main cabin required over 1000 rivets.  
+%   Cleco clips are used to accurately align the panel during riveting.}
+%   \label{fig:panel_revets}
+%\end{figure}
+%
+%Cleaning and corrosion removal continued throughout the cabin. It seems whenever
+%we feel we are near completion, we find more corrosion. We had also planned to
+%finish priming the floor and painting the rear bulkhead before the aircraft
+%returned to the hanger later in the year, but this was delayed by the shutdown
+%of the project while a new Project Manager was hired by the museum. The wood
+%floors and walls still need to be repaired, or new panels made, and all will
+%require painting. 
+%
+%\section{Cabin Liners}
+%\label{sec:liners}
+%
+%All the large pieces of liner have either been repaired or replaced, but with
+%limitations on space, most remaining work on the cabin liners has been
+%postponed. New window surrounds have also been made, but then all liners will
+%require painting and stenciling, and this must wait until there is more room in
+%the shop and the paint booth. Liners were also made for the top of the heater
+%ducts.
+%
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics[scale=0.5]{P1020369-im5.png}
+%   \caption*{\small \em One of the refurbished heater duct sections with a new fabric headliner secured to the top.
 The sides are covered with mica insulation that was removed, cleaned and reattached after repairs to the metal duct.
 These will be installed in the main cabin.}
-   \label{fig:heaters_liners}
-\end{figure}
-
-\section{PNS 2017 Update}
-\label{sec:pnsupdate}
-
-We were able to restart work on the aircraft in late January, following the
-arrival of our new Project manager, Rej Demers. So far, we have completed the
-top cowl panels and one of the radiator flaps, which required a repair patch to
-the outer skin, and the electrical harnesses, which are now waiting for
-painting. Work continues on the air box and the remaining cowl panels.
-
-We have decided not to move the aircraft outside this year. With our efforts
-concentrated on completing engine \#4, it is unlikely we would have the resources
-to also complete the repair work in the cabin needed to allow for priming and
-painting, which is the main reason for moving the aircraft outside in the summer.
+%   \label{fig:heaters_liners}
+%\end{figure}
+%
+%\section{PNS 2017 Update}
+%\label{sec:pnsupdate}
+%
+%We were able to restart work on the aircraft in late January, following the
+%arrival of our new Project manager, Rej Demers. So far, we have completed the
+%top cowl panels and one of the radiator flaps, which required a repair patch to
+%the outer skin, and the electrical harnesses, which are now waiting for
+%painting. Work continues on the air box and the remaining cowl panels.
+%
+%We have decided not to move the aircraft outside this year. With our efforts
+%concentrated on completing engine \#4, it is unlikely we would have the resources
+%to also complete the repair work in the cabin needed to allow for priming and
+%painting, which is the main reason for moving the aircraft outside in the summer.
 

### 2018-01-25 pnsac/newsletter/article_MemberInterview.tex
Incorporated changes from Roger and link to Rej's article.

@@
 navigation and air gunner. I later qualified as an airborne
-interception navigator . My operational experience includes air
+interception navigator. My operational experience includes air
 transport, search and rescue and training. My assignments included
@@
 Navigator and Deputy Commander (COO) of the Air Transport Group. I
-logged 9500 hours , mostly on the following aircraft: Beechcraft,
-Dakota, North Star , Mitchell, Yukon, Hercules, and Boeing 707. My
+logged 9500 hours, mostly on the following aircraft: Beechcraft,
+Dakota, North Star, Mitchell, Yukon, Hercules, and Boeing 707. My
 staff assignments in Europe included movement control (traffic
@@
 proposal for volunteers to restore the North Star.
+
 %\begin{figure}[htbp]

### 2018-01-25 pnsac/newsletter/article_PresReport.tex
Incorporated changes from Roger and link to Rej's article.

@@
 have indicated a willingness to do the necessary work and we are hoping
-to have the site working and up to date by the time this issue of the
-Nstar Chronicle is published.
+to have the site working and up to date in the near future.
 

### 2018-01-25 pnsac/newsletter/article_ProgReport.tex
Incorporated changes from Roger and link to Rej's article.

@@
 volunteers, I recommend you read my conservation report on the subject.
-It is available online: \textcolor{blue}{\texttt{\url{http://www.projectnorthstar.ca}}}
+It is available online: \textcolor{blue}{\texttt{\url{https://documents.techno-science.ca/documents/NorthStarConservationReportOctober2017.pdf}}}
 or by email: \textcolor{blue}{\email{rdemers@techno-science.ca}}

### 2018-07-19 pnsac/newsletter/article_EditorsNotes.tex
Newsletter source file changes for July 2018.

--- 2017-03-13 pnsac/newsletter/article_EditorsNotes.tex
@@
 
-As the President has noted the Association has, through its volunteers,
-expended thousands of hours on the North Star's restoration since 2004. In the
-last couple of years, however, various developments at CASM have resulted in a
-diminution in the amount of restoration work we have been able to do. The result
-has been a decline in the number of volunteers working on the aircraft, and a
-corresponding decline in our membership. This in turn has led to an overall
-decline in the Association's activities, including the publication of this
-newsletter. CASM has made some major changes to its personnel, including those
-directly responsible for the restoration of the North Star. The Association is
-already experiencing a positive change in its relationship with the Museum which
-can only be to the good. We look forward to a renewed partnership with CASM to
-our mutual benefit.
+Our readers will appreciate from the articles and photographs in this edition that real progress
+being made on the restoration of the aircraft.
+A number of new restoration volunteers have signed on this year bringing with them diverse skills and experience.
+Whoever thought that someone with a mountain climbing background would be able to put his special skills to work on an aircraft.
+Believe it or not but see the "Conservator's Corner."  As I have previously commented: When work on the aircraft is moving ahead it encourages enthusiasm at all levels of the organization.
+One of the biggest cheerleaders of our Association is its president, Richard Lodge.
+Richard wears many hats including that of a restoration volunteer so he knows of what he speaks.
+In this edition Richard discusses how he came to be involved in Project North Star and joins the illustrious members of the Association who have contributed to "Our Members" column.
+But I would be remiss if I didn't thank all of our contributors to this edition including our publisher Drew Hodge.
 
-%\vspace{12pt}
+\vspace{12pt}
 

### 2018-07-19 pnsac/newsletter/article_MemberInterview.tex
Newsletter source file changes for July 2018.

@@
 \title{Our Members}
-\author{Interview with Austin J Timmins}
+\author{Interview with Richard Lodge}
 
@@
 
-1.\textit{What is your background in aviation?}
+1.\textit{What is your background in aviation?}\\
 
-My aviation experience includes 36 years service in the RCAF and nine
-years with the Canadian Aeronautics and Space Institute. In the RCAF I
-completed radio navigator training, which included communications,
-navigation and air gunner. I later qualified as an airborne
-interception navigator. My operational experience includes air
-transport, search and rescue and training. My assignments included
-Navigation Leader, Aircrew Leader, Squadron Commander, Air Command
-Navigator and Deputy Commander (COO) of the Air Transport Group. I
-logged 9500 hours, mostly on the following aircraft: Beechcraft,
-Dakota, North Star, Mitchell, Yukon, Hercules, and Boeing 707. My
-staff assignments in Europe included movement control (traffic
-management) for 1 Air Division and a tour at The NATO Defense College
-in Rome.
+I trained in England as a Chartered Accountant, now often known as a bean
+counter. After I obtained my designation, I wanted to work in industry rather
+than in the accountancy profession. Following several interviews, I was offered
+a job with the Aero Engine Division of Rolls-Royce Ltd. in Derby, England.
 
-On retirement from the RCAF I participated in a Transport
-Canada A Base review of several airports. I then secured full
-employment with Canadian Aeronautics and Space Institute as Executive
-Director, retiring in 1995.
+Until I joined Rolls-Royce I had very little interest aviation. I had always
+been interested in transportation and had a major passion for railways which was
+the mode of transport I normally used in Britain in 1950s and 1960s.
 
-2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}
+I became interested in aviation at Rolls-Royce and was fascinated by the
+complexity of designing and building aero engines.
 
-My involvement with efforts to conserve North Star 17515 predates
-Project North Star. Ottawa area members of the 426 Thunderbird Squadron
-Association, mostly retired North Star aircrew, met with the National
-Aviation Museum staff to express their concern about the state of the
-North Star and proposed volunteer involvement in a preservation effort.
-These expressions of concern and interest led to a meeting between the
-Association representatives, where General Adamson presented a proposal
-for the restoration of the aircraft. It was rejected but assurance was
-given that the North Star would be first priority when the new hangar
-was completed Also a North Star Trust Fund would be established. Robert
-Holmgren, an observer at the meeting, then proceeded to develop a
-proposal for volunteers to restore the North Star.
+After leaving Rolls-Royce in 1966, I was not involved with aviation until I came
+to Canada in 1978 and returned to the accountancy profession and started doing
+tax returns for pilots and other aviation personnel located around Dorval,
+Qu\`{e}bec.\\
+
+2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}\\
+
+One of the driving forces behind the foundation of Project North Star was
+Robert Holmgren who was among a group who persuaded the National Aviation and
+Museum, as it was known then, that a group of a volunteers should be allowed to
+work on the North Star and bring it back to Museum display standards. The
+Project North Star Association of Canada (PNSAC) was incorporated in 2003 with
+Robert as its first president. Robert had worked for Air Canada on the
+engineering side of the airline in Dorval and I got to know him when doing the
+accounting for his wife's antiques business. At the time of the introduction of
+the Québec language laws (Bills 22 and 101), many Anglophones left the Montr\'{e}al
+area and moved to Ottawa. Both Robert and I moved in the early 1980's. After
+transferring my business to Ottawa, I remained in contact with Robert
+professionally and when PNSAC was formed he asked me to become the Association's
+first Treasurer.\\
 
@@
 
-3. \textit{What has been the history of your involvement to date?}
+3. \textit{What has been the history of your involvement to date?}\\
 
-I have been involved from the beginning of efforts to establish PNS,
-first as a member of a Steering Committee, led by Robert Holmgren,
-later, Vice President and then President of PNSAC. My contacts in
-aviation provided much needed advice and support, including access to
-initial funding for PNSAC.
+I remained Treasurer until 2010 when I was elected President after our second
+President, Tim Timmins, stood down. My election as President coincided with my
+partial retirement from the accountancy profession. Apart from my interest in
+railways, I had also always been interested in engineering and retirement gave
+me the opportunity to become a Restoration Volunteer on the North Star. Working
+on a Merlin engine and becoming part of the Engine Shop crew was an obvious
+place for me to work. I have enjoyed developing some of the limited skills I
+learned when working on pre-WW2 cars in England, which mostly used similar
+technology to the Merlin engines.\\
 
@@
 
-4. \textit{What has been the highlight of your involvement?}
+4. \textit{What has been the highlight of your involvement?}\\
 
-The highlight of my involvement has to be the establishment of PNS and
-PNSAC.
+In the 1950s I was one of the early volunteers working on the first preserved
+railway in the world, the Talyllyn Railway in Wales. During this I learned much
+about volunteering and how such an organization could work. Helping to organize
+and build PNSAC over the last 14 years has both been challenging and very
+interesting. We are now seeing the fruits of our work both in the actual
+achievements in the restoration of the North Star and the enthusiasm of our
+Association members.\\
 
@@
 
-5. \textit{What has been the most challenging part of your involvement?}
+5. \textit{What has been the most challenging part of your involvement?}\\
 
-PNS is now in its 16th year. It has been a work in progress with many
-bumps along the way. I spent a lot of time and effort, along with other
-members of the Executive, evening out these bumps. Establishing and
-maintaining effective channels of communication with CASM and CSTM
-required constant attention. Much effort was directed to fund raising,
-publicity, and, finding volunteers. In the beginning, every date
-brought a new challenge.
+Volunteers working on an aircraft owned by the Government of Canada in a
+national Federal museum was in the early 2000s an entirely new concept. Until
+that time all work on the aircraft collection had been carried out by Museum
+permanent staff. It was not easy to persuade the Museum that volunteers could be
+trusted to carry out restoration work to the high standards of the Museum and
+that as a group they would have the tenacity to continue working for many years.
+Now 14 years later and after over 80,000 hours of volunteer restoration work,
+PNSAC is going strong and will definitely see the end of the North Star
+restoration in the years to come.
+
+Starting a new project is not easy. There is always skepticism as to whether the
+new organization will succeed and in some cases resistance to the concept of the
+new organization. Over the years we have had to overcome all these problems and
+avoid being discouraged when we were unable to do things the way we initially
+wanted to do them. Our acceptance by the Museum as a trusted partner is a
+testament to the patience and hard work of all our volunteers and the support of
+key members of the Museum's permanent staff.
 

### 2018-07-19 pnsac/newsletter/article_PresReport.tex
Newsletter source file changes for July 2018.

@@
 
-Since the last issue of the NStar Chronicle in March of this year we
-have continued to reorganize our volunteer operations under the
-leadership of the Museum's Special Projects Manager, R\'{e}jean (Rej)
-Demers. We have started a recruitment process for new restoration
-volunteers and to date two new people have started work. During the
-summer two more permanent Museum staff joined Rej to replace people who
-have retired. It is obviously taking time for the new Museum staff to
-get to know the complex operations of the Aviation Museum, to get to
-know the Project North Star volunteers and assess their various talents
-and areas of expertise. Consequently, our summer has been quiet but
-steadily productive.
+These are exciting times to be a member of the Project North Star Association.
+We  have recently taken part in the 2018 Doors Open Ottawa. This is a major
+event for the Museum and PNSAC volunteers were able to make a significant
+contribution to the success of the two open days. Doors Open has become second
+only to Canada Day in its importance to the Museum. (more on Open Doors 2018
+later in this issue).
 
-For several reasons it became necessary to reorganize the location of
-several major Museum artifacts. This affected the North Star aircraft
-which spent the summer outside the Reserve Hanger resulting in some
-problems with water ingress and further bird activity. I am pleased to
-say that the aircraft is now back inside the hangar.
+A few days before Canada Day the aircraft was brought outside. Unfortunately,
+the day itself was extremely hot and the number of visitors was well down from
+previous years. Nonetheless our volunteers did a yeoman service in talking to
+the public about our restoration work.
 
-The Museum is actively supporting Project North Star and trying to
-facilitate our restoration work wherever possible and is promoting our
-work and future activities. Our Treasurer, Paul Labranche, and I
-recently met with the Museum Foundation. The Foundation is the umbrella
-organization with responsibility for fundraising and business
-development at the three museums, comprising the Museum Corporation,
-(Aviation, Science \& Tech and Agriculture). We had a very productive
-meeting, which among other things has resulted in the Foundation
-deciding to highlight Project North Star in its Fall fundraising mail
-out. This type of collaboration is very good for our Association and we
-hope as good for the Foundation.
+Although showing the aircraft to the public is very important to us, our main
+purpose is to restore the North Star to museum condition. Much of our work is
+detailed, painstaking and unspectacular such as restoring a small part of the
+aircraft which can take many days or weeks. At other times making progress is
+very visible. We are in such a situation now.
 
-Our biggest challenge at present is to get our website properly
-operational again. A couple of our volunteers with website knowledge
-have indicated a willingness to do the necessary work and we are hoping
-to have the site working and up to date in the near future.
+At the end of May we had one of our regular meetings with Chris Kitzan (Director
+General of CASM) together with Neil Raynor (VP of PNSAC), Cedric St-Amour (CASM
+Volunteer Coordinator) and Reg Demers (Our CASM supervisor There was a full
+exchange of ideas, suggestions for bringing  the public into the Museum and ways
+in which PNSAC could be involved in these initiatives and events. The meeting
+ended with me having a full sheet of notes covering future planning, major
+restoration work and how we can make sure we recruit younger people both as
+members of PNSAC and where possible active Restoration Volunteers.
 
-On December 2nd we held a Members' meeting at the Museum when we were
-fortunate to have Jim Mason join us to talk about the restoration of an
-early Lockheed aircraft flown by TCA. The meeting was well attended in
-a new venue of the Avro Arrow room instead of the Bush Theatre.
-Following Jim's interesting presentation, the group moved to the
-Museum's Restoration Hangar where Rej talked about the progress of the
-North Star restoration and the plans for work in 2018.
+A meeting like the one above only happens when an organization is thriving, and
+everybody is excited about the future. During the last 15 years PNSAC has
+demonstrated that it is capable delivering very high-quality restoration work.
+We have already accumulated over 80,000 hours of work and during this time we
+have become a trusted partner of the Museum. Our website is now operational
+again and in the next few weeks we expect to be able to update it and make it
+more relevant to where we are with the project now. One of the major initiatives
+taking place is to develop a diagrammatic history of the work done on the North
+Star restoration and a timeline for future work on the aircraft.
 
-One of the initiatives currently underway is to create a visual
-timeline of the work already completed on the North Star restoration
-and the planned future work to bring the project to completion. It is
-anticipated that this visual timeline will be mounted on a display
-board for our members and the public to view.
+In closing, I once again urge Association members to become active volunteers.
+There are many volunteering opportunities both for those with major aviation
+experience to those who can only volunteer with us on occasions like Canada Day
+or can provide help for our many behind-the-scenes activities.
 
-In closing, I am optimistic that 2018 will see increased activity both
-in the aircraft restoration work and also within our Association.
+Our next Members' Meeting is scheduled for
+Saturday, September 29, 2018. Do reserve this date in your calendar if you can
+be in the Ottawa area on that day. The meeting will be held at the Canada
+Aviation and Space Museum. More details will follow later.
 

### 2018-07-19 pnsac/newsletter/article_ProgReport.tex
Newsletter source file changes for July 2018.

--- 2018-01-14 pnsac/newsletter/article_ProgReport.tex
@@
 %\subtitle{PNS 2016 Status Update}
-\author{R\'{e}jean Demers}
+\author{R\'{e}jean Demers\\ Conservator\\ Canadian Aviation and Space Museum}
 
@@
 
-\textit{R\'{e}jean Demers (Rej, to those who work with him) is a Transport Canada
-licensed Aircraft Maintenance Engineer (AME). R\'{e}jean was a small
-business owner in Gatineau QC for over 10 years, working for a wide
-variety of private and commercial aircraft owners and operators.
-Following a stint in Northern Ontario working on turpoprop freighters
-and bulk fuel haulers, he took up his present position as a
-Conservator, Special Projects Manager with Ingenium, Canada's Museums
-of Science and Innovation in January 2017. His main project is the
-restoration of the Canadair North Star DC-4GM at the Canadian Aviation
-and Space Museum. He recounts below his introduction to his new job.}
+\textbf{Stabs and Tail Feathers}
 
-Leaving Pickle Lake was like getting shot out of a cannon in a snow
-storm. At that time, Christmas of 2016, I was the last person on my
-crew to leave the YPL base, having seen my colleagues off for the end
-of our tour and the holidays. After a full day of flying, the fleet had
-to be tucked away in the barn ready for service. On Boxing Day, a fresh
-crew would fly in and start another two week tour of line maintenance
-with the company we worked for. However, on that Christmas, I was
-quitting my job.
+Some tasks are easier than others, this is not a simple one. North Star S/N:
+122, registration 17515, was delivered to the National Aeronautical collection
+in 1965, intact and airworthy. After spending 39 years parked on the tarmac,
+awaiting restoration, the condition of its flight control surfaces had become
+less a statement of endurance and more a sad result of the elements.
 
-Nobody knew I was leaving until I got back to the bunkhouse and shared
-Christmas dinner with the remaining pilots and ops personnel. I refer
-to them in these terms now, but then and there, I could say they were
-my friends. In the weeks leading to my departure, no one knew I was
-leaving. No one speaks of breaking out of prison when they are behind
-bars. The difference being the only bars I saw were the Pickle Lake
-Hotel and a watering hole in Thunder Bay called the Apollo.
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{IMG_0705-scaled.png}
+   \caption*{\small \em Scaffold erected next to North Star to support removal of the rudder and vertical stablizer.}
+   \label{fig:stab-one}
+\end{figure}
 
-Driving down the 599 in that snow storm with a truck I had bought for
-\$100 was a white knuckle ride. Knowing this would probably be the last
-time I drove the 2000km distance between Northern Ontario and home in
-Gatineau. The stakes were high: A flight to catch in Ottawa (T-minus
-48hrs), New Year's eve in New York city with my family waiting there
-for me and a new job starting at CASM on the day of my wedding
-anniversary. Could 2017 start with anything less? Arriving to work that
-January 3rd morning was a good feeling, until I got startled by "Manny"
-the mannequin sitting in my predecessor's office chair. If a sense of
-humour was a job requirement, I hoped to get the joke. I would later
-find Manny in other compromising positions, some of which I refuse to
-admit my participation in writing.
+By the time restoration had begun in 2004, the flight control surfaces had
+sustained the full range of what mother nature had to offer over the years. In
+the fourteen years that followed, work has been carried out removing what
+assemblies could be reached from the ground. Ailerons and flaps, elevators and
+horizontal stabilizers had been removed and now, the vertical stabilizer and
+rudder in 2018.
 
-The momentum I have gained in the twelve months since my escape from
-Pickle Lake is only part of what keeps me moving forward. I also count
-on the support of others, namely co-workers and volunteers. My position
-as Project Manager of the North Star restoration relies on the good
-work and cooperation of all hands involved. As the work itself proves
-to be technical and highly detailed, with wide ranges of abilities
-involved, people are always the driving force.
+This article will focus on the year long process that occupied roughly 2-4
+volunteers, twice a week, working at heights on scaffolding around the empennage
+("tail assembly") of 515. Work commenced following the museum's last big
+shuffle, in the fall of 2017. One side of the storage hangar was emptied to
+accommodate the acquisition of the C130E Hercules and Convair 580. The removal
+of CP-107 Canadair Argus from storage was a necessary trade-off for the new
+stacking, that left Project North Star with just enough space to work on.
 
-It was suggested my article for the North Star Chronicle be focused on
-the restoration of a certain four engined artifact \#1967.0645. For
-details on the work that has been carried out by our dedicated group of
-volunteers, I recommend you read my conservation report on the subject.
-It is available online: \textcolor{blue}{\texttt{\url{http://www.projectnorthstar.ca}}}
-or by email: \textcolor{blue}{\email{rdemers@techno-science.ca}}
+%\begin{figure}[httb]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics[scale=1.25]{DSC_1156-scaled.png}
+%   \caption*{\small \em \color{red} CAPTION NEEDED}
+%   \label{fig:stab-one}
+%\end{figure}
+
+The work involved in removing flight control surfaces from the ship starts by
+disconnecting cables, push rods and torque tubes. These relay pilot input
+directly to the airfoil. There are secondary components such as panels,
+fairings, plumbing for de-icing boots, grounding straps and electrical conduits.
+Finally, when the control surface or stabiliser is ready to be pulled, hinge
+bearings and spar bolts must be freed. Support must be provided to the component
+once the weight of the assembly is relieved from the airframe. Once
+disconnected, the "tail feather" or "stab" is lowered into a cradle on the
+ground for cleaning, condition assessment and preparation for restoration. The
+approach being that a component is more accessible at ground level than 10-20
+feet overhead. Eventually, the outboard wing sections will follow the same
+treatment: Not a light undertaking.
+
+Volunteers on task are comprised of a mix of experienced members and new
+additions to the crew. Setting up the scaffold was the first requirement in
+gaining access. We had an ex roughneck, Will Assad, building the first tower.
+Chris McGuffin, an experienced mountaineer, built the second stage to a full
+height of 30ft. Bruce Gemmill donated blood removing the fin retaining hardware,
+a modest 64 bolts through 4in. access panels. Peter Trobridge spent some hours
+in the hell hole, fitting a tail skid extension jack and running leaders for
+control cables. The heat of lights and summer humidity had a therapeutic effect,
+much like a sauna.
+
+Years of battering had taken its toll on the rudder, shearing the torque tube
+and stripping the rudder of it's fabric. During this time, generations of
+starlings had been introduced to multi-level condo dwelling. Three garbage bags
+of nesting material were removed just to gain access to the pulleys in the
+vertical stabiliser alone. Starlings are neither good nest makers nor parents to
+their young. I will spare you the graphic descriptions of what was removed from
+those access panels.
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.55]{IMG_20180529_132544212-scaled.png}
+   \caption*{\small \em Volunteer Bruce Gemmill removing bird nests from the vertical stabilizer.}
+   \label{fig:stab-one}
+\end{figure}
+
+The horizontal stabilisers were removed earlier, yet there still remained some
+nesting material to clear. These assemblies were brought to the restoration
+shops for a thorough clean-out and condition assessment. Phil Chrysler, Bill
+Cole,  Richard Houle and John Makadi lent their hands at pulling more stacks of
+hay out of the two structures. Access proved difficult necessitating  the use of
+a hinge rods, in the form of hooked tools for retrieving deep tufts of clumped
+nesting. The individual stabilisers were flipped over, blown out, pressure
+washed, and oiled, eventually rendering an accessible, stable assembly ready for
+restoration. We discovered a large patch on the leading edge of the port
+stabiliser. It covered a blunt force impact form which was probably caused by
+the boom of a forklift in a former life.
+
+At the time of going to press, the rudder has been removed from the aircraft and
+the vertical stabiliser is slung and ready to pull. The North Star is prepped to
+roll out of storage for Canada day and the vertical stab removal is planned for
+July 5th. A 30T crane has been rented for the job, with all hands on deck
+confirmed for that day's shift schedule. Following this important milestone, the
+maple leaf will be lowered from North Star 515, until such time as the
+restoration nears its completion. I hope to see the day, dear reader, when the
+maple leaf is flown from those heights again.
 

### 2018-07-19 pnsac/newsletter/article_SpecialEvents.tex
Source files for Dec13 newsletter.
Newsletter source file changes for July 2018.

@@
 
-\title{Quarterly Meeting Report}
-\author{Drew Hodge}
+\title{Doors Open Ottawa 2018}
+\author{Roger Button}
 
@@
 
-On Saturday 7th December 2013, Project North Star Association of
-Canada (PNSAC) held a quarterly meeting in the auditorium of the
-Canadian Aviation and Space Museum (CASM) in Ottawa. It was a
-memorable meeting with over 200 people in attendance. Those in the
-audience that day were able to hear the story of the Gimli Glider told
-in person by Captain (retd.) Robert (Bob) Pearson, the pilot of Air
-Canada Boeing 767-200, fin number 604, when it made an unscheduled
-stop at Gimli Manitoba, thirty years ago.
+Doors Open Ottawa gives the general public an opportunity to view facilities for free which they would
+normally not have access. Evidently, it is the second largest event of its kind in North America and one of
+the largest in the world with over one million visitors since its inception in 2002. In the case of the
+Canadian Aviation and Space Museum its participation meant opening the doors of the reserve hangar.
+On June 2nd and 3rd the Museum welcomed some 3000 visitors. The visits of the reserve hangar took
+place by way of a self-guided tour supported by many of the Museum's volunteers coordinated by Cedric
+St. Amour the volunteer coordinator for the Museums Corporation. Project North Star was ably
+represented by Charles Baril, Claire Cameron, Garry Dupont, Ted Devey, Chris McGuffin, Neil Raynor,
+Jacques Roy, Nelson Smith, Tim Timmins, and Gary Whitten who ably answered questions about the
+aircraft. Rejean Demers, the Conservator Special Projects Manager was also present.
 
-The meeting started with an introduction by Bill Tate, Project North
-Star's Vice President.  Bill began by welcoming the audience, which,
-as well as PNSAC volunteers, included members of Vintage Wings of
-Canada, Ottawa Airport Watch, the Canadian Aviation Historical
-Society, the College of Professional Pilots of Canada, CASM volunteers
-and staff, Air Canada and WestJet pilots, and visitors from the
-general public, one of whom had travelled to Ottawa from Vancouver for
-the meeting. After describing the goals of Project North Star and
-asking PNSAC volunteers to stand and be recognized, Bill invited the
-audience to visit the world class exhibits in the Museum after the
-meeting.
-
-Bill's acquaintance with Bob Pearson goes back thirty-five years, when
-Bill was Captain Pearson's Second Officer on Boeing 727s based in
-Montreal. His duties in the cockpit back then gave Bill ample
-opportunities to watch Captain Pearson at work. Bill described Captain
-Pearson as a "highly professional pilot" who, "wearing his command
-lightly", demonstrated "highly polished interpersonal skills and a
-delightful sense of humour".  He then invited Captain Pearson to tell
-the story of flying unscheduled into Gimli.
-
-Captain Pearson spoke for a little over an hour with no notes and with
-reference to a few appropriate slides.  He described eloquently how,
-because of errors in the conversion of metric measurements, his Boeing
-767-200 ran out of fuel and lost both engines over Manitoba on the way
-from Ottawa to Edmonton. The full story is well documented in the book
-"Freefall: From 41,000 feet to zero - a true story, William and
-Marilyn Hoffer, Simon \& Schuster, 1989", but to hear it told in
-person by the pilot in command that Saturday, 23rd July, 1983, was a
-rare privilege.
-
-\begin{figure}[htbp]
+\begin{figure}[httb]
    \vspace{2em}
    \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{Pearson004.eps}
-   %caption of the figure 
-   \caption*{\small \em Captain (retd.) Robert Pearson.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:pearson004}
+   \includegraphics[scale=0.275]{"PNSAC DOO 2018-scaled".png}
+  \caption*{\small \em Volunteers Roger Button, Richard Lodge, Chris McGuffin, Charles Baril and Garry Dupont on hand to explain our project at Doors Open Ottawa June 3rd.}
+   \label{fig:stab-one}
 \end{figure}
 
-The port engine stopped first at 41,000 feet prompting the decision to
-divert to Winnipeg.  At 28,000 feet the starboard engine ran down too,
-and Maurice Quintal, Captain Pearson's First Officer, made
-calculations that showed they couldn't reach Winnipeg.  Quintal
-advised Captain Pearson to try for the nearest runway on the former
-RCAF aerodrome at Gimili, twelve miles away.  Some eight or ten
-minutes after losing the first engine, Captain Pearson side-slipped
-the big Boeing down to an old runway that was being used as a racing
-track by members of the Winnipeg Sports Car Club. The nose wheel
-collapsed on landing, but the aircraft came to a stop and no one was
-seriously injured. The racers helped to put out a minor fire in the
-nose.
+This event is also an opportunity to do a little fundraising for the Association. The merchandise desk was
+managed by Richard Lodge, Drew Hodge, and Roger Button. Sales of memorabilia and clothing were
+steady throughout the two days. A few memberships were also sold and donations received. The
+merchandise table was well positioned near the exit to catch visitors as they left. Unfortunately, the
+aircraft was not next to the merchandise table and it was clear that some visitors did not make the
+connection and others left without seeing the aircraft. It is hoped that these issues will be addressed for
+future viewings of the aircraft. Thanks to all the volunteers who devoted their time to this successful
+event.
 
-Captain Pearson talked about the inquiry that followed the incident
-and its eventual exoneration of both pilots.  The aircraft was flown
-to Winnipeg just two days after the Gimli landing and saw service with
-Air Canada for another twenty-five years.  When the Boeing was retired
-to the Mohave desert in 2008, Captain Pearson was at the controls to
-make that last landing in Arizona.
-
-Following his talk, Captain Pearson answered questions from the
-audience.  Richard Lodge, PNSAC President, thanked Captain Pearson and
-presented him with a PNSAC golf shirt, a baseball cap, a coffee mug,
-and finally a certificate conferring upon him an Honorary Life
-Membership of the Project North Star Association.  Captain Pearson
-then graciously agreed to make presentations of volunteer-hours
-certificates and 10th Anniversary photos to PNSAC volunteers.  The
-following volunteers received certificates for the hours shown:
-
-\begin{itemize}
-  \item Bruce Gemmill (7000 hrs)
-  \item Ted Devey (4000 hrs)
-  \item Bill Tate (3000 hrs)
-  \item Richard Lodge (1000 hrs)
-  \item Bruce Grant (1000 hrs)
-  \item Robert D\'{e}sjardins (1000 hrs)
-  \item Peter Trowbridge (1000 hrs)
-  \item Charles Baril (1000 hrs)
-\end{itemize}
-
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{Pearson007.eps}
-   %caption of the figure 
-   \caption*{\small \em Richard Lodge presenting Captain Pearon with a
-     certificate for Honorary Life Membership in PNSAC.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:engine3propcowl}
-\end{figure}
-
-To conclude the meeting, Captain Pearson presented signed copies of
-the book "Freefall: From 41,000 feet to zero - a true story" to three
-audience members whose ticket numbers he had drawn at random.
-
-The Association's sincere thanks go to Bill Tate for organizing this
-quarterly meeting and the talk by Captain Pearson -- it was largely
-Bill's efforts that made the meeting the great success that it was.
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.9]{Pearson004.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Captain (retd.) Robert Pearson.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:pearson004}
+%\end{figure}
+%
+%
+%
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.9]{Pearson007.eps}
+%   %caption of the figure 
+%   \caption*{\small \em Richard Lodge presenting Captain Pearson with a
+%     certificate for Honorary Life Membership in PNSAC.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:engine3propcowl}
+%\end{figure}
 

### 2019-01-13 pnsac/newsletter/article_EditorsNotes.tex
January 2019 changes and additions.

--- 2017-03-13 pnsac/newsletter/article_EditorsNotes.tex
@@
 
-As the President has noted the Association has, through its volunteers,
-expended thousands of hours on the North Star's restoration since 2004. In the
-last couple of years, however, various developments at CASM have resulted in a
-diminution in the amount of restoration work we have been able to do. The result
-has been a decline in the number of volunteers working on the aircraft, and a
-corresponding decline in our membership. This in turn has led to an overall
-decline in the Association's activities, including the publication of this
-newsletter. CASM has made some major changes to its personnel, including those
-directly responsible for the restoration of the North Star. The Association is
-already experiencing a positive change in its relationship with the Museum which
-can only be to the good. We look forward to a renewed partnership with CASM to
-our mutual benefit.
+In this issue we are recognizing the 70th anniversary of a very special event
+in Canadian aviation history and one which involved a North Star aircraft. The
+first non-stop transcontinental flight took place on January 15, 1949 and
+readers will find an article dealing with that event in this issue. In
+addition, your Association, in partnership with the Canadian Aviation and Space
+Museum will be holding a members meeting, and a public viewing of the North
+Star on January 19, 2019 to mark this event.
 
-%\vspace{12pt}
+We are also recognizing another one of our long standing members in this
+issue. Ted Devey has been active in our organization since the very early
+days. We are indeed fortunate that we have a number of other members who have
+been with us since the beginning. In 2003 the Ottawa Citizen predicted "a
+bright future" for our nascent organization. A copy of the Citizen article is
+included in this issue. Not surprisingly, given the challenges involved in
+making this project work, the "future" did not always appear as "bright" as
+Polaris. However, our ever optimistic President reports that 2018 was
+characterized by positive developments in the Project, and our Association's
+relationship with CASM.
+
+The Museum's conservator, in the "Conservator's Corner" has contributed a
+thoughtful article based on the classic 70's book, \textit{Zen and the Art of
+Motorcycle Maintenance} which I am sure
+that our readers will enjoy. Good
+reading.
+
+%Our Association was incorporated in 2003. We are fortunate that a
+%number of the original members of the organization, such as Tim
+%Timmins, are still involved as they can provide an historical
+%perspective. As Tim says, in the article "Our Members," there have been
+%many bumps in the road, and the fact that the aircraft is still a long
+%way from completion after 16 years speaks for itself. In January 2017,
+%R\'{e}jean (Rej) Demers took up his position as CASM's Conservation Project
+%Manager. Rej's first contribution to the Chronicle can be found in the
+%"Conservator's Corner." The restoration of the North Star is a priority
+%for Rej. He has spent the first year of his mandate gathering
+%information and developing systems and procedures which will, in the
+%long run, move the restoration of the aircraft forward in a coherent
+%manner. I have previously stated the obvious and that is the health and
+%vigour of the Association is dependent on the restoration status of the
+%aircraft. In the last few years there have been as many bumps in the
+%road as there are potholes in Ottawa in the spring. The President is
+%very optimistic we have reached a real turning point (see the
+%"President's Notes") and wouldn't that be great.
+
+\vspace{12pt}
 

### 2019-01-13 pnsac/newsletter/article_Feature1.tex
January 2019 changes and additions.

@@
 
-\title{Reunion of Giants}
+\title{Volunteers Assure a Bright Future for Classic Plane}
 %\subtitle{Part 4}
-\author{Bruce Grant}
+\author{Randall Denley}
 
@@
 
-% \textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
-%  Sqn, RCAF.}\\
+\end{multicols}
 
-The "Reunion of Giants" video came to the Canadian Aviation and Space Museum's
-theatre on May 14th, courtesy of the Canadian Warplane Heritage Museum in
-Hamilton. CWHM sent along the pilot Captain Leon Evans (retired Air Canada), and
-Dave Rohrer, CEO, as well as Morgan Elliott the videographer. Also present were
-two World War Two decorated Lancaster pilots Don McKechnie and Bill Button. With
-almost a full house of 200 it was a very successful Project North Star public
-event.
+\begin{figure*}[htbp]
+	\vspace{2em}
+	\centering
+	%name of the graphic, without the path AND in EPS format:
+	\includegraphics[scale=0.8]{holmgren-timmin-scaled.png}
+	%caption of the figure 
+	\caption*{\small \em TIm Timmins, left, and Robert Holmgren are forming a
+group of skilled volunteers to restore a 55-year-old North Star, the first
+Canadian-made plane capable of transcontinental flight.}
+	%label of the figure, which has to correspond to \ref{}:
+	\label{fig:timmins-holmgren}
+\end{figure*}
 
-The outstanding feature of the story is the audacity of CWHM to fly Canada's
-only airworthy Lancaster to England for a reunion with the only other flying
-Lancaster in the world. The Lanc is the flagship of the CWHM, its biggest
-attraction and an irreplaceable historic artifact. In a typical year it flies
-only about fifty hours, attending aviation events in Canada and every move is
-calculated some years in advance.
+\begin{multicols}{2}
 
-In England, the arrival over the North Atlantic route of our Lanc was embraced
-as a tribute to the quality of the machine and the courage of its crew. It
-reinforced the link with Canada which provided so much support to Great Britain
-during the war.
+\textit{This article was published in the Ottawa Citizen on February 1, 2003
+under the byline of Randall Denley. It was brought to our attention recently by
+Drew Hodge. The article is republished with the express permission of the
+Ottawa Citizen, a division of Postmedia Network Inc. (ed.).
+}\\
 
-While on a final approach in England, the supercharger on engine \#3 went to
-pieces and the plane landed with a trail of smoke. A replacement engine would
-have to be found. The replacement that turned up was a Rolls Royce Merlin that
-would have to be substituted for the Packard Merlin in our Canadian-built
-Lancaster. This presented a challenge in sorting out the elements of QEC with
-incompatible Whitworth and SAE fasteners. The new engine would be a "bitsa"
-(bitsa this and bitsa that). They did it in about three days and got the plane
-back in the air.
+The world's last classic Canadair North Star sits, forlorn, on the tarmac at
+the Canada Aviation Museum. It's been parked there since 1966, along with half
+a dozen other orphans of the museum, planes that it has no money to restore and
+that are too big to bring inside the crowded building.
 
-On the return flight, stronger than anticipated headwinds raised the risk of
-running out of gas over the Davis Strait. They diverted to Narsarsuaq in
-Greenland.
+The paint on the North Star is faded and peeling.Some gaps in the wings are
+closed with wire to keep out the birds that have made it home for years.The
+engines are protected with pieces of plywood. The interior requires a complete
+renovation.
 
-Following the showing of the film there was a spirited question and answer
-session with CWHM's representatives moderately by Mike Pearson (retired Air
-Canada pilot of the Gimli glider fame). There is no doubt that all those in
-attendance enjoyed themselves thoroughly and gained some understanding of what
-it meant for those involved in this project.
+It's a shame to see an important part of Canada's aviation history slide into
+this kind of ruin, Thanks to the efforts of two Ottawa men, however, the North
+Star finally has a bright future.
 
-At Project North Star, we have some appreciation idea of the scale of CWHM's
-accomplishment. As our members know we are in the twelfth year of our project to
-restore the last surviving North Star. But, we are restoring it only as a museum
-artifact, not an airworthy machine. The video is available for sale at
-www.warplane.com and the trailer can be viewed ath www.suddenlyseemore.com .
+Robert Holmgren and Tim Timmins are forming a group of skilled volunteers to
+restore the North Star. This kind of volunteer effort will be a first for the
+museum, says director general Anthony Smyth, who is supporting the effort,
+along with aviation companies including Air Canada and Boeing.
+
+The museum is showing commendable flexibility because the traditional view in
+the museum world is that restoration is work that can only be done by trained
+experts. Unfortunately, the museum can't afford that kind of work, and that's
+one of the reasons the plane has been sitting there for more than 35 years.
+
+This is no small undertaking. The 55-year-old North Star 1-ST is a sizeable
+plane, nearly 30 metres long and with a wing span of 35 metres. Restoring it
+will be a yearlong process that would cost \$1 million if it were done by paid
+staff.
+
+The North Star is an important plane in Canada's aviation history, the first
+Canadian-,made plane capable of transcontinental flight. It's considered the
+plane that launched the postwar Canadian aviation industry; 70 of them were
+built. Brought into service after the Second World War, North Stars carried
+passengers for several airlines and troops for the Canadian military.  North
+Stars served to airlift Canadian troops to Korea and provided transportation
+for VIPs visiting Canada. The plane at the museum belonged to the RCAF.
+
+Holmgren is a retired Air Canada maintenance expert, and Timmins is a former
+RCAF navigator who flew on the North Stars. Timmins's former squadron had taken
+an interest in the North Star, and Holmgren, who was giving his time as a
+volunteer at the museum, pushed it forward.
+
+"We'd like to see it brought back to something like its former self," Holmgren
+says.
+
+A temporary structure, perhaps a bubble like the one over the playing field at
+Lansdowne Park, will be necessary before the restoration can begin. The plane
+is too large to fit inside the museum's existing building and also too big to
+be easily transported elsewhere.
+
+Holmgren and Timmins need about 200 volunteers to make the project a success.
+Holmgren has approached Air Canada, Bombardier and Boeing to get the help of
+past and present employees. About 40 volunteers have already come forward.
+
+The key need is skilled craftsmen with up-to-date large aircraft experience and
+who live in or near Ottawa.
+
+Some people who can come in almost every day to run the project are required. 
+There is also a demand for people to do record-keeping and research.
+
+The group is hoping to attract national attention for the project, both to
+provide volunteer expertise and to help raise money.  This is a plan to help a
+national museum, after all.  Smyth says the museum will be involved with a
+fundraising campaign, but it's early yet to say how much is required. It
+depends partly on how much time and material are donated by corporations. A
+full-time salaried project manager will be hired by the museum.
+
+The project will take a conservation approach, retaining as much of the
+original material as possible, Smyth says. The restored plane will be
+structurally complete and faithful to the original. Returning the plane to
+flyable condition is too expensive to be feasible and the museum wouldn't want
+to take a chance flying the last remaining version of such an important plane.
+
+Restoring the North Star is just the beginning of what Holmgren and his
+associates hope to accomplish. The museum has another half dozen planes outside
+that require restoration. Restoring them all will take nine or ten years. The
+Museum's new addition, which it hopes to have open by December of this year,
+will be large enough to finally bring the North Star indoors.
+
+%If you want to help with the project, contact Robert Holmgren at 748-5972 or
+%robertholmgren@rogers.com. The group also has a Web site,
+%www.projectnorthstar.ca.
+
+%\address{Web site:
+%{\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}}\\
+%	General enquiries:\email{info@projectnorthstar.ca}}
+%
 
@@
 
-%\textit{Reprinted from the March/April 2014 edition of "Legion Magazine," with the kind permission of Jennifer Morse, General Manager of Canvet Publications, and Tom MacGregor.}
-
 \begin{footnotesize}

### 2019-01-13 pnsac/newsletter/article_MemberInterview.tex
January 2019 changes and additions.

--- 2018-01-25 pnsac/newsletter/article_MemberInterview.tex
@@
 \title{Our Members}
-\author{Interview with Austin J Timmins}
+\author{Interview with Ted Devey}
 
@@
 
-My aviation experience includes 36 years service in the RCAF and nine
-years with the Canadian Aeronautics and Space Institute. In the RCAF I
-completed radio navigator training, which included communications,
-navigation and air gunner. I later qualified as an airborne
-interception navigator. My operational experience includes air
-transport, search and rescue and training. My assignments included
-Navigation Leader, Aircrew Leader, Squadron Commander, Air Command
-Navigator and Deputy Commander (COO) of the Air Transport Group. I
-logged 9500 hours, mostly on the following aircraft: Beechcraft,
-Dakota, North Star, Mitchell, Yukon, Hercules, and Boeing 707. My
-staff assignments in Europe included movement control (traffic
-management) for 1 Air Division and a tour at The NATO Defense College
-in Rome.
+My service background was naval. Professionally I was an electrical engineer
+with the Department of Communications specializing in radio spectrum
+engineering. After retirement in 1989 I sought activities that interested me
+such as helping to install a pipe organ in a theatre in Renfrew, working on
+restoring a sailing yacht, which unfortunately stopped due to lack of money. I
+also took courses in welding, machine shop and woodworking at Algonquin College
+to enhance my manual skills.
 
-On retirement from the RCAF I participated in a Transport
-Canada A Base review of several airports. I then secured full
-employment with Canadian Aeronautics and Space Institute as Executive
-Director, retiring in 1995.
+2. \textit{How long have you been involved with Project North Star and how and
+why did you get involved?}
 
-2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}
-
-My involvement with efforts to conserve North Star 17515 predates
-Project North Star. Ottawa area members of the 426 Thunderbird Squadron
-Association, mostly retired North Star aircrew, met with the National
-Aviation Museum staff to express their concern about the state of the
-North Star and proposed volunteer involvement in a preservation effort.
-These expressions of concern and interest led to a meeting between the
-Association representatives, where General Adamson presented a proposal
-for the restoration of the aircraft. It was rejected but assurance was
-given that the North Star would be first priority when the new hangar
-was completed Also a North Star Trust Fund would be established. Robert
-Holmgren, an observer at the meeting, then proceeded to develop a
-proposal for volunteers to restore the North Star.
+I attended a public meeting in the Museum in 2003 in which it was indicated
+that efforts were being made to restore the Museum's North Star to a showroom
+condition. In 2004 the project finally was approved and work started.
+Professional work had to be carried out to remove asbestos and many dead birds
+and their mess. At first I used my welding skills in adapting steps/platforms
+for working on engines on the aircraft. In winter, seats were removed from the
+cockpit and brought into the shop where they were dismantled, treated and
+cleaned, painted and finally spray painted. I found the work interesting and
+satisfying and so I was hooked into the project.
 
@@
 
-I have been involved from the beginning of efforts to establish PNS,
-first as a member of a Steering Committee, led by Robert Holmgren,
-later, Vice President and then President of PNSAC. My contacts in
-aviation provided much needed advice and support, including access to
-initial funding for PNSAC.
+In 2006 I expressed interest in the work on the four Roll-Royce Merlin engines.
+As it turned out I was asked if I would take on the job of Merlin Crew Chief
+and agreed with the proviso that I would turn the task over if someone came
+along who was properly qualified. When Garry Dupont retired from the RCMP as
+head AME at the RCMP Hangar at Uplands, I offered the position to him which he
+took over. From 2006 until the summer 2018 I did only engine work on the four
+engines. Unfortunately my limited mobility prevented me from helping with the
+more recent work that was being done behind the firewall of \#4 engine. So I
+went into the Conservation shop to start doing different work.
 
@@
 
-The highlight of my involvement has to be the establishment of PNS and
-PNSAC.
+I think that the highlight of my involvement was back in 2006. To work on the
+engines we needed two engine stands, one to transfer the entire engine assembly
+from the aircraft for dismantling to the core engine and a second stand to
+mount the core engine and further dismantle it to bare components. This had to
+be a rotary affair so that the engine could be positioned for whatever
+operation was being done. The first stand had to be designed and built from
+scratch. The design was an adaptation of the engine stand located next to the
+Lancaster bomber on the Museum floor with ample room for a person to work
+underneath. A specification was drawn up for the various pieces of steel to be
+cut precisely by the steel supplier and to be delivered to the RCMP Hangar at
+Uplands, as instructed by Garry Dupont. On a Saturday a group of volunteers
+assembled at the RCMP Hangar and assembled the horizontal and vertical sections
+of the stand with Garry doing the welding. Following this the stand was
+delivered to the Museum's welding shop for final assembly, then into the main
+shop where it was finally assembled and painted. Everything fit as intended.
+
+The rotary stand was located in the storage area of the Science and Tech
+Museum. The base was extended to the required length and a new frame was built
+to hold the core engine. That everything fit together so well from the start
+was the highlight of the project for me.
 
@@
 
-PNS is now in its 16th year. It has been a work in progress with many
-bumps along the way. I spent a lot of time and effort, along with other
-members of the Executive, evening out these bumps. Establishing and
-maintaining effective channels of communication with CASM and CSTM
-required constant attention. Much effort was directed to fund raising,
-publicity, and, finding volunteers. In the beginning, every date
-brought a new challenge.
+The most challenging part of the job was working on engine \#1, the port outer.
+Since I was not a mechanic, I had to plan the disassembly of the engine and
+carry it out with the help of crew members under the strict guidance of the
+then Shop Supervisor Mike Irvin. He was a hard taskmaster. This was not an
+engine that had just been removed from an operating aircraft for overhaul. The
+North Star and its engines had been left out in the Great Canadian outdoors for
+39 years without any attempt at protection from the four seasons of weather
+during those years. Corrosion was rampant. In each of the twelve cylinders the
+piston rings were rusted to the cylinder liners. Many hours of soaking the
+pistons with Liquid Wrench, the use of compressed air and other techniques had
+to be employed to remove the pistons from the cylinders. There were other
+issues too that made the overall disassembly of the engine very difficult. The
+foregoing applied to each of the other three engines, so it has taken about 12
+years to process the four Merlins to a preserved non-running condition. The
+first engine I found the most challenging as we had to learn as we went along.
+The remaining engines were very difficult but the experience gained with \#1
+engine was valuable in restoring the others.
 

### 2019-01-13 pnsac/newsletter/article_PresReport.tex
January 2019 changes and additions.

@@
 
-These are exciting times to be a member of the Project North Star Association.
-We  have recently taken part in the 2018 Doors Open Ottawa. This is a major
-event for the Museum and PNSAC volunteers were able to make a significant
-contribution to the success of the two open days. Doors Open has become second
-only to Canada Day in its importance to the Museum. (more on Open Doors 2018
-later in this issue).
+It is always interesting to look back on what was written in previous issues of
+any journal. The NStar Chronicle is no exception and so before starting to
+write this piece I reread what I had said during the year 2018. For once luck
+was with me and as far as I can see I did not write anything that was either
+untrue or not achieved during the year.
 
-A few days before Canada Day the aircraft was brought outside. Unfortunately,
-the day itself was extremely hot and the number of visitors was well down from
-previous years. Nonetheless our volunteers did a yeoman service in talking to
-the public about our restoration work.
+The last paragraph of my January 2018 Notes suggested that the year would see
+increased activity both in the aircraft restoration and within the Association.
+Both have proved correct. We have successfully recruited several new
+Restoration Volunteers and our Association activities have increased
+substantially.
 
-Although showing the aircraft to the public is very important to us, our main
-purpose is to restore the North Star to museum condition. Much of our work is
-detailed, painstaking and unspectacular such as restoring a small part of the
-aircraft which can take many days or weeks. At other times making progress is
-very visible. We are in such a situation now.
+Apart from the occasions when our members either met or have the been actively
+taking part in Museum activities we have made a major step forward in that our
+Facebook page has again become active under the stewardship of Chris McGuffin.
+Chris is a dedicated Restoration Volunteer who can frequently be seen around
+the aircraft at a considerable height complete with hardhat and safety harness.
+When he returns to the ground, he often gets out his camera and takes
+photographs, many of which are now appearing on Facebook. For those of our
+readers who are not Facebook subscribers, I encourage you to join Facebook and
+to "like" the page Project North Star. You will then see the latest updates on
+our restoration progress together with the recent photographs.
 
-At the end of May we had one of our regular meetings with Chris Kitzan (Director
-General of CASM) together with Neil Raynor (VP of PNSAC), Cedric St-Amour (CASM
-Volunteer Coordinator) and Reg Demers (Our CASM supervisor There was a full
-exchange of ideas, suggestions for bringing  the public into the Museum and ways
-in which PNSAC could be involved in these initiatives and events. The meeting
-ended with me having a full sheet of notes covering future planning, major
-restoration work and how we can make sure we recruit younger people both as
-members of PNSAC and where possible active Restoration Volunteers.
+Within the next few days our webmaster, Drew Hodge, will be relaunching our
+website with an updated homepage and better links to other parts of the
+website. Much work has gone into the redesign and we look forward to receiving
+comments from users. The website is designed primarily to be a location for all
+the Association's historical and archival information. We are planning to add
+to the website more PDF versions of previous issues of the NStar Chronicle and
+photographs which are not currently available on the site. Current information
+will be sent to members by email and updates to our Facebook page.
 
-A meeting like the one above only happens when an organization is thriving, and
-everybody is excited about the future. During the last 15 years PNSAC has
-demonstrated that it is capable delivering very high-quality restoration work.
-We have already accumulated over 80,000 hours of work and during this time we
-have become a trusted partner of the Museum. Our website is now operational
-again and in the next few weeks we expect to be able to update it and make it
-more relevant to where we are with the project now. One of the major initiatives
-taking place is to develop a diagrammatic history of the work done on the North
-Star restoration and a timeline for future work on the aircraft.
+Our first big event of 2019 will be held at the Museum on Saturday, January
+19th to celebrate the 70th anniversary of the first non-stop flight across
+Canada between Vancouver and Halifax. The flight was made by a North Star on
+15th January 1949. Members have already been advised of the event by email and
+on Facebook and I hope to see many of you joining us on that day. Although this
+is a Project North Star Association event, we have received considerable help
+with planning from the staff of the Museum. This is a good example of the
+co-operative working relationship between the Association and Museum.
 
-In closing, I once again urge Association members to become active volunteers.
-There are many volunteering opportunities both for those with major aviation
-experience to those who can only volunteer with us on occasions like Canada Day
-or can provide help for our many behind-the-scenes activities.
+During the summer we expect to show off the aircraft to the public, as usual,
+on Canada Day. We will also be showing the North Star during Doors Open Ottawa
+at the beginning of June, if the Museum decides to take part in the event this
+year.
 
-Our next Members' Meeting is scheduled for
-Saturday, September 29, 2018. Do reserve this date in your calendar if you can
-be in the Ottawa area on that day. The meeting will be held at the Canada
-Aviation and Space Museum. More details will follow later.
+During 2018 the Museum organised two very enjoyable events for volunteers.
+Invitations were given all Museum registered volunteers including North Star
+volunteers. In the summer we once again had a Bush Pilots' Breakfast at which
+the Museum staff cook a great breakfast for everyone attending (there is
+nothing so good as seeing the Director General, Chris Kitzan flipping
+pancakes). Shortly before Christmas the Museum organised a new event, a pot
+luck lunch for staff and all the Museum volunteers---another thoroughly good
+overeating occasion.
+
+One of the benefits of becoming an active member of PNSAC is for the
+comradeship with other members who all have an interest in aviation. It is not
+necessary to be a volunteer to enjoy the company of other members. Once a month
+there is an informal breakfast held in a caf\'{e} on Arch St which is relatively
+near the Ottawa General Hospital. The breakfast started about two years ago
+under the leadership of Bill Tate. Each month we chat about aviation, the North
+Star project and whatever else we feel like talking about. I usually give the
+group an update on anything of interest that has happened since the last
+breakfast. Joining this informal group is a good way to become involved with
+the Association and to possibly move on to taking a more active role in the
+Association.
+
+We will be holding our Annual General Meeting at the Museum on Saturday, April
+13. At this meeting we re-elect our Board of Directors and would welcome the
+nomination of any member who might be interested in serving on our Board. At
+present we have five members. This leaves one vacancy to bring us up to our
+normal Board number of six members.
+
+2019 is starting well. We look forward further significant milestones in the
+North Star restoration process such as rehanging of the restored Engine \#4. We
+are also hoping to arrange further events for our Association members.
 

### 2019-01-13 pnsac/newsletter/article_ProgReport.tex
January 2019 changes and additions.

--- 2018-01-14 pnsac/newsletter/article_ProgReport.tex
@@
 \author{R\'{e}jean Demers}
+\author{Restoration Project Manager}
 
@@
 
-\textit{R\'{e}jean Demers (Rej, to those who work with him) is a Transport Canada
-licensed Aircraft Maintenance Engineer (AME). R\'{e}jean was a small
-business owner in Gatineau QC for over 10 years, working for a wide
-variety of private and commercial aircraft owners and operators.
-Following a stint in Northern Ontario working on turpoprop freighters
-and bulk fuel haulers, he took up his present position as a
-Conservator, Special Projects Manager with Ingenium, Canada's Museums
-of Science and Innovation in January 2017. His main project is the
-restoration of the Canadair North Star DC-4GM at the Canadian Aviation
-and Space Museum. He recounts below his introduction to his new job.}
+\section{Zen and the Art of Aircraft Restoration.}
 
-Leaving Pickle Lake was like getting shot out of a cannon in a snow
-storm. At that time, Christmas of 2016, I was the last person on my
-crew to leave the YPL base, having seen my colleagues off for the end
-of our tour and the holidays. After a full day of flying, the fleet had
-to be tucked away in the barn ready for service. On Boxing Day, a fresh
-crew would fly in and start another two week tour of line maintenance
-with the company we worked for. However, on that Christmas, I was
-quitting my job.
+Through my daily experience over the past two years, working with volunteers, I
+have asked myself the following question: What makes a volunteer dedicated to a
+project? The answer is not a straightforward one. This article explores the
+subject, much like Pirsig did in his book about metaphysical motorcycles; Zen
+and the Art of Motorcycle Maintenance.
 
-Nobody knew I was leaving until I got back to the bunkhouse and shared
-Christmas dinner with the remaining pilots and ops personnel. I refer
-to them in these terms now, but then and there, I could say they were
-my friends. In the weeks leading to my departure, no one knew I was
-leaving. No one speaks of breaking out of prison when they are behind
-bars. The difference being the only bars I saw were the Pickle Lake
-Hotel and a watering hole in Thunder Bay called the Apollo.
+We strive for quality. As restoration work continues on the North Star, we get
+absorbed into the weight and breadth of the subject matter. This is the kind of
+project that demands the level of commitment found in monasteries, dojos and
+Paratrooper units. Commitment comes from dedication and reward. Quality is
+obtained through application of commitment, producing results which may be
+enjoyed. The following effort in writing has been made to define such elements,
+those necessary to sustain the project.
 
-Driving down the 599 in that snow storm with a truck I had bought for
-\$100 was a white knuckle ride. Knowing this would probably be the last
-time I drove the 2000km distance between Northern Ontario and home in
-Gatineau. The stakes were high: A flight to catch in Ottawa (T-minus
-48hrs), New Year's eve in New York city with my family waiting there
-for me and a new job starting at CASM on the day of my wedding
-anniversary. Could 2017 start with anything less? Arriving to work that
-January 3rd morning was a good feeling, until I got startled by "Manny"
-the mannequin sitting in my predecessor's office chair. If a sense of
-humour was a job requirement, I hoped to get the joke. I would later
-find Manny in other compromising positions, some of which I refuse to
-admit my participation in writing.
+A voracious appetite for detail. Precise facts and data build foundations.
+Foundations of information that erode over time. Lost to floods, fires, or
+worst; thieves, vandals and neglect. Information is our firmament that keeps us
+grounded in our work. Daily worksheets are filled with steps taken to satisfy a
+condition. Drawings are made to translate reality to paper. These words,
+drawings and tables form hard to translate time lines. Coloured bars overtaking
+each other, running in thousand hour blocks of volunteer contributions. These
+life hours, cannot be overlooked, as a restoration project advances through the
+years.
 
-The momentum I have gained in the twelve months since my escape from
-Pickle Lake is only part of what keeps me moving forward. I also count
-on the support of others, namely co-workers and volunteers. My position
-as Project Manager of the North Star restoration relies on the good
-work and cooperation of all hands involved. As the work itself proves
-to be technical and highly detailed, with wide ranges of abilities
-involved, people are always the driving force.
+Inorganic compounds risk distraction. Focus is important when avoiding the
+ethereal draw of solvents. A volunteer scrubber, washing away years of residue,
+layered upon a finish that once gleamed with pride. Slowing the return to dust,
+so feared and fought by the living. A process to follow and adhere to. The
+methodical dismantling and reconstruction of an object. Return to form, that
+which was almost indiscernible. The result of all his work is not merely a
+shiny piece of aluminium panelling; He has travelled through time.
 
-It was suggested my article for the North Star Chronicle be focused on
-the restoration of a certain four engined artifact \#1967.0645. For
-details on the work that has been carried out by our dedicated group of
-volunteers, I recommend you read my conservation report on the subject.
-It is available online: \textcolor{blue}{\texttt{\url{http://www.projectnorthstar.ca}}}
-or by email: \textcolor{blue}{\email{rdemers@techno-science.ca}}
+More than just an object, we look to a hidden past. Beyond these material
+trappings, lay the history of Canadair North Star 17515 and her sisters.
+Examples of service are easily found in books and photos. What proves a
+challenge is explaining how a bumblebee becomes a preserved specimen in an
+engine, how graffiti made during heavy maintenance is the only evidence of a
+man's past. Where the dirt under floorboards, came from countries that no
+longer bear the name she visited.
 
-%\section{Nr 4 Engine}
-%\label{sec:engine_4}
-%
-%Work has continued on much of the ancillary equipment for the engine. The
-%supercharger and intercooler have been completed and installed, as well as the
-%reduction gearbox. The radiators and header tank have been restored and painted, and
-%two of the three radiator sections have been installed. Restoration work has
-%begun on the air intake system, and many of the fuel and oil lines. 
-%
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   \includegraphics[scale=0.5]{"PNS 007-im2".png}
-%   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
-%   \label{fig:engine_no_4_2}
-%\end{figure}
-%
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   \includegraphics[scale=0.5]{"PNS 003-im3".png}
-%   \caption*{\small \em Volunteers Robert Desjardins, Bruce Gemmill and Michel Cote hold the refinished coolant tank from engine \#4.
-Hours of polishing were required to get the tank to look like new.}
-%   \label{fig:coolant_tank}
-%\end{figure}
-%
-%The starter motor was found to be very corroded, and disassembly and repair not viable. A
-%new unit has been ordered and will be installed once delivered. The fire
-%suppression lines are complete and will be installed along with the cowls when
-%ready. Some of the cowl panels have been restored, although it has been found
-%that this engine has more corrosion than the other three engines, possibly it
-%had been in service longer? This has meant that more work is required to
-%refurbish each of the many panels. We still hope to have this engine ready for
-%installation by next spring. 
-%
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   \includegraphics[scale=0.5]{"PNS 008-im1".png}
-%   \caption*{\small \em Volunteer Garry Dupont working on the re-assembly of engine \#4.}
-%   \label{fig:engine_no_4}
-%\end{figure}
-%
-%\section{Fuselage and Main Cabin}
-%\label{sec:main_cabin}
-%
-%With the aircraft moved outside in March, little work was done in the aircraft
-%over the winter. Once the temperatures warmed up, we continued work in the main
-%cabin to prepare for completion of the painting started last year. Repairs to
-%the rear bulkhead are now complete with the installation of several new pieces
-%of aluminum to replace corroded ones. A new section of flooring was installed
-%where the rear washroom was situated. This required replacing about 1000 rivets!
-%
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   \includegraphics[scale=0.5]{"pns 043-im4".png}
-%   \caption*{\small \em A single new floor panel at the rear of the main cabin required over 1000 rivets.  
-%   Cleco clips are used to accurately align the panel during riveting.}
-%   \label{fig:panel_revets}
-%\end{figure}
-%
-%Cleaning and corrosion removal continued throughout the cabin. It seems whenever
-%we feel we are near completion, we find more corrosion. We had also planned to
-%finish priming the floor and painting the rear bulkhead before the aircraft
-%returned to the hanger later in the year, but this was delayed by the shutdown
-%of the project while a new Project Manager was hired by the museum. The wood
-%floors and walls still need to be repaired, or new panels made, and all will
-%require painting. 
-%
-%\section{Cabin Liners}
-%\label{sec:liners}
-%
-%All the large pieces of liner have either been repaired or replaced, but with
-%limitations on space, most remaining work on the cabin liners has been
-%postponed. New window surrounds have also been made, but then all liners will
-%require painting and stenciling, and this must wait until there is more room in
-%the shop and the paint booth. Liners were also made for the top of the heater
-%ducts.
-%
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   \includegraphics[scale=0.5]{P1020369-im5.png}
-%   \caption*{\small \em One of the refurbished heater duct sections with a new fabric headliner secured to the top.
-The sides are covered with mica insulation that was removed, cleaned and reattached after repairs to the metal duct.
-These will be installed in the main cabin.}
-%   \label{fig:heaters_liners}
-%\end{figure}
-%
-%\section{PNS 2017 Update}
-%\label{sec:pnsupdate}
-%
-%We were able to restart work on the aircraft in late January, following the
-%arrival of our new Project manager, Rej Demers. So far, we have completed the
-%top cowl panels and one of the radiator flaps, which required a repair patch to
-%the outer skin, and the electrical harnesses, which are now waiting for
-%painting. Work continues on the air box and the remaining cowl panels.
-%
-%We have decided not to move the aircraft outside this year. With our efforts
-%concentrated on completing engine \#4, it is unlikely we would have the resources
-%to also complete the repair work in the cabin needed to allow for priming and
-%painting, which is the main reason for moving the aircraft outside in the summer.
+The static, prevents complete appreciation. She's more than 45,000 lbs. of
+metal. Superficial understanding can only consider her 39 years an outdoor
+relic. Likewise, so would a skin deep treatment of a last remaining artefact of
+this type. What impact she had before the stillness of the seasons is not lost
+to those who look further.
+
+A biological experience in personal evolution. The volunteer has developed new
+muscles and sinew adapted to producing polished surfaces. An eye for
+microtexture and grain direction has developed. Sympathetic scars are
+accumulated through careful little sacrifices. Navigating the disruptions of
+capricious human needs such as rest, nourishment and leisure. Through the
+application of labour, a piece of history is preserved.
+
+Disappearing horizons. Meditative expanses of aluminum, exhibiting varieties of
+corrosion rarely encountered in a lifetime's career. Cavernous interior spaces,
+where cargo of all sorts was bound for ports beyond. Flight control surfaces
+that have pressed through all shades of dusk, emerging in the dawn of a new
+day. Now removed from the ship, these weathered, beaten, bruised remnants of a
+glorious past await restoration. A new horizon forms, some many years away,
+known to those who are on-board for the long run.
+
+Healthy obsessions. Volunteer restoration technicians take home an experience.
+They take to heart the results of their labour, in forming a product that is
+also intangible. The mind has strengthened, forming resolves that assist,
+understand and process the elements of restoration work. As historians,
+fanatics and unwitting library wanderers soon discover; The story of air
+transport, in its beginning, is far from just a few obscure paragraphs in
+Canada's ongoing history. It lives through those people who continue writing
+it, preserving the physical, extending beyond.
 

### 2019-01-13 pnsac/newsletter/article_Special.tex
January 2019 changes and additions.

@@
 
-\title{PNSAC Merchandise}
-%% \author{Drew Hodge}
+\title{70th Anniversary of first Canadian Non-stop Transcontinental Flight
+}
+\author{Roger Button}
 
@@
 
+"If some countries have too much history, we have too much geography" said
+former Prime Minister William Lyon Mackenzie King, to the House of Commons in
+1936. He would not have known at the time that some thirteen years later an
+aviation event would take place that would help to shrink this geography.
+
+January 15, 1949 was a noticeable day in the history of Canadian aviation. On
+that day the first non-stop transcontinental flight was made. The aircraft was
+Royal Canadian Air Force North Star, registration number 17512. It was operated
+by 426 Thunderbird Squadron which was stationed at Dorval, QC. The aircraft was
+captained by F/O J.A.F. Jolicoeur and crewed by 14 members of the Experimental
+and Proving Establishment from RCAF Rockcliffe.
+
 \begin{figure}[ht!]
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=1.0]{fleece_YOW7578.eps}
+   \includegraphics[scale=0.5]{nstar-212-scaled.png}
    %caption of the figure 
-   \caption*{\small \em PNSAC winter apparel}
+   \caption*{\small \em North Star 17512}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:merchandise}
+   \label{fig:nstar-17515}
 \end{figure}
 
-The first frost warnings on the Weather Channel are reminders that
-fall and winter are just around the corner. Keep warm on those frosty
-fall days with a PNSAC toque and sweatshirt. We can also take
-individual orders for the micro fleece vests shown in the photo on the
-left, and if you're looking for early Christmas presents, how about a
-coffee mug or a t-shirt?
+While 17512 was relatively new to the RCAF it had been one of the earliest
+aircraft to roll off the Canadair Assembly line at Cartierville, QC in 1946.
+However, instead of going directly into service with the RCAF it, along with
+six other aircraft were loaned to Trans Canada Airlines (later to become Air
+Canada). It was was not until 1948 that the aircraft was taken on strength by
+426 Squadron.
 
-See our Facebook page,
-{\normalfont\color{blue}\texttt{\href{https://www.facebook.com/media/set/?set=a.610706705637878.1073741851.284829008225651&type=1}{PNSAC
-      Merchandise}}}, for details about what we're offering, and place your
-orders by sending an email message to the following address
-\email{pnsac.merchandise@gmail.com}.
+The aircraft had a maximum range of 3,060 miles. This allowed 17512 to make the
+non-stop flight from Vancouver to Halifax a distance of 2,785 miles. The flight
+took eight hours and thirty two minutes. A year later the aircraft bested its
+time by making the same journey in eight hours and twenty five minutes.
 
-For members in the Ottawa-Gatineau region, you can arrange to pick up
-your merchandise at the Canada Aviation and Space Museum any Monday or
-Tuesday, between 10:00 AM and 3:00 PM. We can also ship your purchases
-to you for an extra fee. Please specify museum pickup or shipping with
-your order.
-
-Merchandise is also available during our Quarterly Meetings
-at CASM.
-
-Look forward to hearing from you.
-
-``The Merchandise Committee''
+While the flight was an aviation first, it was also another step in the history
+of transcontinental travel in Canada. This started with arrival of the famous
+Scottish explorer Sir Alexander Mckenzie at the Pacific Ocean on July 22, 1793.
+He became the first European to cross North America north of Mexico. It was
+nearly one hundred years later that the next major leap in transcontinental
+travel took place with the completion of the Canadian Pacific Railway when the
+last spike was driven on November 7, 1885 at Eagle Pass, BC. In October 1920
+the Canadian Air Board staged a cross country mail demonstration but this
+involved a relay of five aircraft and a total flying time of 49 hours from
+Halifax to Vancouver. In 1939 Trans Canada Airlines flew the first commercial
+transcontinental flight from Toronto to Vancouver in 16 hours with four stops.
+The airline, however, didn't launch a non-stop transcontinental flight between
+the same airports until June 1, 1957. The flight also took eight hours and
+thirty two minutes using a Super Constellation. This was same time, to the
+minute, that the North Star took eight years earlier and on a longer route.
 

### 2020-02-24 pnsac/newsletter/article_EditorsNotes.tex
Draft updates for Feb-Mar 2020.

--- 2018-07-19 pnsac/newsletter/article_EditorsNotes.tex
@@
 
-Our readers will appreciate from the articles and photographs in this edition that real progress
-being made on the restoration of the aircraft.
-A number of new restoration volunteers have signed on this year bringing with them diverse skills and experience.
-Whoever thought that someone with a mountain climbing background would be able to put his special skills to work on an aircraft.
-Believe it or not but see the "Conservator's Corner."  As I have previously commented: When work on the aircraft is moving ahead it encourages enthusiasm at all levels of the organization.
-One of the biggest cheerleaders of our Association is its president, Richard Lodge.
-Richard wears many hats including that of a restoration volunteer so he knows of what he speaks.
-In this edition Richard discusses how he came to be involved in Project North Star and joins the illustrious members of the Association who have contributed to "Our Members" column.
-But I would be remiss if I didn't thank all of our contributors to this edition including our publisher Drew Hodge.
+Putting out this edition of the Chronicle has been an example of great
+co-operation from members and non-members alike. Thanks are owed to a
+number of people. Our member Gary Whitten was instrumental in getting
+the necessary permission from Air Force Magazine to publish the
+"Diversion to Alert" article.  Allan Bowes, another member, suggested
+that we should try to get the "Civil Merlins" article into an edition
+of the Chronicle.  Following his suggestion we contacted Ben Dunnell,
+the editor of Aeroplane Magazine. Through Ben's offices we were able to
+secure permission to reprint this article.  The "Civil Merlins" was
+part of a comprehensive 17 page Aeroplane Database article on the
+Merlin. If any members wish to read the entire article it was published
+in the July 2017 issue.
+
+In addition thanks go out to the other contributors, namely Richard
+Lodge, Garry Dupont,  Neil Raynor and Bruce Gemmill. Bruce took the
+lead in putting together the Progress Report.  Finally, I would like to
+thank Drew Hodge our stalwart publisher of this edition. We are always
+on the lookout for content for the Chronicle so please do not hesitate
+to bring forward any ideas that you might have. Enjoy.
 

### 2020-02-24 pnsac/newsletter/article_Feature1.tex
Draft updates for Feb-Mar 2020.

--- 2017-03-13 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{Reunion of Giants}
+\title{Diversion to Alert}
 %\subtitle{Part 4}
-\author{Bruce Grant}
+\author{Malcolm G. Morrison, RCAF F/L Ret.}
 
@@
 
-% \textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
-%  Sqn, RCAF.}\\
+%\end{multicols}
 
-The "Reunion of Giants" video came to the Canadian Aviation and Space Museum's
-theatre on May 14th, courtesy of the Canadian Warplane Heritage Museum in
-Hamilton. CWHM sent along the pilot Captain Leon Evans (retired Air Canada), and
-Dave Rohrer, CEO, as well as Morgan Elliott the videographer. Also present were
-two World War Two decorated Lancaster pilots Don McKechnie and Bill Button. With
-almost a full house of 200 it was a very successful Project North Star public
-event.
+%\begin{figure*}[htbp]
+%	\vspace{2em}
+%	\centering
+%	%name of the graphic, without the path AND in EPS format:
+%	\includegraphics[scale=0.8]{holmgren-timmin-scaled.png}
+%	%caption of the figure 
+%	\caption*{\small \em TIm Timmins, left, and Robert Holmgren are forming a
+%group of skilled volunteers to restore a 55-year-old North Star, the first
+%Canadian-made plane capable of transcontinental flight.}
+%	%label of the figure, which has to correspond to \ref{}:
+%	\label{fig:timmins-holmgren}
+%\end{figure*}
 
-The outstanding feature of the story is the audacity of CWHM to fly Canada's
-only airworthy Lancaster to England for a reunion with the only other flying
-Lancaster in the world. The Lanc is the flagship of the CWHM, its biggest
-attraction and an irreplaceable historic artifact. In a typical year it flies
-only about fifty hours, attending aviation events in Canada and every move is
-calculated some years in advance.
+%\begin{multicols}{2}
 
-In England, the arrival over the North Atlantic route of our Lanc was embraced
-as a tribute to the quality of the machine and the courage of its crew. It
-reinforced the link with Canada which provided so much support to Great Britain
-during the war.
+\textit{This article was published in Air Force Magazine in April 2019. It was
+brought to our attention by Gary Whitten. The article is republished
+with the express permission of Malcolm Morrison, RCAF F/L Ret. and the
+publishers of Air Force Magazine.}\\
 
-While on a final approach in England, the supercharger on engine \#3 went to
-pieces and the plane landed with a trail of smoke. A replacement engine would
-have to be found. The replacement that turned up was a Rolls Royce Merlin that
-would have to be substituted for the Packard Merlin in our Canadian-built
-Lancaster. This presented a challenge in sorting out the elements of QEC with
-incompatible Whitworth and SAE fasteners. The new engine would be a "bitsa"
-(bitsa this and bitsa that). They did it in about three days and got the plane
-back in the air.
+Immediately after New Year 1960, while based at RCAF Stn Trenton, and
+flying North Stars for 426(T)Sqn., I was assigned to be the First
+Officer on a flight to pick up an entertainment troupe of 30 Passengers
+from Montreal. This was planned as an 8 day trip providing
+entertainment to personnel working along the Quebec, Ontario and
+Manitoba sectors of the Mid Canada Detection Line.
 
-On the return flight, stronger than anticipated headwinds raised the risk of
-running out of gas over the Davis Strait. They diverted to Narsarsuaq in
-Greenland.
+At the height of the Cold War, The Mid Canada Line was to be the main
+detection line between Canada and the USSR. It was located
+approximately along the 55N latitude parallel, and stretched from
+Hopedale, Labrador to Dawson Creek, B.C. The system was put into
+service in 1957, and decommissioned in 1964. The more northerly DEW
+radar line began full operation in 1957.
 
-Following the showing of the film there was a spirited question and answer
-session with CWHM's representatives moderately by Mike Pearson (retired Air
-Canada pilot of the Gimli glider fame). There is no doubt that all those in
-attendance enjoyed themselves thoroughly and gained some understanding of what
-it meant for those involved in this project.
+Our 426Sqn crew consisted of Captain – F/O H.(Hank) Zbesheski, First
+Officer – F/O M.G.(Mac or Mo) Morrison, Navigator – F/O Roger Fridel,
+Radio Officer – F/O Peter Dragovich, Flt Engineer – Sgt Miller, Flt
+Engineer – Sgt Sopaz, Trans Tech – LAC Hall.
 
-At Project North Star, we have some appreciation idea of the scale of CWHM's
-accomplishment. As our members know we are in the twelfth year of our project to
-restore the last surviving North Star. But, we are restoring it only as a museum
-artifact, not an airworthy machine. The video is available for sale at
-www.warplane.com and the trailer can be viewed ath www.suddenlyseemore.com .
+It should be noted that, on northern flights such as this, it was
+normal to have two Flight Engineers on the crew, due to the difficulty
+of fueling and servicing the North Star in remote and unsupported
+airfields. Had there been plans to travel north of Churchill, we would
+normally carry an extra navigator as well. Another part of the Squadron
+normal operating procedure was that; providing the First Officer was
+Type qualified and current, the Captain and First Officer would switch
+seats on alternate legs. The North Star did not have nose wheel
+steering available in the co-pilot's position.
+
+Typically, on the day prior to departure from home base the whole crew
+would meet and review the trip Itinerary, and fully discuss any
+foreseen issues that might be expected. We were all aware of the
+anticipated cold weather, the short snow packed runways at most of the
+planned stops, and the lack of any useable hangar facilities that could
+accommodate a North Star, with the exception of RCAF Churchill.
+
+Having introduced the issue of cold weather, I must now digress for a
+moment to explain one of the severe consequences of cold weather on
+aero engines in 1960.   At ambient temperatures below minus 18 deg. C,
+90w oil is below its "pour point"; in other words it would not flow
+through either a filter or a radiator such as the type required to cool
+the oil on the North Star. Each engine had a 25 gal oil tank located in
+the wing immediately behind the engine firewall. To ensure that the oil
+system worked after a prolonged (30 to 40 minute) engine shut down in
+temperatures below minus 18 deg. C, the requirement was to dilute the
+oil with raw gasoline directly from the main fuel system. As an example
+at a temperature of minus 34 C the dilution time was approximately 3
+minutes when the oil temperature had cooled to 40 deg. C. After
+start-up the next morning the gasoline had to be boiled out of the oil
+for approximately 35 minutes once the oil had reached a
+40degC.temperature.
+
+These procedures were called "oil dilution" and oil "boil-off". The
+dilution and boil off times which I have cited are approximations only,
+since I have not had access to the Standard Operating Procedures for
+the North Star for quite some time.
+
+The reason for dwelling on the oil dilution procedures is that, in
+January of 1962, North Star \#17520 operated by 412 Sqn. was lost at Hall
+Beach NWT, during a VIP tour of Dew Line sights. Within minutes of
+takeoff from Hall Beach in minus 40C conditions, the crew experienced 3
+engine failures in rapid succession. Under very difficult conditions,
+they were able to force land the aircraft on the infield of the
+airport. Fortunately, all passengers and crew were able to exit the
+aircraft without injury. On a subsequent trip in April of 1964 doing a
+similar Dew Line site tour in \#17502, I witnessed the remains of 17520
+still sitting in the infield, a stark reminder of oil dilution
+requirements.
+
+To continue with my story, with the exception of St. Hubert and
+Churchill, there were no stairs available to board or deplane our
+passengers on, and thus the passengers were required to climb down a
+rather slim aluminum ladder each time. This process could amount to
+more than a 30-minute exercise each time, and in those temperatures
+this raised the oil dilution issue. We fully briefed on oil dilution
+procedures prior to shut-down, and burn-off times that would be
+required after engine startup and warmup the following morning. It was
+agreed that, in spite of the extended periods, the passengers would be
+required to board and remain seated, during the boil-off period prior
+to take-off, and that we would perform both the dilution and boil-off
+while the passengers were on board the aircraft.
+
+Our planned itinerary was as follows:
+
+\begin{itemize}
+  \item 8 Jan---Trenton to St. Hubert to Sept Iles PQ---overnight at RCAF Stn
+Moisie 
+  \item 9 Jan---Sept Iles to Knob Lake---overnight at RCAF Stn Knob Lake
+  \item 10 Jan---Knob Lake to Great Whale PQ---overnight at RCAF Stn Great
+Whale River 
+  \item 11 Jan---Great Whale to Winisk ON---overnight at RCAF Stn
+Winisk 
+  \item 12 Jan---Winisk to Churchill MB---three overnights at RCAF Stn
+Churchill 
+  \item 15 Jan---Churchill to St Hubert to Trenton
+\end{itemize}
+
+RCAF Stn Moisie was not part of the Mid Canada Line but was part of the
+original Pinetree line. The airport at Sept Iles was not part of RCAF
+Moisie, but was managed by the Federal Department of Transport. The
+airports along the Mid Canada Line, although used by some civilian
+contractors, were all operated by the RCAF.
+
+The schedule, as planned, was to arrive at each of the entertainment
+destinations in the late afternoon so that the Troupe could have supper
+and then prepare for a show at approximately 8pm each evening. The
+exception would be the days at Churchill, during which time the Troupe
+would travel for 4 to 5 hours via train from Churchill to RCAF Stn
+Bird, about 180 mi. to the south, entertain the base personnel, and
+return to Churchill the following day.
+
+As they say "all the best laid plans of mice and men", something is
+bound to cause a SNAFU. The following is an account of what happened,
+in the best chronology that my memory and my log book records can
+account for.\\
+
+\noindent\textbf{Day 1, 8 Jan 1960, North Star \#17502:}\\
+
+The flight from Trenton to St. Hubert was uneventful, and our
+passengers had already arrived, so that there was a minimum of delay in
+boarding. The fuel load was topped up to the maximum that we could
+carry, bearing in mind that fuel at Sept Iles would have to be
+purchased on a DPO (detached  purchase order) at a significantly higher
+cost than normally paid by the RCAF. We also tried to minimize the fuel
+requirements at the Mid Canada sites due to the need to use fuel cache
+supplies from barrels at those destinations.
+
+The 3:05 hour flight to Sept Iles was also uneventful, with the arrival
+being in darkness and in a light snowfall. Our passengers deplaned, via
+the ladder, to an awaiting bus after remaining seated on board the
+aircraft while the crew carried out the mandatory oil dilution
+procedure prior to securing the aircraft for the night.\\
+
+\noindent\textbf{Day 2, 9 Jan 1960, North Star \#17502:}\\
+
+The planned chock to chock time for the leg Sept Iles to Knob Lake was
+1:15 Hrs., but was logged as 2:00 Hrs., because of the engines on to
+engines off time required to accommodate the boil off and dilution
+procedures. The flight was uneventful; however, the landing in calm
+wind conditions on a 4,500 ft. runway and with the extra fuel load was,
+to say the least, challenging. The packed snow condition did not offer
+good braking action.\\
+
+\noindent\textbf{Day 3, 10 Jan 1960, North Star \#17502:}\\
+
+The overnight temperature dropped to minus 40C as was forecast, and by
+noon had not gone up appreciably. The passengers were loaded onto an
+ice-cold aircraft and told that it would be at least 40 minutes before
+we could commence our take off and thus turn on the main cabin heaters.
+They were told that they could come into the crew area 3 or 4 at a time
+and share what heat came off the cockpit heater (which had a blower
+attached to it), and do that until such time as we had finished the
+boil-off prior to take-off.
+
+The engine start-up sequence on the North Star was normally 3-4-2-1.
+Startup in cold weather was always tricky, in that the amount of
+priming time recommended was not clearly defined in the SOP's other
+than to say that, if you over-primed, you may damage the engine on
+start or, if you under-primed, the engine may not start, and the
+starter may be damaged by over-cranking.
+
+We were lucky with the \#3 \& \#4 starts, but when we got to \#2, our luck
+ran out and we had a severe backfire on startup, thus making the engine
+totally unserviceable.
+
+After unloading our passengers and advising them that, if they wished
+to put on another show for the base personnel that afternoon or
+evening, there would be no problem as far as our crew was concerned,
+and that we would advise them later in the day as to what the plan
+would be going ahead. We immediately advised ATCHQ as to our problems,
+and were advised a few hours later that another North Star with a
+replacement engine was being dispatched that afternoon, and would be
+expected in the early evening. The plan was that the engine change
+would take place overnight, and that our crew would carry on the next
+day using the replacement aircraft and complete the Mid Canada Line
+tour.
+
+Sure enough, about 1800Hrs local time North Star \#17512 landed at Knob
+Lake with a few extra technicians to assist with the engine change. The
+next 6 to 7 hours were probably one of the most challenging that our
+flight engineers had ever faced. With a minus 43C temperature, no
+hangar facility, and virtually no support equipment, other than a 6,000
+lb. forklift, these professionals, with some assistance from the rest
+of our two crews, carried out the miracle of a complete engine change,
+and the loading of the damaged engine onto \#17502. They also started
+the replacement engine and carried out an oil dilution on the new
+engine, so as to be safe for an overnight stay.
+
+The reason for our crew and passengers not carrying on with \#17502 was
+the requirement to "air test" the aircraft after an engine replacement.
+This air test could not be done with passengers on board. The crew
+returning 17502 to Trenton with the unserviceable engine on board
+carried out the required "air test" enroute.\\
+
+\noindent\textbf{Day 4, 11 Jan 1960, North Star \#17512}\\
+
+The 2:50 Hr. flight from Knob Lake to Great Whale River was uneventful.
+The passengers survived the 50 min. warm up and boil off period prior
+to takeoff with admirable patience. On arrival at Great Whale, the
+temperature of minus 29C mandated another oil dilution prior to
+shut-down and deplaning.\\
+
+\noindent\textbf{Day 5, 12 Jan 1960, North Star \#17512}\\
+
+A short hop from Great Whale to Winisk of 1 hour flight time was logged
+as 1:50 Hrs. "engines on" to "shutdown". By this time our passengers
+had fully accepted the necessity of the extra wait on the aircraft. The
+flight was uneventful.\\
+
+\noindent\textbf{Day 6, 13 Jan 1960, North Star \#17512}\\
+
+When the crew appeared at the mess hall for breakfast, we were handed a
+very brief "Confidential " message from ATCHQ advising us to advance
+our planned departure as much as possible, proceed to Churchill,
+deplane our passengers, refuel and proceed immediately to Winnipeg
+where we would be given further instructions.
+
+Following a 2:20 hour flight to Churchill we advised our passengers
+that we had to proceed to Winnipeg immediately, but would be back to
+pick them up and return them home in three days. The 3:10 hour flight
+to Winnipeg was uneventful, with arrival being after dark. The aircraft
+was immediately towed into a hanger with the instructions that we could
+refuel in the morning when we had been fully briefed on our future
+mission.
+
+As the crew deplaned we were met by an army major and taken to a
+briefing room, where we were told the "Confidential" details of our
+diversion. The following is a short synopsis of that briefing. We were
+to get a good night's sleep, and first thing in the morning we would be
+given a full briefing on the perils and extreme hazards that a 15 drum
+load of "specially denatured alcohol 3C, 200 proof" would create on an
+"emergency flight" to CFB Alert the next morning.
+
+The planned route was Winnipeg to Churchill, Resolute Bay and Alert,
+unload, and on to USAF base Thule in Greenland for a crew rest. It was
+agreed that Nav. F/O Fridel was well enough versed in grid navigation
+to eliminate the requirement for a second navigation officer.  We were
+advised that sufficient food would be provided to see us through to
+Thule, and that both Churchill and Resolute would be alerted as to the
+urgency of our mission, and would give us all the required assistance
+for quick turn-around at those bases.\\
+
+\noindent\textbf{Day 7, 14 Jan 1960, North Star \#17512}\\
+
+After a hearty breakfast, our crew was driven to an army site a short
+distance from the airport. There we received an extensive talk on the
+hazards of denatured alcohol. They dealt with the fact that it was
+among the most hazardous of products that could be permitted on an
+aircraft, also that its toxicity was extremely high, and that we were
+not to risk exposure to it. We were fitted and provided with full
+facial gas masks and advised that we must have them within arms' reach
+of us throughout the flight. We were told to remove the over-wing
+emergency hatches for the duration of the flight, and even during the
+leg from Alert to Thule, to ensure that all possible residual fumes had
+been evacuated from the aircraft. If there was the slightest hint of
+fumes within the crew compartment, we were immediately to shut down the
+nose heater, and eliminate any possible source of a spark within that
+area. Last, but absolutely not least, we were told that all the airport
+emergency forces along our route would be alerted as to the extreme
+nature of our cargo.
+
+Upon our return to the airport we noted that army personnel were
+already in the process of loading the aircraft. By the time that we had
+completed and filed our flight plan, the army personnel had finished,
+and we boarded, checked the security of our load, and proceeded with
+the rest of our pre-start check.
+
+The takeoff for Churchill was normal in all respects, apart from the
+fact that the huge array of crash trucks and flashing lights did make
+us wonder as to exactly how hazardous our load was. The enroute weather
+to Churchill was forecast to be good, with a high thin overcast, thus
+we opted to cruise at 8,000 ft. On reaching cruising altitude LAC Hall
+went into the main cabin for a load check.  Within seconds he returned
+to the cockpit and grabbed a gas mask as he advised us that we had at
+least two leaky drums in the cargo area. Immediately we shut down the
+nose heater, advised Winnipeg of our situation and commenced a return
+to the airport. The whole crew donned the supplied gas masks. This
+significantly hampered our ability to communicate, both among the crew
+and with Winnipeg tower. The tower gave immediate clearance for a
+straight-in approach from the north. The approach and landing were
+uneventful, other than the discomfort of the gas masks. The aircraft
+was surrounded by a sea of flashing red lights as we cleared the runway
+and taxied to the RCAF side of the airport.
+
+The Army was there immediately to inspect the drums and found the two
+defective ones. There was no evidence of any of the others being
+problematic. A quick decision was made to remove the two drums and not
+replace them, thus reducing the load to 13 drums. This whole process
+consumed valuable crew time, which put us in the position whereby, if
+we to proceed, we would be well in excess of the normal maximum of 18
+hours from briefing to the final engines off time for the day. We
+decided to take a 12 hr. break and tackle it again first thing in the
+morning.\\
+
+\noindent\textbf{Day 8 \& 9, 15--16 Jan 1960, North Star \#17512:}\\
+
+Following a good crew rest and another hearty breakfast, we immediately
+proceeded to flight planning. The aircraft was kept in a hanger by
+itself overnight, witsh just enough heat to eliminate any dilution or
+burn-off requirements. The take-off, climb-out and cruise to Churchill
+all went well. The one issue that we had was the heat comfort level in
+the cockpit. We found that with the escape hatches open, any heat
+generated by the cockpit heater was quickly sucked out through the
+missing main cabin hatches, in spite of the cabin door being firmly
+closed. Landing in Churchill was uneventful. We immediately briefed for
+Resolute with an outlook for Alert. The forecast for Resolute was
+anything but promising with strong crosswinds and blowing snow. We did
+not consider this to be a major problem, as both Hank Zbesheski and I
+had landed in Resolute a number of times during the previous 12 months,
+many of those being in rather adverse conditions. The flight plan was
+filed using Frobisher Bay as an alternate.
+
+The flight to Resolute Bay was as good as could be expected under the
+circumstances, with everyone in the cockpit wearing gloves or mitts to
+keep our hands warm. The approach into Resolute was turbulent with the
+crosswinds blowing from the west at 25 to 30 Knots. Hank executed a
+perfect landing on a snow-packed runway in spite of the problems of
+getting the nose-wheel to steer on the slippery surface. The real
+challenge, was to refuel and restart the engines as quickly as possible
+so as to avoid the requirement to oil dilute. We succeeded to start
+within less than 40 minutes, thanks to our flight engineers, both of
+whom worked on the over-wing refueling process together.
+
+The runway at Resolute Bay is almost true north and south. The
+buildings were at the time located to the west of the runway with the
+bulk of them being located about halfway along the runway and not less
+than 50 yards from the side of the runway. We decided to execute the
+takeoff to the south, since the windsock appeared to give us a slight
+headwind in that direction.
+
+Standard takeoff procedure in a strong crosswind required that the
+pilot in the right-hand seat would handle the control column and ensure
+that there was good nose-wheel contact as well as maintain lateral
+stability on the into-wind wing. The pilot in the left-hand seat would
+handle the throttles up to takeoff power, hand them off to the flight
+engineer, and at the same time control the directional stability of the
+aircraft with a combination of the nose-wheel steering and rudder. At
+the appropriate moment he would take over the control column and
+execute the take-off.
+
+The first half of the take-off run went well. As we approached takeoff
+speed we encountered the full force of the crosswind and I found it
+extremely difficult to maintain direction with the nose-wheel and opted
+to ease it off the runway at about the same time as a heavy gust lifted
+us into the air. The gust immediately subsided and we hit back onto the
+runway rather heavily on the starboard landing gear. We once again hit
+a gust and this time I was able to cleanly pull the aircraft into the
+air. The landing gear retracted without problem, and we climbed away
+toward Alert without further incident or thoughts of the critical
+issues of the take-off.
+
+The flight to Alert was uneventful in spite of the fact that we lost
+all radio contact with Resolute within 10 minutes after take-off. (This
+was not an unusual circumstance in these latitudes). The arrival at
+Alert was in dead calm winds and absolutely clear. The ground
+temperature was minus 43 C. and our plan was to unload as quickly as
+possible, do an external walk around, and depart as quickly as
+possible. The base personnel at Alert obliged and had us offloaded
+within 30 minutes of engine shutdown. It was agreed that one of our
+F/E's would do the external. Shortly after he went down the ladder, he
+was back up and asking the two pilots to return down the ladder with
+him.
+
+We witnessed a rather large pool of red hydraulic fluid under the
+starboard undercarriage leg and noted that the normal 6" to 8" of leg
+extension was not there. There appeared to be no other damage to the
+undercarriage or any other visible issues that would hinder us from
+flying. We immediately advised the ground personnel of our intent to
+depart with haste, so that we could do a much more thorough inspection
+at the USAF base at Thule, Greenland where they had hangars that would
+give us some relief from the extreme cold.
+
+The start-up, takeoff and flight to Thule, 2:10 hours due south of
+Alert, was uneventful, with Hank executing a perfect landing on the
+long smooth runway. After clearing the runway we were immediately
+marshalled directly into a well-lit empty hangar, without shutting down
+until fully inside the hangar.
+
+Once settled inside the hangar we were in a better position to assess
+the condition of the aircraft. The first obvious issue was that of the
+starboard landing gear assembly. There was no visible damage to the
+assembly, other than the fact that we had lost all the oleo leg fluid
+and that the seals had probably failed as well. It would be a major
+undertaking to replace the seals. A second issue was observed by our
+engineers, which may have proved to be more threatening than the
+landing gear. There were significant ripples on the aluminum upper wing
+surface skin on the starboard wing. This condition seemed to be much
+more prevalent on the starboard wing than on the port wing. The
+conversation was joined by two USAF airframe technicians and after a
+long discussion, the consensus was that the starboard main spar may
+have been damaged. If that was in fact the case, the aircraft was no
+longer flyable. Fortunately the communications via a secure phone line
+routed through Washington to Trenton were excellent and we were able to
+relay all the information to ATCHQ. We were told to get to bed and that
+we would receive advice the following day. It had indeed been a long
+day.\\
+
+\noindent\textbf{Days 10 to 13, 17/18/19/20 Jan 1960, Thule, Greenland:}\\
+
+By noon on the 17th we received a message from ATCHQ advising us that a
+North Star would be leaving Trenton on the 18th and routing via
+Winnipeg, Churchill, and Resolute Bay to Thule. The load would include
+a crew of 6 airframe technicians and a complete new main landing gear
+leg assembly. The aircraft, \#17514, captained by F/L Jock Hutchinson,
+arrived late afternoon 19 Jan, offloaded the spare parts and repair
+crew and proceeded on to Alert with additional supplies the following
+day.
+
+The Technical crew got to work immediately and carried out a full
+non-dynamic inspection of the wing and main spar structure. They found
+the wing to be fully sound and the dimensional check to be fully within
+limits. With a huge sigh of relief from all concerned, they then
+proceeded to replace the starboard landing gear strut assembly. Within
+48 hours the repair crew had put the aircraft back in fully serviceable
+shape. Quite a feat when you consider the circumstances and lack of
+main base support that they had. It was thought that the excessive
+ripple effect that had been witnessed on our arrival at Thule was
+caused by a combination of the rapid heating of the upper wing skin
+surface, primarily because of the type of heaters in the hanger, versus
+the extremely cold soaked spar and interior structural frame of the
+wing. By the time the aircraft had been in the hanger for 3 days the
+temperature of the complete structure had equalized and thus the
+ripples had all but disappeared.\\
+
+\noindent\textbf{Days 14 \& 15, 21/22 Jan 1960, North Star \#17512:}\\
+
+By late afternoon we had fueled, flight planned, and were ready to
+depart on an 8:05 hour flight to Sept Isles PQ. This would be our
+closest point for clearing Canadian Customs and Immigration, as
+required by regulations. The landing at Sept Isles was carried out in
+moderate snow, which did not hinder our arrival. The Customs Officer
+was perturbed because of the late hour (after midnight local time) and
+the fact that he had to climb the ladder to access the cabin of the
+aircraft. His inspection totally disregarded the personal baggage on
+board, and was focused only on the large crate that contained the
+undercarriage leg. We almost had to uncrate the leg before he would
+accept the fact as to what the crate contained. After refueling and
+sweeping the accumulated snow off the wings, we were able to restart
+all engines and proceed on our 3:15 hour flight to Trenton. The flight
+and arrival at Trenton was uneventful.
+
+During a rehash of the journey a few days later, it was disclosed to us
+that a short time prior to our arrival at Alert, the base had resolved
+the problems that required the denatured alcohol, but felt that they
+would like to have the product in case of future issues. We never did
+find out how our entertainment troupe made the return trip from
+Churchill to Montreal.
+
+%\address{Web site:
+%{\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}}\\
+%	General enquiries:\email{info@projectnorthstar.ca}}
+%
 

### 2020-02-24 pnsac/newsletter/article_Feature2.tex
Draft updates for Feb-Mar 2020.

--- 2014-05-17 pnsac/newsletter/article_Feature2.tex
@@
 
-\title{The Whispering Giant}
-\author{Dave Lambeth}
-
+\title{The Civil Merlin} 
+\author{Dr. Jakob Whitfield}
 \maketitle
 
-Twinkle, twinkle, old North Star,\\
-With wrinkled skin and creaky
-spar,\\
-For years you’ve always been on
-sked,\\
-As many of our force you’ve sped\\
-to distant lands beyond the skies,\\
-you’ve never failed to see the rise\\
-of distant suns or moonrise glare.\\
-You’ve carried us with tender care\\
--back you came to take us home,\\
-from old Japan or France or Rome\\
-from England, Goose, and Kef,\\
-We cannot hear you.\\
-(‘cause we’re deaf)\\
+\textit{The following article is an extract from a larger piece which appeared
+in the July 2017 edition of Aeroplane magazine under the heading
+"Rolls-Royce Merlin."  We wish to thank Allan Bowes for bringing it to
+our attention. This article is republished with the kind permission of
+the publishers of the magazine and the author Dr. Jakob Whitfield.}\\
 
-\begin{footnotesize}
-    \raggedleft PNSAC\\
+The earliest Merlins to operate in a civil mode were the Merlin T24
+series, developed in 1944. These were single-stage twin-speed units
+similar to the Merlin 24s fitted to the RAF's Lancasters, but were
+modified to improve service life, and were fitted to Transport
+Command's Avro Yorks. Long-range transport operation entailed running
+at relatively low cruise powers for long periods — under these
+conditions the lower cylinder head temperatures caused deposits of lead
+oxide from the fuel, resulting in excessive spark plug fouling. To
+counter this, the Merlin T24/4 incorporated a charge heater to increase
+the inlet temperature. Post-war, the Merlin 500 series essentially
+comprised civil and export versions of the T24, incorporating its
+modifications, though only the Merlin 501 included a charge heater.
+
+The 600 and 700-series engines were two-speed two-stage engines, based
+on the military 100-series. They were fitted with the so-called
+'transport heads and banks', strengthened for greater reliability. Most
+marks had some form of variable intercooling to allow for charge
+heating to reduce plug fouling under cruise conditions. Initially the
+system was plagued by coolant leaks, but as experience showed that zero
+intercooling at cruise allowed enough charge heating to reduce leading,
+a simple stop valve was fitted to the system. This allowed full
+intercooling for take-off and climb, which could then be turned off for
+cruise.
+
+Civil Merlins were fitted to Avro Lancastrians, Yorks and Tudors, but
+their flagship role was on the Canadair DC-4M North Star, a Douglas
+DC-4-derived design intended for Trans-Canada Air Lines (TCA). The
+North Stars used a modified Universal Power Plant installation, as the
+DC-4's nacelle bulkheads were slightly larger than the SBAC standard.
+In practice it turned out the Merlins were not ideally suited for use
+in civil operations: though the North Stars flew higher and faster than
+DC-4s, their time between engine overhauls was around 850 hours,
+roughly a third of that of commercial US radial engines. The engines
+also produced a lot of cabin noise, though this was alleviated somewhat
+by revised 'cross-over' exhausts that ducted the cabin-side stacks
+outboard, developed first by TCA and then by Rolls-Royce themselves.
+
+The extra running costs were mostly borne by Rolls-Royce. When TCA's
+managing director expressed dissatisfaction with the Merlin's
+commercial performance, Hives asked what a reasonable level of
+maintenance cost would be. Being told \$4 per engine hour, he agreed to
+service the fleet's engines for this amount. This early form of 'power
+by the hour' was initially expensive for Rolls-Royce, but by the end of
+the Merlin's life the company had learned enough to supposedly make a
+small profit at this level.
+
+What was undoubtedly true was that Rolls-Royce learned a great deal
+about the harsh realities of commercial operation in a short time.
+Hives is supposed to have said to TCA, "We didn't know the Merlin until
+you started operating it!" In the longer term, this paid off,
+TCA---later Air Canada---selecting Rolls--Royce engines for its future
+fleet: Darts in Viscounts, Tynes in Vanguards, Conways in DC-8s, and
+RB211s in TriStars.
+
+\begin{footnotesize} \raggedleft PNSAC\\
 \end{footnotesize}

### 2020-02-24 pnsac/newsletter/article_Feature3.tex
Draft updates for Feb-Mar 2020.

@@
 
-\title{The Whispering Giant}
-\author{Dave Lambeth}
+\title{A Personal Journey in Support of Homeless Veterans in Canada}
+\author{Neil Raynor}
 
@@
 
-\begin{centering}
-Twinkle, twinkle, old North Star,\\
-With wrinkled skin and creaky
-spar,\\
-For years you've always been on
-sked,\\
-As many of our force you've sped\\
-to distant lands beyond the skies,\\
-you've never failed to see the rise\\
-of distant suns or moonrise glare.\\
-You've carried us with tender care\\
--- back you came to take us home,\\
-from old Japan or France or Rome\\
-from England, Goose, and Kef.\\
-We cannot hear you.\\
-('cause we're deaf)\\
-\end{centering}
+Veterans' House: the Andy Carswell Building is Canada's first "Housing
+First" community built for homeless Veterans. This pioneering project
+specifically targets the needs of the rising number of Veterans who are
+"living rough" in Canada and combines safe housing with essential
+on-site rehabilitation and wrap around support services.
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{jim_jung_2.eps}
-   %caption of the figure 
-   \caption*{\small \em A Whispering Giant (North Star 17515) at
-     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:jim_jung}
-\end{figure}
+Who knew? How could we? Surely if we knew we'd have done something
+about it?
+
+Fact: every night in Ottawa between 50 and 85 Veterans are sleeping on
+the streets, in the parks or under bridges in Ottawa. How could that be
+happening in the 21st Century in a G7 capital?
+
+Seven years ago a group of people came together within the Multifaith
+Housing Initiative (MHI) to do something about that stat. MHI is a
+federally registered charity based in Ottawa which provides affordable
+housing: currently it has about 140 units with 400 residents scattered
+across the city. What started as a desire to provide affordable housing
+has morphed into an ethos to build communities – Veterans' House is the
+latest iteration of that approach.
+
+What MHI and its volunteer members did was: articulate the need to
+address the issue of homelessness among Veterans; identify the need for
+support for these Veterans that went beyond just housing; and together
+with its partners develop a solution that wraps it all within a
+military community initiative.
+
+The physical manifestation of MHI's work is the structure being raised
+on the former CFB Rockcliffe airbase called Veterans' House: the Andy
+Carswell Building, a 40-unit community apartment building named for a
+retired WWII RCAF Lancaster pilot who was shot down over Germany in
+1943 and who remained a POW until liberated by allied troops in 1945.
+
+Uniquely, this "Housing First" initiative will include provision for a
+range of on-site support services focused on the individual needs of
+the residents and provided by MHI's partner organizations who are
+professionals in their fields of work. And being "Housing First" means
+that residents can remain in Veterans' House for an indeterminate
+period and focus on recovery from the underlying issues that
+contributed to their homelessness, be they economic, health, mental
+health or addiction related issues. Moreover, in line with best
+practices in providing affordable housing to this at-risk group, all
+the ground floor bachelor apartments in Veterans' House will be fully
+wheelchair accessible. Again, with a community spirit in mind, the site
+will also include a gym, gardens, service-dog park, and communal
+kitchen and BBQ area for when residents want to share meals together.
+
+Together with our partners – many of them with direct connections to
+the military and Veterans – Veterans' House: the Andy Carswell Building
+will provide a "military-family" structure to residents which they will
+recognize and which research has shown many homeless Veterans miss and
+which contributes to their slide into homelessness. A fully list of our
+partners, progress to date and lots of other information is on the MHI
+Veterans' House website at
+{\normalfont\color{blue}\texttt{\url{https://www.multifaithhousing.ca/veterans-house}}}
+
+The physical structure is rising from the ground this spring. We broke
+ground on September 5th, 2019 and 96-year old Squadron Leader (retd)
+Andy Carswell laid the cornerstone on September 16th. As at February
+1st, 2020 we are up to the second floor with completion planned for
+November 2020. There's lots to do but to date we are pretty much on
+schedule despite building through the winter months.
+
+The total cost of the initiative is about \$15.4 million, including \$3
+million for the land which was transferred from the federal government
+as part of its contribution. There is a balance of \$2.3 million still
+left to raise and we are in full fund-raising mode. If you are willing
+and able there are a number of ways you can help financially: you can
+send a cash or stock-transfer donations to Veterans' House through MHI;
+organize a fund-raising event within your community; or leave a bequest
+in your Will. All donations are gratefully received and will be put to
+good use where they will make a difference.
+
+While Veterans' House: the Andy Carswell Building is the first of its
+kind our plan is to create a robust model that can be replicated across
+Canada to address the national issue of homeless Veterans.
+
+For more general info about Multifaith Housing or to make a donation go
+to: www.multifaithhousing.ca You'll find more information there
+including a recently released report on a 2019 Survey of homeless
+Veterans in the National Capital Region.
+
+To volunteer you can apply online. Go to:
+{\normalfont\color{blue}\texttt{\url{https://www.multifaithhousing.ca/volunteer}}}
+
+%\begin{centering}
+%Twinkle, twinkle, old North Star,\\
+%With wrinkled skin and creaky
+%spar,\\
+%For years you've always been on
+%sked,\\
+%As many of our force you've sped\\
+%to distant lands beyond the skies,\\
+%you've never failed to see the rise\\
+%of distant suns or moonrise glare.\\
+%You've carried us with tender care\\
+%-- back you came to take us home,\\
+%from old Japan or France or Rome\\
+%from England, Goose, and Kef.\\
+%We cannot hear you.\\
+%('cause we're deaf)\\
+%\end{centering}
+
+%\begin{figure}[htbp]
+%   \vspace{2em}
+%   \centering
+%   %name of the graphic, without the path AND in EPS format:
+%   \includegraphics[scale=0.5]{jim_jung_2.eps}
+%   %caption of the figure 
+%   \caption*{\small \em A Whispering Giant (North Star 17515) at
+%     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
+%   %label of the figure, which has to correspond to \ref{}:
+%   \label{fig:jim_jung}
+%\end{figure}
 

### 2020-02-24 pnsac/newsletter/article_MemberInterview.tex
Draft updates for Feb-Mar 2020.

--- 2018-01-25 pnsac/newsletter/article_MemberInterview.tex
@@
 \title{Our Members}
-\author{Interview with Austin J Timmins}
+\author{Interview with Garry Dupont}
 
@@
 
-My aviation experience includes 36 years service in the RCAF and nine
-years with the Canadian Aeronautics and Space Institute. In the RCAF I
-completed radio navigator training, which included communications,
-navigation and air gunner. I later qualified as an airborne
-interception navigator. My operational experience includes air
-transport, search and rescue and training. My assignments included
-Navigation Leader, Aircrew Leader, Squadron Commander, Air Command
-Navigator and Deputy Commander (COO) of the Air Transport Group. I
-logged 9500 hours, mostly on the following aircraft: Beechcraft,
-Dakota, North Star, Mitchell, Yukon, Hercules, and Boeing 707. My
-staff assignments in Europe included movement control (traffic
-management) for 1 Air Division and a tour at The NATO Defense College
-in Rome.
+I graduated from the Confederation College Aircraft Maintenance course
+in the Spring of 1975. After returning to my home town in the Red Lake
+District I gained employment with Red Lake Seaplane Service. After
+completing the course I had to apprentice for 18 months before writing
+the final exam for my Aircraft Maintenance Engineers' License. I
+successfully wrote my final exam in December 1976 and obtained my AME
+certification. In February 1977 I was hired by Perimeter Aviation in
+Winnipeg Manitoba. There I obtained my Turbine Engine and
+Pressurization Aircraft endorsements.  I started my 30 year career with
+the RCMP Air Services Branch in February 1981. My career took me to
+Edmonton Alberta, Goose Bay Labrador, Winnipeg Manitoba then finally to
+Ottawa. I finished my career as Quality Assurance Manager.
 
-On retirement from the RCAF I participated in a Transport
-Canada A Base review of several airports. I then secured full
-employment with Canadian Aeronautics and Space Institute as Executive
-Director, retiring in 1995.
+2. \textit{How long have you been involved with Project North Star and how and
+why did you get involved?}
 
-2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}
-
-My involvement with efforts to conserve North Star 17515 predates
-Project North Star. Ottawa area members of the 426 Thunderbird Squadron
-Association, mostly retired North Star aircrew, met with the National
-Aviation Museum staff to express their concern about the state of the
-North Star and proposed volunteer involvement in a preservation effort.
-These expressions of concern and interest led to a meeting between the
-Association representatives, where General Adamson presented a proposal
-for the restoration of the aircraft. It was rejected but assurance was
-given that the North Star would be first priority when the new hangar
-was completed Also a North Star Trust Fund would be established. Robert
-Holmgren, an observer at the meeting, then proceeded to develop a
-proposal for volunteers to restore the North Star.
+I attended the very first meeting held for Project North Star. This was
+at my wife's suggestion that it would be a good retirement pastime.
 
@@
 
-I have been involved from the beginning of efforts to establish PNS,
-first as a member of a Steering Committee, led by Robert Holmgren,
-later, Vice President and then President of PNSAC. My contacts in
-aviation provided much needed advice and support, including access to
-initial funding for PNSAC.
+While I was employed I worked an evening shift every 4 weeks so I would
+come in to CASM in the morning for a few hours. Upon my retirement Mike
+Irving had me work in the engine shop, which I have been doing since my
+retirement. I also have been on the board of directors for a number of
+years.
 
@@
 
-The highlight of my involvement has to be the establishment of PNS and
-PNSAC.
+The biggest highlight for me was installing number 2 engine. It was
+about half completed when I started in the engine shop. This is the
+largest engine I have ever installed in my entire career.
 
@@
 
-PNS is now in its 16th year. It has been a work in progress with many
-bumps along the way. I spent a lot of time and effort, along with other
-members of the Executive, evening out these bumps. Establishing and
-maintaining effective channels of communication with CASM and CSTM
-required constant attention. Much effort was directed to fund raising,
-publicity, and, finding volunteers. In the beginning, every date
-brought a new challenge.
+One of the most challenging parts was learning how the engine worked
+with all its systems. I found a RCAF training manual in the office and
+took it home. Over several months of reading I started to understand
+what made this monster of old technology tick. Another challenge was
+learning the British Hardware system and tools. A further challenge was
+reading  the old manuals which were written before they were
+standardized by the ATA chapter system which all aircraft manuals
+follow today. Learning the art of conservation was also a  challenge
+but was made easier by the guidance of Mike Irving.
 

### 2020-02-24 pnsac/newsletter/article_PresReport.tex
Draft updates for Feb-Mar 2020.

--- 2018-07-19 pnsac/newsletter/article_PresReport.tex
@@
 
-These are exciting times to be a member of the Project North Star Association.
-We  have recently taken part in the 2018 Doors Open Ottawa. This is a major
-event for the Museum and PNSAC volunteers were able to make a significant
-contribution to the success of the two open days. Doors Open has become second
-only to Canada Day in its importance to the Museum. (more on Open Doors 2018
-later in this issue).
+Since the last issue of the NStar Chronicle we have taken part in the
+70th Anniversary celebrations of the first non-stop transcontinental
+flight across Canada which was flown by a North Star. We partnered with
+the Museum for this event and were very pleased to welcome
+representatives from the RCAF who joined us for the day. After the
+formal part of the celebration, when impressive plaques of the aircraft
+made by our member Chris McGuffin, had been presented to the RCAF and
+the Museum Director General, Chris Kitzan, everybody moved to the
+Reserve Hangar and the aircraft was opened to our association members,
+guests and in the afternoon to 90 members of the public who had
+pre-booked a tour. R\'{e}jean(R\'{e}j)Demers, the Museum Special Project
+Manager, guided the tours assisted by several of our members. Towards
+the end of the tour session someone in the main Museum activated the
+fire alarm. The Museum was promptly evacuated, and the public was
+recommended to go to the Reserve  Hangar until the fire issue had been
+dealt with (it turned out to be a false alarm). Quite unexpectedly R\'{e}j
+and the guides found themselves surrounded by many Museum visitors who
+had not signed up for the North Star tour. Unperturbed by this event,
+Réj and the guides welcomed the unscheduled visitors and gave them an
+impromptu tour of the aircraft. Both our treasurer and our membership
+secretary were pleased with the days' result!
 
-A few days before Canada Day the aircraft was brought outside. Unfortunately,
-the day itself was extremely hot and the number of visitors was well down from
-previous years. Nonetheless our volunteers did a yeoman service in talking to
-the public about our restoration work.
+As I write this the days are getting longer and R\'{e}j is planning work to
+be done on the aircraft when it is moved outside onto the ramp in the
+spring. Each year we do work on the aircraft which cannot be done
+inside the Reserve Hangar during the winter. We also prepare the
+aircraft for display when, once again, we open the aircraft to enable
+the public to view progress made in the last year.
 
-Although showing the aircraft to the public is very important to us, our main
-purpose is to restore the North Star to museum condition. Much of our work is
-detailed, painstaking and unspectacular such as restoring a small part of the
-aircraft which can take many days or weeks. At other times making progress is
-very visible. We are in such a situation now.
+This year we are planning two major display events. The first will be
+on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
+for our members to meet the public, many of whom have an interest in
+aviation. Our second main event, on Canada Day, is enjoyable for all
+the members who sign up to be part of the team meeting the public.
+There are always large crowds, many of them family groups, If any of
+you reading this article would be interested in joining our members for
+either of these events please contact me (my co-ordinates are on the
+back page of this issue of the Chronicle)
 
-At the end of May we had one of our regular meetings with Chris Kitzan (Director
-General of CASM) together with Neil Raynor (VP of PNSAC), Cedric St-Amour (CASM
-Volunteer Coordinator) and Reg Demers (Our CASM supervisor There was a full
-exchange of ideas, suggestions for bringing  the public into the Museum and ways
-in which PNSAC could be involved in these initiatives and events. The meeting
-ended with me having a full sheet of notes covering future planning, major
-restoration work and how we can make sure we recruit younger people both as
-members of PNSAC and where possible active Restoration Volunteers.
+Shortly before Christmas a major milestone, in the  restoration of the
+North Star, was reached. Work on Merlin engine \#4 was completed and it
+was dragged on its cradle by the Museum forklift truck to the
+Restoration hangar for later mounting on the aircraft to join the other
+three restored engines. Chris McGuffin put some video clips onto our
+Facebook page of the engine being dragged through the Museum parking
+lot and into the hangar. Watch for more of our Facebook messages since
+Chris frequently posts new items.
 
-A meeting like the one above only happens when an organization is thriving, and
-everybody is excited about the future. During the last 15 years PNSAC has
-demonstrated that it is capable delivering very high-quality restoration work.
-We have already accumulated over 80,000 hours of work and during this time we
-have become a trusted partner of the Museum. Our website is now operational
-again and in the next few weeks we expect to be able to update it and make it
-more relevant to where we are with the project now. One of the major initiatives
-taking place is to develop a diagrammatic history of the work done on the North
-Star restoration and a timeline for future work on the aircraft.
+In the late fall we welcomed the appointment of Erin Secord as the
+Museum's new Manager Conservation Services. She is responsible also for
+conservation of the collections at the Science and Tech Museum. Erin
+has now been able to join us at two of our Members' meetings.
 
-In closing, I once again urge Association members to become active volunteers.
-There are many volunteering opportunities both for those with major aviation
-experience to those who can only volunteer with us on occasions like Canada Day
-or can provide help for our many behind-the-scenes activities.
+Finally, we are now working to update the Memorandum of Understanding
+between PNSAC and the Museum.  Much has changed since the original
+agreement was signed in May 2006.
 
-Our next Members' Meeting is scheduled for
-Saturday, September 29, 2018. Do reserve this date in your calendar if you can
-be in the Ottawa area on that day. The meeting will be held at the Canada
-Aviation and Space Museum. More details will follow later.
+Our next Members' Meeting is scheduled for Saturday, June 20, 2020
+which will also be our AGM. Do reserve this date in your calendar if
+you can be in the Ottawa area on that day. The meeting will be held at
+the Canada Aviation and Space Museum. More details will follow later.
 

### 2020-02-24 pnsac/newsletter/article_ProgReport.tex
Draft updates for Feb-Mar 2020.

@@
 %\subtitle{PNS 2016 Status Update}
-\author{R\'{e}jean Demers}
-\author{Restoration Project Manager}
+\author{Bruce Gemmill, R\'{e}jean Demers, and Garry Dupont}
+%\author{Restoration Project Manager}
 
@@
 
-\section{Zen and the Art of Aircraft Restoration.}
+%\section{Zen and the Art of Aircraft Restoration.}
 
-Through my daily experience over the past two years, working with volunteers, I
-have asked myself the following question: What makes a volunteer dedicated to a
-project? The answer is not a straightforward one. This article explores the
-subject, much like Pirsig did in his book about metaphysical motorcycles; Zen
-and the Art of Motorcycle Maintenance.
+The past year was a very busy one for the North Star restoration crew. 
+In the Engine Shop, under the careful watch of Crew Chief Garry Dupont,
+members Charles Baril and Richard Lodge worked to fit the final pieces
+on engine 4 including; all the cowl frames and panels, main and
+auxiliary radiator flaps under the engine, the engine air intake duct
+and associated actuators. All the hard work in setting up sheet metal
+repairs to the cowl panels paid off.  However, fitting the large steel
+exhaust panels around the 12 exhaust stacks was particularly
+challenging. It's no longer a secret that ratchet straps and wooden
+2"x4" blocks have become part of the standard North Star tool kit.
 
-We strive for quality. As restoration work continues on the North Star, we get
-absorbed into the weight and breadth of the subject matter. This is the kind of
-project that demands the level of commitment found in monasteries, dojos and
-Paratrooper units. Commitment comes from dedication and reward. Quality is
-obtained through application of commitment, producing results which may be
-enjoyed. The following effort in writing has been made to define such elements,
-those necessary to sustain the project.
+While that work was underway, Jacques Roy, Peter Scartozzi and Charles
+Baril were busy stripping equipment, pipes, hose and cables from inside
+the nacelle where engine 4 will soon be installed.  Will Assad and Neil
+Raynor also helped with this. After equipment removal, many hours were
+spent carefully cleaning and removing oil and corrosion prior to
+painting.  In some areas a small bead blasting gun was used to remove
+rust from nuts and bolt heads that could not be taken out for cleaning.
 
-A voracious appetite for detail. Precise facts and data build foundations.
-Foundations of information that erode over time. Lost to floods, fires, or
-worst; thieves, vandals and neglect. Information is our firmament that keeps us
-grounded in our work. Daily worksheets are filled with steps taken to satisfy a
-condition. Drawings are made to translate reality to paper. These words,
-drawings and tables form hard to translate time lines. Coloured bars overtaking
-each other, running in thousand hour blocks of volunteer contributions. These
-life hours, cannot be overlooked, as a restoration project advances through the
-years.
+Most notably were efforts by Charles Baril, who spent many months
+inside the nacelle; cleaning, inspecting and masking off for painting.
+The work took so long that special heating ducts and insulation needed
+to be positioned around the nacelle to keep it warm enough for Réjean
+Demers and Bruce Gemmill to apply a coat of primer and silver paint.
+(insert photo of North Star with orange tarp). Soon after the paint was
+applied, the North Star was moved inside for the winter. This occurred
+November 8th, just 3 days before our first snow storm of the winter.
 
-Inorganic compounds risk distraction. Focus is important when avoiding the
-ethereal draw of solvents. A volunteer scrubber, washing away years of residue,
-layered upon a finish that once gleamed with pride. Slowing the return to dust,
-so feared and fought by the living. A process to follow and adhere to. The
-methodical dismantling and reconstruction of an object. Return to form, that
-which was almost indiscernible. The result of all his work is not merely a
-shiny piece of aluminium panelling; He has travelled through time.
+While all that work was underway, the Airframe Crew was busy on the
+structures work.  Peter Trobridge, Michel Cote and Robert Desjardins
+fabricated and installed several replacement parts for main cabin
+airframe assemblies that had suffered from years of corrosion.  The
+crew accomplished a complete review of necessary repairs from front to
+back of the main cabin. A hole was patched near the door to the crew
+area, at forward bulkhead (station) 301. This offered Chris McGuffin an
+intro to sheet metal repairs. Similar corrosion damage around the
+windows required parts shaped on wooden jigs made in the shop. Through
+careful planning and creative solutions, repairs were completed at STA:
+360, and 381. New repairs carried out at starboard STA: 460, 480, 501,
+541, 560, 621 and currently, floor repairs in the loading area near the
+cargo doors.
 
-More than just an object, we look to a hidden past. Beyond these material
-trappings, lay the history of Canadair North Star 17515 and her sisters.
-Examples of service are easily found in books and photos. What proves a
-challenge is explaining how a bumblebee becomes a preserved specimen in an
-engine, how graffiti made during heavy maintenance is the only evidence of a
-man's past. Where the dirt under floorboards, came from countries that no
-longer bear the name she visited.
+The Carpentry Crew, led by Chris McGuffin measured, cut and trimmed new
+plywood panels for the interior walls of the main cabin.  These were
+painted green to match the original colour scheme, then set aside until
+for later installation.  Next came the floor panels.  Richard Houle and
+John Makadi assessed these, according to the amount of damage caused by
+wear and tear, as well as leaks around windows and doors.  Some panels
+were salvaged with a few minor repairs, then painted. These are now
+ready for installation.  Several others were so badly damaged that
+completely new panels needed to be made. (Photo of wall and floor
+panel)
 
-The static, prevents complete appreciation. She's more than 45,000 lbs. of
-metal. Superficial understanding can only consider her 39 years an outdoor
-relic. Likewise, so would a skin deep treatment of a last remaining artefact of
-this type. What impact she had before the stillness of the seasons is not lost
-to those who look further.
+Finally, work started on the third group of panels.  These floor panels
+had significant damage to one or two edges.  The damaged areas need to
+be cut away and a new piece added by bevelling the adjoining pieces
+using what is called a "scarf joint" so the adhesive bond is strong.  A
+special jig was constructed to do this work, allowing an electric
+router to run at an angle to produce the joint. A Resorcinol
+Formaldehyde adhesive will assure a long-lasting repair.
 
-A biological experience in personal evolution. The volunteer has developed new
-muscles and sinew adapted to producing polished surfaces. An eye for
-microtexture and grain direction has developed. Sympathetic scars are
-accumulated through careful little sacrifices. Navigating the disruptions of
-capricious human needs such as rest, nourishment and leisure. Through the
-application of labour, a piece of history is preserved.
+Many of the volunteers got to work on the large collection of
+accessories that had been removed from nacelle \#4.  This included
+pumps, plumbing, fluid and air lines, control cables, bell cranks and
+pulleys, electrical harnesses, valves and filter canisters. Each needed
+assessment, cleaning, repair, polishing or painting as well as
+documentation and photographs of the work undertaken. (photo of fuel
+filter, junction box).
 
-Disappearing horizons. Meditative expanses of aluminum, exhibiting varieties of
-corrosion rarely encountered in a lifetime's career. Cavernous interior spaces,
-where cargo of all sorts was bound for ports beyond. Flight control surfaces
-that have pressed through all shades of dusk, emerging in the dawn of a new
-day. Now removed from the ship, these weathered, beaten, bruised remnants of a
-glorious past await restoration. A new horizon forms, some many years away,
-known to those who are on-board for the long run.
+Phil Chrysler and Brian Cole went to work on the many fairings and
+access panels that also required cleaning and polishing to be ready to
+be placed back on the aircraft. A special mention here for Ted Devey,
+who, before retiring from the shop had his hand on a few cowl panels
+and nacelle components.
 
-Healthy obsessions. Volunteer restoration technicians take home an experience.
-They take to heart the results of their labour, in forming a product that is
-also intangible. The mind has strengthened, forming resolves that assist,
-understand and process the elements of restoration work. As historians,
-fanatics and unwitting library wanderers soon discover; The story of air
-transport, in its beginning, is far from just a few obscure paragraphs in
-Canada's ongoing history. It lives through those people who continue writing
-it, preserving the physical, extending beyond.
+Several special projects were also undertaken during the year.  John
+Makadi removed the nose radome to determine what radar equipment had
+been removed, in case we are able to find a source for missing
+equipment.  We removed the upper rotating navigation beacon to fix a
+persistent leak, this necessitated a twin tower scaffold assembly to
+safely reach the top of the fuselage. Neil Raynor and Will Assad
+overhauled a propeller stand the Museum had not used for many years. 
+It got a fresh coat of paint and new wheels and tires, and is now
+holding two North Star propellers, until they can be installed on the
+appropriate engine. (photo of prop stand, two propellers). Michael Hope
+assisted in removal of Prop. \#1 and assembly of soon-to-be-installed
+Prop. \#4. Phil Chrysler and Brian Cole fabricated cables and special
+fittings to secure the landing gear so the aircraft could be safely
+displayed. Other duties  such as; an inventory of replaced / restored
+and spare components were carried out.  All of these were packed and
+stored in a new location of the Reserve Hangar.
+
+Number \#4 engine has now  been moved to the Reserve Hangar. The crew
+are prepping the oil tank for clear coat then it will be assembled and
+ready to install. Once this is completed  the plan is to carry out some
+deep cleaning and renovations in the shop starting  by  relocating the
+varsol cleaning tank. The paint shaker has been repaired and the base
+has been painted ready for assembly. The paint shaker will be relocated
+where the varsol tank was. The old roll away tool box has been replaced
+by a new Snap-On unit. There have been lots of comments on the orange
+colour  but the price was right.
+
+Currently, work is on-going to complete the rear (main) cargo door and
+quilted liner installation. Nacelle \#4 installation of restored
+components is advancing in small steps, due to the complexity of
+interrelated systems and locations. Scaffolding is in place, over the
+forward fuselage area above the cockpit to assess air-frame
+requirements and treatment methods to be used on exterior hull
+surfaces. Restoration records are also being reviewed by Conservation
+Special Project Manager Réjean Demers. The Work Order / Task Card
+format is lending well to Museum information uploading tools. Efforts
+are underway to recruit two to four new shop volunteers, and R\'{e}j
+invites any new applicants referred by PNSAC to apply.
 

### 2020-03-08 pnsac/newsletter/article_EditorsNotes.tex
PNSAC newsletter--Mar.2020.

@@
 the necessary permission from Air Force Magazine to publish the
-"Diversion to Alert" article.  Allan Bowes, another member, suggested
-that we should try to get the "Civil Merlins" article into an edition
-of the Chronicle.  Following his suggestion we contacted Ben Dunnell,
-the editor of Aeroplane Magazine. Through Ben's offices we were able to
-secure permission to reprint this article.  The "Civil Merlins" was
-part of a comprehensive 17 page Aeroplane Database article on the
-Merlin. If any members wish to read the entire article it was published
-in the July 2017 issue.
+"Diversion to Alert" article.
+
+%Allan Bowes, another member, suggested
+%that we should try to get the "Civil Merlins" article into an edition
+%of the Chronicle.  Following his suggestion we contacted Ben Dunnell,
+%the editor of Aeroplane Magazine. Through Ben's offices we were able to
+%secure permission to reprint this article.  The "Civil Merlins" was
+%part of a comprehensive 17 page Aeroplane Database article on the
+%Merlin. If any members wish to read the entire article it was published
+%in the July 2017 issue.
 
 In addition thanks go out to the other contributors, namely Richard
-Lodge, Garry Dupont,  Neil Raynor and Bruce Gemmill. Bruce took the
-lead in putting together the Progress Report.  Finally, I would like to
-thank Drew Hodge our stalwart publisher of this edition. We are always
-on the lookout for content for the Chronicle so please do not hesitate
-to bring forward any ideas that you might have. Enjoy.
+Lodge, Garry Dupont,  Neil Raynor, and R\'{e}jean Demers. A special thanks
+to Bruce Gemmill who took the lead in putting together the progress report,
+which can be found in the Conservator's Corner section.
+
+Finally, I would like to thank Drew Hodge our stalwart publisher of this
+edition. We do appreciate receiving feedback from our readers including
+comments on the articles or suggestions for new articles. We are always on the
+lookout for content for the Chronicle so please do not hesitate to bring
+forward any ideas that you might have. As noted in the Conservator's Corner
+article the restoration of engine number 4 has been completed. Our next
+edition will focus on the engine restoration work so if you have any material
+or comments specific to that aspect of the restoration please let us know.
+Enjoy.
 

### 2020-03-08 pnsac/newsletter/article_Feature1.tex
PNSAC newsletter--Mar.2020.

@@
 Immediately after New Year 1960, while based at RCAF Stn Trenton, and
-flying North Stars for 426(T)Sqn., I was assigned to be the First
+flying North Stars for 426(T) Sqn., I was assigned to be the First
 Officer on a flight to pick up an entertainment troupe of 30 Passengers
@@
 Manitoba sectors of the Mid Canada Detection Line.
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{alet-PC-155.png}
+   \caption*{\small \em The venerable Canadair North Star provided yeoman service to the RCAF---RCAF Photo PC-155}
+   \label{fig:wall-two}
+\end{figure}
 
@@
 
-Our 426Sqn crew consisted of Captain – F/O H.(Hank) Zbesheski, First
-Officer – F/O M.G.(Mac or Mo) Morrison, Navigator – F/O Roger Fridel,
-Radio Officer – F/O Peter Dragovich, Flt Engineer – Sgt Miller, Flt
-Engineer – Sgt Sopaz, Trans Tech – LAC Hall.
+Our 426Sqn crew consisted of Captain---F/O H.(Hank) Zbesheski, First
+Officer---F/O M.G.(Mac or Mo) Morrison, Navigator---F/O Roger Fridel,
+Radio Officer---F/O Peter Dragovich, Flt Engineer---Sgt Miller, Flt
+Engineer---Sgt Sopaz, Trans Tech---LAC Hall.
 
@@
 account for.\\
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{alert-PCP-83-191_cropped.png}
+   \caption*{\small \em North Stars were routinely used in harsh Arctic conditions---RCAF Photo PCP-83-191}
+   \label{fig:wall-two}
+\end{figure}
 
@@
 
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{alert-PL-54044.png}
+   \caption*{\small \em One of the North Stars from the Morrison story---aircraft \#17512---RCAF Photo PL-54044}
+   \label{fig:wall-two}
+\end{figure}
+
 The runway at Resolute Bay is almost true north and south. The
@@
 
+%\begin{figure}[httb]
+%   \vspace{2em}
+%   \centering
+%   \includegraphics [scale=0.5]{alert-PL-54068.png}
+%   \caption*{\small \em The beauty of Canada's Arctic is framed by the nose of a North Star---RCAF Photo PL-54068}
+%   \label{fig:wall-two}
+%\end{figure}
+
 During a rehash of the journey a few days later, it was disclosed to us
@@
 Churchill to Montreal.
+
+\end{multicols}
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics {alert-PL-54068.png}
+   \caption*{\small \em The beauty of Canada's Arctic is framed by the nose of a North Star---RCAF Photo PL-54068}
+   \label{fig:wall-two}
+\end{figure}
 
@@
 
+\begin{multicols}{2}
+
 % End of text.

### 2020-03-08 pnsac/newsletter/article_Feature2.tex
PNSAC newsletter--Mar.2020.

@@
 Command's Avro Yorks. Long-range transport operation entailed running
-at relatively low cruise powers for long periods — under these
+at relatively low cruise powers for long periods---under these
 conditions the lower cylinder head temperatures caused deposits of lead
@@
 modifications, though only the Merlin 501 included a charge heater.
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{merlin-1-feb-2020.jpg}
+   \caption*{\small \em Rolls-Royce Merlin Engine}
+   \label{fig:wall-two}
+\end{figure}
 
@@
 
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics[scale=0.5]{nstar-tca-feb-2020.jpg}
+   \caption*{\small \em TCA Canadair DC-4M North Star}
+   \label{fig:wall-two}
+\end{figure}
+
 The extra running costs were mostly borne by Rolls-Royce. When TCA's

### 2020-03-08 pnsac/newsletter/article_Feature3.tex
PNSAC newsletter--Mar.2020.

@@
 across the city. What started as a desire to provide affordable housing
-has morphed into an ethos to build communities – Veterans' House is the
+has morphed into an ethos to build communities---Veterans' House is the
 latest iteration of that approach.
@@
 
-Together with our partners – many of them with direct connections to
-the military and Veterans – Veterans' House: the Andy Carswell Building
+Together with our partners---many of them with direct connections to
+the military and Veterans---Veterans' House: the Andy Carswell Building
 will provide a "military-family" structure to residents which they will

### 2020-03-08 pnsac/newsletter/article_PresReport.tex
PNSAC newsletter--Mar.2020.

@@
 
+The notes below were written in the middle of February and since that
+time a sad event has taken place in that our old friend and Life Member
+of PNSAC, Ted Devey, has passed away. We have only a few Life Members.
+Such memberships are granted only in circumstances where a member has
+made a special and exceptional contribution to the North Star
+restoration project. Ted Devey was one such member. At 94 years old,
+Ted was our oldest member and the oldest Museum restoration volunteer.
+We will miss Ted, his continuous supply of amusing anecdotes and of
+course his dedication to the North Star restoration project. Ted was
+the subject of "Our Members" section of the January 2019 edition of the
+Chronicle which is available on the Association's website.
+
 Since the last issue of the NStar Chronicle we have taken part in the
 70th Anniversary celebrations of the first non-stop transcontinental
-flight across Canada which was flown by a North Star. We partnered with
-the Museum for this event and were very pleased to welcome
-representatives from the RCAF who joined us for the day. After the
+flight across Canada which was flown by a North Star. Each year work is
+performed on the aircraft which cannot be done outside the Reserve Hangar
+during the winter. The aircraft is also prepared for display to enable the
+public to view the progress made in the last year. After the
 formal part of the celebration, when impressive plaques of the aircraft
@@
 guests and in the afternoon to 90 members of the public who had
-pre-booked a tour. R\'{e}jean(R\'{e}j)Demers, the Museum Special Project
+pre-booked a tour. R\'{e}jean (R\'{e}j) Demers, the Museum Special Project
 Manager, guided the tours assisted by several of our members. Towards
@@
 had not signed up for the North Star tour. Unperturbed by this event,
-Réj and the guides welcomed the unscheduled visitors and gave them an
+R\'{e}j and the guides welcomed the unscheduled visitors and gave them an
 impromptu tour of the aircraft. Both our treasurer and our membership
 secretary were pleased with the days' result!
+
+% We partnered with
+%the Museum for this event and were very pleased to welcome
+%representatives from the RCAF who joined us for the day.
 
@@
 
-This year we are planning two major display events. The first will be
+This year the Museum is planning two major display events. The first will be
 on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
@@
 was dragged on its cradle by the Museum forklift truck to the
-Restoration hangar for later mounting on the aircraft to join the other
+Reserve hangar for later mounting on the aircraft to join the other
 three restored engines. Chris McGuffin put some video clips onto our

### 2020-03-08 pnsac/newsletter/article_ProgReport.tex
PNSAC newsletter--Mar.2020.

@@
 The work took so long that special heating ducts and insulation needed
-to be positioned around the nacelle to keep it warm enough for Réjean
+to be positioned around the nacelle to keep it warm enough for R\'{e}jean
 Demers and Bruce Gemmill to apply a coat of primer and silver paint.
-(insert photo of North Star with orange tarp). Soon after the paint was
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{nacelle-4-paining-outside-feb-2020.png}
+   \caption*{\small \em Painting nacelle \#4 outside}
+   \label{fig:wall-two}
+\end{figure}
+
+Soon after the paint was
 applied, the North Star was moved inside for the winter. This occurred
@@
 
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{New-Floor-Wall-Panels-feb-2020.png}
+   \caption*{\small \em New floor and wall panels}
+   \label{fig:wall-two}
+\end{figure}
+
 The Carpentry Crew, led by Chris McGuffin measured, cut and trimmed new
@@
 ready for installation.  Several others were so badly damaged that
-completely new panels needed to be made. (Photo of wall and floor
-panel)
+completely new panels needed to be made.
 
@@
 assessment, cleaning, repair, polishing or painting as well as
-documentation and photographs of the work undertaken. (photo of fuel
-filter, junction box).
-
-Phil Chrysler and Brian Cole went to work on the many fairings and
+documentation and photographs of the work undertaken. Phil Chrysler and 
+Brian Cole went to work on the many fairings and
 access panels that also required cleaning and polishing to be ready to
-be placed back on the aircraft. A special mention here for Ted Devey,
-who, before retiring from the shop had his hand on a few cowl panels
-and nacelle components.
+be placed back on the aircraft. A special mention should be made for the late
+Ted Devey who had had his hand on a few cowl panels and nacelle components.
 
@@
 holding two North Star propellers, until they can be installed on the
-appropriate engine. (photo of prop stand, two propellers). Michael Hope
+appropriate engine.
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{prop-stand.png}
+   \caption*{\small \em Prop stand}
+   \label{fig:wall-two}
+\end{figure}
+
+Michael Hope
 assisted in removal of Prop. \#1 and assembly of soon-to-be-installed
@@
 stored in a new location of the Reserve Hangar.
+
+\begin{figure}[httb]
+   \vspace{2em}
+   \centering
+   \includegraphics [scale=0.5]{eng-4-pre-move-feb-2020.png}
+   \caption*{\small \em Engine \#4---before it was moved to the Reserve Hangar}
+   \label{fig:wall-two}
+\end{figure}
 
@@
 surfaces. Restoration records are also being reviewed by Conservation
-Special Project Manager Réjean Demers. The Work Order / Task Card
+Special Project Manager R\'{e}jean Demers. The Work Order / Task Card
 format is lending well to Museum information uploading tools. Efforts

### 2020-09-13 pnsac/newsletter/article_Feature1.tex
First draft for Sep 2020 issue.

--- 2019-01-13 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{Volunteers Assure a Bright Future for Classic Plane}
+\title{North Star Propeller Restorations}
 %\subtitle{Part 4}
-\author{Randall Denley}
+\author{Harry Hope, with assistance from Michael Hope.}
 
@@
 
-\end{multicols}
+\textit{Hope Aero has played a major role in the restoration of the North Star.
+The following article provides some background about the company and its work
+on the North Star’s propellers. }\\
 
-\begin{figure*}[htbp]
-	\vspace{2em}
-	\centering
-	%name of the graphic, without the path AND in EPS format:
-	\includegraphics[scale=0.8]{holmgren-timmin-scaled.png}
-	%caption of the figure 
-	\caption*{\small \em TIm Timmins, left, and Robert Holmgren are forming a
-group of skilled volunteers to restore a 55-year-old North Star, the first
-Canadian-made plane capable of transcontinental flight.}
-	%label of the figure, which has to correspond to \ref{}:
-	\label{fig:timmins-holmgren}
-\end{figure*}
+Hope Aero was formally government approved to certify aircraft propellers in
+1992, at their facility in Mississauga, Ontario. Harry Hope however, already
+had 40 years of propeller overhaul experience on many varieties that included
+Hamilton Standard propellers, which were quite common, as well as Hartzell,
+McCauley, and Sensenich. Harry's first restoration to flying condition, was the
+propeller for the Lysander, which was restored by the Air Force in Winnipeg for
+their Centennial project in 1967. This Lysander is now displayed at CASM. Since
+that time, they have restored several propellers, some to flying condition,
+three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
+(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
+display at the Aviation museum in Boda, Norway.
 
-\begin{multicols}{2}
+ Since that time, their capabilities have added– MT, Dowty, Hoffman, and
+Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
+their customer base expanded to include different types of aircraft, the
+requirement for them to service more types became evident. They kept up with
+the required technology.
 
-\textit{This article was published in the Ottawa Citizen on February 1, 2003
-under the byline of Randall Denley. It was brought to our attention recently by
-Drew Hodge. The article is republished with the express permission of the
-Ottawa Citizen, a division of Postmedia Network Inc. (ed.).
-}\\
+Along the way, they also expanded into aircraft Wheel and Brake overhaul,
+sales, service and on wing dynamic balancing of fixed wing and Helicopters.
+Harry retired in 2001, but the third generation is now capably looking after
+the business.
 
-The world's last classic Canadair North Star sits, forlorn, on the tarmac at
-the Canada Aviation Museum. It's been parked there since 1966, along with half
-a dozen other orphans of the museum, planes that it has no money to restore and
-that are too big to bring inside the crowded building.
+In the spring of 2006, Hope Aero took on the task of restoring the four North
+Star propellers, for the "North Star Restoration" project. The first challenge
+was to remove the propellers from the engines as neither the restoration crew
+nor Hope Aero had any recent experience removing this type of propeller.
+Michael Hope made the trip to Ottawa to assist with this operation and to help
+prevent any damage, which turned out to be exceedingly difficult due to the
+corrosion, starting with the spinner and then the propeller. Once the propeller
+was removed the blades were removed for shipping to Hope Aero.
 
-The paint on the North Star is faded and peeling.Some gaps in the wings are
-closed with wire to keep out the birds that have made it home for years.The
-engines are protected with pieces of plywood. The interior requires a complete
-renovation.
+The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
+Aero in 2006. It was completely disassembled, cleaned and inspected. The
+external parts were very badly corroded, with most damage being to the aluminum
+blades and dome shells, and rust to the steel hub and tail shaft, which was
+likely caused by the outside storage environment. The inside parts were in
+relatively good condition. It was established, that both technical and visual
+effort would be required to restore the propeller to an acceptable display
+condition.
 
-It's a shame to see an important part of Canada's aviation history slide into
-this kind of ruin, Thanks to the efforts of two Ottawa men, however, the North
-Star finally has a bright future.
+The restoration work commenced with the blades, the alcohol de-icer feed shoes,
+troughs and paint being removed. The corrosion of the blade surfaces was such
+that removal by grinding was abandoned, as the inner half of the blades are
+shot peened and grinding would remove this and the authenticity. It was decided
+to media blast the complete airfoil surfaces with an aluminum oxide media, thus
+removing the surface corrosion but leaving a good surface for paint to be
+adhered to; an aviation grade grey paint was applied with yellow stripes at the
+tips. New alcohol feed shoes were installed, however the troughs, which direct
+the alcohol to the shoes, were not available. As the troughs are hidden inside
+the spinner, it was determined that they could be omitted.
+The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
+tubes, and dome retaining nut, were extremely rusty. They were glass bead
+blasted to remove the rust; this did not remove the heavy pitting but did leave
+a surface that could be cadmium plated.  After plating, a clear coat material
+was sprayed on these parts to improve the resistance to rust. The internal
+surfaces were in good condition, and they were well oiled.
 
-Robert Holmgren and Tim Timmins are forming a group of skilled volunteers to
-restore the North Star. This kind of volunteer effort will be a first for the
-museum, says director general Anthony Smyth, who is supporting the effort,
-along with aviation companies including Air Canada and Boeing.
+A replacement dome shell in relatively good condition was found, requiring only
+to be sanded, anodized, and be reinstalled with new seals.
 
-The museum is showing commendable flexibility because the traditional view in
-the museum world is that restoration is work that can only be done by trained
-experts. Unfortunately, the museum can't afford that kind of work, and that's
-one of the reasons the plane has been sitting there for more than 35 years.
+The propeller was then completely assembled to confirm that all the refurbished
+parts fit together. The blades were then removed to facilitate being shipping
+back to the museum.
 
-This is no small undertaking. The 55-year-old North Star 1-ST is a sizeable
-plane, nearly 30 metres long and with a wing span of 35 metres. Restoring it
-will be a yearlong process that would cost \$1 million if it were done by paid
-staff.
-
-The North Star is an important plane in Canada's aviation history, the first
-Canadian-,made plane capable of transcontinental flight. It's considered the
-plane that launched the postwar Canadian aviation industry; 70 of them were
-built. Brought into service after the Second World War, North Stars carried
-passengers for several airlines and troops for the Canadian military.  North
-Stars served to airlift Canadian troops to Korea and provided transportation
-for VIPs visiting Canada. The plane at the museum belonged to the RCAF.
-
-Holmgren is a retired Air Canada maintenance expert, and Timmins is a former
-RCAF navigator who flew on the North Stars. Timmins's former squadron had taken
-an interest in the North Star, and Holmgren, who was giving his time as a
-volunteer at the museum, pushed it forward.
-
-"We'd like to see it brought back to something like its former self," Holmgren
-says.
-
-A temporary structure, perhaps a bubble like the one over the playing field at
-Lansdowne Park, will be necessary before the restoration can begin. The plane
-is too large to fit inside the museum's existing building and also too big to
-be easily transported elsewhere.
-
-Holmgren and Timmins need about 200 volunteers to make the project a success.
-Holmgren has approached Air Canada, Bombardier and Boeing to get the help of
-past and present employees. About 40 volunteers have already come forward.
-
-The key need is skilled craftsmen with up-to-date large aircraft experience and
-who live in or near Ottawa.
-
-Some people who can come in almost every day to run the project are required. 
-There is also a demand for people to do record-keeping and research.
-
-The group is hoping to attract national attention for the project, both to
-provide volunteer expertise and to help raise money.  This is a plan to help a
-national museum, after all.  Smyth says the museum will be involved with a
-fundraising campaign, but it's early yet to say how much is required. It
-depends partly on how much time and material are donated by corporations. A
-full-time salaried project manager will be hired by the museum.
-
-The project will take a conservation approach, retaining as much of the
-original material as possible, Smyth says. The restored plane will be
-structurally complete and faithful to the original. Returning the plane to
-flyable condition is too expensive to be feasible and the museum wouldn't want
-to take a chance flying the last remaining version of such an important plane.
-
-Restoring the North Star is just the beginning of what Holmgren and his
-associates hope to accomplish. The museum has another half dozen planes outside
-that require restoration. Restoring them all will take nine or ten years. The
-Museum's new addition, which it hopes to have open by December of this year,
-will be large enough to finally bring the North Star indoors.
-
-%If you want to help with the project, contact Robert Holmgren at 748-5972 or
-%robertholmgren@rogers.com. The group also has a Web site,
-%www.projectnorthstar.ca.
-
-%\address{Web site:
-%{\normalfont\color{blue}\texttt{\url{http://www.projectnorthstar.ca}}}\\
-%	General enquiries:\email{info@projectnorthstar.ca}}
-%
-
-%\begin{quotation}
-%	\textit{You have to be careful opening up the panels covering the 
-%	engines, you never know if a bird or some animal might have made
-%	its home there.}
-%\end{quotation}
-
-%\begin{figure*}[htbp]
-%	\vspace{2em}
-%	\centering
-%	%name of the graphic, without the path AND in EPS format:
-%	\includegraphics[scale=0.5]{op-hawk-one.jpg}
-%	%caption of the figure 
-%	%\caption*{\small \em Volunteers put in countless hours restoring a
-%	North Star aircraft.}
-%	%label of the figure, which has to correspond to \ref{}:
-%	\label{fig:op-hawk-one.eps}
-%\end{figure*}
+All of the four propellers were  found to be in similar condition, and all were
+restored in a similar manner.
 

### 2020-09-13 pnsac/newsletter/article_Feature2.tex
First draft for Sep 2020 issue.

--- 2020-02-24 pnsac/newsletter/article_Feature2.tex
@@
 
-The earliest Merlins to operate in a civil mode were the Merlin T24
-series, developed in 1944. These were single-stage twin-speed units
-similar to the Merlin 24s fitted to the RAF's Lancasters, but were
-modified to improve service life, and were fitted to Transport
-Command's Avro Yorks. Long-range transport operation entailed running
-at relatively low cruise powers for long periods — under these
-conditions the lower cylinder head temperatures caused deposits of lead
-oxide from the fuel, resulting in excessive spark plug fouling. To
-counter this, the Merlin T24/4 incorporated a charge heater to increase
-the inlet temperature. Post-war, the Merlin 500 series essentially
-comprised civil and export versions of the T24, incorporating its
-modifications, though only the Merlin 501 included a charge heater.
+The earliest Merlins to operate in a civil mode were the Merlin T24 series,
+developed in 1944. These were single-stage twin-speed units similar to the
+Merlin 24s fitted to the RAF's Lancasters, but were modified to improve service
+life, and were fitted to Transport Command's Avro Yorks. Long-range transport
+operation entailed running at relatively low cruise powers for long periods —
+under these conditions the lower cylinder head temperatures caused deposits of
+lead oxide from the fuel, resulting in excessive spark plug fouling. To counter
+this, the Merlin T24/4 incorporated a charge heater to increase the inlet
+temperature. Post-war, the Merlin 500 series essentially comprised civil and
+export versions of the T24, incorporating its modifications, though only the
+Merlin 501 included a charge heater.
 
-The 600 and 700-series engines were two-speed two-stage engines, based
-on the military 100-series. They were fitted with the so-called
-'transport heads and banks', strengthened for greater reliability. Most
-marks had some form of variable intercooling to allow for charge
-heating to reduce plug fouling under cruise conditions. Initially the
-system was plagued by coolant leaks, but as experience showed that zero
-intercooling at cruise allowed enough charge heating to reduce leading,
-a simple stop valve was fitted to the system. This allowed full
-intercooling for take-off and climb, which could then be turned off for
-cruise.
+The 600 and 700-series engines were two-speed two-stage engines, based on the
+military 100-series. They were fitted with the so-called 'transport heads and
+banks', strengthened for greater reliability. Most marks had some form of
+variable intercooling to allow for charge heating to reduce plug fouling under
+cruise conditions. Initially the system was plagued by coolant leaks, but as
+experience showed that zero intercooling at cruise allowed enough charge
+heating to reduce leading, a simple stop valve was fitted to the system. This
+allowed full intercooling for take-off and climb, which could then be turned
+off for cruise.
 
-Civil Merlins were fitted to Avro Lancastrians, Yorks and Tudors, but
-their flagship role was on the Canadair DC-4M North Star, a Douglas
-DC-4-derived design intended for Trans-Canada Air Lines (TCA). The
-North Stars used a modified Universal Power Plant installation, as the
-DC-4's nacelle bulkheads were slightly larger than the SBAC standard.
-In practice it turned out the Merlins were not ideally suited for use
-in civil operations: though the North Stars flew higher and faster than
-DC-4s, their time between engine overhauls was around 850 hours,
-roughly a third of that of commercial US radial engines. The engines
-also produced a lot of cabin noise, though this was alleviated somewhat
-by revised 'cross-over' exhausts that ducted the cabin-side stacks
+Civil Merlins were fitted to Avro Lancastrians, Yorks and Tudors, but their
+flagship role was on the Canadair DC-4M North Star, a Douglas DC-4-derived
+design intended for Trans-Canada Air Lines (TCA). The North Stars used a
+modified Universal Power Plant installation, as the DC-4's nacelle bulkheads
+were slightly larger than the SBAC standard. In practice it turned out the
+Merlins were not ideally suited for use in civil operations: though the North
+Stars flew higher and faster than DC-4s, their time between engine overhauls
+was around 850 hours, roughly a third of that of commercial US radial engines.
+The engines also produced a lot of cabin noise, though this was alleviated
+somewhat by revised 'cross-over' exhausts that ducted the cabin-side stacks
 outboard, developed first by TCA and then by Rolls-Royce themselves.
 
-The extra running costs were mostly borne by Rolls-Royce. When TCA's
-managing director expressed dissatisfaction with the Merlin's
-commercial performance, Hives asked what a reasonable level of
-maintenance cost would be. Being told \$4 per engine hour, he agreed to
-service the fleet's engines for this amount. This early form of 'power
-by the hour' was initially expensive for Rolls-Royce, but by the end of
-the Merlin's life the company had learned enough to supposedly make a
-small profit at this level.
+The extra running costs were mostly borne by Rolls-Royce. When TCA's managing
+director expressed dissatisfaction with the Merlin's commercial performance,
+Hives asked what a reasonable level of maintenance cost would be. Being told
+\$4 per engine hour, he agreed to service the fleet's engines for this amount.
+This early form of 'power by the hour' was initially expensive for Rolls-Royce,
+but by the end of the Merlin's life the company had learned enough to
+supposedly make a small profit at this level.
 
-What was undoubtedly true was that Rolls-Royce learned a great deal
-about the harsh realities of commercial operation in a short time.
-Hives is supposed to have said to TCA, "We didn't know the Merlin until
-you started operating it!" In the longer term, this paid off,
-TCA---later Air Canada---selecting Rolls--Royce engines for its future
-fleet: Darts in Viscounts, Tynes in Vanguards, Conways in DC-8s, and
-RB211s in TriStars.
+What was undoubtedly true was that Rolls-Royce learned a great deal about the
+harsh realities of commercial operation in a short time. Hives is supposed to
+have said to TCA, "We didn't know the Merlin until you started operating it!"
+In the longer term, this paid off, TCA — later Air Canada — selecting
+Rolls-Royce engines for its future fleet: Darts in Viscounts, Tynes in
+Vanguards, Conways in DC-8s, and RB211s in TriStars.
 
-\begin{footnotesize} \raggedleft PNSAC\\
+\begin{footnotesize}
+  \raggedleft PNSAC\\
 \end{footnotesize}

### 2020-09-13 pnsac/newsletter/article_Feature3.tex
First draft for Sep 2020 issue.

--- 2014-05-24 pnsac/newsletter/article_Feature3.tex
@@
 
-\title{The Whispering Giant}
-\author{Dave Lambeth}
+\title{North Star Propeller Restorations}
+%\subtitle{Part 4}
+\author{Harry Hope, with assistance from Michael Hope.}
 
@@
 
-\begin{centering}
-Twinkle, twinkle, old North Star,\\
-With wrinkled skin and creaky
-spar,\\
-For years you've always been on
-sked,\\
-As many of our force you've sped\\
-to distant lands beyond the skies,\\
-you've never failed to see the rise\\
-of distant suns or moonrise glare.\\
-You've carried us with tender care\\
--- back you came to take us home,\\
-from old Japan or France or Rome\\
-from England, Goose, and Kef.\\
-We cannot hear you.\\
-('cause we're deaf)\\
-\end{centering}
+\textit{Hope Aero has played a major role in the restoration of the North Star.
+The following article provides some background about the company and its work
+on the North Star's propellers. }\\
 
-\begin{figure}[htbp]
-   \vspace{2em}
-   \centering
-   %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{jim_jung_2.eps}
-   %caption of the figure 
-   \caption*{\small \em A Whispering Giant (North Star 17515) at
-     Resolute Bay NWT in about 1963, taken by Radio Operator Jim Jung.}
-   %label of the figure, which has to correspond to \ref{}:
-   \label{fig:jim_jung}
-\end{figure}
+Hope Aero was formally government approved to certify aircraft propellers in
+1992, at their facility in Mississauga, Ontario. Harry Hope however, already
+had 40 years of propeller overhaul experience on many varieties that included
+Hamilton Standard propellers, which were quite common, as well as Hartzell,
+McCauley, and Sensenich. Harry's first restoration to flying condition, was the
+propeller for the Lysander, which was restored by the Air Force in Winnipeg for
+their Centennial project in 1967. This Lysander is now displayed at CASM. Since
+that time, they have restored several propellers, some to flying condition,
+three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
+(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
+display at the Aviation museum in Boda, Norway.
+
+ Since that time, their capabilities have added– MT, Dowty, Hoffman, and
+Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
+their customer base expanded to include different types of aircraft, the
+requirement for them to service more types became evident. They kept up with
+the required technology.
+
+Along the way, they also expanded into aircraft Wheel and Brake overhaul,
+sales, service and on wing dynamic balancing of fixed wing and Helicopters.
+Harry retired in 2001, but the third generation is now capably looking after
+the business.
+
+In the spring of 2006, Hope Aero took on the task of restoring the four North
+Star propellers, for the "North Star Restoration" project. The first challenge
+was to remove the propellers from the engines as neither the restoration crew
+nor Hope Aero had any recent experience removing this type of propeller.
+Michael Hope made the trip to Ottawa to assist with this operation and to help
+prevent any damage, which turned out to be exceedingly difficult due to the
+corrosion, starting with the spinner and then the propeller. Once the propeller
+was removed the blades were removed for shipping to Hope Aero.
+
+The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
+Aero in 2006. It was completely disassembled, cleaned and inspected. The
+external parts were very badly corroded, with most damage being to the aluminum
+blades and dome shells, and rust to the steel hub and tail shaft, which was
+likely caused by the outside storage environment. The inside parts were in
+relatively good condition. It was established, that both technical and visual
+effort would be required to restore the propeller to an acceptable display
+condition.
+
+The restoration work commenced with the blades, the alcohol de-icer feed shoes,
+troughs and paint being removed. The corrosion of the blade surfaces was such
+that removal by grinding was abandoned, as the inner half of the blades are
+shot peened and grinding would remove this and the authenticity. It was decided
+to media blast the complete airfoil surfaces with an aluminum oxide media, thus
+removing the surface corrosion but leaving a good surface for paint to be
+adhered to; an aviation grade grey paint was applied with yellow stripes at the
+tips. New alcohol feed shoes were installed, however the troughs, which direct
+the alcohol to the shoes, were not available. As the troughs are hidden inside
+the spinner, it was determined that they could be omitted.
+
+The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
+tubes, and dome retaining nut, were extremely rusty. They were glass bead
+blasted to remove the rust; this did not remove the heavy pitting but did leave
+a surface that could be cadmium plated.  After plating, a clear coat material
+was sprayed on these parts to improve the resistance to rust. The internal
+surfaces were in good condition, and they were well oiled.
+
+A replacement dome shell in relatively good condition was found, requiring only
+to be sanded, anodized, and be reinstalled with new seals.
+
+The propeller was then completely assembled to confirm that all the refurbished
+parts fit together. The blades were then removed to facilitate being shipping
+back to the museum.
+
+All of the four propellers were found to be in similar condition, and all were
+restored in a similar manner.
 

### 2020-09-13 pnsac/newsletter/article_MemberInterview.tex
First draft for Sep 2020 issue.

@@
 \title{Our Members}
-\author{Interview with Garry Dupont}
+% \author{Interview with Garry Dupont}
 
@@
 
-\end{multicols}
+% \end{multicols}
 
@@
 
-\begin{multicols}{2}
+% \begin{multicols}{2}
 
-1.\textit{What is your background in aviation?}
+Readers of the Chronicle will know that this item usually refers to one of our
+members. This edition, however, salutes all of our members who worked as
+volunteers on the restoration of the North Star engines namely:
 
-I graduated from the Confederation College Aircraft Maintenance course
-in the Spring of 1975. After returning to my home town in the Red Lake
-District I gained employment with Red Lake Seaplane Service. After
-completing the course I had to apprentice for 18 months before writing
-the final exam for my Aircraft Maintenance Engineers' License. I
-successfully wrote my final exam in December 1976 and obtained my AME
-certification. In February 1977 I was hired by Perimeter Aviation in
-Winnipeg Manitoba. There I obtained my Turbine Engine and
-Pressurization Aircraft endorsements.  I started my 30 year career with
-the RCMP Air Services Branch in February 1981. My career took me to
-Edmonton Alberta, Goose Bay Labrador, Winnipeg Manitoba then finally to
-Ottawa. I finished my career as Quality Assurance Manager.
+\begin{itemize}
+  \item Charles Baril 
+  \item John Corby (deceased)
+  \item Ted Devey (deceased)
+  \item Ed Hogan
+  \item Peter Houston
+  \item Ron Lemieux
+  \item Richard Lodge
+  \item Wib Neal
+  \item Jim Riddoch
+  \item Stan Rideout
+  \item Ted Slack
+  \item John Tasseron
+  \item Bill Tate
+  \item Tim Timmins
+  \item Giuseppe Zanetti
+\end{itemize}
 
-2. \textit{How long have you been involved with Project North Star and how and
-why did you get involved?}
+The following members have provided comments on their personal experiences.
 
-I attended the very first meeting held for Project North Star. This was
-at my wife's suggestion that it would be a good retirement pastime.
+\textbf{Charles Baril}
 
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   %name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
-%   %caption of the figure 
-%   \caption*{\small \em Forward cockpit -- complete.}
-%   %label of the figure, which has to correspond to \ref{}:
-%   \label{fig:fwd_cockpit_complete}
-%\end{figure}
+Working on all 4 Merlin engines has been an extremely rewarding experience that
+has taught me a lot about older technology, an appreciation for history
+involved (and those in the engine crew with that knowledge) and learning new
+skills and techniques. Although most of the parts were auxiliary systems,
+seeing them as part of the full engine gives a sense of accomplishment. It is
+very satisfying to see a part going from the dirty, oil covered, rusty part to
+newly refurbished one that looked like it was overhauled in an RCAF maintenance
+depot. It has been an honour to be part of this project and restoration of
+these pieces of history.
 
-3. \textit{What has been the history of your involvement to date?}
+In particular I remember working on the lower cowling flaps.  Trying to undo
+the inner skin was difficult due to longer rivets through the mounts for the
+Flaps. It took several tries and a few extra holes, but it was done. Fixing
+those holes often meant using original methods which was revealing and a
+commitment to historical methods.
 
-While I was employed I worked an evening shift every 4 weeks so I would
-come in to CASM in the morning for a few hours. Upon my retirement Mike
-Irving had me work in the engine shop, which I have been doing since my
-retirement. I also have been on the board of directors for a number of
-years.
+I also recall the work on the filter ducts.  The filter duct is the last stage
+of the air intake before the supercharger. During its cleaning, there was a
+starling nest on top of one filter. This caused a lot of corrosion damage that
+needed to be removed and treated. Yet, what got me was that starlings could get
+that far into the engine.
 
-%\begin{figure}[htbp]
-%	\vspace{2em}
-%	\centering
-%	%name of the graphic, without the path AND in EPS format:
-%	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
-%	%caption of the figure 
-%	\caption*{\small \em Propeller assembly.}
-%	%label of the figure, which has to correspond to \ref{}:
-%	\label{fig:propeller}
-%\end{figure}
+\textbf{Richard Lodge}
 
-%\begin{figure}[htbp]
-%	\vspace{2em}
-%	\centering
-%	%name of the graphic, without the path AND in EPS format:
-%	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
-%	%caption of the figure 
-%	\caption*{\small \em Forward cockpit -- complete.}
-%	%label of the figure, which has to correspond to \ref{}:
-%	\label{fig:fwd_cockpit_complete}
-%\end{figure}
+As probably the only volunteer who ever worked in the Aero Engine Division of
+Rolls-Royce in Derby, England, I was greatly interested in becoming involved in
+Project North Star, even though, as an accountant, I had never worked on the
+design or building of an aeroengine.  By the time I started work at Rolls-Royce
+in 1963, new build of Merlin engines had ceased, and only spare parts were
+being manufactured.
 
-4. \textit{What has been the highlight of your involvement?}
+I have several strong memories of working on the Merlin engine.  In general,  I
+particularly remember John Tasseron working endlessly on freeing seized piston
+rings.  I learned much from John about patiently working at something which
+appeared to be impossible! Later, when working with  R\'{e}j Demers and Garry
+Dupont, I started to learn how to rivet and how to lock wire. The most
+important lesson for me has been, photograph, photograph, photograph, at each
+stage of dismantling, cleaning and reinstalling an engine part. When sitting in
+an overcrowded office at Rolls-Royce, it never occurred to me that at some time
+in the future I would actually be working on a Merlin engine or working my way
+through an original parts manual to work out how to reassemble an item.  Life
+is certainly full of interesting surprises.
 
-The biggest highlight for me was installing number 2 engine. It was
-about half completed when I started in the engine shop. This is the
-largest engine I have ever installed in my entire career.
+\textbf{John Tasseron}
  
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   %name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
-%   %caption of the figure 
-%   \caption*{\small \em Navigator rack.}
-%   %label of the figure, which has to correspond to \ref{}:
-%   \label{fig:nav-radio-rack}
-%\end{figure}
+The list below is intended to give a brief overview of the various highlights
+that arose during my participation in the restoration of the second, third and
+fourth Merlin engines:
 
-5. \textit{What has been the most challenging part of your involvement?}
+\begin{enumerate}
+  \item It's not every day one gets the opportunity to see up close what makes
+the Rolls-Royce 620 Series Merlin engine tick!
+  \item It's a real challenge to safely lower each "power egg" down off the
+wing, place it in "the big blue stand" and move it in to the workshop.
+  \item Then starts the process of carefully separating about 11,000 individual
+parts before cleaning, inspecting and re-conditioning them as necessary. Do not
+lose any and make sure everything is photographed for museum records!
+  \item When things unexpectedly go wrong, the real expertise kicks in to
+perhaps develop a repair scheme or even manufacture a brand-new replacement
+part. The objective is to bring everything back together again and end up with
+a "like new" engine, ready for preservation.
+  \item It all takes team effort based on effective communication and
+co-operation. There are no dumb questions and the answers are all aimed at the
+challenge to create an "airworthy" product that will never fly.
+  \item Finally, its important and fun to share the project with a curious
+public, so be ready to answer any questions that might be asked about the
+aircraft and its engines!
+\end{enumerate}
 
-One of the most challenging parts was learning how the engine worked
-with all its systems. I found a RCAF training manual in the office and
-took it home. Over several months of reading I started to understand
-what made this monster of old technology tick. Another challenge was
-learning the British Hardware system and tools. A further challenge was
-reading  the old manuals which were written before they were
-standardized by the ATA chapter system which all aircraft manuals
-follow today. Learning the art of conservation was also a  challenge
-but was made easier by the guidance of Mike Irving.
+%1.\textit{What is your background in aviation?}
+%
+%
+%2. \textit{How long have you been involved with Project North Star and how and
+%why did you get involved?}
+%
+%3. \textit{What has been the history of your involvement to date?}
+%
+%4. \textit{What has been the highlight of your involvement?}
+%
+%5. \textit{What has been the most challenging part of your involvement?}
 

### 2020-09-13 pnsac/newsletter/article_PresReport.tex
First draft for Sep 2020 issue.

--- 2020-02-24 pnsac/newsletter/article_PresReport.tex
@@
 
-Since the last issue of the NStar Chronicle we have taken part in the
-70th Anniversary celebrations of the first non-stop transcontinental
-flight across Canada which was flown by a North Star. We partnered with
-the Museum for this event and were very pleased to welcome
-representatives from the RCAF who joined us for the day. After the
-formal part of the celebration, when impressive plaques of the aircraft
-made by our member Chris McGuffin, had been presented to the RCAF and
-the Museum Director General, Chris Kitzan, everybody moved to the
-Reserve Hangar and the aircraft was opened to our association members,
-guests and in the afternoon to 90 members of the public who had
-pre-booked a tour. R\'{e}jean(R\'{e}j)Demers, the Museum Special Project
-Manager, guided the tours assisted by several of our members. Towards
-the end of the tour session someone in the main Museum activated the
-fire alarm. The Museum was promptly evacuated, and the public was
-recommended to go to the Reserve  Hangar until the fire issue had been
-dealt with (it turned out to be a false alarm). Quite unexpectedly R\'{e}j
-and the guides found themselves surrounded by many Museum visitors who
-had not signed up for the North Star tour. Unperturbed by this event,
-Réj and the guides welcomed the unscheduled visitors and gave them an
-impromptu tour of the aircraft. Both our treasurer and our membership
-secretary were pleased with the days' result!
+As always, before I start to write the Notes from the President, I reread my
+previous Notes to remind myself of what I had previously said in order to set
+the record straight if necessary. As I write this at the beginning of
+September, I am now aware of the impact of the COVID-19 pandemic on the
+restoration schedule for the North Star and the operations of your
+Association.
 
-As I write this the days are getting longer and R\'{e}j is planning work to
-be done on the aircraft when it is moved outside onto the ramp in the
-spring. Each year we do work on the aircraft which cannot be done
-inside the Reserve Hangar during the winter. We also prepare the
-aircraft for display when, once again, we open the aircraft to enable
-the public to view progress made in the last year.
+Since March, restoration work on the aircraft has been suspended and the
+activities of the Association have been limited. We have, however, been keeping
+in touch with each other through the weekly Zoom meeting organized by the
+Museum's volunteer co-coordinator Cédric St.-Amour. This virtual meeting takes
+place every Thursday at 10 AM EST; it is open to all our volunteers and any
+other PNSAC members who would like to join in.
 
-This year we are planning two major display events. The first will be
-on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
-for our members to meet the public, many of whom have an interest in
-aviation. Our second main event, on Canada Day, is enjoyable for all
-the members who sign up to be part of the team meeting the public.
-There are always large crowds, many of them family groups, If any of
-you reading this article would be interested in joining our members for
-either of these events please contact me (my co-ordinates are on the
-back page of this issue of the Chronicle)
+As previously reported, the North Star project reached a major milestone with
+the completion of the restoration of Engine \#4 which was moved into the
+reserve hangar on a cold December 2019 morning.
 
-Shortly before Christmas a major milestone, in the  restoration of the
-North Star, was reached. Work on Merlin engine \#4 was completed and it
-was dragged on its cradle by the Museum forklift truck to the
-Restoration hangar for later mounting on the aircraft to join the other
-three restored engines. Chris McGuffin put some video clips onto our
-Facebook page of the engine being dragged through the Museum parking
-lot and into the hangar. Watch for more of our Facebook messages since
-Chris frequently posts new items.
+Since we are unable to physically meet and there has been no restoration work
+carried out, we felt that a special issue of the NSTAR Chronicle devoted
+entirely to the 15 year restoration of all four of the Merlin engines would be
+of interest to members.
 
-In the late fall we welcomed the appointment of Erin Secord as the
-Museum's new Manager Conservation Services. She is responsible also for
-conservation of the collections at the Science and Tech Museum. Erin
-has now been able to join us at two of our Members' meetings.
+Understandably, with a project which has taken many years. the volunteers
+working on the engines have changed over time. Some have retired or stopped
+volunteering and sadly, some have died. Earlier this year we asked Ted Devey to
+contribute a piece but before writing about of his work, he passed away. Ted
+was involved with restoring all the engines and for several of the early years
+was the volunteer leading the restoration team. Ted was a great raconteur and
+our oldest volunteer at 93 years old. His recollections would have added
+considerably to the interest of this special issue of the NSTAR
+Chronicle.
 
-Finally, we are now working to update the Memorandum of Understanding
-between PNSAC and the Museum.  Much has changed since the original
-agreement was signed in May 2006.
+After much consideration, the Museum has decided that volunteering cannot
+restart before January 2021. Until then we will continue with our regular Zoom
+meetings and will also organize a virtual Annual General Meeting for the
+Association. We will also continue to provide updates on the situation and much
+will depend upon how the different levels of government handle the ongoing
+recovery from the pandemic.
 
-Our next Members' Meeting is scheduled for Saturday, June 20, 2020
-which will also be our AGM. Do reserve this date in your calendar if
-you can be in the Ottawa area on that day. The meeting will be held at
-the Canada Aviation and Space Museum. More details will follow later.
+There is an old saying "That it is an ill wind that blows nobody any good."
+This is very true even of this pandemic. One of the very pleasant results is
+having conversations with PNSAC members who don' t reside in the National
+Capital area and have sent membership renewals by mail. Where a renewal has
+been received and confirmation has been delayed as a result of the pandemic, I
+try to call a member on the telephone to explain what is happened and that the
+renewal has not been lost. As a result I have had many interesting phone calls
+with people whom I have never spoken to before. A definite bonus for me.
 

### 2020-09-13 pnsac/newsletter/article_ProgReport.tex
First draft for Sep 2020 issue.

--- 2019-01-13 pnsac/newsletter/article_ProgReport.tex
@@
 %\subtitle{PNS 2016 Status Update}
-\author{R\'{e}jean Demers}
-\author{Restoration Project Manager}
+\author{R\'{e}jean Demers, , AME---Conservator \/ Special Project Manager
+Canada Aviation \& Space Museum}
+%\author{Restoration Project Manager}
 
@@
 
-\section{Zen and the Art of Aircraft Restoration.}
+%\section{Zen and the Art of Aircraft Restoration.}
 
-Through my daily experience over the past two years, working with volunteers, I
-have asked myself the following question: What makes a volunteer dedicated to a
-project? The answer is not a straightforward one. This article explores the
-subject, much like Pirsig did in his book about metaphysical motorcycles; Zen
-and the Art of Motorcycle Maintenance.
+When touring the shops in the Museum’s Conservation wing, the engine shop is
+reserved for what is referred to as, the “Crown Jewel”. Guests are lead through
+the corridor and the lights are flicked on to reveal a shining piece of
+aeronautical engineering and craftsmanship. The Merlin 622 engine, one of four
+such power plants, sits resplendent. Gasps and expressions of delight from
+guests are guaranteed to insight a prideful glance upon this piece of work by
+volunteers and staff alike. Appearing as if the engine had just been delivered
+from Bristol Aero Ltd. is an illusion. What is this strange magic? How can such
+beauty manifested though steel and aluminum even exist? As any good sorcerer
+will tell, illusion is lost upon the dissolution of mystery. Yet this piece of
+work, a restoration of great effort, time and detail, cannot shed the cloak of
+timelessness so easily. The result of thousands of hours culminate through
+unrelenting intrigue, garnered by all who cast eyes upon the Merlin. These
+words simply scatter as embellishments to the enduring design and purpose,
+which cannot be hidden in all evidence of its greatness.
 
-We strive for quality. As restoration work continues on the North Star, we get
-absorbed into the weight and breadth of the subject matter. This is the kind of
-project that demands the level of commitment found in monasteries, dojos and
-Paratrooper units. Commitment comes from dedication and reward. Quality is
-obtained through application of commitment, producing results which may be
-enjoyed. The following effort in writing has been made to define such elements,
-those necessary to sustain the project.
+Referencing an Ottawa Citizen article dated Saturday Feb. 1st, 2003. We read of
+an effort to restore the Canadair North Star to a version of her “former self”
+as PNSAC founder Robert Holmgren put it. A sad, sorry state at the time.
+Destitute to serve as a bird house to the starlings. No mention of any spells
+enacted to summon the aviation gods or mojo hand placed upon the yoke to stir
+the spirits which have tilled the air.  Yet today, she’s a Hangar Queen,
+patiently awaiting final installation of engine \#4. Is this sorcery? No,
+simply the fourth iteration of the Project’s labour of love; Rolls-Royce Merlin
+622, Serial \#309573. The task of accomplishing the finale to a four part, long
+haul repeat restoration of a power plant began in 2014. Looking back at power
+plant restoration cycles, we see the evolution of the Project North Star
+volunteer organization since the early 2000’s. The forming of Engine Shop crews
+and the establishing of Crew Chiefs. The coordination of daily grunt work of
+condition recording, treatment and reassembly of components. Today, members of
+the project continue to develop ideas, adapting to new technology, finding
+different approaches and maintaining a solid work ethic. This spirit remains
+unchanged throughout difficult times and challenges we have encountered.
 
-A voracious appetite for detail. Precise facts and data build foundations.
-Foundations of information that erode over time. Lost to floods, fires, or
-worst; thieves, vandals and neglect. Information is our firmament that keeps us
-grounded in our work. Daily worksheets are filled with steps taken to satisfy a
-condition. Drawings are made to translate reality to paper. These words,
-drawings and tables form hard to translate time lines. Coloured bars overtaking
-each other, running in thousand hour blocks of volunteer contributions. These
-life hours, cannot be overlooked, as a restoration project advances through the
-years.
+The Tech. log states: Engine installed on North Star 17515 on August 23rd, 1965
+at airframe Total Time of 19845.45 Hrs. with 1121 hours remaining in \#4
+position (4th to the right). The engine had previously flown on 17514 until
+April 30th, 1965 when a broken intake valve \@ \#1 cylinder, “B” bank
+necessitated a repair. Flipping through pages, various signatures, authorities,
+stamps and markings show evidence of a storied service life going back to 1948.
+Engines are shared among power plant “skeletons” which in turn, are shared
+across the fleet. This was one of the major design features of the “Power Egg”
+as the promotional material of the day had put it, interchangeability and ease
+of maintenance. A glance at the log tells me this engine serial number served
+on North Stars 506, 510, 508 and 502 to name a few. The last remaining example
+of its type, North Star 17515 holds pieces of the past, belonging to aircraft
+that have long been gone. Like crew members who have also shared their time
+amongst a fleet that traveled all over the globe, with change-overs,
+turn-arounds and plenty of distance in-between.
 
-Inorganic compounds risk distraction. Focus is important when avoiding the
-ethereal draw of solvents. A volunteer scrubber, washing away years of residue,
-layered upon a finish that once gleamed with pride. Slowing the return to dust,
-so feared and fought by the living. A process to follow and adhere to. The
-methodical dismantling and reconstruction of an object. Return to form, that
-which was almost indiscernible. The result of all his work is not merely a
-shiny piece of aluminium panelling; He has travelled through time.
+One of the big pay-offs for members of Project North Star is the opportunity to
+display the work carried out by volunteers. Many occasions come to mind, such
+as Canada Day, Doors Open Ottawa and family visits to the Museum. One
+opportunity, of which I will always be grateful to have participated in, was
+such a moment. Myself and a few volunteers met a retired employee of the
+aforementioned Bristol Aero Industries Ltd. who worked at the overhaul shops
+when this engine was undergoing acceptance runs in the early to mid-sixties.
+Quickly, job experience and familiarity with the object was brought forward. A
+quick look under the cowling to reveal preserved markings and decals, evidence
+of original repair shop work was pointed out. “I probably worked on that job”
+he said. I understood at this point, how someone’s work, preserved in such a
+way, could be of intangible value. The history coming around full-circle before
+my eyes, as the gentleman pointed out the ROLLS-ROYCE cast upon the rocker
+covers, saying “it was one of my first jobs, painting on those red letters”.
 
-More than just an object, we look to a hidden past. Beyond these material
-trappings, lay the history of Canadair North Star 17515 and her sisters.
-Examples of service are easily found in books and photos. What proves a
-challenge is explaining how a bumblebee becomes a preserved specimen in an
-engine, how graffiti made during heavy maintenance is the only evidence of a
-man's past. Where the dirt under floorboards, came from countries that no
-longer bear the name she visited.
+These records, object qualities, historic log entries and our treatment reports
+all add up. We string together these bits of information to form longer chains
+that communicate the history of an artifact. As focal points shift towards
+other areas of interest, or priorities, work carried out in the past helps
+maintain a continued relevance for others to pick-up where we left off. With
+such a long term project, setting the pace is just as important as taking a
+break now and then. Knowing that we had an impact and remembering the
+experiences we’ve shared can carry us through new challenges. It’s a pleasure
+for me to observe a new volunteer work towards acquiring the skills that may be
+passed on to others, like an old wizard speaking to his young apprentice.
 
-The static, prevents complete appreciation. She's more than 45,000 lbs. of
-metal. Superficial understanding can only consider her 39 years an outdoor
-relic. Likewise, so would a skin deep treatment of a last remaining artefact of
-this type. What impact she had before the stillness of the seasons is not lost
-to those who look further.
-
-A biological experience in personal evolution. The volunteer has developed new
-muscles and sinew adapted to producing polished surfaces. An eye for
-microtexture and grain direction has developed. Sympathetic scars are
-accumulated through careful little sacrifices. Navigating the disruptions of
-capricious human needs such as rest, nourishment and leisure. Through the
-application of labour, a piece of history is preserved.
-
-Disappearing horizons. Meditative expanses of aluminum, exhibiting varieties of
-corrosion rarely encountered in a lifetime's career. Cavernous interior spaces,
-where cargo of all sorts was bound for ports beyond. Flight control surfaces
-that have pressed through all shades of dusk, emerging in the dawn of a new
-day. Now removed from the ship, these weathered, beaten, bruised remnants of a
-glorious past await restoration. A new horizon forms, some many years away,
-known to those who are on-board for the long run.
-
-Healthy obsessions. Volunteer restoration technicians take home an experience.
-They take to heart the results of their labour, in forming a product that is
-also intangible. The mind has strengthened, forming resolves that assist,
-understand and process the elements of restoration work. As historians,
-fanatics and unwitting library wanderers soon discover; The story of air
-transport, in its beginning, is far from just a few obscure paragraphs in
-Canada's ongoing history. It lives through those people who continue writing
-it, preserving the physical, extending beyond.
+I encourage Project North Star members to communicate their interest in the
+project to the public. Generate their own records that may contribute to the
+history of the machine and the lives it affected. This newsletter, social media
+and other forms of reaching out to the public are excellent ways to remember
+one of Canada’s great contributions to air transport. With the tools at our
+disposition, I would not hesitate to believe we can create many new and
+engaging experiences for the public to learn about the role this aircraft
+undertook while in service. Now, in continuation with a sense of national
+pride, the North Star remains an important platform for interpretation of the
+work carried out by dedicated volunteers supported by a museum that understands
+the value of such an artifact.
 

### 2020-09-26 pnsac/newsletter/article_EditorsNotes.tex
Sep2020--first draft.

--- 2020-02-24 pnsac/newsletter/article_EditorsNotes.tex
@@
 
-Putting out this edition of the Chronicle has been an example of great
-co-operation from members and non-members alike. Thanks are owed to a
-number of people. Our member Gary Whitten was instrumental in getting
-the necessary permission from Air Force Magazine to publish the
-"Diversion to Alert" article.  Allan Bowes, another member, suggested
-that we should try to get the "Civil Merlins" article into an edition
-of the Chronicle.  Following his suggestion we contacted Ben Dunnell,
-the editor of Aeroplane Magazine. Through Ben's offices we were able to
-secure permission to reprint this article.  The "Civil Merlins" was
-part of a comprehensive 17 page Aeroplane Database article on the
-Merlin. If any members wish to read the entire article it was published
-in the July 2017 issue.
+In this issue we are recognizing the completion of a very important step in the
+restoration of the North Star, namely, its Merlin engines, and our members who
+have volunteered with CASM to carry out the necessary work.  This is a very
+informative issue and I am sure we will all learn a great deal. I was not
+aware, until putting this edition together, that the Merlin designation was in
+keeping with the custom of Rolls Royce to name its aircraft engines after birds
+of prey.
 
-In addition thanks go out to the other contributors, namely Richard
-Lodge, Garry Dupont,  Neil Raynor and Bruce Gemmill. Bruce took the
-lead in putting together the Progress Report.  Finally, I would like to
-thank Drew Hodge our stalwart publisher of this edition. We are always
-on the lookout for content for the Chronicle so please do not hesitate
-to bring forward any ideas that you might have. Enjoy.
+The article, "Four Merlins" provides the reader with a detailed history of the
+restoration of these engines.   Due to its length, the article will be divided
+into two editions;  starting with engines number one and two in this edition;
+and three and four in the next edition. We are fortunate to have members who
+have had  lengthy involvement with this part of the project. Sadly one of those
+members passed away recently. Inside you will find a tribute to Tim Timmins by
+the President of our Association.
+
+We look forward to receiving your comments and suggestions regarding this
+edition and for future editions whether good, bad or ugly. Please use my email
+address located on the back page.
+
+Good reading.
 

### 2020-09-26 pnsac/newsletter/article_Feature1.tex
Sep2020--first draft.

@@
 
-\title{North Star Propeller Restorations}
+\title{The Four Merlins: The First Two}
 %\subtitle{Part 4}
-\author{Harry Hope, with assistance from Michael Hope.}
+\author{Bruce Gemmill}
 
@@
 
-\textit{Hope Aero has played a major role in the restoration of the North Star.
-The following article provides some background about the company and its work
-on the North Star’s propellers. }\\
+When North Star 17515 was retired from service with the RCAF, it was delivered
+to the National Aviation Collection at Rockcliffe  in 1966, where it sat
+exposed to the elements for many years. When a volunteer group was formed in
+2001 to restore the aircraft, the priority was to move the aircraft inside the
+newly constructed Reserve Hangar to prevent further deterioration.  This was
+accomplished on August 11, 2005 when the North Star was moved inside for the
+first time.
 
-Hope Aero was formally government approved to certify aircraft propellers in
-1992, at their facility in Mississauga, Ontario. Harry Hope however, already
-had 40 years of propeller overhaul experience on many varieties that included
-Hamilton Standard propellers, which were quite common, as well as Hartzell,
-McCauley, and Sensenich. Harry's first restoration to flying condition, was the
-propeller for the Lysander, which was restored by the Air Force in Winnipeg for
-their Centennial project in 1967. This Lysander is now displayed at CASM. Since
-that time, they have restored several propellers, some to flying condition,
-three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
-(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
-display at the Aviation museum in Boda, Norway.
+\begin{figure}[H]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-1.png}
+   %caption of the figure 
+   \caption*{\small \em North Star being moved inside hangar 11 August 2005 .}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
 
- Since that time, their capabilities have added– MT, Dowty, Hoffman, and
-Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
-their customer base expanded to include different types of aircraft, the
-requirement for them to service more types became evident. They kept up with
-the required technology.
+The restoration project attracted many aviation enthusiasts and the early group
+of volunteers were eager to work on restoring the four Rolls Royce engines. The
+Canada Aviation Museum (as it was then called) agreed and set this as the top
+priority for the volunteers.  Work started in earnest in 2006 under the
+supervision of the Museum's Project Coordinator, Mike Irvin.
 
-Along the way, they also expanded into aircraft Wheel and Brake overhaul,
-sales, service and on wing dynamic balancing of fixed wing and Helicopters.
-Harry retired in 2001, but the third generation is now capably looking after
-the business.
+This is the detailed  history of the restoration work on the first two Merlin
+engines.
 
-In the spring of 2006, Hope Aero took on the task of restoring the four North
-Star propellers, for the "North Star Restoration" project. The first challenge
-was to remove the propellers from the engines as neither the restoration crew
-nor Hope Aero had any recent experience removing this type of propeller.
-Michael Hope made the trip to Ottawa to assist with this operation and to help
-prevent any damage, which turned out to be exceedingly difficult due to the
-corrosion, starting with the spinner and then the propeller. Once the propeller
-was removed the blades were removed for shipping to Hope Aero.
+\paragraph*{Engine 1}\
 
-The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
-Aero in 2006. It was completely disassembled, cleaned and inspected. The
-external parts were very badly corroded, with most damage being to the aluminum
-blades and dome shells, and rust to the steel hub and tail shaft, which was
-likely caused by the outside storage environment. The inside parts were in
-relatively good condition. It was established, that both technical and visual
-effort would be required to restore the propeller to an acceptable display
-condition.
+The removal, restoration and assembly of Engine Nr 1 began in June 2006 and was
+completed in Feb 2010 with the installation of the engine back on the North
+Star.  This work included a complete teardown and re-assembly of all major
+components including the Merlin V12 engine, propeller, supercharger,
+intercooler and auxiliary gearbox, as well as the engine frame, cowlings,
+radiators, pipes and accessories.
 
-The restoration work commenced with the blades, the alcohol de-icer feed shoes,
-troughs and paint being removed. The corrosion of the blade surfaces was such
-that removal by grinding was abandoned, as the inner half of the blades are
-shot peened and grinding would remove this and the authenticity. It was decided
-to media blast the complete airfoil surfaces with an aluminum oxide media, thus
-removing the surface corrosion but leaving a good surface for paint to be
-adhered to; an aviation grade grey paint was applied with yellow stripes at the
-tips. New alcohol feed shoes were installed, however the troughs, which direct
-the alcohol to the shoes, were not available. As the troughs are hidden inside
-the spinner, it was determined that they could be omitted.
-The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
-tubes, and dome retaining nut, were extremely rusty. They were glass bead
-blasted to remove the rust; this did not remove the heavy pitting but did leave
-a surface that could be cadmium plated.  After plating, a clear coat material
-was sprayed on these parts to improve the resistance to rust. The internal
-surfaces were in good condition, and they were well oiled.
+Before the engine could be removed, the propeller and spinner cover needed to
+be removed.  The spinners were popular nesting sites for birds, and the bird
+droppings had caused extensive corrosion to the propeller hub.  To remove the
+hub, a large wrench was fabricated to fit over the large, threaded ring that
+held the hub firmly in place.  This was struck repeatedly with a 5 lb sledge
+hammer before the ring began to move.  This process was repeated each time a
+propeller was removed.  The propeller was then disassembled and sent to Hope
+Aero in Mississauga, Ontario for restoration.
 
-A replacement dome shell in relatively good condition was found, requiring only
-to be sanded, anodized, and be reinstalled with new seals.
+Engine 1 was removed and mounted in a mobile assembly stand and moved to the
+conservation shop for inspection disassembly. Since this was the first Merlin
+engine any of the volunteers had worked on, much time was taken photographing
+the engine from all angles, and studying the Repair and Overhaul manuals that
+the RCAF and Rolls Royce had provided.  The digital photographs were also an
+important conservation record for the museum.
 
-The propeller was then completely assembled to confirm that all the refurbished
-parts fit together. The blades were then removed to facilitate being shipping
-back to the museum.
+Each engine is contained within a Quick Engine Change (QEC) assembly or power
+pack. This consisted of a rigid frame to hold the cowl panels which surround
+the engine, and also hold such vital parts as the two radiator flaps, the three
+piece radiator itself, and support the fire suppression system and various
+pipes and hoses.  This frame also holds the four attachment points to secure
+the engine assembly to the aircraft.
 
-All of the four propellers were  found to be in similar condition, and all were
-restored in a similar manner.
+The cowl panels were removed, along with the support frame, to expose the
+engine.  Large amounts of nesting material were removed, even though museum
+staff had attempted to "bird-proof" the aircraft while outside.  Old engine oil
+was drained from the crankcase, and the volunteers began familiarizing
+themselves with the numerous parts of the wonderful V-12 Merlin engine they
+would work on for the next four years.
+
+Along with learning the intricacies of the Merlin, a substantial number of
+special tools and jigs needed to be found or made.  This began with a full set
+of British sockets and wrenches for the Whitworth and BA nuts and bolts used
+throughout the engine.
+
+When it came time to remove the cylinder head, the volunteers found the pistons
+were seized in place. To free up the pistons, large amounts of penetrating oil
+was applied, then each piston struck with a large sledge hammer and block of
+wood.
+
+While this work was going on at the museum, Hope Aerospace in Mississauga was
+restoring the first of the four Hamilton propellers sent to them under
+contract. The first propeller was returned from Hope Aero after restoration in
+2008, and Mike Hope volunteered his time to help assemble it.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-2.png}
+   %caption of the figure 
+   \caption*{\small \em  Mike Hope and Bruce Gemmill assembled Prop 1ready to
+install on Engine 1 October 26, 2008.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+By early 2009, most of the main engine components had been cleaned, and painted
+or polished, ready to be assembled on the completed engine block and cylinder
+banks one and two.
+
+Reassembly took place on the rotary engine stand, which aided installation of
+the crankshaft and other lower engine components. Below is a photo of the
+nearly completed engine, showing the cylinder heads and spark plug cables in
+place, before moving to the mobile engine stand for final assembly. Note the
+highly polished brass expansion tank for the main radiator, nestled behind the
+propeller cowl.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.55]{four-merlins-3.png}
+   %caption of the figure 
+   \caption*{\small \em Engine 1 assembled in the Conservation shop Feb 2,
+2010.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+Once the engine was moved to the assembly stand where the engine frame and
+large radiator awaited, the wiring and various pipes and hoses could be fitted,
+along with the already assembled wheelcase and supercharger. Lowering this very
+large assembly into the tight space at the rear of the engine was a very
+delicate manoeuvre, but the reward was a nearly completed Merlin!
+
+After installing the massive air intake and intercooler, the various cowl
+panels were fitted to the outside of the engine frame, to complete the QEC
+assembly.
+
+While this work was going on, final preparations were being made at the
+aircraft to receive Engine 1.  This required cleaning and polishing the
+firewall and various engine fittings, and installing the auxiliary gearbox that
+drives the electrical generator and other accessories on the engine.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.55]{four-merlins-4.png}
+   %caption of the figure 
+   \caption*{\small \em Firewall ready, auxiliary gearbox installed Jan 7,
+2010.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.55]{four-merlins-5.png}
+   %caption of the figure 
+   \caption*{\small \em Engine 1 complete at the Restoration Shop Feb 9, 2010
+.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.45]{four-merlins-6.png}
+   %caption of the figure 
+   \caption*{\small \em Engine 1 installed Feb 24, 2010. Shown top left to
+bottom right, engine
+supervisor Ted Devey,  Bill Tate, Ron Lemieux, Giuseppe Zanetti, John Corby,
+Garnet Chapman, Volunteer Project Manager Jim Riddoch, John Tasseron, Bruce
+Gemmill, Museum Project Coordinator Mike Irvin, Charles Baril, and then
+Association President Tim Timmins.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+The engine was reinstalled on February 24, 2010 and the crew could take a short
+break before beginning work on engine 2.
+
+\paragraph*{Engine 2}\
+
+Work started on the removal of Engine 2 in early April, beginning with removal
+of all the cowl panels, draining any fluids left in the engine and radiators,
+and disconnecting all pipes, hoses and electrical fittings at the firewall. 
+Care was taken to take better photos of the removal of parts so reassembly
+would not be as much a challenge second time around.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-8.png}
+   %caption of the figure 
+   \caption*{\small \em Engine 2 with cowl panels removed, April 6, 2010.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+To remove the engine, a sling was attached to the lifting points on the engine,
+and a fork lift took the weight while four attachment bolts were removed from
+the firewall.  Keeping the engine level while it was slowly backed away from
+the aircraft was important, to avoid damage.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-9.png}
+   %caption of the figure 
+   \caption*{\small \em Engine 2 removed April 14, 2010.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+The engine team made a quick job of disassembling the number 2 engine.  All
+major assemblies such as the intercooler and supercharger were removed, along
+with all pipes and hoses.  The engine was removed from the frame and installed
+on the rotatable engine stand.  The cylinder heads, cylinder blocks and pistons
+were removed, as well as the crankshaft.  These major engine components were
+cleaned so that the task of reassembling the engine could begin.  Disassembly
+was not easy.  After years outside exposed to the elements, several of the
+pistons were seized inside the cylinders.  Freeing these required a lot of
+muscle and Liquid Wrench.  Still, the disassembly of Nr 2 went much faster than
+the first engine.
+
+A lot of corrosion was removed by a process called glass beading, but
+volunteers needed to be careful not to damage aluminum surfaces or contaminate
+seals.  Most aluminum components were clearcoated to protect from further
+corrosion, while other parts were often painted with silver aluminum paint to
+restore the original look.
+
+As with Engine 1, this engine was assembled in the rotating stand.  Shown
+below, volunteer Stan Ride is adjusting the torque on the main bearing caps.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-10.png}
+   %caption of the figure 
+   \caption*{\small \em Stan Rideout working to reassemble Engine 2 in engine
+shop, Oct 7, 2010.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+By December 2010 many major components such as the engine block, crankshaft and
+pistons had been refurbished and re-assembled.  The experience gained from
+Number 1 engine paid off in spades.
+
+The wiring harness was refurbished, but would not be installed until after the
+engine was in place.  This was a key lesson learned from engine Nr 1.  Some
+assemblies were installed too early, and had to be removed again so the engine
+was accessible.  This time, the crew assembled the Quick Engine Change  module
+from the inside out.  Pipes and hoses would only go on after major parts were
+in place.
+
+By March 2011 Nr 2 frame had been refurbished, and work was progressing on
+cowlings, radiators and accessories.  In the meantime all four propellers had
+been restored, by Hope Aero along with the four spinners. Three were installed
+back on the aircraft, while the fourth was stored until all four engines were
+completed.
+
+The Nr 2 engine was  re-assembled by April 2011. The crankshaft and pistons had
+been installed, along with the cylinder heads and valve train. Work was then
+take to restore the ignition wiring and the auxiliary gearbox.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-11.png}
+   %caption of the figure 
+   \caption*{\small \em Austin (Tim) Timmins.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+Although engine Nr 1 was installed on the North Star in February, 2010, it 
+continued to serve as a model for what engine Nr 2 should look like when
+completed.  Volunteers made numerous trips to the storage hangar to compare
+engine Nr 1 to the items currently then being restored.  This allowed the crew
+to solve several re-assembly problems quickly. The engine block was complete
+and the crankshaft, cylinder heads and pistons had  been installed by June
+2011. Many engine accessories and sub-assemblies had also been treated and were
+ready to be reattached.
+
+The engine frame also received some attention.  The rear cowl ring that held
+several of the external panels was completely stripped and repaired before
+painting and re-assembly. The three radiators for the engine, intercooler and
+oil cooler were dismantled, cleaned and painted.  These three units were
+assembled and attached to the front of the engine frame.  The radiators were
+then filled with inhibiting oil to prevent corrosion.  There were many
+accessories that needed to be refurbished before being installed on the engine
+frame.  When completed, most of these items were stored until after the engine
+was installed in the frame.  This made assembly much easier.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-12.png}
+   %caption of the figure 
+   \caption*{\small \em Jim Riddoch and Bill Tate fill the radiators with
+inhibiting oil.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+The auxiliary gearbox, which is a complex piece of machinery, was reassembled,
+but the reduction gear that connects the propeller shaft to the engine proved
+to be difficult, due to excessive corrosion.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.7]{four-merlins-13.png}
+   %caption of the figure 
+   \caption*{\small \em Engine 2 on rotating assembly stand.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+The cowl panels were rebuilt, as severe corrosion had affected many of the
+steel stiffeners riveted to the edges of the panels.  These had to be removed
+and refinished.  Several were so badly gone that new ones were fabricated and
+installed.  Once riveted in place, the stiffeners were painted and the outside
+of the panels polished to a high shine.  Again, these were stored until the
+engine assembly was complete, and then installed once all other assembly work
+was finished.
+
+By December 2011 the engine was nearly complete, with only a few sub-assemblies
+to be added.  The engine was installed in the engine frame and this allowed
+many pipes and hoses and electrical components to be added.  The supercharger
+was installed, which was another major milestone for the engine crew.  This
+went much quicker than installation on engine  Nr.1.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.45]{four-merlins-14.png}
+   %caption of the figure 
+   \caption*{\small \em  Engine Nr 2 being re-assembled in the engine shop.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+Engine Nr. 2 was completed by March 2012, with all assemblies and most cowl
+panels in place.  A lot of detail work was needed to ensure all pipes, hoses,
+clamps and other items were all installed exactly where they were originally. 
+Quality control was important to ensure accuracy of the final assembly.  The
+auxiliary gearbox was installed on the firewall, so final preparations could be
+made  to install engine  Nr. 2 and the propeller, before the aircraft was moved
+outside for Canada Day.
+
+It is worth noting that while Engine 1 required more than four years to
+complete the restoration, Engine 2 was completed in just 2 years.
+
+Finally, after six years of work by a host of dedicated volunteers, both
+restores engines were once again fitted to the North Star by May 2012 as shown
+in the photo below.
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{four-merlins-15.png}
+   %caption of the figure 
+  % \caption*{\small \em .}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+In our next edition we plan to tell the story of the restoration of engines 3
+and 4.
 

### 2020-09-26 pnsac/newsletter/article_Feature2.tex
Sep2020--first draft.

@@
 life, and were fitted to Transport Command's Avro Yorks. Long-range transport
-operation entailed running at relatively low cruise powers for long periods —
-under these conditions the lower cylinder head temperatures caused deposits of
+operation entailed running at relatively low cruise powers for long
+periods---under these conditions the lower cylinder head temperatures caused
+deposits of
 lead oxide from the fuel, resulting in excessive spark plug fouling. To counter
@@
 have said to TCA, "We didn't know the Merlin until you started operating it!"
-In the longer term, this paid off, TCA — later Air Canada — selecting
+In the longer term, this paid off, TCA---later Air Canada---selecting
 Rolls-Royce engines for its future fleet: Darts in Viscounts, Tynes in

### 2020-09-26 pnsac/newsletter/article_MemberInterview.tex
Sep2020--first draft.

@@
   \item Bill Tate
-  \item Tim Timmins
+  \item Tim Timmins (deceased)
   \item Giuseppe Zanetti

### 2020-09-26 pnsac/newsletter/article_PresReport.tex
Sep2020--first draft.

@@
 in touch with each other through the weekly Zoom meeting organized by the
-Museum's volunteer co-coordinator Cédric St.-Amour. This virtual meeting takes
+Museum's volunteer co-coordinator C\'edric St.-Amour. This virtual meeting
+takes
 place every Thursday at 10 AM EST; it is open to all our volunteers and any
@@
 
+\end{multicols}
+
+\begin{figure}[htbp]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.78]{north_star_welcome_1.png}
+   %caption of the figure 
+%   \caption*{\small \em  Mike Hope and Bruce Gemmill assembled Prop 1ready to
+%install on Engine 1 October 26, 2008.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:tim}
+\end{figure}
+
+\begin{multicols}{2}
+
 % End of text.

### 2020-09-26 pnsac/newsletter/article_ProgReport.tex
Sep2020--first draft.

@@
 %\subtitle{PNS 2016 Status Update}
-\author{R\'{e}jean Demers, , AME---Conservator \/ Special Project Manager
+\author{R\'{e}jean Demers, AME---Conservator \/ Special Project Manager
 Canada Aviation \& Space Museum}
@@
 
-When touring the shops in the Museum’s Conservation wing, the engine shop is
-reserved for what is referred to as, the “Crown Jewel”. Guests are lead through
+When touring the shops in the Museum's Conservation wing, the engine shop is
+reserved for what is referred to as, the "Crown Jewel". Guests are lead through
 the corridor and the lights are flicked on to reveal a shining piece of
@@
 Referencing an Ottawa Citizen article dated Saturday Feb. 1st, 2003. We read of
-an effort to restore the Canadair North Star to a version of her “former self”
+an effort to restore the Canadair North Star to a version of her "former self"
 as PNSAC founder Robert Holmgren put it. A sad, sorry state at the time.
@@
 enacted to summon the aviation gods or mojo hand placed upon the yoke to stir
-the spirits which have tilled the air.  Yet today, she’s a Hangar Queen,
+the spirits which have tilled the air.  Yet today, she's a Hangar Queen,
 patiently awaiting final installation of engine \#4. Is this sorcery? No,
-simply the fourth iteration of the Project’s labour of love; Rolls-Royce Merlin
+simply the fourth iteration of the Project's labour of love; Rolls-Royce Merlin
 622, Serial \#309573. The task of accomplishing the finale to a four part, long
@@
 plant restoration cycles, we see the evolution of the Project North Star
-volunteer organization since the early 2000’s. The forming of Engine Shop crews
+volunteer organization since the early 2000's. The forming of Engine Shop crews
 and the establishing of Crew Chiefs. The coordination of daily grunt work of
@@
 position (4th to the right). The engine had previously flown on 17514 until
-April 30th, 1965 when a broken intake valve \@ \#1 cylinder, “B” bank
+April 30th, 1965 when a broken intake valve \@ \#1 cylinder, "B" bank
 necessitated a repair. Flipping through pages, various signatures, authorities,
 stamps and markings show evidence of a storied service life going back to 1948.
-Engines are shared among power plant “skeletons” which in turn, are shared
-across the fleet. This was one of the major design features of the “Power Egg”
+Engines are shared among power plant "skeletons" which in turn, are shared
+across the fleet. This was one of the major design features of the "Power Egg"
 as the promotional material of the day had put it, interchangeability and ease
@@
 quick look under the cowling to reveal preserved markings and decals, evidence
-of original repair shop work was pointed out. “I probably worked on that job”
-he said. I understood at this point, how someone’s work, preserved in such a
+of original repair shop work was pointed out. "I probably worked on that job"
+he said. I understood at this point, how someone's work, preserved in such a
 way, could be of intangible value. The history coming around full-circle before
 my eyes, as the gentleman pointed out the ROLLS-ROYCE cast upon the rocker
-covers, saying “it was one of my first jobs, painting on those red letters”.
+covers, saying "it was one of my first jobs, painting on those red letters".
 
@@
 break now and then. Knowing that we had an impact and remembering the
-experiences we’ve shared can carry us through new challenges. It’s a pleasure
+experiences we've shared can carry us through new challenges. It's a pleasure
 for me to observe a new volunteer work towards acquiring the skills that may be
@@
 and other forms of reaching out to the public are excellent ways to remember
-one of Canada’s great contributions to air transport. With the tools at our
+one of Canada's great contributions to air transport. With the tools at our
 disposition, I would not hesitate to believe we can create many new and

### 2020-09-26 pnsac/newsletter/article_Special.tex
Sep2020--first draft.

--- 2013-10-17 pnsac/newsletter/article_Special.tex
@@
 
-\title{PNSAC Merchandise}
-%% \author{Drew Hodge}
+\title{Remembering Austin (Tim) Timmins 1928-2020}
+\author{Richard Lodge}
 
@@
 
-%\end{multicols}
-
-\begin{figure}[ht!]
+\begin{figure}[htbp]
    \vspace{2em}
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=1.0]{fleece_YOW7578.eps}
+   \includegraphics[scale=0.75]{tim-timmins.png}
    %caption of the figure 
-   \caption*{\small \em PNSAC winter apparel}
+   \caption*{\small \em Austin (Tim) Timmins.}
    %label of the figure, which has to correspond to \ref{}:
-   \label{fig:merchandise}
+   \label{fig:tim}
 \end{figure}
 
-The first frost warnings on the Weather Channel are reminders that
-fall and winter are just around the corner. Keep warm on those frosty
-fall days with a PNSAC toque and sweatshirt. We can also take
-individual orders for the micro fleece vests shown in the photo on the
-left, and if you're looking for early Christmas presents, how about a
-coffee mug or a t-shirt?
+We were all very sorry to hear that our second President and also Life Member,
+Tim Timmins, had passed away on September 7th. Tim was a founding member of the
+Project North Star Association of Canada and one of its original directors. He
+was an active member of the small group of aviation enthusiasts who worked for
+many months with the Canada Aviation \& Space Museum to negotiate a
+groundbreaking arrangement to restore Canada's sole remaining North Star
+aircraft to a standard suitable for display in the Museum.
 
-See our Facebook page,
-{\normalfont\color{blue}\texttt{\href{https://www.facebook.com/media/set/?set=a.610706705637878.1073741851.284829008225651&type=1}{PNSAC
-      Merchandise}}}, for details about what we're offering, and place your
-orders by sending an email message to the following address
-\email{pnsac.merchandise@gmail.com}.
+Tim became a dirty hands volunteer in the early years of the project and was
+involved in the often difficult and usually very dirty initial restoration work
+on the aircraft. He will have taken part in "bird nesting" when volunteers
+spent many hours removing years of bird nest debris from various parts of the
+aircraft.
 
-For members in the Ottawa-Gatineau region, you can arrange to pick up
-your merchandise at the Canada Aviation and Space Museum any Monday or
-Tuesday, between 10:00 AM and 3:00 PM. We can also ship your purchases
-to you for an extra fee. Please specify museum pickup or shipping with
-your order.
+Following the untimely death of our first President, Robert Holmgren in October
+2007, Tim agreed to become our second President. He held this position until
+June 2000 during which time he was instrumental in helping the Association move
+from a fledgling operation to a respected part of the Aviation Museum's
+volunteer team.
 
-Merchandise is also available during our Quarterly Meetings
-at CASM.
-
-Look forward to hearing from you.
-
-``The Merchandise Committee''
+Many of us will remember Tim's cheery smile and the long hours he spent talking
+to members of the public during the Museum's open days and public events. He
+always had anecdotes and stories to tell of his time in the RCAF as a navigator
+on the North Star.
 
@@
 \end{footnotesize}
-
-%\begin{multicols}{2}
 

### 2021-08-13 pnsac/newsletter/article_EditorsNotes.tex
Changes for issue 17-1

@@
 
-In this issue we are recognizing the completion of a very important step in the
-restoration of the North Star, namely, its Merlin engines, and our members who
-have volunteered with CASM to carry out the necessary work.  This is a very
-informative issue and I am sure we will all learn a great deal. I was not
-aware, until putting this edition together, that the Merlin designation was in
-keeping with the custom of Rolls Royce to name its aircraft engines after birds
-of prey.
+Unfortunately COVID-19 and its restrictions continue to hamper the restoration
+of the North Star. So it's a good time to look at what has been accomplished to
+date by the restoration team which consists of both Museum staff and volunteers.
+This edition will continue the focus on the Merlin engines. As many of you know
+many of the volunteers are also members of our organization.
 
-The article, "Four Merlins" provides the reader with a detailed history of the
-restoration of these engines.   Due to its length, the article will be divided
-into two editions;  starting with engines number one and two in this edition;
-and three and four in the next edition. We are fortunate to have members who
-have had  lengthy involvement with this part of the project. Sadly one of those
-members passed away recently. Inside you will find a tribute to Tim Timmins by
-the President of our Association.
+You will find photos of some of those volunteers in the technical article on
+engines three and four. You will also find a interview with Garry Dupont who has
+had a leading role in the engine shop work.
 
-We look forward to receiving your comments and suggestions regarding this
-edition and for future editions whether good, bad or ugly. Please use my email
-address located on the back page.
+In the next edition we expect to have more articles about our members
+experiences in the restoration project.
 
-Good reading.
+In the meantime if you have any comments on this edition or suggestions for
+future editions please send them to me.
 

### 2021-08-13 pnsac/newsletter/article_Feature1.tex
Changes for issue 17-1

--- 2020-09-13 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{North Star Propeller Restorations}
+\title{The Four Merlins: The Second Two}
 %\subtitle{Part 4}
-\author{Harry Hope, with assistance from Michael Hope.}
+\author{Based on Bruce Gemmill's notes; reviewed by Garry Dupont, and with editing assistance from Richard Lodge}
 
@@
 
-\textit{Hope Aero has played a major role in the restoration of the North Star.
-The following article provides some background about the company and its work
-on the North Star’s propellers. }\\
+% \paragraph*{Engine 3}
 
-Hope Aero was formally government approved to certify aircraft propellers in
-1992, at their facility in Mississauga, Ontario. Harry Hope however, already
-had 40 years of propeller overhaul experience on many varieties that included
-Hamilton Standard propellers, which were quite common, as well as Hartzell,
-McCauley, and Sensenich. Harry's first restoration to flying condition, was the
-propeller for the Lysander, which was restored by the Air Force in Winnipeg for
-their Centennial project in 1967. This Lysander is now displayed at CASM. Since
-that time, they have restored several propellers, some to flying condition,
-three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
-(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
-display at the Aviation museum in Boda, Norway.
+After its removal in the Spring of 2012, engine \#3 was completely stripped of
+all accessories, pipes and hoses, and disassembled.  The block, cylinders and
+pistons were cleaned, and work was begun to reassemble this engine.  It was at
+this point that a lack of volunteers started to slow down the restoration
+process.
 
- Since that time, their capabilities have added– MT, Dowty, Hoffman, and
-Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
-their customer base expanded to include different types of aircraft, the
-requirement for them to service more types became evident. They kept up with
-the required technology.
+The engine rebuild was nearing completion by April 2013, with magnetos and a few
+other accessories left to be installed.  The basic engine frame was back
+together.  The three radiator sections were cleaned and painted, then attached
+to the front of the frame.  The rear cowl ring had been disassembled, and new
+mounting hardware was fitted before the parts were painted and reassembled.  The
+front cowl assembly (radiator cover) was also cleaned, repaired and polished,
+ready to be installed.  Other cowl panels were then cleaned, repaired and
+polished, ready to install once the complete engine was in place.  Work was also
+underway on a host of pipes, cables, accessories, and fittings needed to
+complete the assembly of the engine and the Quick Engine Change (QEC) prior to
+installation on the aircraft.
 
-Along the way, they also expanded into aircraft Wheel and Brake overhaul,
-sales, service and on wing dynamic balancing of fixed wing and Helicopters.
-Harry retired in 2001, but the third generation is now capably looking after
-the business.
+By September 2013 engine \#3 was almost completely reassembled, with only the
+supercharger and inter cooler preheat assemblies left to install on the engine,
+which had been moved to the QEC engine frame in early summer.  The work slowed,
+due to summer down time for staff and volunteers. The auxiliary gearbox was
+removed from the engine firewall and was then being disassembled.  This was the
+last major assembly that had to be overhauled before the engine could be placed
+back on the aircraft.
 
-In the spring of 2006, Hope Aero took on the task of restoring the four North
-Star propellers, for the "North Star Restoration" project. The first challenge
-was to remove the propellers from the engines as neither the restoration crew
-nor Hope Aero had any recent experience removing this type of propeller.
-Michael Hope made the trip to Ottawa to assist with this operation and to help
-prevent any damage, which turned out to be exceedingly difficult due to the
-corrosion, starting with the spinner and then the propeller. Once the propeller
-was removed the blades were removed for shipping to Hope Aero.
+Engine \#3 was nearing completion by December 2013. The supercharger and inter
+cooler preheat assemblies had been installed, along with the propeller reduction
+gearbox.  While this work was being carried out by the engine crew, the
+remainder of the volunteer workforce busied themselves completing the numerous
+cowl panels, pipes, hose and other pieces needed to complete the engine.  Two
+items that seemed to take forever were the large steel exhaust shrouds.  A lot
+of hammering was needed to pound out years of dents, then numerous rivets and
+all the Dzus fasteners and springs needed to be replaced.  The complex shape of
+the exhaust shroud made this challenging work.  Further delaying completion was
+the search for suitable clear coat to protect the steel from corrosion.  The old
+clear coat used successfully on engines \#1 and \#2 was no longer available.  A
+suitable replacement was located, the shrouds were clear coated and serial
+numbers were stenciled on.
 
-The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
-Aero in 2006. It was completely disassembled, cleaned and inspected. The
-external parts were very badly corroded, with most damage being to the aluminum
-blades and dome shells, and rust to the steel hub and tail shaft, which was
-likely caused by the outside storage environment. The inside parts were in
-relatively good condition. It was established, that both technical and visual
-effort would be required to restore the propeller to an acceptable display
-condition.
+The auxiliary gearbox was disassembled, and significant corrosion was found on
+the bearings inside.  Bearings had to be ordered from England.  Meanwhile, the
+gearbox was cleaned and painted.  At the same time the large DC generator was
+dismantled, repainted and reassembled, as well as the air pump, and the
+tachometer generator and propeller synchronizer.  Once the new bearings were
+installed, the gearbox was completed, and the generator, and air pump were
+attached.  Finally, the gearbox was installed on the firewall awaiting the
+installation of the engine.
 
-The restoration work commenced with the blades, the alcohol de-icer feed shoes,
-troughs and paint being removed. The corrosion of the blade surfaces was such
-that removal by grinding was abandoned, as the inner half of the blades are
-shot peened and grinding would remove this and the authenticity. It was decided
-to media blast the complete airfoil surfaces with an aluminum oxide media, thus
-removing the surface corrosion but leaving a good surface for paint to be
-adhered to; an aviation grade grey paint was applied with yellow stripes at the
-tips. New alcohol feed shoes were installed, however the troughs, which direct
-the alcohol to the shoes, were not available. As the troughs are hidden inside
-the spinner, it was determined that they could be omitted.
-The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
-tubes, and dome retaining nut, were extremely rusty. They were glass bead
-blasted to remove the rust; this did not remove the heavy pitting but did leave
-a surface that could be cadmium plated.  After plating, a clear coat material
-was sprayed on these parts to improve the resistance to rust. The internal
-surfaces were in good condition, and they were well oiled.
+In April 2013 \#3 engine frame was completed after the installation of the engine
+and supercharger.  Fitting the many cowl panels proved difficult, due to the
+extensive rework that many of these items went through, causing some alignment
+difficulties when fitting these panels onto the engine frame.  This work was
+done in July 2014, and the completed engine was successfully installed on the
+aircraft.
 
-A replacement dome shell in relatively good condition was found, requiring only
-to be sanded, anodized, and be reinstalled with new seals.
+\begin{figure}[H]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{img-1-merlins-aug2021.png}
+   %caption of the figure 
+   \caption*{\small \em Volunteers Bill Tate and Garry Dupont discussing work on Engine 3.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:img1}
+\end{figure}
 
-The propeller was then completely assembled to confirm that all the refurbished
-parts fit together. The blades were then removed to facilitate being shipping
-back to the museum.
+% \paragraph*{Engine 4}
 
-All of the four propellers were  found to be in similar condition, and all were
-restored in a similar manner.
+The last engine to undergo restoration was removed iand work
+immediately began in the Engine Shop to disassemble the engine.  All cowl
+panels, pipes, hoses and accessories were removed and stored until space was
+available to restore these items. The engine was moved to the yellow rotating
+stand at this point. The priority was to disassemble the engine.]. The crank
+case was cleaned and clear coated before installing back into the yellow stand.
+The refurbished crankshaft was installed in the engine block, and the connecting
+rods and pistons cleaned and polished. On this engine we were able to remove the
+propeller shaft from the reduction gear box, unlike engine \#3 the shaft could
+not be removed.  Work was then begun on the more complex cylinder heads and
+valve train assemblies.
+
+Work then proceeded on much of the ancillary equipment for the engine. Once the
+frame was completed and placed on the blue stand, the engine block was removed
+from the yellow stand and installed in the frame. The supercharger \& inter
+cooler were completed and installed, as well as the reduction gearbox. The
+radiators and header tank were restored and painted, and two of the three
+radiator sections were installed. Restoration work was then begun on the air
+intake system, and many of the fuel and oil lines.  The starter motor was found
+to be very corroded and disassembly and repair not viable. A replacement unit
+was located and installed. The fire suppression lines were complete and would be
+installed along with the cowls when ready. Some of the cowl panels were
+restored, although it was found that this engine had more corrosion than the
+other three engines. Possibly it had been in service longer.  This meant that
+more work was required to refurbish each of the many panels.
+
+After restarting work on the aircraft in late January 2018, the conservation
+effort focused on the completion of engine number \#4 accessories and cowl
+panels.   Many of the cowl panels required extensive repairs, including making
+many new panel stiffeners, since the old ones were often badly corroded.
+Besides painting and polishing, the refurbished panels also required new part
+number and serial number stencils, and this led to investigating missing
+stencils from the same panels on earlier engines.
+
+Before final assembly could take place, the radiator cowl panel had to be fitted
+which was done by December. The panels were fitted and installed with some
+fasteners being replaced. The front support was installed for transportation.
+
+\begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{img-2-merlins-aug2021.png}
+    %caption of the figure 
+    \caption*{\small \em Engine 4 completed and ready to be moved to the Storage Hangar.}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img2}
+ \end{figure}
+
+ \begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{img-3-merlins-aug2021.png}
+    %caption of the figure 
+    \caption*{\small \em Some of the crew who worked on the restoration of Engine 4 (Richard Lodge, R\'{e}j Demers, Garry DuPont, Ted Devey, Charles Baril).}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img3}
+ \end{figure}
+
+On December 19, 2019, the engine, still on its blue stand, was towed by the
+Museum forklift, from the Restoration Shop to the Storage Hangar where it now
+sits next to the North Star prior to re-installation on the aircraft.
+
+\begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{img-4-merlins-aug2021.png}
+    %caption of the figure 
+    \caption*{\small \em R\'{e}j Demers (Museum Special Projects Manager) very carefully bringing the Museum forklift to lift one end of Engine 4 ready for moving from the Engine Shop to the Storage Hangar on 13 December 2019.}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img4}
+ \end{figure}
+
+ \begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{img-5-merlins-aug2021.png}
+    %caption of the figure 
+    \caption*{\small \em Engine 4 starts it long, slow, cold 20 minute journey from the Engine Shop to the Storage Hanger to join the North Star aircraft.}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img5}
+ \end{figure}
+
+ \begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{img-6-merlins-aug2021.png}
+    %caption of the figure 
+    \caption*{\small \em The Museum’s forklift, driven with great skill by  R\'{e}j Demers, hauls Engine 4 towards the Storage Hangar.
+The blue engine stand and its tiny wheels certainly was not made for racing!}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img6}
+ \end{figure}
+
+2020 Finally early in 2020, before the lock down,  propeller \#4 was assembled
+and stored on the refurbished prop stand in the Storage Hangar ready to mount on
+engine \#4.
+
+% \begin{figure}[H]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-1.png}
+%    %caption of the figure 
+%    \caption*{\small \em North Star being moved inside hangar 11 August 2005 .}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-2.png}
+%    %caption of the figure 
+%    \caption*{\small \em  Mike Hope and Bruce Gemmill assembled Prop 1ready to
+% install on Engine 1 October 26, 2008.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.55]{four-merlins-3.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 1 assembled in the Conservation shop Feb 2,
+% 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.55]{four-merlins-4.png}
+%    %caption of the figure 
+%    \caption*{\small \em Firewall ready, auxiliary gearbox installed Jan 7,
+% 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.55]{four-merlins-5.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 1 complete at the Restoration Shop Feb 9, 2010
+% .}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.45]{four-merlins-6.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 1 installed Feb 24, 2010. Shown top left to
+% bottom right, engine
+% supervisor Ted Devey,  Bill Tate, Ron Lemieux, Giuseppe Zanetti, John Corby,
+% Garnet Chapman, Volunteer Project Manager Jim Riddoch, John Tasseron, Bruce
+% Gemmill, Museum Project Coordinator Mike Irvin, Charles Baril, and then
+% Association President Tim Timmins.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \paragraph*{Engine 2}\
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-8.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 2 with cowl panels removed, April 6, 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-9.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 2 removed April 14, 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-10.png}
+%    %caption of the figure 
+%    \caption*{\small \em Stan Rideout working to reassemble Engine 2 in engine
+% shop, Oct 7, 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-11.png}
+%    %caption of the figure 
+%    \caption*{\small \em Austin (Tim) Timmins.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-12.png}
+%    %caption of the figure 
+%    \caption*{\small \em Jim Riddoch and Bill Tate fill the radiators with
+% inhibiting oil.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.7]{four-merlins-13.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 2 on rotating assembly stand.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.45]{four-merlins-14.png}
+%    %caption of the figure 
+%    \caption*{\small \em  Engine Nr 2 being re-assembled in the engine shop.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-15.png}
+%    %caption of the figure 
+%   % \caption*{\small \em .}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
 

### 2021-08-13 pnsac/newsletter/article_MemberInterview.tex
Changes for issue 17-1

--- 2020-02-24 pnsac/newsletter/article_MemberInterview.tex
@@
 \title{Our Members}
-\author{Interview with Garry Dupont}
+\author{This is the edited transcript of an interview with Garry Dupont by Richard Lodge.}
 
@@
 
-\end{multicols}
+\textit{Garry why did you join Project North Star?}
+
+I joined it when I was working full time and it was a discussion between me and
+my wife as I knew retirement was looming and it was what I was going to do. I
+had seen the article that was in the Ottawa Citizen with Robert Holmgren and Tim
+Timmins requesting people to attend a meeting starting the project, and my wife
+suggested I go to it, which I did and it gave me something to look forward to.
+Even when I was employed I was able to physically contribute by coming in the
+odd day and helping out with special projects and overhaul a few components and
+stuff like that.\\
+
+\noindent\textit{So you worked only in the engine shop?}
+
+No.  When initially I was working, I worked on various other things and only
+after I fully retired, I showed up and said Mike Irvin "I am here. I'll be here
+for four days a week".  That's when he asked me to go into the engine shop
+because they needed help in there.\\
+
+\noindent\textit{Could you have worked anywhere but you chose to work in the engine shop?}
+
+Actually, Mike chose for me.  Mike was the one who said go to the engine shop.\\
+
+\noindent\textit{When you started with Project North Star would you have felt
+that you had most experience of aviation work on engines, or could you gone
+anywhere on the plane?}
+
+I could have gone anywhere on the plane. Of this particular engine I had no
+knowledge. I had never worked on these liquid cooled engines before.  I never
+worked on a piston engine of that size before and actually I hadn't worked on
+piston engines for quite a number of years because with my employment most of my
+airplanes were with turbines.  We had some small piston engined airplanes of
+which I had limited experience.\\
+
+\noindent\textit{So quite a bit of your work in the engine shop was a learning
+experience apart from learning about how the Merlin 622 was assembled, you were
+also learning a bit about actually working on a large piston engine aircraft?}
+
+That's correct and the systems on this particular engine were completely
+different from anything I had ever worked on and I came across  an RCAF training
+manual, which somebody had donated and I took it home and read it from one end
+to the other over a couple of weeks. I was able to learn a considerable amount
+about the engine and understand the systems and it started making sense to me,
+so when you start assembling and disassembling and you look at a line you know
+what that the line is for. You know what the electrical connections are for. You
+know what system they belong to so it makes that part of the process much
+easier.\\
+
+\noindent\textit{I know from when we first started working together, you said
+you worked for Perimeter Airlines in Winnipeg.  When you were working for
+Perimeter, you were working mostly on Metroliners. Correct?}
+
+Yes.  We had 2 Metroliners when I was there, but we also had 4 piston operated
+Beechcraft Queenairs, we had 6 other Beechcraft.\\
+
+\noindent\textit{When you worked at Perimeter, were you working on any part of
+the plane and you weren't specialising in any particular part, so moving into
+solely engines was really a new area of work for you?}
+
+That's correct.\\
+
+\noindent\textit{When you first started working in the engine shop, who were you
+working with?}
+
+There was Ted Devey, John Tasseron, Michel Lacasse and two other
+fellows, the names escape me.\\
+
+\noindent\textit{I asked a few minutes ago.  If you were only interested in
+working on the engines.  Where else would you like to work on the plane? So you worked only in the engine shop?}
+
+Anything actually, I don't do well with sheet metal work, I have done some,
+minor bits. I could develop those skills a bit better.  Hydraulic systems; I
+have a fairly good interest in, landing gear, the flaps, the hydraulics and all
+those systems.  I have learned where the pumps are and how all that works now.
+I have also read up little bit on the landing gear to understand it a bit better
+and how it works and that would be interesting.\\
+
+\noindent\textit{So you could, if they had been doing the landing gear at the
+time you started have been quite interested in working on that side of things?}
+
+Yes.\\
+
+\noindent\textit{What would you say in retrospect was the most difficult part of
+the work you were having to do on engines 3 and 4?  Was it lack of paperwork,
+was it corrosion or what was it?}
+
+The most difficult part was dealing with corrosion because of the amount of
+corrosion in the engine and also the decision-making between my career and what
+I would do and what the museum wanted in restoration. The hardest learning part
+was learning about restoration and how to do it properly. I have had some good
+mentors.  and I have learned fairly well on what is required to do and what to
+look forward to in the future when planning the project; how to tackle it from a
+restoration perspective rather than a maintenance perspective.\\
+
+\noindent\textit{Was Mike Irvin one of your mentors helping you to know the
+various procedures needed for restoration?}
+
+Yes, Mike was the main mentor.  He was the one that took me under his wing and
+showed me the requirements and the basis of doing restoration.\\
+
+\noindent\textit{When you started working there, you weren't really in charge of
+the engine because it was a combination of you and Ted Devey who was still
+working there and who had worked on engines 1 and 2.  So you were working with
+him under Mike Irvin?}
+
+Yes.  That's right.\\
+
+\noindent\textit{You feel one of the most difficult parts was to make the shift
+from keeping an aircraft flying to restoring and conserving everything?}
+
+That's right, yes.\\
+
+\noindent\textit{Having done that and got into that sort of mindset, what would
+you feel was the most difficult part of doing the engine, of restoring the
+engines that you had to be involved with, or you had to work with.  Which part,
+I am thinking of?}
+
+The biggest part with the engines was the cleaning.  Believe it or not, the
+engines and the old oil had been sitting there for 35 years, it was very
+congealed and very difficult to remove.  Cleaning with solvents is not a
+pleasant thing to do and that was about the only method we could use was to
+clean and put it in the parts washer.  Once you put a part in the parts washer
+you also had to be very careful because if it was steel, when it was removed
+from the part washer it would flash freeze almost immediately, so it had to be
+removed and treated straightaway with some sort of preserving oil to prevent the
+surface from oxidizing.\\
+
+\noindent\textit{I very much remember John Tasseron working on the pistons and
+all work of getting the pistons out of the block  and then John working on the
+cleaning the piston rings.  Were you very much involved in that side of things?}
+
+Yes.  Cleaning up the piston rings was a real tedious job where you could spend
+a whole week on one piston ring trying to free it up by scraping away the
+corrosion, by tapping it to free the rings.  Sadly, a lot of them broke.  There
+was nothing we could do to free them up.  We tried various chemicals, soaking
+procedures, vibration procedures, putting them in sonic cleaners so that the
+vibration would free up the dirt, but nothing seemed to work really well, except
+getting in there with a little scraper and trying to scrape them clean.\\
+
+\noindent\textit{And when you did manage to free the parts and get them clean,
+you couldn't use a glass bead machine to do any cleaning because the parts were
+aluminum so we had to use Scotch Brite and other things to clean them?}
+
+Correct.  The purpose of glass beading is to remove corrosion and not to clean.
+If you glass bead a surface, it changes the outer texture of it considerably. In
+the engine shop we had at Perimeter, the only time they used glass beading was
+for removing the outside surface of the engine block and then they could prepare
+it for painting.\\
+
+\noindent\textit{Now that you have finished the main part of the work on engines
+3 and 4, what are the main additional parts or work still to be done on engines
+3 and 4 and maybe on engines 1 and 2?}
+
+R\'{e}j [Demers] wants to remove the engines so we will reinstall engine 4 when
+it is ready and then we will go back and will work on each engine successively
+and then work on the firewall and behind the firewall with the engine removed.
+At the time when we do this we will inspect the engines to see if corrosion has
+come back in some areas and I notice on some of the engines during Canada Day
+when we removed parts and cowlings to display the engine that there was
+corrosion showing up in certain areas where it had not been treated properly.
+So when we remove the engines, and R\'{e}j agrees with me, we will put them back
+in the engine shop and go over them thoroughly.  We have also found some issues
+with some of the cowl parts just not being adjusted properly.  So in the end, I
+would like to have all four engines installed on the wings properly treated so
+that we don't have to worry about them again and also the cowls adjusted and
+done properly so that at any time we can remove the cowl parts and reinstall
+them with ease.  So that in the future, if you want to display the engine again,
+you could effectively remove all the cowls on all four engines with ease and
+then reinstall them without making any big issue of it.\\
+
+\noindent\textit{Effectively you have got to remove any distortions in the cowl
+panels. So those are the things to be done?  When you are finished working on
+the Merlin engines, what would you like to do?  This is purely a personal thing
+and nothing to do with the Museum.  More engines on a different plane or working
+on the airframe?}
+
+It really doesn't matter to me, whatever the Museum sees fit as a project, I am
+willing to tackle it.
+
+% \end{multicols}
 
@@
 
-\begin{multicols}{2}
+% \begin{multicols}{2}
 
-1.\textit{What is your background in aviation?}
-
-I graduated from the Confederation College Aircraft Maintenance course
-in the Spring of 1975. After returning to my home town in the Red Lake
-District I gained employment with Red Lake Seaplane Service. After
-completing the course I had to apprentice for 18 months before writing
-the final exam for my Aircraft Maintenance Engineers' License. I
-successfully wrote my final exam in December 1976 and obtained my AME
-certification. In February 1977 I was hired by Perimeter Aviation in
-Winnipeg Manitoba. There I obtained my Turbine Engine and
-Pressurization Aircraft endorsements.  I started my 30 year career with
-the RCMP Air Services Branch in February 1981. My career took me to
-Edmonton Alberta, Goose Bay Labrador, Winnipeg Manitoba then finally to
-Ottawa. I finished my career as Quality Assurance Manager.
-
-2. \textit{How long have you been involved with Project North Star and how and
-why did you get involved?}
-
-I attended the very first meeting held for Project North Star. This was
-at my wife's suggestion that it would be a good retirement pastime.
-
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   %name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
-%   %caption of the figure 
-%   \caption*{\small \em Forward cockpit -- complete.}
-%   %label of the figure, which has to correspond to \ref{}:
-%   \label{fig:fwd_cockpit_complete}
-%\end{figure}
-
-3. \textit{What has been the history of your involvement to date?}
-
-While I was employed I worked an evening shift every 4 weeks so I would
-come in to CASM in the morning for a few hours. Upon my retirement Mike
-Irving had me work in the engine shop, which I have been doing since my
-retirement. I also have been on the board of directors for a number of
-years.
-
-%\begin{figure}[htbp]
-%	\vspace{2em}
-%	\centering
-%	%name of the graphic, without the path AND in EPS format:
-%	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
-%	%caption of the figure 
-%	\caption*{\small \em Propeller assembly.}
-%	%label of the figure, which has to correspond to \ref{}:
-%	\label{fig:propeller}
-%\end{figure}
-
-%\begin{figure}[htbp]
-%	\vspace{2em}
-%	\centering
-%	%name of the graphic, without the path AND in EPS format:
-%	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
-%	%caption of the figure 
-%	\caption*{\small \em Forward cockpit -- complete.}
-%	%label of the figure, which has to correspond to \ref{}:
-%	\label{fig:fwd_cockpit_complete}
-%\end{figure}
-
-4. \textit{What has been the highlight of your involvement?}
-
-The biggest highlight for me was installing number 2 engine. It was
-about half completed when I started in the engine shop. This is the
-largest engine I have ever installed in my entire career.
-
-%\begin{figure}[htbp]
-%   \vspace{2em}
-%   \centering
-%   %name of the graphic, without the path AND in EPS format:
-%   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
-%   %caption of the figure 
-%   \caption*{\small \em Navigator rack.}
-%   %label of the figure, which has to correspond to \ref{}:
-%   \label{fig:nav-radio-rack}
-%\end{figure}
-
-5. \textit{What has been the most challenging part of your involvement?}
-
-One of the most challenging parts was learning how the engine worked
-with all its systems. I found a RCAF training manual in the office and
-took it home. Over several months of reading I started to understand
-what made this monster of old technology tick. Another challenge was
-learning the British Hardware system and tools. A further challenge was
-reading  the old manuals which were written before they were
-standardized by the ATA chapter system which all aircraft manuals
-follow today. Learning the art of conservation was also a  challenge
-but was made easier by the guidance of Mike Irving.
+%1.\textit{What is your background in aviation?}
+%
+%
+%2. \textit{How long have you been involved with Project North Star and how and
+%why did you get involved?}
+%
+%3. \textit{What has been the history of your involvement to date?}
+%
+%4. \textit{What has been the highlight of your involvement?}
+%
+%5. \textit{What has been the most challenging part of your involvement?}
 

### 2021-08-13 pnsac/newsletter/article_PresReport.tex
Changes for issue 17-1

--- 2020-09-13 pnsac/newsletter/article_PresReport.tex
@@
 
-As always, before I start to write the Notes from the President, I reread my
-previous Notes to remind myself of what I had previously said in order to set
-the record straight if necessary. As I write this at the beginning of
-September, I am now aware of the impact of the COVID-19 pandemic on the
-restoration schedule for the North Star and the operations of your
-Association.
+Once again, we are experiencing summer with Covid restrictions in place and
+suspended North Star volunteer operations.  As I write this piece in the middle
+of July, we are seeing hopeful signs.  The Museum has reopened with reduced
+numbers of visitors and we are expecting to arrange an in person coffee morning
+in the middle of August for North Star volunteers outside or under the very
+large door of the Restoration hangar if the weather is bad.
 
-Since March, restoration work on the aircraft has been suspended and the
-activities of the Association have been limited. We have, however, been keeping
-in touch with each other through the weekly Zoom meeting organized by the
-Museum's volunteer co-coordinator Cédric St.-Amour. This virtual meeting takes
-place every Thursday at 10 AM EST; it is open to all our volunteers and any
-other PNSAC members who would like to join in.
+The Museum is reopening cautiously, and it will be at least September before the
+North Star restoration can recommence.  Assuming that the Ontario Covid
+infections remain low and the Ontario government rules allow it, we would also
+like to have our AGM indoors in the fall.
 
-As previously reported, the North Star project reached a major milestone with
-the completion of the restoration of Engine \#4 which was moved into the
-reserve hangar on a cold December 2019 morning.
+This issue of the Chronicle is mainly about the completion of restoration of
+engines 3 and 4. Some of our volunteers, however, have been busy during the
+spring and early summer locating and acquiring a used North Star radar.  This
+has now been delivered to the Museum.  In the next issue of the Chronicle, we
+hope to have the story of how John Makadi and Chris McGuffin negotiated the
+purchase of the radar and arranged for its transportation to Ottawa from
+Springfield, Missouri.  Quite a story.
 
-Since we are unable to physically meet and there has been no restoration work
-carried out, we felt that a special issue of the NSTAR Chronicle devoted
-entirely to the 15 year restoration of all four of the Merlin engines would be
-of interest to members.
+The crate containing the radar has not yet been opened.  We are expecting to do
+this at the same time as we gather for our in person coffee morning in August.
+We will be donating the radar to the Museum since it will be installed in the
+restored North Star.
 
-Understandably, with a project which has taken many years. the volunteers
-working on the engines have changed over time. Some have retired or stopped
-volunteering and sadly, some have died. Earlier this year we asked Ted Devey to
-contribute a piece but before writing about of his work, he passed away. Ted
-was involved with restoring all the engines and for several of the early years
-was the volunteer leading the restoration team. Ted was a great raconteur and
-our oldest volunteer at 93 years old. His recollections would have added
-considerably to the interest of this special issue of the NSTAR
-Chronicle.
+The used radar and its transportation costs have been financed in a very
+pleasant but unexpected way.  Earlier in the year, our Membership Secretary,
+Bruce Gemmill, at the request of the Board of Directors sent a message to all
+our members saying that for the year 2020/2021 the annual membership dues of
+\$25 would be cancelled and anyone who was a paid-up member in 2020 would
+continue their membership for another year.  The Board decided to do this
+because our members have received very little during the pandemic in return for
+their support and membership.
 
-After much consideration, the Museum has decided that volunteering cannot
-restart before January 2021. Until then we will continue with our regular Zoom
-meetings and will also organize a virtual Annual General Meeting for the
-Association. We will also continue to provide updates on the situation and much
-will depend upon how the different levels of government handle the ongoing
-recovery from the pandemic.
+Members will have been surprised to receive the letter from Bruce, which as
+always included a form for making donations.  Bruce decided that he would use a
+large number of unused small denomination postage stamps he had collected over
+many years.  As a result, all the letters had a rainbow of stamps affixed to
+them.  Bruce will have had a very dry mouth when he finished attaching all the
+stamps. A really nice idea by Bruce.
 
-There is an old saying "That it is an ill wind that blows nobody any good."
-This is very true even of this pandemic. One of the very pleasant results is
-having conversations with PNSAC members who don' t reside in the National
-Capital area and have sent membership renewals by mail. Where a renewal has
-been received and confirmation has been delayed as a result of the pandemic, I
-try to call a member on the telephone to explain what is happened and that the
-renewal has not been lost. As a result I have had many interesting phone calls
-with people whom I have never spoken to before. A definite bonus for me.
+Although Bruce's letter did not specifically ask for donations, many of our
+members decided to make donations to the project.  As a result we received over
+\$2000 in donations we were not expecting.  The receipt of these donations
+coincided with the decision to purchase the used radar from Missouri.  Although
+the final cost of the purchase and transportation of the radar has not been
+determined at the time of writing these notes, we are sure that the donations
+received as a result of Bruce's letter will probably cover the full cost of the
+acquisition.
+
+Throughout this pandemic, Cedric St Amour, our Volunteer Coordinator, has every
+Thursday morning hosted a Zoom virtual coffee morning.  This is normally
+attended by six or seven people and remains open to all volunteers and members.
+Anybody who has not been onto this meeting and would like to do so can contact
+me at \email{president@projectnorthstar.ca}. I will then send the necessary
+login information to you.
+
+As we gradually get ready for recommencing volunteer work on the Northstar, we
+will be talking with R\'{e}j Demers, CASM Special Project Manager, to organise a
+workplan for the next few months, particularly in view of the fact that
+restoration work on all four engines has mainly been completed.
 
@@
 \end{footnotesize}
+
+\end{multicols}
+
+\begin{multicols}{2}
 

### 2022-04-08 pnsac/newsletter/article_Feature1.tex
First set of changes for 18-1 issue, Apr.2022.

--- 2020-09-13 pnsac/newsletter/article_Feature1.tex
@@
 
-\title{North Star Propeller Restorations}
-%\subtitle{Part 4}
-\author{Harry Hope, with assistance from Michael Hope.}
+\title{APS 42 Radar -- Another Piece of the North Star Puzzle}
+
+%\author{Based on Bruce Gemmill's notes; reviewed by Garry Dupont, and with editing assistance from Richard Lodge}
 
@@
 
-\textit{Hope Aero has played a major role in the restoration of the North Star.
-The following article provides some background about the company and its work
-on the North Star’s propellers. }\\
+% \paragraph*{Engine 3}
 
-Hope Aero was formally government approved to certify aircraft propellers in
-1992, at their facility in Mississauga, Ontario. Harry Hope however, already
-had 40 years of propeller overhaul experience on many varieties that included
-Hamilton Standard propellers, which were quite common, as well as Hartzell,
-McCauley, and Sensenich. Harry's first restoration to flying condition, was the
-propeller for the Lysander, which was restored by the Air Force in Winnipeg for
-their Centennial project in 1967. This Lysander is now displayed at CASM. Since
-that time, they have restored several propellers, some to flying condition,
-three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
-(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
-display at the Aviation museum in Boda, Norway.
+\textit{The article below was written by our members John Makadi and Chris McGuffin.
+Chris and John played key roles in the acquisition of the APS 42 Radar for the North Star aircraft.
+In 2019 John was approached by Rejean Demers, CASM's project manager for the North Star restoration to see if he could locate the radar unit which was missing from the aircraft as explained below.
+John was successful and as a result in 2001  Rej asked Chris if  our association  could arrange for the acquisition of the unit.
+John and Chris coordinated its acquisition and what follows is a very fascinating story about what may be the only unit of its kind still in existence.
+The authors have been very modest about their contributions in dealing with a very challenging process made more difficult by the pandemic Covid-19.}
 
- Since that time, their capabilities have added– MT, Dowty, Hoffman, and
-Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
-their customer base expanded to include different types of aircraft, the
-requirement for them to service more types became evident. They kept up with
-the required technology.
+The North Star was the RCAF’s first strategic lift platform.
+It was dispatched across the country, and bridged the Atlantic and Pacific in service to Canada.
+In the late 1950s, the RCAF began upgrading the fleet with the addition of the APS-42 military navigation/weather radar.
+This radar was still sensitive equipment in 1966 when North Star 17515 was donated to the Canada Aviation and Space Museum.
+The radar was stripped out of the airframe along with the radios and other sensitive equipment.
 
-Along the way, they also expanded into aircraft Wheel and Brake overhaul,
-sales, service and on wing dynamic balancing of fixed wing and Helicopters.
-Harry retired in 2001, but the third generation is now capably looking after
-the business.
+Project North Star is frequently described as an effort to return 17515 to the condition she was in on her last day of service with the RCAF.
+The hunt was on.
+In 2019 Project North Star Association of Canada began searching for an APS-42 radar.
+Initial results were discouraging.
+US aviation museums with USAF aircraft had no components to offer.
+Further research of US military logistics documents revealed that US Department of Defense disposal instructions for APS-42 radar was to “Destroy” them due to their sensitive nature at the time.
 
-In the spring of 2006, Hope Aero took on the task of restoring the four North
-Star propellers, for the "North Star Restoration" project. The first challenge
-was to remove the propellers from the engines as neither the restoration crew
-nor Hope Aero had any recent experience removing this type of propeller.
-Michael Hope made the trip to Ottawa to assist with this operation and to help
-prevent any damage, which turned out to be exceedingly difficult due to the
-corrosion, starting with the spinner and then the propeller. Once the propeller
-was removed the blades were removed for shipping to Hope Aero.
+\begin{figure}[H]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{radar_the_unit.png}
+   %caption of the figure 
+   \caption*{\small \em Radar -- The Unit.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:img1}
+\end{figure}
 
-The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
-Aero in 2006. It was completely disassembled, cleaned and inspected. The
-external parts were very badly corroded, with most damage being to the aluminum
-blades and dome shells, and rust to the steel hub and tail shaft, which was
-likely caused by the outside storage environment. The inside parts were in
-relatively good condition. It was established, that both technical and visual
-effort would be required to restore the propeller to an acceptable display
-condition.
+We were disheartened by that discovery but certainly not defeated.
+Over the years, PNSAC volunteers have been resourceful about rescuing parts and assemblies.
+The search continued.
+In the spring of 2021, while continuing on-line research John stumbled across an eBay listing for an APS-42 transmitter out of Cross Timbers, Missouri (Ozarks).
+The seller, a military communications enthusiast, acquired the radar in a lot purchase of US Government surplus equipment.
+After negotiating an acceptable price for the radar, we just needed to get it to CASM.
 
-The restoration work commenced with the blades, the alcohol de-icer feed shoes,
-troughs and paint being removed. The corrosion of the blade surfaces was such
-that removal by grinding was abandoned, as the inner half of the blades are
-shot peened and grinding would remove this and the authenticity. It was decided
-to media blast the complete airfoil surfaces with an aluminum oxide media, thus
-removing the surface corrosion but leaving a good surface for paint to be
-adhered to; an aviation grade grey paint was applied with yellow stripes at the
-tips. New alcohol feed shoes were installed, however the troughs, which direct
-the alcohol to the shoes, were not available. As the troughs are hidden inside
-the spinner, it was determined that they could be omitted.
-The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
-tubes, and dome retaining nut, were extremely rusty. They were glass bead
-blasted to remove the rust; this did not remove the heavy pitting but did leave
-a surface that could be cadmium plated.  After plating, a clear coat material
-was sprayed on these parts to improve the resistance to rust. The internal
-surfaces were in good condition, and they were well oiled.
+\begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{radar_the_aps_42.png}
+    %caption of the figure 
+    \caption*{\small \em Radar -- The APS-42.}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img2}
+ \end{figure}
 
-A replacement dome shell in relatively good condition was found, requiring only
-to be sanded, anodized, and be reinstalled with new seals.
+Shipping freight is usually pretty straight forward.
+In the summer of 2021, there were “unusual” circumstances.
+Transportation was a bottle neck to many industries and costs rose across the carriers.
+We were unable to find a freight company willing to collect the radar in the seller’s remote location.
+Our seller was willing to deliver to Springfield, Missouri (for a fee) but would not provide a crate.
+Furthermore, the freight companies that offered custom crating services were prohibitively high.
+Then we made contact with the “Air and Military Museum of the Ozarks” (AMMO) – a small volunteer-based museum in Springfield.
+The volunteers at AMMO were sympathetic to our cause and enthusiastic to help.
+They agreed to receive our radar from the seller, build a custom crate and transfer it to the freight carrier.
+We were grateful to pay them a small honorarium which wasn’t much more than the cost of lumber for the crate.
+This was a wonderful example of museums cooperating across borders to preserve aviation heritage.
+AMMO ended up hosting the crated radar for several weeks.
+The hurdles of NAFTA attestations and the export of US military technology took additional bureaucratic kung-fu but the radar arrived safely in Ottawa on 10 June, 2021.
 
-The propeller was then completely assembled to confirm that all the refurbished
-parts fit together. The blades were then removed to facilitate being shipping
-back to the museum.
+ \begin{figure}[H]
+    \vspace{2em}
+    \centering
+    %name of the graphic, without the path AND in EPS format:
+    \includegraphics[scale=0.5]{radar_opening_the_unit.png}
+    %caption of the figure 
+    \caption*{\small \em Radar -- Opening the Unit.}
+    %label of the figure, which has to correspond to \ref{}:
+    \label{fig:img3}
+ \end{figure}
 
-All of the four propellers were  found to be in similar condition, and all were
-restored in a similar manner.
+After 17 months of social distancing, Project North Star volunteers and CASM staff gathered in the parking lot to celebrate the arrival of a missing radar.
+Customs seals were removed and a wood crate was opened to reveal our beautifully preserved APS-42.
+We look forward to seeing this new artifact in the nose wheel well of 17515.
+
+% \begin{figure}[H]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-1.png}
+%    %caption of the figure 
+%    \caption*{\small \em North Star being moved inside hangar 11 August 2005 .}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-2.png}
+%    %caption of the figure 
+%    \caption*{\small \em  Mike Hope and Bruce Gemmill assembled Prop 1ready to
+% install on Engine 1 October 26, 2008.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.55]{four-merlins-3.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 1 assembled in the Conservation shop Feb 2,
+% 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.55]{four-merlins-4.png}
+%    %caption of the figure 
+%    \caption*{\small \em Firewall ready, auxiliary gearbox installed Jan 7,
+% 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.55]{four-merlins-5.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 1 complete at the Restoration Shop Feb 9, 2010
+% .}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.45]{four-merlins-6.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 1 installed Feb 24, 2010. Shown top left to
+% bottom right, engine
+% supervisor Ted Devey,  Bill Tate, Ron Lemieux, Giuseppe Zanetti, John Corby,
+% Garnet Chapman, Volunteer Project Manager Jim Riddoch, John Tasseron, Bruce
+% Gemmill, Museum Project Coordinator Mike Irvin, Charles Baril, and then
+% Association President Tim Timmins.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \paragraph*{Engine 2}\
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-8.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 2 with cowl panels removed, April 6, 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-9.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 2 removed April 14, 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-10.png}
+%    %caption of the figure 
+%    \caption*{\small \em Stan Rideout working to reassemble Engine 2 in engine
+% shop, Oct 7, 2010.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-11.png}
+%    %caption of the figure 
+%    \caption*{\small \em Austin (Tim) Timmins.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-12.png}
+%    %caption of the figure 
+%    \caption*{\small \em Jim Riddoch and Bill Tate fill the radiators with
+% inhibiting oil.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.7]{four-merlins-13.png}
+%    %caption of the figure 
+%    \caption*{\small \em Engine 2 on rotating assembly stand.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.45]{four-merlins-14.png}
+%    %caption of the figure 
+%    \caption*{\small \em  Engine Nr 2 being re-assembled in the engine shop.}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
+
+% \begin{figure}[htbp]
+%    \vspace{2em}
+%    \centering
+%    %name of the graphic, without the path AND in EPS format:
+%    \includegraphics[scale=0.5]{four-merlins-15.png}
+%    %caption of the figure 
+%   % \caption*{\small \em .}
+%    %label of the figure, which has to correspond to \ref{}:
+%    \label{fig:tim}
+% \end{figure}
 

### 2022-04-08 pnsac/newsletter/article_PresReport.tex
First set of changes for 18-1 issue, Apr.2022.

--- 2020-09-13 pnsac/newsletter/article_PresReport.tex
@@
 \title{Notes from the President}
-\author{Richard Lodge}
+\author{Chris McGuffin}
 
@@
 
-As always, before I start to write the Notes from the President, I reread my
-previous Notes to remind myself of what I had previously said in order to set
-the record straight if necessary. As I write this at the beginning of
-September, I am now aware of the impact of the COVID-19 pandemic on the
-restoration schedule for the North Star and the operations of your
-Association.
+I showed up for the freebie, got an education and made some friends.
+That was my start in aviation.
+In 1985 I signed up for Air Cadets for something to do.
+One could say my mother made me.
+Certainly, she encouraged.
+The Cadets played sports and hosted camps; for free.
+I had no idea about the flight scholarships and other opportunities.
+Certainly, I was enamored with aviation.
+I built model airplanes and made a wind tunnel for a science fair.
+That wasn't much different from my sister's engagement with visual art or my brother's fascination with insects.
+My real passion was fixing things, improving them or making new things.
+My mum probably just wanted me to get out of the basement and socialize.
 
-Since March, restoration work on the aircraft has been suspended and the
-activities of the Association have been limited. We have, however, been keeping
-in touch with each other through the weekly Zoom meeting organized by the
-Museum's volunteer co-coordinator Cédric St.-Amour. This virtual meeting takes
-place every Thursday at 10 AM EST; it is open to all our volunteers and any
-other PNSAC members who would like to join in.
+Aviation was exciting but I never considered earning a living from it.
+I studied electrical engineering at the Royal Military College.
+I thought a little military service would be a precursor to advancement in the civilian tech industry.
+Instead of Nortel I ended up serving 28 years with the Canadian Forces.
+Many of my friends wore the blue uniform.
+I wore army green or tan depending on the season.
+When I flew it was usually to participate in a meeting, attend training or visit a forward operating base.
+Still, there can be a persistence to childhood interest.
 
-As previously reported, the North Star project reached a major milestone with
-the completion of the restoration of Engine \#4 which was moved into the
-reserve hangar on a cold December 2019 morning.
+My interest in wood and metal working matured as a hobby.
+I have a particularly thoughtful wife who accommodated a larger work space with every move.
+Nonetheless, a residential workshop can be a solitary and confined space.
+When I retired from military service in 2017 it was time to look for different ways to pursue my interests.
+It would be fun to learn from new people and with tools I didn't have access to at home.
+I could not imagine a better opportunity than an aircraft restoration.
 
-Since we are unable to physically meet and there has been no restoration work
-carried out, we felt that a special issue of the NSTAR Chronicle devoted
-entirely to the 15 year restoration of all four of the Merlin engines would be
-of interest to members.
+My timing lined up with Project North Star, a comprehensive restoration effort: metal fabrication, engine repair, refinishing, reupholstery, overhauling mechanical systems, woodworking and more.
+Some volunteers contribute decades of skill and expertise while others offer enthusiasm and commitment to learn.
+Everyone benefits from cross training and camaraderie.
+Progress is important but not more than the historical value and preservation of the artefact.
+Each step is supervised by our project manager, an aircraft maintenance engineer employed by the Museum.
 
-Understandably, with a project which has taken many years. the volunteers
-working on the engines have changed over time. Some have retired or stopped
-volunteering and sadly, some have died. Earlier this year we asked Ted Devey to
-contribute a piece but before writing about of his work, he passed away. Ted
-was involved with restoring all the engines and for several of the early years
-was the volunteer leading the restoration team. Ted was a great raconteur and
-our oldest volunteer at 93 years old. His recollections would have added
-considerably to the interest of this special issue of the NSTAR
-Chronicle.
+Initially, I was assigned to reproduce and restore the wood paneling that lined the interior of our North Star.
+Occasionally I helped with other assignments.
+I watched as replacement parts were fabricated, engine \#4 was reassembled and bird nests were pulled out of the tail section.
+In 2018 I began publishing progress and event updates on the Project Facebook and Instagram pages.
+Many people enjoy seeing airplanes up close.
+Airshows are popular for good reasons.
+I certainly look forward to the time I spend at the aviation museum, particularly the areas normally hidden from public view.
+Taking a few photos to share with our social media audience was a natural fit for me.
+Those updates have been few during the pandemic but they will resume imminently.
 
-After much consideration, the Museum has decided that volunteering cannot
-restart before January 2021. Until then we will continue with our regular Zoom
-meetings and will also organize a virtual Annual General Meeting for the
-Association. We will also continue to provide updates on the situation and much
-will depend upon how the different levels of government handle the ongoing
-recovery from the pandemic.
-
-There is an old saying "That it is an ill wind that blows nobody any good."
-This is very true even of this pandemic. One of the very pleasant results is
-having conversations with PNSAC members who don' t reside in the National
-Capital area and have sent membership renewals by mail. Where a renewal has
-been received and confirmation has been delayed as a result of the pandemic, I
-try to call a member on the telephone to explain what is happened and that the
-renewal has not been lost. As a result I have had many interesting phone calls
-with people whom I have never spoken to before. A definite bonus for me.
+2022 is going to be a year of renewal in many ways.
+In January I assumed the presidency of the Association.
+Our previous president, Rodger Lodge, held the position for nearly 12 years.
+Roger has been involved with the Association for 20 years and I am thankful that he has agreed to stay on the Board of Directors.
+The Canada Aviation and Space Museum is actively renovating the conservation workshop.
+Plans for a return of volunteers to the Museum are being developed and restoration activity should resume in the months ahead.
+I am grateful for the encouragement of our supporters and the tenacity of our volunteers.
+We'll be back to the restoration effort soon.
 

### 2022-05-20 pnsac/newsletter/article_EditorsNotes.tex
Updated for final May 2022 issue.

@@
 
-Unfortunately COVID-19 and its restrictions continue to hamper the restoration
-of the North Star. So it's a good time to look at what has been accomplished to
-date by the restoration team which consists of both Museum staff and volunteers.
-This edition will continue the focus on the Merlin engines. As many of you know
-many of the volunteers are also members of our organization.
-
-You will find photos of some of those volunteers in the technical article on
-engines three and four. You will also find a interview with Garry Dupont who has
-had a leading role in the engine shop work.
-
-In the next edition we expect to have more articles about our members
-experiences in the restoration project.
-
-In the meantime if you have any comments on this edition or suggestions for
-future editions please send them to me.
+This edition marks a change of leadership of our association.
+Richard Lodge, our longtime president, retired at the end of 2021.
+His replacement is Chris McGuffin.
+Richard's reminiscences of his time as president can be found in the Our Members column.
+While Richard's story talks about his past experiences, Chris' Notes from the President sets out his vision for the future of the organization.
+R\'{e}jean Demers, CASM's Project Manger for the North Star also addresses, in the Conservator's Corner column,  how he envisages the Project restarting this year.
+Finally, there is a fascinating article by John Makadi and Christ McGuffin about how they found and acquired an APS 42 radar unit for the aircraft.
+Good reading.
 
@@
 
+\noindent As always comments and suggestions for future editions are welcomed.
+
 \begin{footnotesize} \raggedleft PNSAC\\
 \end{footnotesize}
+
+\vspace{60pt}
 

### 2022-05-20 pnsac/newsletter/article_Feature1.tex
Updated for final May 2022 issue.

@@
 Chris and John played key roles in the acquisition of the APS 42 Radar for the North Star aircraft.
-In 2019 John was approached by Rejean Demers, CASM's project manager for the North Star restoration to see if he could locate the radar unit which was missing from the aircraft as explained below.
-John was successful and as a result in 2001  Rej asked Chris if  our association  could arrange for the acquisition of the unit.
+In 2019 John was approached by R\'{e}jean Demers, CASM's project manager for the North Star restoration to see if he could locate the radar unit which was missing from the aircraft as explained below.
+John was successful and as a result in 2001  R\'{e}j asked Chris if  our association  could arrange for the acquisition of the unit.
 John and Chris coordinated its acquisition and what follows is a very fascinating story about what may be the only unit of its kind still in existence.
@@
    %name of the graphic, without the path AND in EPS format:
-   \includegraphics[scale=0.5]{radar_the_unit.png}
+   \includegraphics[scale=0.45]{radar_the_unit.png}
    %caption of the figure 

### 2022-05-20 pnsac/newsletter/article_PresReport.tex
Updated for final May 2022 issue.

@@
 In January I assumed the presidency of the Association.
-Our previous president, Rodger Lodge, held the position for nearly 12 years.
-Roger has been involved with the Association for 20 years and I am thankful that he has agreed to stay on the Board of Directors.
+Our previous president, Richard Lodge, held the position for nearly 12 years.
+Richard has been involved with the Association for 20 years and I am thankful that he has agreed to stay on the Board of Directors.
 The Canada Aviation and Space Museum is actively renovating the conservation workshop.

### 2023-05-09 pnsac/newsletter/article_MemberInterview.tex

--- 2020-09-13 pnsac/newsletter/article_MemberInterview.tex
@@
 \title{Our Members}
-% \author{Interview with Garry Dupont}
+\author{Richard Lodge.}
 
 \maketitle
+
+Last New Year's Eve, December 2021, I became the third past president of PNSAC, after 12 years in the position and following several years at treasurer.
+I was delighted when Chris McGuffin agreed take on the role of president following the unanimous support of the board of directors.
+
+Before starting to write this article, I made a list of some of the more important and memorable events of my presidency.
+This article is my recollection of the things that have stuck in my mind over the years.
+During my term in office, PNSAC changed from being an experimental volunteer organization which still had to prove itself to the Canadian Aviation and Space Museum (CASM), to being part of the CASM family where PNSAC members are well respected and trusted by the museum staff and management.
+
+Several events stand out when I think about my years as president.
+The first was on Canada Day 2017 shortly after R\'{e}j Demers had arrived at CASM to take over the position of Special Project Manager.
+The ex RCAF Hercules C130 had arrived in April 2016 and was parked on the apron outside the Storage Hangar.
+A large crowd gathered around the aircraft when the day was at its it hottest.
+Standing in the middle of the cargo ramp was R\'{e}j where for half an hour or so he entertained the crowd.
+The event was unplanned, and his comments were unscripted.
+The crowd just stood there enjoying the informality of listening to what he was saying and answering questions shouted to him.
+
+The second event was in December 2019 when Engine \#4 was moved from the Engine Shop to be located in front of the North Star in the Storage Hangar.
+It was a cold early winter day with snow flurries.
+On this occasion the engine could only be moved on its stand which had four wheels and no steering.
+It was towed by R\'{e}j driving CASM's forklift from the staff parking lot along the side of the building and into the Storage Hanger.
+It was a very slow process taking something like half an hour.
+By the time we got the engine safely to bed the three engine shop volunteers (me, Garry Dupont, and Charles Baril) as well as other volunteers who were walking behind the engine were really not enjoying the beginning of a Canadian winter.
+R\'{e}j who was sitting in the heated cab of the forklift, with a nice smile on his face, was actually quite warm!
+
+\begin{figure}[H]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.75]{engin_4.png}
+   %caption of the figure 
+   \caption*{\small \em From left to right: Garry Dupont, R\'{e}j Demers (in the cab), and Chris McGuffin.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:img1}
+\end{figure}
+
+The next event was when I was MC at the showing of the film  made in 2014 when the  Lancaster bomber, part of the collection of the Canadian Warplane Heritage Museum, flew to England to join the only other airworthy Lancaster in the world for a summer tour of the airshows in Britain.
+The epic flight was filmed throughout its three-day journey to England, during its time in England and on its return.
+Bill Tate, one of our very active members, arranged for the film to be shown in the auditorium of the CASM to about 250 members of the general public.
+The film itself was very interesting, but what made it even more interesting was that the flight crew agreed to come to Ottawa and talk to the audience after the showing of the film.
+Hearing the flight crew talk of the preparations and difficulties of the tour and knowing how successful it had been was particularly interesting to me.
+I had managed to attend the first airshow in England when the two Lancasters, a Hurricane and two Spitfires had flown together off the south coast of England, near where they would regularly pass on bombing missions during World War 2.
+The sound of 11 Merlin engines flying together reminded me of my boyhood in England when I would hear the bombers leaving their bases in Yorkshire to bomb Germany – quite an emotional sound all these years later.
+
+There are always difficulties and problems to deal with when one is the head of any organization, be it in a paid position or as a volunteer.
+PNSAC has experienced a  number of challenges during it existence.
+We have had two major closures and a lock down and closure caused by the Covid 19 pandemic.
+The Covid 19 closure needs no explaining but the previous two were necessitated by CASM staging major events in 2013 and 2016 around the Storage Hangar.
+While the closures were disruptive to the Project North Star operations, the presentation of the Star Wars and Star Trek exhibits allowed CASM to reach a broader audience and raise awareness of CASM.
+The exhibits made it impossible for the restoration team to work on the North Star aircraft for many weeks in both years.
+The challenge was to try to keep the restoration volunteers together when some volunteers were finding other ways to occupy their time.
+Many meetings and discussions took place, and another side of my role was to help the restoration volunteers understand that these closures were only temporary and that as soon as the spectacle events were finished, restoration work would start on the aircraft again.
+Fortunately, we did not lose many restoration volunteers although our actual non-volunteer membership dropped considerably.
+
+During the summer of 2018 it was decided that the vertical fin of the North Star should be removed.
+This involved weeks of preparatory work inside the aircraft, loosening the many nuts and bolts securing the fin to the aircraft.
+On the appointed day a large mobile crane arrived to remove the vertical fin.
+The rental cost was financed by PNSAC donations received from members.
+It was a hot day, and the crane was duly stationed beside the aircraft.
+However, several of the securing bolts had been left in place and needed to be removed once the crane had been attached to the fin.
+Several volunteers assisted in the attaching of the crane hook to the fin, but one volunteer, Charles Baril, offered to crawl inside the plane and remove the last remaining bolts.
+The inside of the plane was like an oven with no ventilation and no proper lighting.
+The only way Charles could communicate with the crew outside was by shouting.
+After about an hour of very difficult work, Charles shouted that all the nuts and bolts had been undone and the actual lifting of the fin could proceed.
+After the fin had been lifted by the crane, it was lowered into a specially prepared cradle.
+Charles emerged from the aircraft to the applause of all people outside, looking as if he was about to melt and thoroughly tired out by his exertions.
+
+\begin{figure}[H]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.5]{nstar_tail_fin.png}
+   %caption of the figure 
+   \caption*{\small \em The North Star with its tail fin removed.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:img1}
+\end{figure}
+
+ As president, I was involved in several other memorable events organized by Bill Tate.
+Bill has an amazing ability to organize events and had a substantial list of contacts.
+I was primarily involved as the MC of the trips while in the buses, also providing a backup for Bill should he encounter any problems.
+Twice we visited RCAF Trenton and once we went to Montréal and Mirabel to see the Canadair/Bombardier production line in action.
+The most memorable was the weekend bus trip to Hamilton to attend the airshow organized by the Canadian Warplane Heritage Museum on the Saturday and on the Sunday, a visit to the Air Canada control centre in Toronto.
+On the Friday evening, Bill organized a visit to the vineyards in Niagara-on-the-Lake.
+Sunday morning saw us all back in the bus heading for downtown Toronto, where we breakfasted in the brewery district before our Air Canada visit.
+Many members will have great memories of these times, which all took place prior to the closures I referred to earlier in this article.
+
+\end{multicols}
+
+\begin{figure}[H]
+   \vspace{2em}
+   \centering
+   %name of the graphic, without the path AND in EPS format:
+   \includegraphics[scale=0.4]{july2012.png}
+   %caption of the figure 
+   \caption*{\small \em Hamilton trip.}
+   %label of the figure, which has to correspond to \ref{}:
+   \label{fig:img1}
+\end{figure}
+
+\begin{multicols}{2}
+
+One of the best aspects of assuming responsibility as a volunteer is that the more one puts into it, the more pleasure and interest one gets from the work.
+As president, I regularly met all the senior management of CASM and many other people outside CASM.
+Often, I was involved in discussions about the future direction of PNSAC, which of course involved discussing future and sometimes confidential CASM plans and operations.
+
+A very small incident, probably four or five years ago, really demonstrated to me how as a volunteer association we had "arrived".
+My predecessor presidents, Bob Holmgren and Tim Timmins had to overcome the skepticism of CASM management as to whether it would be possible for us to sustain a continuing volunteer effort.
+By the time I became president in 2010, CASM management had accepted that our members would work to a very high standard of restoration work and would continue to put in several thousand hours of painstaking and sometimes difficult work every year.
+The incident I am remembering best is a particular day when I went upstairs to the office level  at CASM and saw a group of CASM staff talking in the corridor.
+I stopped and said hello to them and then started to excuse myself by saying that they were having a discussion and I would get out of the way.
+Chris Kitzan, the CASM Director General, looked at me and said "There was no need to go away Richard, you are part of the family".
+This for me was a watershed moment and one I shall always appreciate.
+
+Looking towards the future, as the pandemic begins to be contained, we must get the restoration operation going again and try to recruit more members and restoration volunteers.
+I feel very confident that Chris McGuffin and the other directors of PNSAC will be able to do this and continue to help CASM in volunteer work in addition to North Star restoration.
+I look forward to continuing volunteering, but definitely not climbing around the top of the aircraft in a full safety harness!
 
@@
 % \begin{multicols}{2}
-
-Readers of the Chronicle will know that this item usually refers to one of our
-members. This edition, however, salutes all of our members who worked as
-volunteers on the restoration of the North Star engines namely:
-
-\begin{itemize}
-  \item Charles Baril 
-  \item John Corby (deceased)
-  \item Ted Devey (deceased)
-  \item Ed Hogan
-  \item Peter Houston
-  \item Ron Lemieux
-  \item Richard Lodge
-  \item Wib Neal
-  \item Jim Riddoch
-  \item Stan Rideout
-  \item Ted Slack
-  \item John Tasseron
-  \item Bill Tate
-  \item Tim Timmins
-  \item Giuseppe Zanetti
-\end{itemize}
-
-The following members have provided comments on their personal experiences.
-
-\textbf{Charles Baril}
-
-Working on all 4 Merlin engines has been an extremely rewarding experience that
-has taught me a lot about older technology, an appreciation for history
-involved (and those in the engine crew with that knowledge) and learning new
-skills and techniques. Although most of the parts were auxiliary systems,
-seeing them as part of the full engine gives a sense of accomplishment. It is
-very satisfying to see a part going from the dirty, oil covered, rusty part to
-newly refurbished one that looked like it was overhauled in an RCAF maintenance
-depot. It has been an honour to be part of this project and restoration of
-these pieces of history.
-
-In particular I remember working on the lower cowling flaps.  Trying to undo
-the inner skin was difficult due to longer rivets through the mounts for the
-Flaps. It took several tries and a few extra holes, but it was done. Fixing
-those holes often meant using original methods which was revealing and a
-commitment to historical methods.
-
-I also recall the work on the filter ducts.  The filter duct is the last stage
-of the air intake before the supercharger. During its cleaning, there was a
-starling nest on top of one filter. This caused a lot of corrosion damage that
-needed to be removed and treated. Yet, what got me was that starlings could get
-that far into the engine.
-
-\textbf{Richard Lodge}
-
-As probably the only volunteer who ever worked in the Aero Engine Division of
-Rolls-Royce in Derby, England, I was greatly interested in becoming involved in
-Project North Star, even though, as an accountant, I had never worked on the
-design or building of an aeroengine.  By the time I started work at Rolls-Royce
-in 1963, new build of Merlin engines had ceased, and only spare parts were
-being manufactured.
-
-I have several strong memories of working on the Merlin engine.  In general,  I
-particularly remember John Tasseron working endlessly on freeing seized piston
-rings.  I learned much from John about patiently working at something which
-appeared to be impossible! Later, when working with  R\'{e}j Demers and Garry
-Dupont, I started to learn how to rivet and how to lock wire. The most
-important lesson for me has been, photograph, photograph, photograph, at each
-stage of dismantling, cleaning and reinstalling an engine part. When sitting in
-an overcrowded office at Rolls-Royce, it never occurred to me that at some time
-in the future I would actually be working on a Merlin engine or working my way
-through an original parts manual to work out how to reassemble an item.  Life
-is certainly full of interesting surprises.
-
-\textbf{John Tasseron}
- 
-The list below is intended to give a brief overview of the various highlights
-that arose during my participation in the restoration of the second, third and
-fourth Merlin engines:
-
-\begin{enumerate}
-  \item It's not every day one gets the opportunity to see up close what makes
-the Rolls-Royce 620 Series Merlin engine tick!
-  \item It's a real challenge to safely lower each "power egg" down off the
-wing, place it in "the big blue stand" and move it in to the workshop.
-  \item Then starts the process of carefully separating about 11,000 individual
-parts before cleaning, inspecting and re-conditioning them as necessary. Do not
-lose any and make sure everything is photographed for museum records!
-  \item When things unexpectedly go wrong, the real expertise kicks in to
-perhaps develop a repair scheme or even manufacture a brand-new replacement
-part. The objective is to bring everything back together again and end up with
-a "like new" engine, ready for preservation.
-  \item It all takes team effort based on effective communication and
-co-operation. There are no dumb questions and the answers are all aimed at the
-challenge to create an "airworthy" product that will never fly.
-  \item Finally, its important and fun to share the project with a curious
-public, so be ready to answer any questions that might be asked about the
-aircraft and its engines!
-\end{enumerate}
 

### 2023-05-09 pnsac/newsletter/article_ProgReport.tex

@@
 
-When touring the shops in the Museum's Conservation wing, the engine shop is
-reserved for what is referred to as, the "Crown Jewel". Guests are lead through
-the corridor and the lights are flicked on to reveal a shining piece of
-aeronautical engineering and craftsmanship. The Merlin 622 engine, one of four
-such power plants, sits resplendent. Gasps and expressions of delight from
-guests are guaranteed to insight a prideful glance upon this piece of work by
-volunteers and staff alike. Appearing as if the engine had just been delivered
-from Bristol Aero Ltd. is an illusion. What is this strange magic? How can such
-beauty manifested though steel and aluminum even exist? As any good sorcerer
-will tell, illusion is lost upon the dissolution of mystery. Yet this piece of
-work, a restoration of great effort, time and detail, cannot shed the cloak of
-timelessness so easily. The result of thousands of hours culminate through
-unrelenting intrigue, garnered by all who cast eyes upon the Merlin. These
-words simply scatter as embellishments to the enduring design and purpose,
-which cannot be hidden in all evidence of its greatness.
+National Volunteer Week was held recently. It was a timely call to action. Let's focus our collective strengths and resources to prepare for Project North Star RESTART 2022.
 
-Referencing an Ottawa Citizen article dated Saturday Feb. 1st, 2003. We read of
-an effort to restore the Canadair North Star to a version of her "former self"
-as PNSAC founder Robert Holmgren put it. A sad, sorry state at the time.
-Destitute to serve as a bird house to the starlings. No mention of any spells
-enacted to summon the aviation gods or mojo hand placed upon the yoke to stir
-the spirits which have tilled the air.  Yet today, she's a Hangar Queen,
-patiently awaiting final installation of engine \#4. Is this sorcery? No,
-simply the fourth iteration of the Project's labour of love; Rolls-Royce Merlin
-622, Serial \#309573. The task of accomplishing the finale to a four part, long
-haul repeat restoration of a power plant began in 2014. Looking back at power
-plant restoration cycles, we see the evolution of the Project North Star
-volunteer organization since the early 2000's. The forming of Engine Shop crews
-and the establishing of Crew Chiefs. The coordination of daily grunt work of
-condition recording, treatment and reassembly of components. Today, members of
-the project continue to develop ideas, adapting to new technology, finding
-different approaches and maintaining a solid work ethic. This spirit remains
-unchanged throughout difficult times and challenges we have encountered.
+It has been roughly 20 years now, since the beginning of the North Star restoration project.
+Over the past two years, we've had to endure the effects of the CoVID-19 global pandemic.
+This spring, we have an opportunity to make an attempt to (gradually) return to normal operations.
+Ingenium is currently initiating actions that will permit the return of Project North Star volunteers.
+This will require a concerted effort on behalf of all stakeholders.
 
-The Tech. log states: Engine installed on North Star 17515 on August 23rd, 1965
-at airframe Total Time of 19845.45 Hrs. with 1121 hours remaining in \#4
-position (4th to the right). The engine had previously flown on 17514 until
-April 30th, 1965 when a broken intake valve \@ \#1 cylinder, "B" bank
-necessitated a repair. Flipping through pages, various signatures, authorities,
-stamps and markings show evidence of a storied service life going back to 1948.
-Engines are shared among power plant "skeletons" which in turn, are shared
-across the fleet. This was one of the major design features of the "Power Egg"
-as the promotional material of the day had put it, interchangeability and ease
-of maintenance. A glance at the log tells me this engine serial number served
-on North Stars 506, 510, 508 and 502 to name a few. The last remaining example
-of its type, North Star 17515 holds pieces of the past, belonging to aircraft
-that have long been gone. Like crew members who have also shared their time
-amongst a fleet that traveled all over the globe, with change-overs,
-turn-arounds and plenty of distance in-between.
+My intent is to have crews receive training, gradually integrating in numbers, until we could finally support full daily compliments based on operational capacities.
+Now, with the demands imposed by Ingenium's recent acquisition approval of an RCAF CC-115 Buffalo, we must act upon this news in conjunction with a collection re-org.
+This signifies an earlier than expected requirement for Project North Star volunteer support, due to the increased scope of work.
 
-One of the big pay-offs for members of Project North Star is the opportunity to
-display the work carried out by volunteers. Many occasions come to mind, such
-as Canada Day, Doors Open Ottawa and family visits to the Museum. One
-opportunity, of which I will always be grateful to have participated in, was
-such a moment. Myself and a few volunteers met a retired employee of the
-aforementioned Bristol Aero Industries Ltd. who worked at the overhaul shops
-when this engine was undergoing acceptance runs in the early to mid-sixties.
-Quickly, job experience and familiarity with the object was brought forward. A
-quick look under the cowling to reveal preserved markings and decals, evidence
-of original repair shop work was pointed out. "I probably worked on that job"
-he said. I understood at this point, how someone's work, preserved in such a
-way, could be of intangible value. The history coming around full-circle before
-my eyes, as the gentleman pointed out the ROLLS-ROYCE cast upon the rocker
-covers, saying "it was one of my first jobs, painting on those red letters".
+CASM Conservation is preparing a work schedule reliant upon this volunteer support, in order to mobilise the North Star from the Reserve Hangar.
+Moving forward, through consultations with all parties, obtaining guidelines and limitations, as well as setting goals and expectations so that we may provide an integrated work plan for 2022-2023.
 
-These records, object qualities, historic log entries and our treatment reports
-all add up. We string together these bits of information to form longer chains
-that communicate the history of an artifact. As focal points shift towards
-other areas of interest, or priorities, work carried out in the past helps
-maintain a continued relevance for others to pick-up where we left off. With
-such a long term project, setting the pace is just as important as taking a
-break now and then. Knowing that we had an impact and remembering the
-experiences we've shared can carry us through new challenges. It's a pleasure
-for me to observe a new volunteer work towards acquiring the skills that may be
-passed on to others, like an old wizard speaking to his young apprentice.
+High on our list of immediate concern is rendering 17515 mobile so that we may conduct a massive hangar shuffle.
+Collection objectives are driving CASM Conservation Services (and Project North Star) into a 24 month action plan which will eventually see us dispose of some artifacts and acquire one big yellow beast (115452).
+Thankfully, I have been able to secure accommodation within the storage hangar for the North Star between brief stints outdoors during fair weather movements, in amongst the storage hangar and the main museum.
 
-I encourage Project North Star members to communicate their interest in the
-project to the public. Generate their own records that may contribute to the
-history of the machine and the lives it affected. This newsletter, social media
-and other forms of reaching out to the public are excellent ways to remember
-one of Canada's great contributions to air transport. With the tools at our
-disposition, I would not hesitate to believe we can create many new and
-engaging experiences for the public to learn about the role this aircraft
-undertook while in service. Now, in continuation with a sense of national
-pride, the North Star remains an important platform for interpretation of the
-work carried out by dedicated volunteers supported by a museum that understands
-the value of such an artifact.
+Project North Star has an important role in reinvigorating the museum ecosystem, providing a necessary boost of youthful vigour and effervescence.
+You may be asking yourself if during this prolonged absence, a time machine or temporal portal was made available for us to return to a simpler time, a golden age of aviation perhaps?
+Disappointingly, such a device has not yet surfaced beyond the top secret labs of the NRC.
+We must rely upon more tangible resources, such as students, cadets and young urban professionals.
+Diversification and integration of alternative human resources are being tapped, like springtime sugar maples.
+
+It's been a long winter, or two.
+Attempts to grasp fleeting moments of human connection leave us questioning what sort of catharsis may be obtained from a grocery store checkout aisle.
+I myself, have taken steps towards awkward avenues of discussion by making cold-calls to random organizations which may prove to be valuable partners in the advancement of the project.
+Recognizing the North Star as an important platform for discussion and broadening the scope of Canadian heritage interpretation.
+
+With this recent streak of renewal, and a strong and healthy relationship, Project North Star and in turn, Ingenium may remain viable organizations as we continue our recovery.
+I look forward to hearing back from everyone, so that we may set a date for Project North Star RESTART 2022.
 
